### 2019-11-28 manuscript/manuscript.tex
First commit with Snakefile to run scaling. Added manuscript

\documentclass{article}

\usepackage{url}
\usepackage{nicefrac}
\usepackage{amssymb,amsfonts,amsmath,amsthm}
\usepackage{mathtools}
\usepackage{adjustbox}
\usepackage{bm}
\usepackage{bbold}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1

\usepackage{xcolor}
\definecolor{RED}{HTML}{EB6231}
\definecolor{YELLOW}{HTML}{E29D26}
\definecolor{BLUE}{HTML}{5D80B4}
\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
\definecolor{GREEN}{HTML}{8FB03E}
\definecolor{PURPLE}{HTML}{BE1E2D}
\definecolor{BROWN}{HTML}{A97C50}
\definecolor{PINK}{HTML}{DA1C5C}

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\dnds}{dN/dS}
\newcommand{\indice}{l}
\newcommand{\indiceexp}{^{(\indice)}}
\newcommand{\angstrom}{\text{\normalfont\AA}}

% Time, effective population size and mutation rate.
\newcommand{\Ne}{N_{\mathrm{e}}}
\newcommand{\Nsite}{S}
\newcommand{\site}{\text{s}}
% DNA

\newcommand{\ci}{{i}}
\newcommand{\cj}{{j}}
\newcommand{\itoj}{\ci, \cj}
\newcommand{\nucitoj}{\ci \rightarrow \cj}
\newcommand{\submatrix}{Q}

\begin{document}

\part*{Simulations}

\section{Simulation with independent fitness profiles}
\begin{center}
	\includegraphics[width=165mm] {artworks/ModelSimuDiv.pdf}
\end{center}

\section{Simulation in Wright-Fisher context}
\begin{center}
	\includegraphics[width=165mm] {artworks/ModelSimuPoly.pdf}
\end{center}

\section{Simulation with Fisher geometric landscape}
\begin{center}
	\includegraphics[width=165mm] {artworks/ModelSimuGeo.pdf}
\end{center}

\section{Simulation with protein folding probability}
We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites) as in Goldstein \& Pollock (2017).
From a DNA sequence $\mathbb{S}^t$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}(\mathbb{S}^t)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
\begin{equation}
G_{\mathrm{F}}(\mathbb{S}^t) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\mathbb{S}^t(\site), \mathbb{S}^t(r)  \right),
\end{equation}
where $I(a,b)$ is the pair-wise contact energies between amino-acid $a$ and $b$, using contact potentials estimated by Miya-zawa and Jernigan, and $\mathcal{N}(\site)$ are the neighbor residues of site $\site$ (closer than $7\angstrom$) in the $3$D structure.\\

The free energy of unfolded states $G_{\mathrm{U}}(\mathbb{S}^t)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
\begin{equation}
G_{\mathrm{U}}(\mathbb{S}^t) = \langle G(\mathbb{S}^t) \rangle - kT \log (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G(\mathbb{S}^t)^2 \rangle - \langle G(\mathbb{S}^t) \rangle^2\right] }{kT}
\end{equation}
where the average $\langle . \rangle$ runs other the $55$ decoy $3$D structures, and $k$ is the Boltzmann constant and $T$ the temperature in Kelvin.\\

From the energy of folded and unfolded states, we can compute the difference in free energy between the states:
\begin{equation}
\Delta G^t = G_{\mathrm{F}}(\mathbb{S}^t) - G_{\mathrm{U}}(\mathbb{S}^t)
\end{equation}

And the fitness is defined as the probability of our protein to be in the folded state: 
\begin{equation}
f(\Delta G^{t}) = \dfrac{P_{\mathrm{F}}}{P_{\mathrm{F}} + P_{\mathrm{U}}} = \dfrac{e^{-\beta G_{\mathrm{F}}(\mathbb{S}^t) }}{e^{-\beta G_{\mathrm{F}} (\mathbb{S}^t) } + e^{-\beta G_{\mathrm{U}}(\mathbb{S}^t) }} = \dfrac{1}{1 + e^{\beta \Delta G(\mathbb{S}^t) }}, 
\end{equation}
where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
\begin{center}
	\includegraphics[width=165mm] {artworks/ModelSimuFold.pdf}
\end{center}
For each possible mutant ($t+1$ substitutions), we compute $\Delta G^{t+1}$ from the updated sequence $\mathbb{S}^{t+1}$, and subsequently the selection coefficient of the mutant:
\begin{equation}
s \left( \Delta G^{t}, \Delta G^{t+1} \right) = \dfrac{f(\Delta G^{t+1}) - f(\Delta G^{t})}{f(\Delta G^{t})}.
\end{equation}
The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
\begin{equation}
{\submatrix_{\itoj}} = \mu_{\itoj} \dfrac{4 \Ne s \left( \Delta G^{t}, \Delta G^{t+1} \right)}{{1 - \e^{-4 \Ne s \left( \Delta G^{t}, \Delta G^{t+1} \right)} }}, 
\end{equation}
where ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.

\end{document}

### 2019-11-28 DataSimulated/simulated_experiment.py

@@ -1,0 +1,50 @@
+#!python3
+import argparse
+import os
+from subprocess import run
+
+
+def create_experiment(experiment, screen, sbatch, nbr_cpu):
+    exp_path = os.getcwd() + '/Experiments/' + experiment
+
+    os.makedirs(exp_path, exist_ok=True)
+    os.system('cp config.yaml {0}'.format(exp_path))
+    os.remove(exp_path + "/Snakefile") if os.path.exists(exp_path + "/Snakefile") else None
+    os.symlink(os.getcwd() + "/Snakefile", exp_path + "/Snakefile")
+    run_file = exp_path + "/snakeslurm.sh"
+    with open(run_file, 'w') as w:
+        w.write("#!/usr/bin/env bash\n")
+        run_str = 'snakemake '
+        if sbatch:
+            run_str += '-j 99 --cluster "sbatch -J {0} -p long -N 1 ' \
+                       '-o {1}/slurm.%x.%j.out -e {1}/slurm.%x.%j.err '.format(experiment, exp_path)
+            run_str += '--cpus-per-task={params.threads} --mem={params.mem} -t {params.time}"\n'
+        else:
+            run_str += "-j {0} --printshellcmds".format(nbr_cpu)
+        w.write(run_str)
+    os.system("chmod 755 " + run_file)
+    cmd = 'cd ' + exp_path + ' && ./snakeslurm.sh'
+    screen_cmd = 'screen -dmS ' + experiment + ' bash -c "' + cmd + '"'
+    with open(exp_path + "/screen.sh", 'w') as w:
+        w.write("#!/usr/bin/env bash\n")
+        w.write(screen_cmd)
+    if screen:
+        print(screen_cmd)
+        run(screen_cmd, shell=True)
+    else:
+        print(cmd)
+        run(cmd, shell=True)
+
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-e', '--experiment', required=False, type=str, default="selcoeff", dest="experiment")
+    parser.add_argument('-s', '--screen', required=False, type=bool, default=False, dest="screen")
+    parser.add_argument('-b', '--sbatch', required=False, type=bool, default=False, dest="sbatch")
+    parser.add_argument('-c', '--nbr_cpu', required=False, type=int, default=4, dest="nbr_cpu")
+    args = parser.parse_args()
[…]

### 2019-11-28 run_all.sh

@@ -1,0 +1,9 @@
+#!/usr/bin/env bash
+for EXPERIMENT in ./DataSimulated/Experiments/*; do
+  NAME=$(basename "${EXPERIMENT}")
+  echo "${NAME}"
+  cd ${EXPERIMENT}
+  snakemake --unlock
+  snakemake --printshellcmds --rerun-incomplete -j 4
+  cd ../../..
+done

### 2019-11-28 scripts/scaling_analysis.py

@@ -1,0 +1,47 @@
+#!python3
+import os
+import argparse
+import numpy as np
+import pandas as pd
+import matplotlib
+
+matplotlib.rcParams['font.family'] = 'monospace'
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
+    parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
+    args = parser.parse_args()
+    print(args.input, args.output)
+    files = dict()
+    for filepath in args.input:
+        if not os.path.isfile(filepath):
+            continue
+        for param, vals in pd.read_csv(filepath, sep='\t').items():
+            if param not in files:
+                files[param] = list()
+            files[param].append({'x': float(filepath.split("_")[-2]), 'y': vals})
+
+    for param, list_x_y in files.items():
+        sorted_x_y = sorted(list_x_y, key=lambda i: i['x'])
+        x = [i['x'] for i in sorted_x_y]
+        y_array = [i['y'] for i in sorted_x_y]
+        my_dpi = 256
+        fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
+        plt.plot(x, [np.mean(a) for a in y_array], linewidth=3)
+        plt.fill_between(x, [np.nanpercentile(a, 10) for a in y_array], [np.nanpercentile(a, 90) for a in y_array], alpha=0.3)
+        plt.xlabel(r'$\mathrm{N_{e}}$')
+        plt.xscale("log")
+        plt.ylabel(param)
+        minimum = min([min(a) for a in y_array])
+        if "dnd" in param:
+            plt.ylim((0, 1))
+        if minimum > 0 and ("dnd" not in param):
+            plt.yscale("log")
+        plt.tight_layout()
+        name = param.replace("<", "").replace(">", "").replace("|", "-")
+        plt.savefig("{0}/{1}.svg".format(args.output, name), format="svg")
+        plt.clf()
+        plt.close('all')

### 2019-11-28 scripts/scaling_merge.py

@@ -1,0 +1,13 @@
+#!python3
+import os
+import argparse
+import pandas as pd
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
+    parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
+    args = parser.parse_args()
+    rows = [{k: v[0] for k, v in pd.read_csv(filepath + '.tsv', sep='\t').items()} for filepath in args.input if
+            os.path.isfile(filepath + '.tsv')]
+    pd.DataFrame(rows).to_csv(args.output, index=False, header=rows[0].keys(), sep="\t")

### 2019-11-28 scripts/simurelax_analysis.py

@@ -1,0 +1,69 @@
+from glob import glob
+from csv import reader
+from math import isfinite
+from numpy import nanmean, nanstd
+import matplotlib.pyplot as plt
+
+folder_path = "/home/thibault/PolyMutSel/DataSimulated/SimuRelax"
+
+for tsv_path in sorted(glob("{0}/*.tsv".format(folder_path))):
+    with open(tsv_path, 'r') as tsvfile:
+        print(tsv_path)
+        tsvin = list(reader(tsvfile, delimiter='\t'))
+
+        params = tsv_path.split("/")[-1].replace('.tsv', '').split("_")
+        population_size = int(params[0])
+        k = int(params[1])
+        mu = float(params[2])
+        r = float(params[3])
+        n = int(params[4])
+        m = int(params[5])
+        a = float(params[6])
+        q = float(params[7])
+        title = "$Ne={0}$, $k={1}$, $\\mu={2}$, $r={3}$, $c={4}$, $p={5}$, $a={6}$, $Q={7}$"
+        title = title.format(population_size, k, mu, r, n, m, a, q)
+
+        time = int([row[1] for row in tsvin if row[0] == "t"][0])
+
+        data = [row for row in tsvin if len(row) == 4 and row[1] != "0"]
+        dict_num_label_data = {}
+        for row in data:
+            if not row[1] in dict_num_label_data:
+                dict_num_label_data[row[1]] = {}
+            if not row[2] in dict_num_label_data[row[1]]:
+                dict_num_label_data[row[1]][row[2]] = {"x": [], "y": []}
+
+            if isfinite(float(row[3])):
+                dict_num_label_data[row[1]][row[2]]["x"].append(float(row[0]))
+                dict_num_label_data[row[1]][row[2]]["y"].append(float(row[3]))
+
+        my_dpi = 92
+        rc_size = len(dict_num_label_data)
+        plt.figure(figsize=(1920 / my_dpi, 360 * rc_size / my_dpi), dpi=my_dpi)
+        for i, (num, dict_label_data) in enumerate(dict_num_label_data.items()):
+            for index in [False, True]:
+                plt.subplot(rc_size, 2, i * 2 + index + 1)
+                maximum = 0.0
+                minimum = 0.0
+                for label, dict_data in dict_label_data.items():
+                    if index:
+                        index = len([t for t in dict_data["x"] if t < time])
+                    plt.plot(dict_data["x"][index:], dict_data["y"][index:], label=label)
[…]

### 2019-11-28 scripts/simurelax_runs.py

@@ -1,0 +1,77 @@
+import os
+from subprocess import run
+
+cluster = False
+nbr_cpu = 1
+
+if cluster:
+    current_dir = "/beegfsdata/tlatrill/SimuVsInfer"
+    cmd = "qsub"
+else:
+    current_dir = "/home/thibault/PolyMutSel"
+    cmd = "sh"
+
+os.makedirs(current_dir, exist_ok=True)
+os.makedirs(current_dir + "/qsub", exist_ok=True)
+data_path = "{0}/DataSimulated/SimuRelax".format(current_dir)
+os.makedirs(data_path, exist_ok=True)
+
+
+def runs(pop_size, k, mu, r, n, m, a, q):
+    id_prefix = "{0}_{1}_{2}_{3}_{4}_{5}_{6}_{7}".format(pop_size, k, mu, r, n, m, a, q)
+    qsub_path = "{0}/qsub/{1}.pbs".format(current_dir, id_prefix)
+
+    qsub_str = "#!/bin/bash\n"
+    qsub_str += "#\n"
+    qsub_str += "#PBS -q q1day\n"
+    qsub_str += "#PBS -l nodes=1:ppn={0},mem=1gb\n".format(nbr_cpu)
+    qsub_str += "#PBS -o /pandata/tlatrill/read/read_out_{0}\n".format(id_prefix)
+    qsub_str += "#PBS -e /pandata/tlatrill/read/read_err_{0}\n".format(id_prefix)
+    qsub_str += "#PBS -j oe\n"
+    qsub_str += "#PBS -W umask=022\n"
+    qsub_str += "#PBS -r n\n"
+
+    simu_evol_cmd = "{0}/SimuEvol/build/SimuRelax".format(current_dir)
+    simu_evol_cmd += " --pop_size {0} --chromosomes {1} --mutation_rate_per_generation {2} --radius {3} --complexity {4} --pleiotropy {5} --peakness {6} --epistasis {7} --output {8}"
+    simu_evol_cmd = simu_evol_cmd.format(pop_size, k, mu, r, n, m, a, q, data_path)
+    qsub_str += simu_evol_cmd
+
+    qsub = open(qsub_path, 'w')
+    qsub.write(qsub_str)
+    qsub.close()
+
+    print("Running " + qsub_path)
+    run("{0} {1}".format(cmd, qsub_path), shell=True)
+    print("Finished running")
+
+
[…]

### 2020-01-21 scripts/heatmap_analysis.py
Heatmap analysis along 2 parameters

@@ -1,0 +1,49 @@
+#!python3
+import os
+import argparse
+import numpy as np
+import pandas as pd
+import itertools
+import matplotlib
+
+matplotlib.rcParams['font.family'] = 'monospace'
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
+    parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
+    parser.add_argument('-x', '--x_param', required=True, type=str, dest="x_param")
+    parser.add_argument('-y', '--y_param', required=True, type=str, dest="y_param")
+    args = parser.parse_args()
+    files = dict()
+    for filepath in args.input:
+        if not os.path.isfile(filepath):
+            continue
+        for param, vals in pd.read_csv(filepath, sep='\t').items():
+            if param not in files:
+                files[param] = dict()
+            files[param][(float(filepath.split("_")[-3]), float(filepath.split("_")[-2]))] = np.mean(vals)
+    
+    for param, x_y_z in files.items():
+        x_axis = sorted(set([k[0] for k in x_y_z.keys()]))
+        y_axis = sorted(set([k[1] for k in x_y_z.keys()]))
+        y_array, x_array = np.meshgrid(y_axis, x_axis)
+        z_array = np.zeros((len(x_axis), len(y_axis)))
+        for (i, x), (j, y) in itertools.product(enumerate(x_axis), enumerate(y_axis)):
+            z_array[i, j] = x_y_z[(x, y)]
+        my_dpi = 256
+        fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
+        c = plt.pcolor(x_array, y_array, z_array)
+        plt.title(param)
+        plt.colorbar(c)
+        plt.xlabel(args.x_param)
+        plt.xscale("log")
+        plt.ylabel(args.y_param)
+        plt.yscale("log")
+        plt.tight_layout()
+        name = param.replace("<", "").replace(">", "").replace("|", "-")
+        plt.savefig("{0}/{1}.svg".format(args.output, name), format="svg")
+        plt.clf()
+        plt.close('all')

### 2020-01-21 manuscript/manuscript.tex
Manuscript added fisher geometric landscape

@@
 
+\newcommand{\phenoDiv}{S}
+\newcommand{\phenoGeo}{\bm{P}}
+\newcommand{\phenoFold}{\Delta G}
+
 \begin{document}
@@
 \section{Simulation with independent fitness profiles}
+
+Overall, the protein phenotype is computed as the sum of site-specific selection coefficient, obtained by accessing the amino-acid present at each site of the protein.
+From a DNA sequence $\mathbb{S}^t$ after $t$ substitutions, the protein's phenotype is given by:
+\begin{equation}
+\phenoGeo\left(\mathbb{S}^{t}\right) = \sum_{1 \leq \site \leq \Nsite} \phenoGeo_{\site} \left(\mathbb{S}^t(\site) \right),
+\end{equation}
+where $\phenoGeo_{\site}$ is the genotype-phenotype map at site $\site$.\\
+
+And the fitness of $\mathbb{S}^t$ is : 
+\begin{equation}
+w\left( \phenoGeo\left(\mathbb{S}^{t}\right) \right) = e^{-\alpha \left| \phenoGeo\left(\mathbb{S}^{t}\right) \right|^{\beta}},
+\end{equation}
+where strength ($\alpha > 0$) and epistasis ($\beta$) are parameters of the fitness function.
+
+The Malthusian fitness selection coefficients $f$ at site $\site$, are obtained by taking the logarithm of the fitness profile assigned to this site:
+\begin{equation}
+\label{eq:sitefitness}
+f = \ln \left( g \right)
+\end{equation}
+
 \begin{center}
@@
 
+For each possible mutant ($t+1$ substitutions), we compute $\phenoGeo\left(\mathbb{S}^{t+1}\right)$ from the updated sequence $\mathbb{S}^{t+1}$, and subsequently the selection coefficient of the mutant:
+\begin{equation}
+s \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right) = \ln 
+\end{equation}
+The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
+\begin{equation}
+{\submatrix_{\itoj}} = \mu_{\itoj} \dfrac{4 \Ne s \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right)}{{1 - \e^{-4 \Ne \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right)} }}, 
+\end{equation}
+where ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
+
 \section{Simulation in Wright-Fisher context}
+
+The evolutionary dynamics was formalized as a Wright-Fisher model with mutation, selection and drift.
+The population is assumed to be panmictic, with constant size $N_e$ and with non-overlapping generations.
+
+The sites mutate at rate $u$ per generation. Each non-synonymous mutation produces a new functional PRDM9 variant
+
 \begin{center}
@@
 \section{Simulation with Fisher geometric landscape}
+
+We simulated substitutions in a protein using an adaptation of Fisher's geometric landscape.
+In the original context, the phenotype is a vector ($\phenoGeo$) in a multidimensional space, where the number of dimensions is often termed complexity.
+From a phenotype, the fitness is a monotonously decreasing function of the phenotype distance to $0$.
+The exact functional phenotype-fitness map depend on $2$ external parameters controlling for strength ($\alpha$) and epistasis ($\beta$).
+If the phenotype-fitness map is explicit, the genotype-phenotype map is more pervasive.
+Mutations are seen has displacement of the phenotype in the multidimensional space.
+Beneficial mutations are moving the phenotype closer to $0$, whereas deleterious mutations are moving the phenotype further away.
+In such original context, the distribution of mutational effects is not dependent on the current genotype, but this can be relaxed using a genotype-phenotype map.\\
+
+In a protein context, the genotype-phenotype map can be defined by assigning to each of the $20$ amino-acid a vector in the multidimensional space.
+Since different sites of the protein do not have the same physico-chimical properties, we can define a specific genotype-phenotype map for each position of the sequence.
+Overall, the protein phenotype is computed as the sum of site-specific multidimensional vectors, obtained by accessing the amino-acid present at each site of the protein.
+From a DNA sequence $\mathbb{S}^t$ after $t$ substitutions, the protein's phenotype is given by:
+\begin{equation}
+\phenoGeo\left(\mathbb{S}^{t}\right) = \sum_{1 \leq \site \leq \Nsite} \phenoGeo_{\site} \left(\mathbb{S}^t(\site) \right),
+\end{equation}
+where $\phenoGeo_{\site}$ is the genotype-phenotype map at site $\site$.\\
+
+And the Wrightian fitness of $\mathbb{S}^t$ is : 
+\begin{equation}
+w\left( \phenoGeo\left(\mathbb{S}^{t}\right) \right) = e^{-\alpha \left| \phenoGeo\left(\mathbb{S}^{t}\right) \right|^{\beta}},
+\end{equation}
+where strength ($\alpha > 0$) and epistasis ($\beta$) are parameters of the fitness function.
 \begin{center}
@@
 \end{center}
+For each possible mutant ($t+1$ substitutions), we compute $\phenoGeo\left(\mathbb{S}^{t+1}\right)$ from the updated sequence $\mathbb{S}^{t+1}$, and subsequently the selection coefficient of the mutant:
+\begin{equation}
+s \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right) = \dfrac{ w\left( \phenoGeo\left(\mathbb{S}^{t+1}\right) \right) - w\left( \phenoGeo\left(\mathbb{S}^{t}\right) \right)}{w\left( \phenoGeo\left(\mathbb{S}^{t}\right) \right)}.
+\end{equation}
+The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
+\begin{equation}
+{\submatrix_{\itoj}} = \mu_{\itoj} \dfrac{4 \Ne s \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right)}{{1 - \e^{-4 \Ne \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right)} }}, 
+\end{equation}
+where ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
 
@@
 We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites) as in Goldstein \& Pollock (2017).
-From a DNA sequence $\mathbb{S}^t$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}(\mathbb{S}^t)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
+From a DNA sequence $\mathbb{S}^t$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}\left(\mathbb{S}^{t}\right)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
 \begin{equation}
-G_{\mathrm{F}}(\mathbb{S}^t) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\mathbb{S}^t(\site), \mathbb{S}^t(r)  \right),
+G_{\mathrm{F}}\left(\mathbb{S}^{t}\right) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\mathbb{S}^t(\site), \mathbb{S}^t(r) \right),
 \end{equation}
@@
 
-The free energy of unfolded states $G_{\mathrm{U}}(\mathbb{S}^t)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
+The free energy of unfolded states $G_{\mathrm{U}}\left(\mathbb{S}^{t}\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
 \begin{equation}
-G_{\mathrm{U}}(\mathbb{S}^t) = \langle G(\mathbb{S}^t) \rangle - kT \log (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G(\mathbb{S}^t)^2 \rangle - \langle G(\mathbb{S}^t) \rangle^2\right] }{kT}
+G_{\mathrm{U}}\left(\mathbb{S}^{t}\right) = \langle G\left(\mathbb{S}^{t}\right) \rangle - kT \ln (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G\left(\mathbb{S}^{t}\right)^2 \rangle - \langle G\left(\mathbb{S}^{t}\right) \rangle^2\right] }{kT}
 \end{equation}
@@
 \begin{equation}
-\Delta G^t = G_{\mathrm{F}}(\mathbb{S}^t) - G_{\mathrm{U}}(\mathbb{S}^t)
+\phenoFold\left(\mathbb{S}^{t}\right) = G_{\mathrm{F}}\left(\mathbb{S}^{t}\right) - G_{\mathrm{U}}\left(\mathbb{S}^{t}\right)
 \end{equation}
 
-And the fitness is defined as the probability of our protein to be in the folded state: 
+Wrightian fitness is defined as the probability of our protein to be in the folded state: 
 \begin{equation}
-f(\Delta G^{t}) = \dfrac{P_{\mathrm{F}}}{P_{\mathrm{F}} + P_{\mathrm{U}}} = \dfrac{e^{-\beta G_{\mathrm{F}}(\mathbb{S}^t) }}{e^{-\beta G_{\mathrm{F}} (\mathbb{S}^t) } + e^{-\beta G_{\mathrm{U}}(\mathbb{S}^t) }} = \dfrac{1}{1 + e^{\beta \Delta G(\mathbb{S}^t) }},
+w(\phenoFold\left(\mathbb{S}^{t}\right)) = \dfrac{P_{\mathrm{F}}\left(\mathbb{S}^{t}\right)}{P_{\mathrm{F}}\left(\mathbb{S}^{t}\right) + P_{\mathrm{U}}\left(\mathbb{S}^{t}\right)} = \dfrac{e^{-\beta G_{\mathrm{F}}\left(\mathbb{S}^{t}\right) }}{e^{-\beta G_{\mathrm{F}} \left(\mathbb{S}^{t}\right) } + e^{-\beta G_{\mathrm{U}}\left(\mathbb{S}^{t}\right) }} = \dfrac{1}{1 + e^{\beta \phenoFold\left(\mathbb{S}^{t}\right) }},
 \end{equation}
@@
 \end{center}
-For each possible mutant ($t+1$ substitutions), we compute $\Delta G^{t+1}$ from the updated sequence $\mathbb{S}^{t+1}$, and subsequently the selection coefficient of the mutant:
+For each possible mutant ($t+1$ substitutions), we compute $\phenoFold^{t+1}$ from the updated sequence $\mathbb{S}^{t+1}$, and subsequently the selection coefficient of the mutant:
 \begin{equation}
-s \left( \Delta G^{t}, \Delta G^{t+1} \right) = \dfrac{f(\Delta G^{t+1}) - f(\Delta G^{t})}{f(\Delta G^{t})}.
+s \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right) = \dfrac{ w\left( \phenoFold\left(\mathbb{S}^{t+1}\right) \right) - w\left( \phenoFold\left(\mathbb{S}^{t}\right) \right)}{w\left( \phenoFold\left(\mathbb{S}^{t}\right) \right)}.
 \end{equation}
@@
 \begin{equation}
-{\submatrix_{\itoj}} = \mu_{\itoj} \dfrac{4 \Ne s \left( \Delta G^{t}, \Delta G^{t+1} \right)}{{1 - \e^{-4 \Ne s \left( \Delta G^{t}, \Delta G^{t+1} \right)} }}, 
+{\submatrix_{\itoj}} = \mu_{\itoj} \dfrac{4 \Ne s \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right)}{{1 - \e^{-4 \Ne \left( \mathbb{S}^{t},\mathbb{S}^{t+1}\right)} }}, 
 \end{equation}

### 2020-01-23 scripts/heatmap_analysis.py
Compute the half-life to reach the new dNdS. Heatmap along composite parameters in x & y

@@ -43,7 +43,7 @@ if __name__ == '__main__':
         plt.ylabel(args.y_param)
         plt.yscale("log")
         plt.tight_layout()
-        name = param.replace("<", "").replace(">", "").replace("|", "-")
+        name = param.replace("/", "").replace("|", "")
         plt.savefig("{0}/{1}.svg".format(args.output, name), format="svg")
         plt.clf()
         plt.close('all')

### 2020-01-23 scripts/relaxation_analysis.py

@@ -1,7 +1,6 @@
 #!python3
 import os
 import argparse
-import numpy as np
 import pandas as pd
 import matplotlib
 

### 2020-01-23 scripts/scaling_merge.py

@@ -2,12 +2,55 @@
 import os
 import argparse
 import pandas as pd
+import matplotlib
+
+matplotlib.rcParams['font.family'] = 'monospace'
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
 
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
     parser.add_argument('-o', '--output', required=True, type=str, dest="output")
     parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
     args = parser.parse_args()
-    rows = [{k: v[0] for k, v in pd.read_csv(filepath + '.tsv', sep='\t').items()} for filepath in args.input if
-            os.path.isfile(filepath + '.tsv')]
-    pd.DataFrame(rows).to_csv(args.output, index=False, header=rows[0].keys(), sep="\t")
+    dict_df = {filepath: {k: v[0] for k, v in pd.read_csv(filepath + '.tsv', sep='\t').items()} for filepath in
+               args.input if os.path.isfile(filepath + '.tsv')}
+
+    cols = ["<dN>/<dN0>", "<dN/dN0>"]
+    for col in cols:
+        my_dpi = 128
+        fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
+        max_y = 0.0
+        for filepath in dict_df.keys():
+            assert (os.path.isfile(filepath + ".substitutions.tsv"))
+            df = pd.read_csv(filepath + ".substitutions.tsv", sep='\t',
+                             usecols=["NodeName", "AbsoluteStartTime", "EndTime"] + cols)
+            max_y = max((max(df[col]), max_y))
+            base_line, = plt.plot(df["AbsoluteStartTime"], df[col], linewidth=3)
+
+            end_value = 1.0
+            for node_name in set(df["NodeName"]):
+                df_filt = df[df['NodeName'] == node_name]
+                plt.axvline(x=df_filt.iloc[0]["AbsoluteStartTime"], linewidth=3, color='black')
+                start_value = df_filt[col].values[0]
+                if node_name != "0":
+                    dict_df[filepath]["Node{0}_incr{1}".format(node_name, col)] = df_filt[col].values[-1] / end_value
+                end_value = df_filt[col].values[-1]
+                half_value = (start_value + end_value) / 2
+                if half_value < start_value:
[…]

### 2020-03-04 manuscript/supp-mat.tex
First version of supp. mat.

\documentclass{article}

\usepackage{url}

\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{xcolor}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage{pgfplots, pgf,tikz}
\usepackage[margin=60pt]{geometry}
\pdfinclusioncopyfonts=1

\definecolor{RED}{HTML}{EB6231}
\definecolor{YELLOW}{HTML}{E29D26}
\definecolor{BLUE}{HTML}{5D80B4}
\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
\definecolor{GREEN}{HTML}{8FB03E}
\definecolor{PURPLE}{HTML}{BE1E2D}
\definecolor{BROWN}{HTML}{A97C50}
\definecolor{PINK}{HTML}{DA1C5C}
\captionsetup{width=0.85\textwidth}

\newcommand{\specialcell}[2][c]{%
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\DeclareMathOperator{\E}{\mathbb{E}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}

% Time, effective population size and mutation rate.
\newcommand{\Ne}{N_{\mathrm{e}}}
\newcommand{\dnds}{\omega}
\newcommand{\Nsite}{\text{n}}
\newcommand{\Nstate}{\text{K}}

\newcommand{\x}{x}
\newcommand{\eq}{^{*}}
\newcommand{\dx}{\delta \x}

\begin{document}
\part*{Supplementary materials}

\section*{Genotype to phenotype map}
Define $\Nsite$ as the number of sites in the genotype sequence.
Each site can be in one of $\Nstate \geq 2$ states, where only $1$ state is defined the stable state, and $\Nstate - 1$ states are unstable.
For a given genotype sequence, define phenotype $0 \leq \x \leq 1$ as the current proportion of sites in the unstable state.
After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
Define $\rho_{\x}(\dx)$ as the probability to get a change of phenotype equal to $\dx$, if the current phenotype is $\x$:\\
\begin{gather}
\begin{cases}
\dx &\text{ with probability } \rho_{\x}(\dx) = 1-x, \\
0 &\text{ with probability } \rho_{\x}(0) = \x \left[1 - \frac{1}{\Nstate - 1}\right], \\
-\dx &\text{ with probability } \rho_{\x}(-\dx) = \frac{x}{\Nstate - 1}.\\
\end{cases} \label{eq:proba}
\end{gather}

\section*{Selection coefficient}
$s(\x, \dx)$ is the selection coefficient of an effect $\dx$ if the current phenotype is $\x$:
\begin{align}
s(\x, \dx) &= \frac{f(\x + \dx) - f(\x)}{f( \x)}, \\
 &\simeq \frac{1}{f(\x)}\frac{\der f( \x)}{\der \x} \dx, \\
 &\simeq \frac{\der \ln (f(\x))}{\der \x} \dx, \label{eq:s_from_fitness}
\end{align}
where $f( \x)$ is the Wrightian fitness of phenotype $x$. And thus we also have:
\begin{gather}
s(\x, -\dx) \simeq - s(\x, \dx) \text{ from eq. \ref{eq:s_from_fitness}} \label{eq:s_minus_deltax}, \\
\iff S(\x, -\dx) \simeq - S(\x, \dx), \label{eq:S_minus_deltax}
\end{gather}
where $S(\x\eq, \dx) = 4\Ne s (\x\eq, \dx)$ is the scaled selection coefficient.

\section*{Probability of fixation}
The probability of fixation of a mutation with effect $\dx$, for a resident phenotype $\x$ is :
\begin{align}
p_{\text{fix}}(\x, \dx) &= \frac{ 2 s(\x, \dx)}{1 - \e^{-4\Ne s(\x, \dx)}}, \\
 &= \frac{ 2 s(\x, \dx)}{1 - \e^{-S(\x, \dx)}}. \label{eq:pfix}
\end{align}
And in the case of neutral mutations, the probability of fixation is:
\begin{gather}
p_{\text{fix}}(\x, 0) = \frac{ 1}{2 \Ne} \label{eq:pfix0}.
\end{gather}
And the ratio of probability of fixation between selected and neutral mutations is:
\begin{align}
\frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} & = \frac{ 2 \Ne 2 s(\x, \dx)}{1 - \e^{-S(\x, \dx)}},\\
& = \frac{S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \label{eq:pfixratio}.
\end{align}
\section*{Equilibrium phenotype}
At equilibrium phenotype $\x\eq$, the expected selection coefficient of mutation that reached fixation must be $0$:
\begin{gather}
 0 = \E_{\dx} \left[ s(\x\eq, \dx) p_{\text{fix}}(\x\eq, \dx) \right], \\
\iff 0 = \E_{\dx} \left[ s(\x\eq, \dx)\frac{ 2 s(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}} \right], \\
\iff 0 = s(\x\eq, \dx)\frac{ 2s(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) + s(\x\eq, 0) \frac{ \rho_{\x\eq}(0)}{2\Ne} + s(\x\eq, -\dx)\frac{ 2s(\x\eq, -\dx)}{1 - \e^{-S(\x\eq, -\dx)}}   \rho_{\x\eq}(-\dx), \\
\Longrightarrow s(\x\eq, \dx) \frac{ 2s(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) \simeq s(\x\eq, \dx) \frac{ - 2s(\x\eq, \dx)}{1 - \e^{S(\x\eq, \dx)}}   \rho_{\x\eq}(-\dx) \text{ from eq. \ref{eq:S_minus_deltax}}, \\
\iff \frac{ \rho_{\x\eq}(\dx)}{\rho_{\x\eq}(-\dx)} \simeq \e^{-S(\x\eq, \dx)} \frac{ \e^{-S(\x\eq, \dx)} - 1}{ \e^{-S(\x\eq, \dx)} \left( 1 - \e^{S(\x\eq, \dx)} \right)}, \\
\iff \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq -S(\x\eq, \dx) \text{ from eq. \ref{eq:proba}} \label{eq:equilibrium}, \\
\iff \x\eq \simeq \frac{1}{1 + \frac{\e^{-S(\x\eq, \dx)}}{\Nstate-1}}.
\end{gather}

\section*{Substitution rate ($\bm{\dnds}$)}
The substitution rate of selected mutations relative to neutral mutations  ($\dnds$) is by definition:
\begin{align}
\dnds & = \E_{\dx} \left[ \frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} \right], \\
 & = \E_{\dx} \left[ \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \right], \\
 & = (1 - \x) \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} + \x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) + \frac{\x}{\Nstate-1} \frac{ S(\x, -\dx)}{1 - \e^{-S(\x, -\dx)}} \text{ from eq. \ref{eq:proba}}, \\
 & = (1 - \x) \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} - \frac{\x}{\Nstate-1}  \frac{ S(\x, \dx)}{1 - \e^{S(\x, \dx)}} +  \x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) \text{ from eq. \ref{eq:S_minus_deltax}}.
\end{align}
$\dnds\eq$ at equilibrium is then determined by the phenotype at equilibrium $\x\eq$:
\begin{align}
\dnds\eq &= (1 - \x\eq) \frac{ S(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}} - \frac{\x\eq}{\Nstate-1} \frac{ S(\x\eq, \dx)}{1 - \e^{S(\x\eq, \dx)}} + \x\eq \left(\frac{\Nstate - 2}{\Nstate - 1}\right) , \\
 &= \x\eq \left[ \frac{2 (\x\eq-1)  \left(\ln (\Nstate-1)+\ln \left(\frac{1 -\x\eq}{\x\eq}\right)\right)}{\Nstate (\x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right] \text{ from eq. \ref{eq:equilibrium}}.
\end{align}
\begin{center}
	\begin{tikzpicture}
	\begin{axis}[
	ylabel={$\dnds\eq$},
	xlabel={$\x\eq$},
	domain=0:1.1,
	ymin=0.0, ymax=1.05,
	samples=50,
	legend entries={$\Nstate=2$, $\Nstate=3$, $\Nstate=4$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.6,0.34)},anchor=north west}
	]
	\addplot[line width=2.0pt, color=BLUE]{ (ln(2-1) + ln((1-x) / x)) * x * 2 * (x - 1) / (2 * (x - 1) + 1) + x * (2-2) / (2-1) };
	\addplot[line width=2.0pt, color=YELLOW]{ (ln(3-1) + ln((1-x) / x)) * x * 2 * (x - 1) / (3 * (x - 1) + 1) + x * (3-2) / (3-1)};
	\addplot[line width=2.0pt, color=GREEN]{ (ln(4-1) + ln((1-x) / x)) * x * 2 * (x - 1) / (4 * (x - 1) + 1) + x * (4-2) / (4-1)};
	\end{axis}
	\end{tikzpicture}
\end{center}
And the derivative of $\dnds\eq$ w.r.t to $\x\eq$ is: 
\begin{gather}
\frac{\der \dnds\eq}{\der \x\eq} = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \left[ \ln (\Nstate-1) + \ln \left(\frac{1 - \x\eq}{\x\eq} \right) \right]}{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1} \label{eq:dw_dx}.
\end{gather}

\begin{center}
	\begin{tikzpicture}
	\begin{axis}[
	ylabel={$\frac{\der \dnds\eq}{\der \x\eq}$},
	xlabel={$\x\eq$},
	domain=0.0:1.0,
	ymin=0, ymax=6,
	samples=50,
	legend entries={$\Nstate=2$, $\Nstate=3$, $\Nstate=4$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.5,0.95)},anchor=north west}
	]
	\addplot[line width=2.0pt, color=BLUE]{2*(2*(x-1)+1 + (2*(x-1)^2+2*x-1)*(ln(2-1) + ln((1-x)/x)))/(2*(x - 1) + 1)^2 + (2-2) / (2-1) };
	\addplot[line width=2.0pt, color=YELLOW]{2*(3*(x-1)+1 + (3*(x-1)^2+2*x-1)*(ln(3-1) + ln((1-x)/x)))/(3*(x - 1) + 1)^2 + (3-2) / (3-1) };
	\addplot[line width=2.0pt, color=GREEN]{2*(4*(x-1)+1 + (4*(x-1)^2+2*x-1)*(ln(4-1) + ln((1-x)/x)))/(4*(x - 1) + 1)^2 + (4-2) / (4-1)};
	\addplot[dashed,line width=0.5pt, color=black]{0};
	\end{axis}
	\end{tikzpicture}
\end{center}

\section*{Substitution rate ($\bm{\dnds}$) for the special case \Nstate=2}
In the special of $2$ states ($1$ stable and $1$ unstable), $\dnds\eq$ at equilibrium is:
\begin{align}
\dnds\eq & = \ln \left( \frac{1 - \x\eq}{\x\eq} \right) \frac{2 (\x\eq -1) \x\eq }{2 \x\eq - 1},\\
& = \begin{cases}
 - 2 \x\eq \ln (\x\eq) + \mathcal{O}({\x\eq}^2) \text{ as } \x\eq \rightarrow 0, \\
 1 - \frac{8}{3} (\x\eq - 0.5)^2 + \mathcal{O}((\x\eq - 0.5)^3) \text{ as } \x\eq \rightarrow 0.5. \\
\end{cases}
\end{align}
\begin{center}
	\begin{tikzpicture}
	\begin{axis}[
	ylabel={$\dnds\eq$},
	xlabel={$\x\eq$},
	domain=0:0.52,
	ymin=0.0, ymax=1.05,
	samples=100,
	legend entries={
		$\ln( \frac{1 - \x\eq}{\x\eq}) \frac{2 (\x\eq -1) \x\eq }{2 \x\eq - 1}$,
		$- 2 \x\eq \ln (\x\eq)$,
		$ 1 - \frac{8}{3} (\x\eq - 0.5)^2$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.5,0.34)},anchor=north west}
	]
	\addplot[line width=2.0pt, color=BLUE]{ ln((1-x) / x) * x * 2 * (x - 1) / (2 * x - 1)};
	\addplot[dashed,line width=2.0pt, color=YELLOW]{ -2 * x * ln(x) };
	\addplot[dashed,line width=2.0pt, color=GREEN]{ 1 - 8 * (x - 0.5)^2 / 3 };
	\end{axis}
	\end{tikzpicture}
\end{center}
And the derivative of $\dnds\eq$ w.r.t to $\x\eq$ is: 
\begin{align}
\frac{\der \dnds\eq}{\der \x\eq} &= \frac{\ln \left( \frac{1 - \x\eq}{\x\eq} \right)(2 - 4 \x\eq + 4 {\x\eq}^2) + 4\x\eq - 2}{(2\x\eq -1)^2} \label{eq:dw_dx_Nstate2}, \\
&= 
\begin{cases}
 - (\x\eq + 1)(2\ln(\x\eq) + 2) - 2\x\eq + \mathcal{O}({\x\eq}^2) \text{ as } \x\eq \rightarrow 0, \\
 \frac{8 - 16 \x\eq}{3} + \mathcal{O}((\x\eq - 0.5)^2) \text{ as } \x\eq \rightarrow 0.5.
\end{cases} \label{eq:dw_dx_approx_Nstate2}
\end{align}

\begin{center}
	\begin{tikzpicture}
	\begin{axis}[
	ylabel={$\frac{\der \dnds(\x\eq)}{\der \x\eq}$},
	xlabel={$\x\eq$},
	domain=0.0:0.51,
	ymin=0.0, ymax=6,
	samples=30,
	legend entries={
		$\frac{\ln ( \frac{1 - \x\eq}{\x\eq})(2 - 4 \x\eq + 4 {\x\eq}^2) + 4\x\eq - 2}{(2\x\eq -1)^2}$,
		$- (\x\eq + 1)(2\ln(\x\eq) + 2) - 2\x\eq$,
		$\frac{8 - 16 \x\eq}{3}$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.25,0.95)},anchor=north west}
	]
	\addplot[line width=2.0pt, color=BLUE]{ (-2 + 4 * x + (2 - 4 * x + 4 * x^2) * ln((1-x) / x))/(1 - 2 * x)^2 };
	\addplot[dashed,line width=2.0pt, color=YELLOW]{ -(x+1)*(2*ln(x) + 2) -2*x };
	\addplot[dashed,line width=2.0pt, color=GREEN]{ 8 * ( 1 - 2*x)/3};
	\end{axis}
	\end{tikzpicture}
\end{center}

\section*{Change of equilibrium phenotype ($\bm{\x\eq}$) after a change in $\bm{\Ne}$}
Given the equilibrium equation (eq. \ref{eq:equilibrium}), $\x\eq$ is implicitly a function of $\Ne$, and the change of phenotype at equilibrium ($\der \x\eq$) after a change of effective population size ($\der \Ne$) can be obtain as:
\begin{gather}
G(\x, \Ne ) \equiv \ln \left( \frac{1 - \x}{\x} \right) + \ln (\Nstate-1) + 4 \Ne s(\x, \dx), \\
\iff G(\x\eq(\Ne), \Ne ) = 0 \text{ from eq. \ref{eq:equilibrium}}, \\
\Longrightarrow \frac{\partial G(\x\eq, \Ne ) }{\partial \x\eq }\frac{ \der \x\eq}{\der \Ne} + \frac{ \partial G(\x\eq, \Ne )}{\partial \Ne} = 0, \\
\iff \left[ \frac{\partial \ln (1-\x\eq) }{\partial \x\eq } - \frac{\partial \ln (\x\eq) }{\partial \x\eq}  + 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right]\frac{ \der \x\eq}{\der \Ne} + 4 s(\x\eq, \dx) = 0, \\
\iff \left[ \frac{1}{(1 - \x\eq) \x\eq} + 4 \Ne \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}\dx \right]\frac{ \der \x\eq}{\der \Ne}  = - 4 \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}\dx \text{ from eq. \ref{eq:s_from_fitness}}, \\
\iff 4 \dx \left[ \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2} \right] \Ne \frac{ \der \x\eq}{\der \Ne}  = - 4 \dx \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
\iff \frac{ \der \x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}  \label{eq:dx_dlnNe}.
\end{gather}
\section*{Change of equilibrium substitution rate ($\bm{\dnds\eq}$) after a change in $\bm{\Ne}$}
Together, the change of substitution rate at equilibrium ($\der \dnds\eq$) after a change of effective population size ($\der \Ne$) can be obtain as:
\begin{align}
\frac{\der \dnds\eq}{\der \ln (\Ne)} & = \frac{\der \dnds\eq}{\der \x\eq} \frac{ \der \x\eq}{\der \ln (\Ne)}, \\
 &= - \frac{\der \dnds\eq}{\der \x\eq} \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}} \text{ from eq. \ref{eq:dx_dlnNe}} \label{eq:dw_dlnNe}.
\end{align}
\section*{Phenotype to fitness mapping}
From a given phenotype $\x$, define the fitness as :
\begin{gather}
 f(\x) = \frac{1}{1 + \e^{\beta(\alpha + \kappa \x)}}, \label{eq:fitness}
\end{gather}
where $\alpha = \Delta G_0 < 0$ is the difference in free energy between stable and unstable protein when all sites are stable. $\kappa$ is the expected change in $\Delta G$ when all sites are unstable. By definition, $\kappa \dx = \sfrac{\kappa}{\Nsite} = \Delta \Delta G $\\
The derivative of fitness w.r.t to phenotype is:
\begin{align}
 \frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\alpha + \kappa \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
 & = - \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \label{eq:df_dx}.
\end{align}
And the second derivative of fitness w.r.t to phenotype is:
\begin{align}
\frac{\partial^2 \ln (f(\x))}{\partial {\x}^2} & = - \beta \kappa \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \right) \text{ from eq. \ref{eq:df_dx}}, \\
 & = - \beta \kappa  \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{\left( 1 + \e^{\beta(\alpha + \kappa \x)}\right)^2}, \\
 & = \frac{\beta \kappa}{1 + \e^{\beta(\alpha + \kappa \x)}} \frac{ \partial \ln f(\x) }{\partial {\x}} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}.
\end{align}
The equilibrium phenotype ($\x\eq$) is :
\begin{gather}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) = 4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x\eq)}}{1 + \e^{\beta(\alpha + \kappa \x\eq)}}  \dx   \text{ from eq. \ref{eq:equilibrium} and \ref{eq:df_dx}}.
\end{gather}
Using $\Ne=10^4$, $\beta=1.686$, $\alpha = -118$, $\Nsite=300$, $\kappa=300$, we have the following:
\begin{center}
\begin{tikzpicture}
	\begin{axis}[
	xlabel={$\x$},
	ymin=0.0, ymax=5.0,
	samples=100,
	legend entries={
		$\ln(\frac{1 - \x}{\x}) + \ln(\Nstate-1) \text{, }\Nstate=2$,
		$\ln ( \frac{1 - \x}{\x})+ \ln(\Nstate-1) \text{, }\Nstate=4$,
		$\ln ( \frac{1 - \x}{\x})+ \ln(\Nstate-1) \text{, }\Nstate=8$,
		$\ln ( \frac{1 - \x}{\x})+ \ln(\Nstate-1) \text{, }\Nstate=16$,
		$4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \dx$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.5,0.95)},anchor=north west}
	]
	\addplot[domain=0.15:0.5,line width=2.0pt, color=RED]{ ln((1-x)/x) };
	\addplot[domain=0.15:0.68,line width=2.0pt, color=YELLOW]{ ln((1-x)/x) + ln(4-2) };
	\addplot[domain=0.15:0.8,line width=2.0pt, color=GREEN]{ ln((1-x)/x) + ln(8-2) };
	\addplot[domain=0.15:0.8,line width=2.0pt, color=LIGHTGREEN]{ ln((1-x)/x) + ln(16-2) };
	\addplot[domain=0.15:0.38, line width=2.0pt, color=BLUE]{ 4 * 1000 * 1.686 *exp(1.686 * (-118 + 300 *x))};
	\end{axis}
\end{tikzpicture}
\end{center}
\section*{Approximations}
We assume that $\Ne \gg 1$, $\alpha + \kappa \gg 1$ and $\kappa \dx \simeq 1$.
\begin{gather}
-S(\x\eq, \dx) \simeq 1, \\
\iff s(\x\eq, \dx) \simeq -\frac{1}{4 \Ne}, \\
\iff \frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{1}{4 \Ne \dx}, \\
\iff \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \simeq \frac{1}{4 \Ne \dx}, \\
\iff \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \simeq \frac{1}{4 \Ne \dx \beta \kappa}, \\
\Longrightarrow \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \ll 1, \\
\Longrightarrow \e^{\beta(\alpha + \kappa \x)} \ll 1, \\
\Longrightarrow \frac{\e^{\beta(\alpha + \kappa \x\eq)}}{1 + \e^{\beta(\alpha + \kappa \x\eq)}} \simeq \e^{\beta(\alpha + \kappa \x\eq)}.
\end{gather}
Moreover, we can get the following:
\begin{gather}
\frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{1}{4 \Ne \dx}, \\
\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{1}{4 \Ne \dx \beta \kappa }, \\
\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{(1 - \x\eq)\x\eq }{\beta \kappa} \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \dx \frac{(1 - \x\eq)\x\eq }{\beta } \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
\Longrightarrow \left| \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \right| \gg \left| \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} \right|. \\
\end{gather}
Together, these approximations leads to:
\begin{align}
\frac{ \der \x\eq}{\der \ln (\Ne)} & = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
& \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
& \simeq - \frac{1 + \e^{\beta(\alpha + \kappa \x\eq)}}{\beta \kappa}, \\
& \simeq - \frac{1}{\beta \kappa}.
\end{align}
Finally,
\begin{align}
\frac{ \der \dnds\eq}{\der \ln (\Ne)}  & \simeq - \frac{ \der \dnds\eq}{\der \x\eq} \frac{1}{\beta \kappa}, \\
  & = - \frac{1}{\beta \kappa}.
\end{align}
\section*{Numerical results}
\begin{center}
\begin{tikzpicture}
	\begin{axis}[
	xlabel={$\x$},
	ymin=0.0, ymax=1.0,
	samples=100,
	legend entries={$\ln(\frac{1 - \x}{\x})$,
		$\alpha = -118\text{, }\kappa = 300\text{, }\Ne = 10^{4}$,
		$\alpha = -118\text{, }\kappa = 300\text{, }\Ne = 10^{5}$,
		$\alpha = -16.7\text{, }\kappa = 30\text{, }\Ne = 10^{4}$,
		$\alpha = -16.7\text{, }\kappa = 30\text{, }\Ne = 10^{5}$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.5,0.95)},anchor=north west}
	]
	\addplot[domain=0.3:0.6,line width=2.0pt, color=BLUE]{ ln((1-x)/x) };		\addplot[domain=0.3:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
	\addplot[dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
	\addplot[domain=0.3:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-16.71+30*x))};
	\addplot[dashed,domain=0.3:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-16.71+30*x))};
	\end{axis}
\end{tikzpicture}
\end{center}
\end{document}

### 2020-03-04 scripts/scaling_analysis.py

@@ -1,47 +1,0 @@
-#!python3
-import os
-import argparse
-import numpy as np
-import pandas as pd
-import matplotlib
-
-matplotlib.rcParams['font.family'] = 'monospace'
-matplotlib.use('Agg')
-import matplotlib.pyplot as plt
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
-    parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
-    args = parser.parse_args()
-    print(args.input, args.output)
-    files = dict()
-    for filepath in args.input:
-        if not os.path.isfile(filepath):
-            continue
-        for param, vals in pd.read_csv(filepath, sep='\t').items():
-            if param not in files:
-                files[param] = list()
-            files[param].append({'x': float(filepath.split("_")[-2]), 'y': vals})
-
-    for param, list_x_y in files.items():
-        sorted_x_y = sorted(list_x_y, key=lambda i: i['x'])
-        x = [i['x'] for i in sorted_x_y]
-        y_array = [i['y'] for i in sorted_x_y]
-        my_dpi = 256
-        fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
-        plt.plot(x, [np.mean(a) for a in y_array], linewidth=3)
-        plt.fill_between(x, [np.nanpercentile(a, 10) for a in y_array], [np.nanpercentile(a, 90) for a in y_array], alpha=0.3)
-        plt.xlabel(r'$\mathrm{N_{e}}$')
-        plt.xscale("log")
-        plt.ylabel(param)
-        minimum = min([min(a) for a in y_array])
-        if "dnd" in param:
-            plt.ylim((0, 1))
-        if minimum > 0 and ("dnd" not in param):
-            plt.yscale("log")
-        plt.tight_layout()
-        name = param.replace("<", "").replace(">", "").replace("|", "-")
-        plt.savefig("{0}/{1}.svg".format(args.output, name), format="svg")
-        plt.clf()
-        plt.close('all')

### 2020-03-04 manuscript/manuscript.tex
Updated manuscript and the scripts

--- 2019-11-28 manuscript/manuscript.tex
@@
 \usepackage{url}
-\usepackage{nicefrac}
-\usepackage{amssymb,amsfonts,amsmath,amsthm}
-\usepackage{mathtools}
+
+\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
 \usepackage{adjustbox}
-\usepackage{bm}
-\usepackage{bbold}
+\usepackage{float}
+\usepackage{caption}
+\usepackage{mdframed}
+\usepackage{lmodern}
+\usepackage{bm,bbold}
+\usepackage{xfrac, nicefrac}
+\usepackage{xcolor}
+\usepackage{lmodern}
+\usepackage{enumitem}
+\usepackage{pgfplots, pgf,tikz}
 \usepackage[margin=60pt]{geometry}
@@
 
-\usepackage{xcolor}
 \definecolor{RED}{HTML}{EB6231}
@@
 \definecolor{PINK}{HTML}{DA1C5C}
+\pgfplotsset{every axis/.append style={line width=1pt}}
+\pgfplotscreateplotcyclelist{colors}{LIGHTGREEN\\YELLOW\\RED\\GREEN\\BLUE\\}
+\captionsetup{width=0.85\textwidth}
 
@@
 
+\DeclareMathOperator{\E}{\mathbb{E}}
 \newcommand{\der}{\mathrm{d}}
 \newcommand{\e}{\mathrm{e}}
-\newcommand{\dnds}{dN/dS}
 \newcommand{\indice}{l}
@@
 \newcommand{\Ne}{N_{\mathrm{e}}}
-\newcommand{\Nsite}{S}
-\newcommand{\site}{\text{s}}
+\newcommand{\dnds}{\omega}
+\newcommand{\Nsite}{\text{n}}
+\newcommand{\site}{\text{i}}
 % DNA
 
-\newcommand{\ci}{{i}}
-\newcommand{\cj}{{j}}
+\newcommand{\ci}{\mathbb{S}_{t}}
+\newcommand{\cj}{\mathbb{S}_{t}'}
 \newcommand{\itoj}{\ci, \cj}
-\newcommand{\nucitoj}{\ci \rightarrow \cj}
-\newcommand{\submatrix}{Q}
+\newcommand{\setNeighbors}{\mathcal{M}\left(\ci\right)}
+\newcommand{\setNonSynNeighbors}{\mathcal{N}\left(\ci\right)}
+\newcommand{\setSynNeighbors}{\mathcal{S}\left(\ci\right)}
+\newcommand{\submatrix}{q}
+
+\newcommand{\phenoDiv}{S}
+\newcommand{\s}{s}
+\newcommand{\pheno}{x}
+\newcommand{\dpheno}{\delta \pheno}
+\newcommand{\phenoFold}{\Delta G}
+\newcommand{\phenoFoldMin}{\alpha}
+\newcommand{\dphenoFold}{\gamma}
 
@@
 
-\part*{Simulations}
+\title{Substitution rate responses to changes in effective population size.}
 
-\section{Simulation with independent fitness profiles}
+\author{T. Latrille, N. Lartillot}
+
+\maketitle
+
+\abstract{
+The substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) had been theoretically expected to decrease with effective population size ($\Ne$).
+Mechanistically, populations with high $\Ne$ would have stronger purifying selection, hence lower $\dnds$, due to the decrease of random drift.
+Such theoretical prediction had been observed in empirical data across many clades.
+However, several studies failed to observe such response of $\dnds$ to changes in $\Ne$, or with weak strength or direction.
+Computational model of protein folding have shown that $\dnds$ can be independent of $\Ne$, which can mathematical be proven under certain assumptions.
+Moreover, non-equilibrium properties can implies that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
+Together, assumptions about the mapping of sequence to fitness can display a variety of behaviours in the $\dnds$ responses to changes in $\Ne$.
+Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
+We provide a simple mechanistic model that can display a wide range of behaviours depending on the parametrization.
+Our result show that models assuming site-specific fitness landscapes (without epistasis) is at risk of over-estimating the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$, but paradoxically also over-estimating the equilibrium $\dnds$ response to changes in $\Ne$.
+}
+
+\section{Introduction}
+Differences in DNA sequences of a gene in separate species are due to the particular history of DNA substitutions along each specie's respective lineage.
+These substitutions  at the level of the population are the result of point mutations at the level of individuals, subject to evolutionary forces such as selection and random drift.
+In protein-coding DNA sequences, non-synonymous mutations are changing the amino-acid sequence, and by comparing the non-synonymous substitutions rate relative to the synonymous substitution rate (supposedly neutral), a statistic called $\dnds$, we can estimate the global strength of selection and random drift.\\
+
+For a given gene, it has been empirically observed changes of $\dnds$ along lineages of a phylogeny \cite{Yang2001, Zhang2004}.
+More importantly, changes in substitution rate correlate with changes in life-history-traits (longevity, body mass, ...) such as shown by the molecular comparative framework \cite{Lartillot2011,Weber2014}.
+Such covariation are compatible with the theoretical prediction of the nearly-neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
+Indeed, populations with low $\Ne$ would have both a large body size and a high $\dnds$ (weaker selection) due to the increase of random drift.
+If $\dnds$ is empirically correlated to life-history traits, such as body mass and longevity, the direction of correlation could not be replicated across all experiments \cite{Figuet2016}.\\
+
+Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects \cite{Ohta1972, Welch2008}, and also been shown in context of a fixed fitness landscape, where each amino-acid have different fitnesses \cite{Spielman2015a, DosReis2015}.
+However, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly-neutral theory of evolution \cite{Lanfear2014}.
+In the example of a bio-chemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ is independent of $\Ne$ \cite{Goldstein2013}.
+Theoretically, a more general case is obtained whenever the fitness is a concave function of a trait, and such trait is equimutable \cite{Cherry1998}.
+Equimutabibilty meaning a mutation has the same effect on the trait whatever the current trait of the individual.
+Moreover, even in the case of a $\dnds$ response to $\Ne$, this response is far from being $1$-to-$1$ due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$, and subsequentially a decrease in $\dnds$ \cite{Jones2016}.\\
+
+Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
+This question is important if we want to extract changes in effective population size from substitution history, as well as to determine the fitness landscape under which protein-coding DNA sequences are evolving.
+
+\section{Results}
+\subsection{$\bm{\dnds}$ at equilibrium as a function of $\bm{\Ne}$}
+
+We provide a simple toy model, with $2$ parameters that shows a variety of behaviours. 
+This model map the amino-acid sequence to the free energy (kcal/mol) between folded and unfolded state of the whole protein.
+Each amino-acid site in the protein can be in the unstable or stable state.
+A protein is optimal ($\phenoFold = \phenoFoldMin$) if all amino-acids are in the stable state.
+Each amino-acid destabilizing the protein brings an increase of $\phenoFold$ equal to $\dphenoFold$.
+\begin{equation}
+\phenoFold = \phenoFoldMin + \dphenoFold * d, 
+\end{equation}
+where $d$ is the number of unstable amino-acid sites in the sequence, $\alpha$ and $\gamma$ are in kcal/mol.
+Wrightian fitness is defined as the probability of our protein to be in the folded state, given by Boltzmann equation: 
+\begin{equation}
+w(\phenoFold) = \dfrac{1}{1 + e^{\beta \phenoFold }}, 
+\end{equation}
+where $\beta=1.686$ mol/kcal at room temperature.\\
+
+In the original studies \cite{Williams2006, Goldstein2011, Pollock2012}, free energy of the folded state is computed using the $3$D structure of the folded state and pair-wise contact energies between neighboring amino-acid residues using Miya-zawa \& Jernigan contact potentials \cite{Miyazawa1985}.
+The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).\\
+
+The parameters $\alpha$ and $\gamma$ of our toy model are set to match the average values observed in the original model.
+$\alpha=-118$ kcal/mol, as it is the $\phenoFold$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
+$\gamma=1.0$ kcal/mol, as the average observed increase in $\phenoFold$ per destabilizing mutation \cite{Zeldovich2007}.
+
+\begin{figure*}[htb!]
+\begin{mdframed}
+	\centering
+	\includegraphics[width=\textwidth] {artworks/XPopSizeYOmegaEq.pdf}
+	\vspace*{0.2em}
+	\caption{
+		\textbf{$\bm{\dnds}$ response to change in $\bm{\Ne}$}.
+		Panel A.
+Our toy model reproduces the observations of Goldstein \cite{Goldstein2013} that $\dnds$ at equilibrium is independent of $\Ne$.
+Indeed, whenever the parameters $\alpha$ and $\gamma$ are taken to match the average values observed in the original model, we observed no response of $\dnds$ to $\Ne$ (pink and brown lines).
+		However, decreasing both $\alpha$ and $\gamma$ together implies a strong negative relationship between $\Ne$ and $\dnds$, as predicted under the nearly-neutral theory of evolution.
+		Panel B.
+By reducing the number of sites of the protein, thus by effectively reducing the potential number of sites that can compensate for the change in $\Ne$, the $\dnds$ response to changes in $\Ne$ is stronger.
+	}
+	\label{fig:EqdndsNe}
+\end{mdframed}
+\end{figure*}
+
+DNA sequence evolution are simulated under an origin-fixation model \cite{McCandlish2014}, where the rate of substitution of a new mutation with selection coefficient $s$ is:
+\begin{equation}
+p_{\text{fix}}(s) = \mu \dfrac{4 \Ne s}{1 + e^{- 4 \Ne s }}, 
+\end{equation}
+where $\mu$ is the mutation rate.\\
+
+Our toy model reproduces the observations that $\dnds$ at equilibrium is independent of $\Ne$.
+Indeed, whenever the parameters $\alpha$ and $\gamma$ are taken to match the average values observed in the original model, we observed no response of $\dnds$ to $\Ne$.
+However, decreasing both $\alpha$ and $\gamma$ together implies a strong negative relationship between $\Ne$ and $\dnds$, as predicted under the nearly-neutral theory of evolution.
+Moreover, by reducing the number of sites of the protein, thus by effectively reducing the potential number of sites that can compensate for the change in $\Ne$, the $\dnds$ at equilibrium after changes in $\Ne$ is stronger \textit{Necessary mechanistic explanation for this fact} (see Figure \ref{fig:EqdndsNe}).
+
+\subsection{Epistasis determines the time to relaxation}
+
+Although the equilibrium value of $\dnds$ after changes in $\Ne$ is an important feature of the $\dnds$-$\Ne$ relationship, an other parameter that is often neglected is the relaxation time to reach the new equilibrium $\dnds$ \cite{Jones2016}.
+\begin{figure*}[htb!]
+\begin{mdframed}
+	\centering
+	\includegraphics[width=\textwidth]{artworks/SimuStabExponentialDecay.png}
+	\vspace*{0.2em}
+	\caption{
+		\textbf{$\bm{\dnds}$ relaxation time to changes in $\bm{\Ne}$}.
+		Panel A.
+Our result show that with low epistasis, the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
+		Panel B. With high epistasis, the relaxation time is short.
+	}
+	\label{fig:relaxStability}
+\end{mdframed}
+\end{figure*}
+In the case of a fitness landscape with epistasis, after a change in $\Ne$, any site that is used to climb either up or down the fitness landscape will have diminishing return in other site of the sequences.
+Not taking into account epistasis have the consequence of over-estimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$ \ref{fig:relaxStability}).
+
+Using a site-specific fitness landscape, where each amino-acid have different fitnesses, thus with no epistasis, the relaxation time is long since every site has to adapt to the new change in $\Ne$.
+However, using a distribution of fitness effect (DFE) does not suffer from under-estimating the relaxation rate, since the effect is instantaneous (see Figure \ref{fig:relaxProfileDFE}).
+
+\begin{figure*}[htb!]
+\begin{mdframed}
+	\centering
+	\includegraphics[width=\textwidth]{artworks/SimuExponentialDecay.png}
+	\vspace*{0.2em}
+	\caption{
+		\textbf{$\bm{\dnds}$ relaxation time under different fitness models.}.
+		Panel A.
+In context of a fixed fitness landscape, where each amino-acid have different fitnesses (site-specific profile), the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
+		Panel B. In the context of a distribution of fitness effects, the relaxation time is non-existent.
+	}
+	\label{fig:relaxProfileDFE}
+\end{mdframed}
+\end{figure*}
+
+\section{Discussion}
+
+Goldstein \& Pollock argued that model of protein evolution should consider how the acceptance of mutations depends on the sequence context in which they arise (epistasis), as to correctly represent resultant rate heterogeneity along the sequence, to predict the role of compensatory substitutions in protein evolution, predict which of the 10\% of deleterious mutations in humans are harmless in other species, or to accurately represent the rate and time dependence of convergence and homoplasy \cite{Goldstein2017}.
+We argue that not taking into account epistasis also have the consequence of over-estimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$.
+Paradoxically, not taking into account epistasis means over-estimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.\\
+
+If using a distribution of fitness effect (DFE) is mathematically convenient and does not suffer from under-estimating the relaxation rate, it is also over-estimating the response of $\dnds$ to changes in $\Ne$ at equilibrium (\textit{should show proof and mechanistic explanation that the selection coefficient shall not be independent of $\Ne$}).
+As such, developing inference of fitness landscape will require to assume site-interdependent fitness landscapes.\\
+
+Our model does not display site variation in $\dnds$, which is a strong limitation, as it has observed in most protein \cite{Echave2017}.
+\section{Materials \& Methods}
+
+\subsection{Simulation with protein folding probability}
+\label{MatMet:folding}
+From a DNA sequence $\ci$ after $t$ substitutions, the protein's difference in free energy between folded and unfolded state is given by:
+\begin{equation*}
+\phenoFold\left(\ci\right) = \phenoFoldMin + \dphenoFold * d\left(\ci\right), 
+\end{equation*}
+where $d\left(\ci\right)$ is the number of non-optimal amino-acid sites in the sequence.
+Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Boltzmann equation: 
+\begin{equation}
+w(\phenoFold\left(\ci\right)) = \dfrac{P_{\mathrm{F}}\left(\ci\right)}{P_{\mathrm{F}}\left(\ci\right) + P_{\mathrm{U}}\left(\ci\right)} = \dfrac{e^{-\beta G_{\mathrm{F}}\left(\ci\right) }}{e^{-\beta G_{\mathrm{F}} \left(\ci\right) } + e^{-\beta G_{\mathrm{U}}\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \phenoFold\left(\ci\right) }},
+\end{equation}
+where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
+
+From the resident sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
+For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codon that are one mutation away and that are not stop codon.
+For each mutant sequence $\cj \in \setNeighbors$, we compute $\phenoFold^{t+1}$ from the updated sequence $\cj$, and subsequently the selection coefficient of the mutant:
+\begin{equation}
+s \left( \ci,\cj\right) = \dfrac{ w\left( \phenoFold\left(\cj\right) \right) - w\left( \phenoFold\left(\ci\right) \right)}{w\left( \phenoFold\left(\ci\right) \right)}.
+\end{equation}
+The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
+\begin{equation}
+\submatrix_{\itoj} = \mu_{\itoj} \dfrac{4 \Ne s \left( \ci,\cj\right)}{1 - \e^{-4 \Ne s \left( \ci,\cj\right)}}, 
+\end{equation}
+where $\mu_{\itoj}$ is the mutation rate between $\ci$ and $\cj$, determined by the underlying $4x4$ nucleotide mutation rate matrix, and ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
+
+\subsection{Simulation with independent fitness profiles}
+A fitness profile give a fitness for each amino-acid (vector of size $20$).
+Each site of the protein has a specific amino-acid fitness profile.
+Overall, the protein phenotype is computed as the sum of site-specific selection coefficient, obtained by accessing the amino-acid present at each site of the protein.
+The selection coefficient of the mutant $\cj$ is:
+\begin{equation}
+s \left( \ci,\cj\right) = \sum_{1 \leq \site \leq \Nsite} \ln \left( \dfrac{G_{\site} \left(\cj(\site) \right)}{G_{\site} \left(\ci(\site) \right)} \right) ,
+\end{equation}
+where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \cite{Bloom2017}.
 \begin{center}
@@
 
-\section{Simulation in Wright-Fisher context}
-\begin{center}
-	\includegraphics[width=165mm] {artworks/ModelSimuPoly.pdf}
-\end{center}
+The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
 
-\section{Simulation with Fisher geometric landscape}
-\begin{center}
-	\includegraphics[width=165mm] {artworks/ModelSimuGeo.pdf}
-\end{center}
+\subsection{Simulation with distribution of fitness effects}
+The selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
+\begin{equation}
+- s \left( \ci,\cj\right) \sim \text{Gamma} \left( \bar{|s|}, \beta \right)
+\end{equation}
+The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
 
-\section{Simulation with protein folding probability}
-We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites) as in Goldstein \& Pollock (2017).
-From a DNA sequence $\mathbb{S}^t$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}(\mathbb{S}^t)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
+\subsection{$\bm{\dnds}$ along the simulation}
+From the set of mutant $\setNeighbors$ that are one nucleotide away from $\ci$, we define the subsets $\setNonSynNeighbors$ and $\setSynNeighbors$ that are respectively the set of non-synonymous and synonymoys mutants, where  $\setNonSynNeighbors \cup \setSynNeighbors = \setNeighbors$.
+As in previous works \cite{Spielman2015a, DosReis2015, Jones2016}, the ratio of non-synonymous over synonymous substitution rates of the sequence is defined as :
+\begin{align}
+\dnds(t) &= \dfrac{\sum_{\cj \in \setNonSynNeighbors} \submatrix_{\itoj}}{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj}} \left( \dfrac{\sum_{\cj \in \setSynNeighbors} \submatrix_{\itoj}}{\sum_{\cj \in \setSynNeighbors} \mu_{\itoj}} \right)^{-1}\\
+ &= \dfrac{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj} \dfrac{4 \Ne s \left( \ci,\cj\right)}{{1 - \e^{-4 \Ne \left( \ci,\cj\right)} }}}{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj}} 
+\end{align}
+\subsection{Reproducibility}
+The simulators written in C++ are publicly available under MIT license at \url{https://github.com/ThibaultLatrille/SimuEvol}.
+The scripts and instructions necessary to reproduce the experiments are available at \url{https://github.com/ThibaultLatrille/GenotypePhenotypeFitness}.
+
+\bibliographystyle{apalike}
+\bibliography{refs-codons,refs-cds}
+
+\newpage
+\part*{Supplementary materials}
+
+\section*{Probability of folding as in Goldstein \& Pollock}
+We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
+From a DNA sequence $\ci$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}\left(\ci\right)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
 \begin{equation}
-G_{\mathrm{F}}(\mathbb{S}^t) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\mathbb{S}^t(\site), \mathbb{S}^t(r)  \right),
+G_{\mathrm{F}}\left(\ci\right) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\ci(\site), \ci(r) \right),
 \end{equation}
@@
 
-The free energy of unfolded states $G_{\mathrm{U}}(\mathbb{S}^t)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
+The free energy of unfolded states $G_{\mathrm{U}}\left(\ci\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
 \begin{equation}
-G_{\mathrm{U}}(\mathbb{S}^t) = \langle G(\mathbb{S}^t) \rangle - kT \log (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G(\mathbb{S}^t)^2 \rangle - \langle G(\mathbb{S}^t) \rangle^2\right] }{kT}
+G_{\mathrm{U}}\left(\ci\right) = \langle G\left(\ci\right) \rangle - kT \ln (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G\left(\ci\right)^2 \rangle - \langle G\left(\ci\right) \rangle^2\right] }{kT}
 \end{equation}
@@
 \begin{equation}
-\Delta G^t = G_{\mathrm{F}}(\mathbb{S}^t) - G_{\mathrm{U}}(\mathbb{S}^t)
+\phenoFold\left(\ci\right) = G_{\mathrm{F}}\left(\ci\right) - G_{\mathrm{U}}\left(\ci\right)
 \end{equation}
 
-And the fitness is defined as the probability of our protein to be in the folded state: 
-\begin{equation}
-f(\Delta G^{t}) = \dfrac{P_{\mathrm{F}}}{P_{\mathrm{F}} + P_{\mathrm{U}}} = \dfrac{e^{-\beta G_{\mathrm{F}}(\mathbb{S}^t) }}{e^{-\beta G_{\mathrm{F}} (\mathbb{S}^t) } + e^{-\beta G_{\mathrm{U}}(\mathbb{S}^t) }} = \dfrac{1}{1 + e^{\beta \Delta G(\mathbb{S}^t) }},
-\end{equation}
-where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
 \begin{center}
@@
 \end{center}
-For each possible mutant ($t+1$ substitutions), we compute $\Delta G^{t+1}$ from the updated sequence $\mathbb{S}^{t+1}$, and subsequently the selection coefficient of the mutant:
-\begin{equation}
-s \left( \Delta G^{t}, \Delta G^{t+1} \right) = \dfrac{f(\Delta G^{t+1}) - f(\Delta G^{t})}{f(\Delta G^{t})}.
-\end{equation}
-The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
-\begin{equation}
-{\submatrix_{\itoj}} = \mu_{\itoj} \dfrac{4 \Ne s \left( \Delta G^{t}, \Delta G^{t+1} \right)}{{1 - \e^{-4 \Ne s \left( \Delta G^{t}, \Delta G^{t+1} \right)} }}, 
-\end{equation}
-where ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
+
+\section*{Equimutability}
+We define $\pheno$ as the current phenotype and $\dpheno$ as the phenotypic effect of a mutation.
+$\s \left( \pheno, \dpheno \right)$ is the selection coefficient of a phenotypic effect $\dpheno$ if the current phenotype is $\pheno$:
+\begin{equation*}
+\s \left( \pheno, \dpheno \right) = \dfrac{w \left( \pheno + \dpheno \right) - w \left( \pheno \right)}{w \left( \pheno \right)} \simeq w'\left( \pheno \right) \dpheno, 
+\end{equation*}
+where $w \left( \pheno \right)$ is the Wrightian fitness of phenotype $x$.
+We also define $\rho_{\pheno}(\dpheno)$ as the distribution of phenotypic effects, at phenotype $\pheno$.
+At equilibrium, the expected selection coefficient $ 4 \Ne \s \left( \pheno, \dpheno \right)$ of mutations that reached fixation must be $0$:
+\begin{align*}
+0 & = \E_{\dpheno} \left[ 4 \Ne \s\left( \pheno, \dpheno \right) \dfrac{ 4 \Ne \s\left( \pheno, \dpheno \right) }{1 - \e^{4\Ne\s\left( \pheno, \dpheno \right)}} \right] \\ & 
+\simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \pheno \right)  \right]^2}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno} }  \dpheno^2  \rho_{\pheno}(\dpheno) \der \dpheno 
+\end{align*}
+And the substitution rate is :
+\begin{align*}
+\omega(\Ne) & = \E_{\dpheno} \left[ \dfrac{ 4 \Ne \s \left( \pheno, \dpheno \right)}{1 - \e^{4\Ne \s \left( \pheno, \dpheno \right)}} \right] \simeq \E_{\dpheno} \left[ \dfrac{ 4 \Ne f'\left( \pheno \right) \dpheno}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno}} \right] \\
+& \simeq \int_{-\infty}^{\infty} \dfrac{4 \Ne f'\left( \pheno \right) \dpheno}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno } } \rho_{\pheno}(\dpheno) \der \dpheno \\
+\end{align*}
+If $\rho_{\pheno}(\dpheno) = \rho(\dpheno)$ is independent of $\pheno$:
+\begin{align*}
+0 & \simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \pheno \right)  \right]^2}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno} } \dpheno^2 \rho(\dpheno)  \der \dpheno \\
+\Longrightarrow & \ 4 \Ne f'\left( \pheno \right) = K \in \mathbb{R}^{*} \text{ and } \s\left( \pheno, \dpheno \right) \simeq \dfrac{K \dpheno}{4\Ne} \\
+\Longrightarrow & \ \omega( \Ne ) \simeq \int_{-\infty}^{\infty} \dfrac{ K  }{1 - \e^{K \dpheno } } \dpheno \rho(\dpheno)  \der \dpheno  \text{ is independent of } \Ne
+\end{align*}
 

### 2020-03-04 scripts/heatmap_analysis.py

@@ -5,6 +5,7 @@ import numpy as np
 import pandas as pd
 import itertools
 import matplotlib
+
 label_size = 16
 my_dpi = 96
 matplotlib.rcParams['ytick.labelsize'] = label_size
@@ -36,6 +37,9 @@ if __name__ == '__main__':
     for param, x_y_z in mean_values.items():
         name = param.replace("/", "").replace("|", "")
         x_axis = sorted(set([k[0] for k in x_y_z.keys()]))
+        if len(x_axis) < 2:
+            continue
+
         y_axis = sorted(set([k[1] for k in x_y_z.keys()]))
         y_array, x_array = np.meshgrid(y_axis, x_axis)
         z_mean_array = np.zeros((len(x_axis), len(y_axis)))
@@ -45,7 +49,10 @@ if __name__ == '__main__':
             z_std_array[i, j] = 1.96 * std_values[param][(x, y)]
 
         for (j, y) in enumerate(y_axis):
-            plt.plot(x_axis, z_mean_array[:, j], label="{0}={1:.3g}".format(args.y_param, y), linewidth=3)
+            if args.y_param != "":
+                plt.plot(x_axis, z_mean_array[:, j], label="{0}={1:.3g}".format(args.y_param, y), linewidth=3)
+            else:
+                plt.plot(x_axis, z_mean_array[:, j], linewidth=3)
             plt.fill_between(x_axis, z_mean_array[:, j] - z_std_array[:, j], z_mean_array[:, j] + z_std_array[:, j],
                              alpha=0.3)
         plt.xlabel(args.x_param, fontsize=label_size)

### 2020-03-04 scripts/scaling_merge.py

@@ -16,13 +16,14 @@ import matplotlib.pyplot as plt
 
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-f', '--fitting', required=False, default=False, type=bool, dest="fitting")
     parser.add_argument('-o', '--output', required=True, type=str, dest="output")
     parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
     args = parser.parse_args()
     dict_df = {filepath: {k: v[0] for k, v in pd.read_csv(filepath + '.tsv', sep='\t').items()} for filepath in
                args.input if os.path.isfile(filepath + '.tsv')}
 
-    cols = ["<dN>/<dN0>", "<dN/dN0>"]
+    cols = ["<dN/dN0>"]
     fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
     for col in cols:
         max_y = 0.0
@@ -38,15 +39,18 @@ if __name__ == '__main__':
             for node_name in set(df["NodeName"]):
                 df_filt = df[df['NodeName'] == node_name]
                 t = df_filt["AbsoluteStartTime"].values
-                y = df_filt[col].values
                 plt.axvline(x=t[0], linewidth=3, color='black')
+                if not args.fitting:
+                    continue
 
                 def exponential(x, r, a, b):
-                    return a * np.exp(r * (x - t[0])) + b
+                    return a * np.exp(-r * (x - t[0])) + b
+
+                y = df_filt[col].values
 
                 p_0 = [np.log(2) * 2.0 / (t[-1] - t[0]), y[0] - y[-1], y[-1]]
-                [rate, gap, final], pcov = curve_fit(exponential, t, df_filt[col], p0=p_0, maxfev=10000,
-                                                     bounds=([0, -np.inf, -np.inf], [np.inf, np.inf, np.inf]))
+                [rate, gap, final], pcov = curve_fit(exponential, t, df_filt[col], p0=p_0, maxfev=50000,
+                                                     bounds=([0., -np.inf, 0.], [np.inf, np.inf, np.inf]))
 
                 plt.plot(t, [exponential(x, rate, gap, final) for x in t], linewidth=3, color=base_line.get_color())
 
@@ -57,6 +61,7 @@ if __name__ == '__main__':
                 last_final = final
 
[…]

### 2020-03-04 scripts/theoretical_equilibrium.py

@@ -1,0 +1,86 @@
+#!python3
+import argparse
+import pandas as pd
+import numpy as np
+from scipy.optimize import brentq
+import matplotlib
+
+label_size = 24
+my_dpi = 96
+matplotlib.rcParams['ytick.labelsize'] = label_size
+matplotlib.rcParams['xtick.labelsize'] = label_size
+matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
+    parser.add_argument('-g', '--exon_size', required=True, type=int, dest="n")
+    parser.add_argument('-p', '--population_size', required=True, type=float, dest="population_size")
+    parser.add_argument('-a', '--alpha', required=True, type=float, dest="alpha")
+    parser.add_argument('-k', '--kappa', required=True, type=float, dest="kappa")
+    parser.add_argument('-b', '--beta', required=True, type=float, dest="beta")
+    args, unknown = parser.parse_known_args()
+    dict_df = dict()
+
+
+    def delta_g(x, alpha):
+        return alpha + args.kappa * x
+
+
+    def sel_coeff(x, alpha):
+        edg = np.exp(args.beta * delta_g(x, alpha))
+        return args.kappa * args.beta * edg / ((1 + edg) * args.n)
+
+
+    def scaled_sel_coeff(x, alpha):
+        return 4 * args.population_size * sel_coeff(x, alpha)
+
+
+    def mut_bias(x):
+        if x == 0.:
+            return float("inf")
+        elif x == 1.0:
+            return -float("inf")
+        return np.log((1 - x) / x)
+
+
+    def self_consistent_eq(x, alpha):
+        return mut_bias(x) - scaled_sel_coeff(x, alpha)
+
+
+    x_eq = brentq(lambda x: self_consistent_eq(x, args.alpha), 0.0, 1.0, full_output=True)[0]
+    assert (x_eq <= 0.5)
+    s = sel_coeff(x_eq, args.alpha)
+    S = 4 * args.population_size * s
+    assert ((S - mut_bias(x_eq)) < 1e-5)
[…]

### 2020-03-05 manuscript/supp-mat.tex
Added Scaling experiments, define function in supp. mat.

@@
 \begin{align}
-\frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} & = \frac{ 2 \Ne 2 s(\x, \dx)}{1 - \e^{-S(\x, \dx)}},\\
+\frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} & = \frac{ 2 \Ne 2 s(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \text{ from eq. \ref{eq:pfix} and \ref{eq:pfix0}},\\
 & = \frac{S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \label{eq:pfixratio}.
@@
  0 = \E_{\dx} \left[ s(\x\eq, \dx) p_{\text{fix}}(\x\eq, \dx) \right], \\
-\iff 0 = \E_{\dx} \left[ s(\x\eq, \dx)\frac{ 2 s(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}} \right], \\
-\iff 0 = s(\x\eq, \dx)\frac{ 2s(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) + s(\x\eq, 0) \frac{ \rho_{\x\eq}(0)}{2\Ne} + s(\x\eq, -\dx)\frac{ 2s(\x\eq, -\dx)}{1 - \e^{-S(\x\eq, -\dx)}}   \rho_{\x\eq}(-\dx), \\
-\Longrightarrow s(\x\eq, \dx) \frac{ 2s(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) \simeq s(\x\eq, \dx) \frac{ - 2s(\x\eq, \dx)}{1 - \e^{S(\x\eq, \dx)}}   \rho_{\x\eq}(-\dx) \text{ from eq.
+\iff 0 = \frac{ 2 s(\x\eq, \dx)^2}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) + s(\x\eq, 0) \frac{ \rho_{\x\eq}(0)}{2\Ne} + \frac{ 2s(\x\eq, -\dx)^2}{1 - \e^{-S(\x\eq, -\dx)}} \rho_{\x\eq}(-\dx)\text{ from eq.
+\ref{eq:pfix} and \ref{eq:pfix0}}, \\
+\Longrightarrow \frac{ 2s(\x\eq, \dx)^2}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) \simeq \frac{ - 2s(\x\eq, \dx)^2}{1 - \e^{S(\x\eq, \dx)}}   \rho_{\x\eq}(-\dx) \text{ from eq.
 \ref{eq:S_minus_deltax}}, \\
@@
 \iff \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq -S(\x\eq, \dx) \text{ from eq. \ref{eq:proba}} \label{eq:equilibrium}, \\
-\iff \x\eq \simeq \frac{1}{1 + \frac{\e^{-S(\x\eq, \dx)}}{\Nstate-1}}.
+\iff \lambda_{K}(\x\eq) \simeq -S(\x\eq, \dx) \label{eq:equilibrium_lambda},
 \end{gather}
-
-\section*{Substitution rate ($\bm{\dnds}$)}
+where  $\lambda_{K}(\x\eq) = \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1)$.
+\section*{Substitution rate ($\bm{\dnds}$) at equilibrium}
 The substitution rate of selected mutations relative to neutral mutations  ($\dnds$) is by definition:
@@
 \dnds & = \E_{\dx} \left[ \frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} \right], \\
- & = \E_{\dx} \left[ \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \right], \\
  & = (1 - \x) \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} + \x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) + \frac{\x}{\Nstate-1} \frac{ S(\x, -\dx)}{1 - \e^{-S(\x, -\dx)}} \text{ from eq.
-\ref{eq:proba}}, \\
+\ref{eq:proba}, \ref{eq:pfix} and \ref{eq:pfix0}}, \\
  & = (1 - \x) \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} - \frac{\x}{\Nstate-1}  \frac{ S(\x, \dx)}{1 - \e^{S(\x, \dx)}} +  \x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) \text{ from eq.
@@
 \dnds\eq &= (1 - \x\eq) \frac{ S(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}} - \frac{\x\eq}{\Nstate-1} \frac{ S(\x\eq, \dx)}{1 - \e^{S(\x\eq, \dx)}} + \x\eq \left(\frac{\Nstate - 2}{\Nstate - 1}\right) , \\
- &= \x\eq \left[ \frac{2 (\x\eq-1)  \left(\ln (\Nstate-1)+\ln \left(\frac{1 -\x\eq}{\x\eq}\right)\right)}{\Nstate (\x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right] \text{ from eq.
-\ref{eq:equilibrium}}.
+ &= \x\eq \left[ \frac{2 (\x\eq-1)  \lambda_{K}(\x\eq) }{\Nstate (\x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right] \text{ from eq. \ref{eq:equilibrium}}.
 \end{align}
 \begin{center}
-	\begin{tikzpicture}
+	\begin{tikzpicture}[
+	declare function={
+		omega(\x,\k)= \x*(2*(\x-1)*(ln(\k-1)+ln((1-\x)/\x))/(\k*(\x-1)+1)+ (\k-2)/(\k-1));
+	},]
 	\begin{axis}[
@@
 	xlabel={$\x\eq$},
-	domain=0:1.1,
-	ymin=0.0, ymax=1.05,
-	samples=50,
-	legend entries={$\Nstate=2$, $\Nstate=3$, $\Nstate=4$},
+	domain=0:1.0,
+	ymin=0.0, ymax=1.1,
+	samples=100,
+	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
 	legend cell align=left,
@@
 	]
-	\addplot[line width=2.0pt, color=BLUE]{ (ln(2-1) + ln((1-x) / x)) * x * 2 * (x - 1) / (2 * (x - 1) + 1) + x * (2-2) / (2-1) };
-	\addplot[line width=2.0pt, color=YELLOW]{ (ln(3-1) + ln((1-x) / x)) * x * 2 * (x - 1) / (3 * (x - 1) + 1) + x * (3-2) / (3-1)};
-	\addplot[line width=2.0pt, color=GREEN]{ (ln(4-1) + ln((1-x) / x)) * x * 2 * (x - 1) / (4 * (x - 1) + 1) + x * (4-2) / (4-1)};
+	\addplot[line width=2.0pt, color=BLUE]{omega(x, 2)};
+	\addplot[line width=2.0pt, color=YELLOW]{omega(x, 4)};
+	\addplot[line width=2.0pt, color=GREEN]{omega(x, 8)};
+	\addplot[line width=2.0pt, color=RED]{omega(x, 16)};
 	\end{axis}
@@
 \begin{gather}
-\frac{\der \dnds\eq}{\der \x\eq} = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \left[ \ln (\Nstate-1) + \ln \left(\frac{1 - \x\eq}{\x\eq} \right) \right]}{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1} \label{eq:dw_dx}.
+\frac{\der \dnds\eq}{\der \x\eq} = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1} \label{eq:dw_dx}.
 \end{gather}
-
 \begin{center}
-	\begin{tikzpicture}
+	\begin{tikzpicture}[
+	declare function={
+		domega(\x,\k)= 2*(\k*(\x-1)+1 + (\k*(\x-1)^2+2*\x-1)*(ln(\k-1) + ln((1-\x)/\x)))/(\k*(\x - 1) + 1)^2 + (\k-2) / (\k-1);
+	},]
 	\begin{axis}[
@@
 	samples=50,
-	legend entries={$\Nstate=2$, $\Nstate=3$, $\Nstate=4$},
+	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
 	legend cell align=left,
@@
 	]
-	\addplot[line width=2.0pt, color=BLUE]{2*(2*(x-1)+1 + (2*(x-1)^2+2*x-1)*(ln(2-1) + ln((1-x)/x)))/(2*(x - 1) + 1)^2 + (2-2) / (2-1) };
-	\addplot[line width=2.0pt, color=YELLOW]{2*(3*(x-1)+1 + (3*(x-1)^2+2*x-1)*(ln(3-1) + ln((1-x)/x)))/(3*(x - 1) + 1)^2 + (3-2) / (3-1) };
-	\addplot[line width=2.0pt, color=GREEN]{2*(4*(x-1)+1 + (4*(x-1)^2+2*x-1)*(ln(4-1) + ln((1-x)/x)))/(4*(x - 1) + 1)^2 + (4-2) / (4-1)};
-	\addplot[dashed,line width=0.5pt, color=black]{0};
+	\addplot[line width=2.0pt, color=BLUE]{domega(x, 2)};
+	\addplot[line width=2.0pt, color=YELLOW]{domega(x, 4)};
+	\addplot[line width=2.0pt, color=GREEN]{domega(x, 8)};
+	\addplot[line width=2.0pt, color=RED]{domega(x, 16)};
 	\end{axis}
@@
 
-\section*{Substitution rate ($\bm{\dnds}$) for the special case \Nstate=2}
-In the special of $2$ states ($1$ stable and $1$ unstable), $\dnds\eq$ at equilibrium is:
-\begin{align}
-\dnds\eq & = \ln \left( \frac{1 - \x\eq}{\x\eq} \right) \frac{2 (\x\eq -1) \x\eq }{2 \x\eq - 1},\\
-& = \begin{cases}
- - 2 \x\eq \ln (\x\eq) + \mathcal{O}({\x\eq}^2) \text{ as } \x\eq \rightarrow 0, \\
- 1 - \frac{8}{3} (\x\eq - 0.5)^2 + \mathcal{O}((\x\eq - 0.5)^3) \text{ as } \x\eq \rightarrow 0.5. \\
-\end{cases}
-\end{align}
-\begin{center}
-	\begin{tikzpicture}
-	\begin{axis}[
-	ylabel={$\dnds\eq$},
-	xlabel={$\x\eq$},
-	domain=0:0.52,
-	ymin=0.0, ymax=1.05,
-	samples=100,
-	legend entries={
-		$\ln( \frac{1 - \x\eq}{\x\eq}) \frac{2 (\x\eq -1) \x\eq }{2 \x\eq - 1}$,
-		$- 2 \x\eq \ln (\x\eq)$,
-		$ 1 - \frac{8}{3} (\x\eq - 0.5)^2$},
-	legend cell align=left,
-	minor tick num=2,
-	axis x line=bottom,
-	axis y line=left,
-	legend style={at={(0.5,0.34)},anchor=north west}
-	]
-	\addplot[line width=2.0pt, color=BLUE]{ ln((1-x) / x) * x * 2 * (x - 1) / (2 * x - 1)};
-	\addplot[dashed,line width=2.0pt, color=YELLOW]{ -2 * x * ln(x) };
-	\addplot[dashed,line width=2.0pt, color=GREEN]{ 1 - 8 * (x - 0.5)^2 / 3 };
-	\end{axis}
-	\end{tikzpicture}
-\end{center}
-And the derivative of $\dnds\eq$ w.r.t to $\x\eq$ is: 
-\begin{align}
-\frac{\der \dnds\eq}{\der \x\eq} &= \frac{\ln \left( \frac{1 - \x\eq}{\x\eq} \right)(2 - 4 \x\eq + 4 {\x\eq}^2) + 4\x\eq - 2}{(2\x\eq -1)^2} \label{eq:dw_dx_Nstate2}, \\
-&= 
-\begin{cases}
- - (\x\eq + 1)(2\ln(\x\eq) + 2) - 2\x\eq + \mathcal{O}({\x\eq}^2) \text{ as } \x\eq \rightarrow 0, \\
- \frac{8 - 16 \x\eq}{3} + \mathcal{O}((\x\eq - 0.5)^2) \text{ as } \x\eq \rightarrow 0.5.
-\end{cases} \label{eq:dw_dx_approx_Nstate2}
-\end{align}
-
-\begin{center}
-	\begin{tikzpicture}
-	\begin{axis}[
-	ylabel={$\frac{\der \dnds(\x\eq)}{\der \x\eq}$},
-	xlabel={$\x\eq$},
-	domain=0.0:0.51,
-	ymin=0.0, ymax=6,
-	samples=30,
-	legend entries={
-		$\frac{\ln ( \frac{1 - \x\eq}{\x\eq})(2 - 4 \x\eq + 4 {\x\eq}^2) + 4\x\eq - 2}{(2\x\eq -1)^2}$,
-		$- (\x\eq + 1)(2\ln(\x\eq) + 2) - 2\x\eq$,
-		$\frac{8 - 16 \x\eq}{3}$},
-	legend cell align=left,
-	minor tick num=2,
-	axis x line=bottom,
-	axis y line=left,
-	legend style={at={(0.25,0.95)},anchor=north west}
-	]
-	\addplot[line width=2.0pt, color=BLUE]{ (-2 + 4 * x + (2 - 4 * x + 4 * x^2) * ln((1-x) / x))/(1 - 2 * x)^2 };
-	\addplot[dashed,line width=2.0pt, color=YELLOW]{ -(x+1)*(2*ln(x) + 2) -2*x };
-	\addplot[dashed,line width=2.0pt, color=GREEN]{ 8 * ( 1 - 2*x)/3};
-	\end{axis}
-	\end{tikzpicture}
-\end{center}
-
-\section*{Change of equilibrium phenotype ($\bm{\x\eq}$) after a change in $\bm{\Ne}$}
-Given the equilibrium equation (eq.
-\ref{eq:equilibrium}), $\x\eq$ is implicitly a function of $\Ne$, and the change of phenotype at equilibrium ($\der \x\eq$) after a change of effective population size ($\der \Ne$) can be obtain as:
+\section*{Response of equilibrium after a change in $\bm{\Ne}$}
+Define the function $G(\x, \Ne )$ as:
 \begin{gather}
-G(\x, \Ne ) \equiv \ln \left( \frac{1 - \x}{\x} \right) + \ln (\Nstate-1) + 4 \Ne s(\x, \dx), \\
-\iff G(\x\eq(\Ne), \Ne ) = 0 \text{ from eq. \ref{eq:equilibrium}}, \\
+G(\x, \Ne ) \equiv \ln \left( \frac{1 - \x}{\x} \right) + \ln (\Nstate-1) + 4 \Ne s(\x, \dx), 
+\end{gather}
+The equilibrium equation (eq. \ref{eq:equilibrium}) states that $G(\x\eq, \Ne )=0$, meaning that $\x\eq$ is implicitly a function of $\Ne$:
+\begin{gather}
+G(\x\eq(\Ne), \Ne ) = 0, \\
 \Longrightarrow \frac{\partial G(\x\eq, \Ne ) }{\partial \x\eq }\frac{ \der \x\eq}{\der \Ne} + \frac{ \partial G(\x\eq, \Ne )}{\partial \Ne} = 0, \\
@@
 \end{gather}
-\section*{Change of equilibrium substitution rate ($\bm{\dnds\eq}$) after a change in $\bm{\Ne}$}
-Together, the change of substitution rate at equilibrium ($\der \dnds\eq$) after a change of effective population size ($\der \Ne$) can be obtain as:
+Giving the equation for the response of phenotype at equilibrium after a change of effective population size.
+Together, the response of substitution rate at equilibrium, after a change of effective population size can be obtain as:
 \begin{align}
@@
 \begin{gather}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) = 4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x\eq)}}{1 + \e^{\beta(\alpha + \kappa \x\eq)}}  \dx   \text{ from eq.
-\ref{eq:equilibrium} and \ref{eq:df_dx}}.
+\lambda_{K}(\x\eq) = 4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x\eq)}}{1 + \e^{\beta(\alpha + \kappa \x\eq)}}  \dx   \text{ from eq. \ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
 \end{gather}
@@
 \begin{center}
-\begin{tikzpicture}
+\begin{tikzpicture}[
+	declare function={
+		lambda(\x,\k)= ln(\k-1) + ln((1-\x)/\x);
+	},]
 	\begin{axis}[
@@
 	legend entries={
-		$\ln(\frac{1 - \x}{\x}) + \ln(\Nstate-1) \text{, }\Nstate=2$,
-		$\ln ( \frac{1 - \x}{\x})+ \ln(\Nstate-1) \text{, }\Nstate=4$,
-		$\ln ( \frac{1 - \x}{\x})+ \ln(\Nstate-1) \text{, }\Nstate=8$,
-		$\ln ( \frac{1 - \x}{\x})+ \ln(\Nstate-1) \text{, }\Nstate=16$,
-		$4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \dx$},
+		$4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \dx$,
+		$\lambda_{2}(\x)$,
+		$\lambda_{4}(\x)$,
+		$\lambda_{8}(\x)$,
+		$\lambda_{16}(\x)$},
 	legend cell align=left,
@@
 	]
-	\addplot[domain=0.15:0.5,line width=2.0pt, color=RED]{ ln((1-x)/x) };
-	\addplot[domain=0.15:0.68,line width=2.0pt, color=YELLOW]{ ln((1-x)/x) + ln(4-2) };
-	\addplot[domain=0.15:0.8,line width=2.0pt, color=GREEN]{ ln((1-x)/x) + ln(8-2) };
-	\addplot[domain=0.15:0.8,line width=2.0pt, color=LIGHTGREEN]{ ln((1-x)/x) + ln(16-2) };
-	\addplot[domain=0.15:0.38, line width=2.0pt, color=BLUE]{ 4 * 1000 * 1.686 *exp(1.686 * (-118 + 300 *x))};
+	\addplot[domain=0.0:0.38, line width=2.0pt, color=BLUE]{ 4 * 1000 * 1.686 *exp(1.686 * (-118 + 300 *x))};
+	\addplot[domain=0.0:0.5,line width=2.0pt, color=YELLOW]{ lambda(x,2) };
+	\addplot[domain=0.0:0.8,line width=2.0pt, color=GREEN]{ lambda(x,4) };
+	\addplot[domain=0.0:0.9,line width=2.0pt, color=RED]{ lambda(x,8) };
+	\addplot[domain=0.0:1.0,line width=2.0pt, color=LIGHTGREEN]{ lambda(x,16) };
 	\end{axis}
@@
 \begin{gather}
--S(\x\eq, \dx) \simeq 1, \\
-\iff s(\x\eq, \dx) \simeq -\frac{1}{4 \Ne}, \\
-\iff \frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{1}{4 \Ne \dx}, \\
-\iff \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \simeq \frac{1}{4 \Ne \dx}, \\
-\iff \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \simeq \frac{1}{4 \Ne \dx \beta \kappa}, \\
-\Longrightarrow \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \ll 1, \\
-\Longrightarrow \e^{\beta(\alpha + \kappa \x)} \ll 1, \\
-\Longrightarrow \frac{\e^{\beta(\alpha + \kappa \x\eq)}}{1 + \e^{\beta(\alpha + \kappa \x\eq)}} \simeq \e^{\beta(\alpha + \kappa \x\eq)}.
+-S(\x\eq, \dx) \simeq \lambda_{\Nstate}(\x\eq)\text{ from eq. \ref{eq:equilibrium_lambda}}, \\
+\iff s(\x\eq, \dx) \simeq -\frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne}, \\
+\iff \frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:s_from_fitness}}, \label{eq:approx_df_dx}\\
+\iff \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)} } \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:df_dx}}, \\
+\iff \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx \beta \kappa}, \\
+\Longrightarrow \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \ll 1\text{ given that $\Ne \gg 1$ and $\kappa \dx \simeq 1$}, \\
+\Longrightarrow \e^{\beta(\alpha + \kappa \x)} \ll 1. \label{eq:approx_exp}
 \end{gather}
@@
 \begin{gather}
-\frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{1}{4 \Ne \dx}, \\
-\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{1}{4 \Ne \dx \beta \kappa }, \\
-\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{(1 - \x\eq)\x\eq }{\beta \kappa} \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
-\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \dx \frac{(1 - \x\eq)\x\eq }{\beta } \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
-\Longrightarrow \left| \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \right| \gg \left| \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} \right|. \\
+\frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:approx_df_dx}}, \\
+\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)\left( 1 + \e^{\beta(\alpha + \kappa \x)} \right)}{4 \Ne \dx \beta \kappa }\text{ from eq.
+\ref{eq:df2_dx2}}, \\
+\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx \beta \kappa }\text{ from eq. \ref{eq:approx_exp}}, \\
+\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \lambda_{\Nstate}(\x\eq) \frac{(1 - \x\eq)\x\eq }{\beta \kappa} \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
+\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \lambda_{\Nstate}(\x\eq) \dx \frac{(1 - \x\eq)\x\eq }{\beta } \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}\text{ given that $\kappa \dx \simeq 1$}, \\
+\Longrightarrow \left| \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \right| \gg \left| \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} \right|\text{ given that $\dx \ll 1$}. \label{eq:approx_df2_df} \\
 \end{gather}
-Together, these approximations leads to:
+Together, these approximations leads to the following response in equilibrium phenotype at change in $\Ne$ as:
 \begin{align}
-\frac{ \der \x\eq}{\der \ln (\Ne)} & = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
-& \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
-& \simeq - \frac{1 + \e^{\beta(\alpha + \kappa \x\eq)}}{\beta \kappa}, \\
-& \simeq - \frac{1}{\beta \kappa}.
+\frac{ \der \x\eq}{\der \ln (\Ne)} & = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\text{ from eq.
+\ref{eq:dw_dlnNe}}, \\
+& \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\text{ from eq. \ref{eq:approx_df2_df}}, \\
+& \simeq - \frac{1 + \e^{\beta(\alpha + \kappa \x\eq)}}{\beta \kappa}\text{ from eq. \ref{eq:df2_dx2}}, \\
+& \simeq - \frac{1}{\beta \kappa}\text{ from eq. \ref{eq:approx_exp}}. \label{eq:approx_dx_dlnNe}
+\end{align}
+Moreover, given that the number of state if large enough $\Nstate \gg 1$, the response in equilibrium $\dnds$ due to change in phenotype can be approximated as:
+\begin{align}
+\frac{ \der \dnds\eq}{\der \x\eq}  & = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}\text{ from eq.
+\ref{eq:approx_exp}}, \\
+& \simeq \frac{2 \lambda_{K}(\x\eq)}{\Nstate} + 1, \\
+& \simeq 1. \label{eq:approx_domega_dx}
 \end{align}
@@
 \begin{align}
-\frac{ \der \dnds\eq}{\der \ln (\Ne)}  & \simeq - \frac{ \der \dnds\eq}{\der \x\eq} \frac{1}{\beta \kappa}, \\
-  & = - \frac{1}{\beta \kappa}.
+\frac{ \der \dnds\eq}{\der \ln (\Ne)}  & = \frac{ \der \dnds\eq}{\der \x\eq}  \frac{ \der \x\eq}{\der \ln (\Ne)}, \\
+  & \simeq - \frac{1}{\beta \kappa} \text{from eq. \ref{eq:approx_dx_dlnNe} and \ref{eq:approx_domega_dx}}.
 \end{align}

### 2020-03-05 Scaling/simulated_experiment.py

@@ -1,0 +1,50 @@
+#!python3
+import argparse
+import os
+from subprocess import run
+
+
+def create_experiment(experiment, screen, sbatch, nbr_cpu):
+    exp_path = os.getcwd() + '/Experiments/' + experiment
+
+    os.makedirs(exp_path, exist_ok=True)
+    os.system('cp config.yaml {0}'.format(exp_path))
+    os.remove(exp_path + "/Snakefile") if os.path.exists(exp_path + "/Snakefile") else None
+    os.symlink(os.getcwd() + "/Snakefile", exp_path + "/Snakefile")
+    run_file = exp_path + "/snakeslurm.sh"
+    with open(run_file, 'w') as w:
+        w.write("#!/usr/bin/env bash\n")
+        run_str = 'snakemake '
+        if sbatch:
+            run_str += '-j 99 --cluster "sbatch -J {0} -p long -N 1 ' \
+                       '-o {1}/slurm.%x.%j.out -e {1}/slurm.%x.%j.err '.format(experiment, exp_path)
+            run_str += '--cpus-per-task={params.threads} --mem={params.mem} -t {params.time}"\n'
+        else:
+            run_str += "-j {0} --printshellcmds".format(nbr_cpu)
+        w.write(run_str)
+    os.system("chmod 755 " + run_file)
+    cmd = 'cd ' + exp_path + ' && ./snakeslurm.sh'
+    screen_cmd = 'screen -dmS ' + experiment + ' bash -c "' + cmd + '"'
+    with open(exp_path + "/screen.sh", 'w') as w:
+        w.write("#!/usr/bin/env bash\n")
+        w.write(screen_cmd)
+    if screen:
+        print(screen_cmd)
+        run(screen_cmd, shell=True)
+    else:
+        print(cmd)
+        run(cmd, shell=True)
+
+
+if __name__ == '__main__':
+    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('-e', '--experiment', required=False, type=str, default="selcoeff", dest="experiment")
+    parser.add_argument('-s', '--screen', required=False, type=bool, default=False, dest="screen")
+    parser.add_argument('-b', '--sbatch', required=False, type=bool, default=False, dest="sbatch")
+    parser.add_argument('-j', '--nbr_cpu', required=False, type=int, default=6, dest="nbr_cpu")
+    args = parser.parse_args()
[…]

### 2020-03-31 manuscript/manuscript.tex
Moved the notations into new-commands.tex

@@
 \pdfinclusioncopyfonts=1
+\captionsetup{width=0.85\textwidth}
 
-\definecolor{RED}{HTML}{EB6231}
-\definecolor{YELLOW}{HTML}{E29D26}
-\definecolor{BLUE}{HTML}{5D80B4}
-\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
-\definecolor{GREEN}{HTML}{8FB03E}
-\definecolor{PURPLE}{HTML}{BE1E2D}
-\definecolor{BROWN}{HTML}{A97C50}
-\definecolor{PINK}{HTML}{DA1C5C}
+\include{new-commands}
 \pgfplotsset{every axis/.append style={line width=1pt}}
 \pgfplotscreateplotcyclelist{colors}{LIGHTGREEN\\YELLOW\\RED\\GREEN\\BLUE\\}
-\captionsetup{width=0.85\textwidth}
-
-\newcommand{\specialcell}[2][c]{%
-	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-\DeclareMathOperator{\E}{\mathbb{E}}
-\newcommand{\der}{\mathrm{d}}
-\newcommand{\e}{\mathrm{e}}
-\newcommand{\indice}{l}
-\newcommand{\indiceexp}{^{(\indice)}}
-\newcommand{\angstrom}{\text{\normalfont\AA}}
-
-% Time, effective population size and mutation rate.
-\newcommand{\Ne}{N_{\mathrm{e}}}
-\newcommand{\dnds}{\omega}
-\newcommand{\Nsite}{\text{n}}
-\newcommand{\site}{\text{i}}
-% DNA
-
-\newcommand{\ci}{\mathbb{S}_{t}}
-\newcommand{\cj}{\mathbb{S}_{t}'}
-\newcommand{\itoj}{\ci, \cj}
-\newcommand{\setNeighbors}{\mathcal{M}\left(\ci\right)}
-\newcommand{\setNonSynNeighbors}{\mathcal{N}\left(\ci\right)}
-\newcommand{\setSynNeighbors}{\mathcal{S}\left(\ci\right)}
-\newcommand{\submatrix}{q}
-
-\newcommand{\phenoDiv}{S}
-\newcommand{\s}{s}
-\newcommand{\pheno}{x}
-\newcommand{\dpheno}{\delta \pheno}
-\newcommand{\phenoFold}{\Delta G}
-\newcommand{\phenoFoldMin}{\alpha}
-\newcommand{\dphenoFold}{\gamma}
 
@@
 
-We provide a simple toy model, with $2$ parameters that shows a variety of behaviours. 
-This model map the amino-acid sequence to the free energy (kcal/mol) between folded and unfolded state of the whole protein.
+We provide a toy model of protein evolution under mutation-selection-drift equilibrium, reproducing the observation of more complex models, where we derive mathematical relationship using population-genetic equations.
+This model map the amino-acid sequence to the difference in free energy ($\deltaG$ in kcal/mol) between folded and unfolded state of the whole protein.
 Each amino-acid site in the protein can be in the unstable or stable state.
-A protein is optimal ($\phenoFold = \phenoFoldMin$) if all amino-acids are in the stable state.
-Each amino-acid destabilizing the protein brings an increase of $\phenoFold$ equal to $\dphenoFold$.
 \begin{equation}
-\phenoFold = \phenoFoldMin + \dphenoFold * d, 
+\deltaG = \deltaGMin + \Nsite * \gamma * \x, 
 \end{equation}
-where $d$ is the number of unstable amino-acid sites in the sequence, $\alpha$ and $\gamma$ are in kcal/mol.
+where $\x$ is the proportion of unstable amino-acid sites in the sequence, $\deltaGMin$ and $\gamma$ are in kcal/mol.
+An optimal protein has a $\deltaG = \deltaGMin$, and pessimal protein has a $\deltaG = \deltaGMin + \Nsite \gamma$.
 Wrightian fitness is defined as the probability of our protein to be in the folded state, given by Boltzmann equation: 
 \begin{equation}
-w(\phenoFold) = \dfrac{1}{1 + e^{\beta \phenoFold }}, 
+f(\deltaG) = \dfrac{1}{1 + e^{\beta \deltaG }}, 
 \end{equation}
-where $\beta=1.686$ mol/kcal at room temperature.\\
+where the fixed parameter $\beta=1.686$ mol/kcal at room temperature.\\
 
-In the original studies \cite{Williams2006, Goldstein2011, Pollock2012}, free energy of the folded state is computed using the $3$D structure of the folded state and pair-wise contact energies between neighboring amino-acid residues using Miya-zawa \& Jernigan contact potentials \cite{Miyazawa1985}.
-The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).\\
-
-The parameters $\alpha$ and $\gamma$ of our toy model are set to match the average values observed in the original model.
-$\alpha=-118$ kcal/mol, as it is the $\phenoFold$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
-$\gamma=1.0$ kcal/mol, as the average observed increase in $\phenoFold$ per destabilizing mutation \cite{Zeldovich2007}.
+We compare our model to the original studies \cite{Williams2006, Goldstein2011, Pollock2012}, where free energy of the folded state is computed using the $3$D structure of the folded state and pair-wise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
+The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
 
@@
 \centering
-	\includegraphics[width=\textwidth] {artworks/XPopSizeYOmegaEq.pdf}
-	\vspace*{0.2em}
+	\includegraphics[width=0.6\textwidth] {artworks/GoldsteinPollock/dnds_event_tot.pdf}\\
+	\includegraphics[width=0.6\textwidth] {artworks/N300/dnds_event_tot.pdf}\\
+	\includegraphics[width=0.6\textwidth] {artworks/N300Variance/dnds_event_tot.pdf}\\
 	\caption{
 		\textbf{$\bm{\dnds}$ response to change in $\bm{\Ne}$}.
-		Panel A.
-Our toy model reproduces the observations of Goldstein \cite{Goldstein2013} that $\dnds$ at equilibrium is independent of $\Ne$.
-Indeed, whenever the parameters $\alpha$ and $\gamma$ are taken to match the average values observed in the original model, we observed no response of $\dnds$ to $\Ne$ (pink and brown lines).
-		However, decreasing both $\alpha$ and $\gamma$ together implies a strong negative relationship between $\Ne$ and $\dnds$, as predicted under the nearly-neutral theory of evolution.
-		Panel B.
-By reducing the number of sites of the protein, thus by effectively reducing the potential number of sites that can compensate for the change in $\Ne$, the $\dnds$ response to changes in $\Ne$ is stronger.
+		Top panel, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ in the model of Goldstein \& Pollock, but is not independent as claimed in \cite{Goldstein2013}.
+		For each population size, $200$ simulation were performed and the average (blue solid line) and standard deviation (blue filling) are shown.
+		This weak correlation matches the prediction of our toy model that the linear relation (blue dashed line) has a slope equal to $(\beta  \Nsite \gamma)^{-1} \simeq 0.0019$.
+		Middle panel, $\alpha=-118$ and $\Nsite \gamma=300\E[\gamma]=300$.
+		Bottom panel, $\alpha=-118$ and each site has it's own normally distributed $\gamma$ with mean $1$ and standard deviation of $1$.
+		The original $3$D model of Goldstein \& Pollock and our simple toy model provide qualitatively and quantitatively the same linear relationship between $\dnds$ and log-$\Ne$, whenever the parameters $\alpha$ and $\Nsite \gamma$ are taken to match the average values observed in the original $3$D model.
+		\label{fig:GoldsteinVsToy}
+	}
+\end{mdframed}
+\end{figure*}
+
+\begin{figure*}[htb!]
+\begin{mdframed}
+	\centering
+	\includegraphics[width=0.8\textwidth] {artworks/ScaleAlphaGammaStdN300/SimuStab/dnds_event_tot.pdf}\\
+	\caption{
+		\textbf{$\bm{\dnds}$ linear response to change in $\bm{\Ne}$}.
+		The parameter $\Nsite \gamma$ is controlling the $\dnds$-$\Ne$ relationship, decreasing $\Nsite \gamma$ implies a stronger negative linear relationship between $\dnds$ and log-$\Ne$.
+The intersect is a function of both $\alpha$ and $\Nsite \gamma$.
 	}
@@
 
-DNA sequence evolution are simulated under an origin-fixation model \cite{McCandlish2014}, where the rate of substitution of a new mutation with selection coefficient $s$ is:
+Protein sequence evolution are simulated under an origin-fixation model \cite{McCandlish2014}, where the rate of substitution of a new mutation with selection coefficient $s$ is:
 \begin{equation}
@@
 
-Our toy model reproduces the observations that $\dnds$ at equilibrium is independent of $\Ne$.
-Indeed, whenever the parameters $\alpha$ and $\gamma$ are taken to match the average values observed in the original model, we observed no response of $\dnds$ to $\Ne$.
+Simulations of protein evolution with both models display the same $\dnds$-$\Ne$ relationship, whenever the parameters $\alpha$ and $\Nsite \gamma$ of our model are set to match the average values of the $3$D model (see Figure \ref{fig:EqdndsNe}, panel A).
+In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
+The parameter $\Nsite \gamma=300.0$ kcal/mol, since the average observed increase in $\deltaG$ per destabilizing mutation is $\deltadeltaG = 1.0$ kcal/mol per site and our protein is $300$ sites \cite{Zeldovich2007}.
+The original works claimed that under the $3$D model, $\dnds$ is independent of $\Ne$ because $\deltaG$ is an equimutable parameter.
+We argue instead that $\dnds$ is approximatively linear with log-$\Ne$, and that $\deltaG$ is not equimutable (see Figure \ref{fig:EqdndsNe}, panel B).
+More precisely, this mathematical equation can be derived (see Supplementary Materials) and the slope is inversely proportional to $\Nsite \gamma$ :
+\begin{equation}
+\dnds \propto -\dfrac{1}{\beta \Nsite \gamma} \ln(\Ne).
+\end{equation}
+Meaning that under a value of $\Nsite \gamma=300.0$ kcal/mol, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is $0.002$.
+Thus if we consider $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations, the slope is determined solely by the number of sites in the protein. 
 However, decreasing both $\alpha$ and $\gamma$ together implies a strong negative relationship between $\Ne$ and $\dnds$, as predicted under the nearly-neutral theory of evolution.
 Moreover, by reducing the number of sites of the protein, thus by effectively reducing the potential number of sites that can compensate for the change in $\Ne$, the $\dnds$ at equilibrium after changes in $\Ne$ is stronger \textit{Necessary mechanistic explanation for this fact} (see Figure \ref{fig:EqdndsNe}).
+
+\begin{figure*}[htb!]
+\begin{mdframed}
+\centering
+	\begin{tikzpicture}
+	\begin{axis}[
+	xlabel={$\x$},
+	ymin=0.0, ymax=1.0,
+	samples=100,
+	legend entries={$\ln(\frac{1 - \x}{\x})$,
+		$S_{\deltaGMin = -118,\Nsite \gamma = 300,\Ne = 10^{4}}(\x)$,
+		$S_{\deltaGMin = -118,\Nsite \gamma = 300,\Ne = 10^{5}}(\x)$,
+		$S_{\deltaGMin = -16.7,\Nsite \gamma = 30,\Ne = 10^{4}}(\x)$,
+		$S_{\deltaGMin = -16.7,\Nsite \gamma = 30,\Ne = 10^{5}}(\x)$,
+	},
+	legend cell align=left,
+	minor tick num=2,
+	axis x line=bottom,
+	axis y line=left,
+	legend style={at={(0.5,0.95)},anchor=north west}
+	]
+	\addplot[domain=0.3:0.6,line width=2.0pt, color=BLUE]{ ln((1-x)/x) };
+	\addplot[name path=A, domain=0.3:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
+	\addplot[name path=B, dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
+	\addplot[name path=C,domain=0.3:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-16.71+30*x))};
+	\addplot[name path=D,dashed,domain=0.3:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-16.71+30*x))};
+	\addplot[fill=RED, opacity=0.2] fill between[ of = A and B];
+	\addplot[fill=GREEN, opacity=0.2] fill between[ of = C and D];
+	\end{axis}
+	\end{tikzpicture}
+	\caption{
+	\textbf{$\bm{\dnds}$ response to change in $\bm{\Ne}$}.
+	The equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=S_{\deltaGMin,\Nsite \gamma,\Ne}(\x\eq)$ (see Supplementary Materials).
+$S$ is proportional to $\Ne$, but increase exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate.
+	}
+\label{fig:NeChangeInfluence}
+\end{mdframed}
+\end{figure*}
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{artworks/SimuStabExponentialDecay.png}
-	\vspace*{0.2em}
+	\includegraphics[width=\textwidth]{artworks/RelaxationScaleAlphaGammaN100/14_1_dNdN0.png}
+	\includegraphics[width=\textwidth]{artworks/RelaxationScaleAlphaGammaN100/100_1_dNdN0.png}\\
 	\caption{
@@
 	\centering
-	\includegraphics[width=\textwidth]{artworks/SimuExponentialDecay.png}
-	\vspace*{0.2em}
+	\includegraphics[width=\textwidth]{artworks/SimuExponentialDecay.png}\\
 	\caption{
@@
 \begin{equation*}
-\phenoFold\left(\ci\right) = \phenoFoldMin + \dphenoFold * d\left(\ci\right), 
+\deltaG\left(\ci\right) = \deltaGMin + \Nsite \gamma * \x\left(\ci\right), 
 \end{equation*}
-where $d\left(\ci\right)$ is the number of non-optimal amino-acid sites in the sequence.
+where $\x\left(\ci\right)$ is the proportion of non-optimal amino-acid sites in the sequence.
 Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Boltzmann equation: 
 \begin{equation}
-w(\phenoFold\left(\ci\right)) = \dfrac{P_{\mathrm{F}}\left(\ci\right)}{P_{\mathrm{F}}\left(\ci\right) + P_{\mathrm{U}}\left(\ci\right)} = \dfrac{e^{-\beta G_{\mathrm{F}}\left(\ci\right) }}{e^{-\beta G_{\mathrm{F}} \left(\ci\right) } + e^{-\beta G_{\mathrm{U}}\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \phenoFold\left(\ci\right) }},
+f(\deltaG\left(\ci\right)) = \dfrac{P_{\mathrm{F}}\left(\ci\right)}{P_{\mathrm{F}}\left(\ci\right) + P_{\mathrm{U}}\left(\ci\right)} = \dfrac{e^{-\beta G_{\mathrm{F}}\left(\ci\right) }}{e^{-\beta G_{\mathrm{F}} \left(\ci\right) } + e^{-\beta G_{\mathrm{U}}\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }},
 \end{equation}
@@
 For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codon that are one mutation away and that are not stop codon.
-For each mutant sequence $\cj \in \setNeighbors$, we compute $\phenoFold^{t+1}$ from the updated sequence $\cj$, and subsequently the selection coefficient of the mutant:
+For each mutant sequence $\cj \in \setNeighbors$, we compute $\deltaG^{t+1}$ from the updated sequence $\cj$, and subsequently the selection coefficient of the mutant:
 \begin{equation}
-s \left( \ci,\cj\right) = \dfrac{ w\left( \phenoFold\left(\cj\right) \right) - w\left( \phenoFold\left(\ci\right) \right)}{w\left( \phenoFold\left(\ci\right) \right)}.
+s \left( \ci,\cj\right) = \dfrac{ f\left( \deltaG \left(\cj\right) \right) - f\left( \deltaG \left(\ci\right) \right)}{f\left( \deltaG \left(\ci\right) \right)}.
 \end{equation}
@@
 
-\newpage
-\part*{Supplementary materials}
-
-\section*{Probability of folding as in Goldstein \& Pollock}
-We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
-From a DNA sequence $\ci$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}\left(\ci\right)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
-\begin{equation}
-G_{\mathrm{F}}\left(\ci\right) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\ci(\site), \ci(r) \right),
-\end{equation}
-where $I(a,b)$ is the pair-wise contact energies between amino-acid $a$ and $b$, using contact potentials estimated by Miya-zawa and Jernigan, and $\mathcal{N}(\site)$ are the neighbor residues of site $\site$ (closer than $7\angstrom$) in the $3$D structure.\\
-
-The free energy of unfolded states $G_{\mathrm{U}}\left(\ci\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
-\begin{equation}
-G_{\mathrm{U}}\left(\ci\right) = \langle G\left(\ci\right) \rangle - kT \ln (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G\left(\ci\right)^2 \rangle - \langle G\left(\ci\right) \rangle^2\right] }{kT}
-\end{equation}
-where the average $\langle . \rangle$ runs other the $55$ decoy $3$D structures, and $k$ is the Boltzmann constant and $T$ the temperature in Kelvin.\\
-
-From the energy of folded and unfolded states, we can compute the difference in free energy between the states:
-\begin{equation}
-\phenoFold\left(\ci\right) = G_{\mathrm{F}}\left(\ci\right) - G_{\mathrm{U}}\left(\ci\right)
-\end{equation}
-
-\begin{center}
-	\includegraphics[width=165mm] {artworks/ModelSimuFold.pdf}
-\end{center}
-
-\section*{Equimutability}
-We define $\pheno$ as the current phenotype and $\dpheno$ as the phenotypic effect of a mutation.
-$\s \left( \pheno, \dpheno \right)$ is the selection coefficient of a phenotypic effect $\dpheno$ if the current phenotype is $\pheno$:
-\begin{equation*}
-\s \left( \pheno, \dpheno \right) = \dfrac{w \left( \pheno + \dpheno \right) - w \left( \pheno \right)}{w \left( \pheno \right)} \simeq w'\left( \pheno \right) \dpheno, 
-\end{equation*}
-where $w \left( \pheno \right)$ is the Wrightian fitness of phenotype $x$.
-We also define $\rho_{\pheno}(\dpheno)$ as the distribution of phenotypic effects, at phenotype $\pheno$.
-At equilibrium, the expected selection coefficient $ 4 \Ne \s \left( \pheno, \dpheno \right)$ of mutations that reached fixation must be $0$:
-\begin{align*}
-0 & = \E_{\dpheno} \left[ 4 \Ne \s\left( \pheno, \dpheno \right) \dfrac{ 4 \Ne \s\left( \pheno, \dpheno \right) }{1 - \e^{4\Ne\s\left( \pheno, \dpheno \right)}} \right] \\ & 
-\simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \pheno \right)  \right]^2}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno} }  \dpheno^2  \rho_{\pheno}(\dpheno) \der \dpheno 
-\end{align*}
-And the substitution rate is :
-\begin{align*}
-\omega(\Ne) & = \E_{\dpheno} \left[ \dfrac{ 4 \Ne \s \left( \pheno, \dpheno \right)}{1 - \e^{4\Ne \s \left( \pheno, \dpheno \right)}} \right] \simeq \E_{\dpheno} \left[ \dfrac{ 4 \Ne f'\left( \pheno \right) \dpheno}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno}} \right] \\
-& \simeq \int_{-\infty}^{\infty} \dfrac{4 \Ne f'\left( \pheno \right) \dpheno}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno } } \rho_{\pheno}(\dpheno) \der \dpheno \\
-\end{align*}
-If $\rho_{\pheno}(\dpheno) = \rho(\dpheno)$ is independent of $\pheno$:
-\begin{align*}
-0 & \simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \pheno \right)  \right]^2}{1 - \e^{4\Ne f'\left( \pheno \right) \dpheno} } \dpheno^2 \rho(\dpheno)  \der \dpheno \\
-\Longrightarrow & \ 4 \Ne f'\left( \pheno \right) = K \in \mathbb{R}^{*} \text{ and } \s\left( \pheno, \dpheno \right) \simeq \dfrac{K \dpheno}{4\Ne} \\
-\Longrightarrow & \ \omega( \Ne ) \simeq \int_{-\infty}^{\infty} \dfrac{ K  }{1 - \e^{K \dpheno } } \dpheno \rho(\dpheno)  \der \dpheno  \text{ is independent of } \Ne
-\end{align*}
-
 \end{document}

### 2020-03-31 manuscript/supp-mat.tex
Moved the notations into new-commands.tex

@@
 \usepackage{xfrac, nicefrac}
-\usepackage{xcolor}
 \usepackage{lmodern}
 \usepackage{enumitem}
-\usepackage{pgfplots, pgf,tikz}
 \usepackage[margin=60pt]{geometry}
-\pdfinclusioncopyfonts=1
 
-\definecolor{RED}{HTML}{EB6231}
-\definecolor{YELLOW}{HTML}{E29D26}
-\definecolor{BLUE}{HTML}{5D80B4}
-\definecolor{LIGHTGREEN}{HTML}{6ABD9B}
-\definecolor{GREEN}{HTML}{8FB03E}
-\definecolor{PURPLE}{HTML}{BE1E2D}
-\definecolor{BROWN}{HTML}{A97C50}
-\definecolor{PINK}{HTML}{DA1C5C}
-\captionsetup{width=0.85\textwidth}
-
-\newcommand{\specialcell}[2][c]{%
-	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
-
-\DeclareMathOperator{\E}{\mathbb{E}}
-\newcommand{\der}{\mathrm{d}}
-\newcommand{\e}{\mathrm{e}}
-
-% Time, effective population size and mutation rate.
-\newcommand{\Ne}{N_{\mathrm{e}}}
-\newcommand{\dnds}{\omega}
-\newcommand{\Nsite}{\text{n}}
-\newcommand{\Nstate}{\text{K}}
-
-\newcommand{\x}{x}
-\newcommand{\eq}{^{*}}
-\newcommand{\dx}{\delta \x}
+\include{new-commands}
 
@@
 0 &\text{ with probability } \rho_{\x}(0) = \x \left[1 - \frac{1}{\Nstate - 1}\right], \\
--\dx &\text{ with probability } \rho_{\x}(-\dx) = \frac{x}{\Nstate - 1}.\\
+-\dx &\text{ with probability } \rho_{\x}(-\dx) = \frac{\x}{\Nstate - 1}.\\
 \end{cases} \label{eq:proba}
@@
 \begin{gather}
- f(\x) = \frac{1}{1 + \e^{\beta(\alpha + \kappa \x)}}, \label{eq:fitness}
+ f(\x) = \frac{1}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}, \label{eq:fitness}
 \end{gather}
-where $\alpha = \Delta G_0 < 0$ is the difference in free energy between stable and unstable protein when all sites are stable.
-$\kappa$ is the expected change in $\Delta G$ when all sites are unstable.
-By definition, $\kappa \dx = \sfrac{\kappa}{\Nsite} = \Delta \Delta G $\\
+where $\deltaGMin = \Delta G_0 < 0$ is the difference in free energy between stable and unstable protein when all sites are stable.
+$\Nsite \gamma$ is the expected change in $\Delta G$ when all sites are unstable.
+By definition, $\Nsite \gamma \dx = \sfrac{\Nsite \gamma}{\Nsite} = \Delta \Delta G $\\
 The derivative of fitness w.r.t to phenotype is:
 \begin{align}
- \frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\alpha + \kappa \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
- & = - \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \label{eq:df_dx}.
+ \frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
+ & = - \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \label{eq:df_dx}.
 \end{align}
@@
 \begin{align}
-\frac{\partial^2 \ln (f(\x))}{\partial {\x}^2} & = - \beta \kappa \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \right) \text{ from eq.
+\frac{\partial^2 \ln (f(\x))}{\partial {\x}^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \right) \text{ from eq.
 \ref{eq:df_dx}}, \\
- & = - \beta \kappa  \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{\left( 1 + \e^{\beta(\alpha + \kappa \x)}\right)^2}, \\
- & = \frac{\beta \kappa}{1 + \e^{\beta(\alpha + \kappa \x)}} \frac{ \partial \ln f(\x) }{\partial {\x}} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}.
+ & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{\left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}\right)^2}, \\
+ & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \frac{ \partial \ln f(\x) }{\partial {\x}} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}.
 \end{align}
@@
 \begin{gather}
-\lambda_{K}(\x\eq) = 4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x\eq)}}{1 + \e^{\beta(\alpha + \kappa \x\eq)}}  \dx   \text{ from eq. \ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
+\lambda_{K}(\x\eq) = 4\Ne \beta \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x\eq)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x\eq)}}  \text{ from eq.
+\ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
 \end{gather}
-Using $\Ne=10^4$, $\beta=1.686$, $\alpha = -118$, $\Nsite=300$, $\kappa=300$, we have the following:
+Using $\Ne=10^4$, $\beta=1.686$, $\deltaGMin = -118$, $\Nsite=300$, $\gamma=1$, we have the following:
 \begin{center}
@@
 	legend entries={
-		$4\Ne \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \dx$,
+		$4\Ne \beta \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}$,
 		$\lambda_{2}(\x)$,
@@
 \section*{Approximations}
-We assume that $\Ne \gg 1$, $\alpha + \kappa \gg 1$ and $\kappa \dx \simeq 1$.
+We assume that $\Ne \gg 1$, $\deltaGMin + \Nsite \gamma \gg 1$ and $ \gamma \simeq 1$.
 \begin{gather}
@@
 \iff \frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:s_from_fitness}}, \label{eq:approx_df_dx}\\
-\iff \beta \kappa \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)} } \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:df_dx}}, \\
-\iff \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx \beta \kappa}, \\
-\Longrightarrow \frac{\e^{\beta(\alpha + \kappa \x)}}{1 + \e^{\beta(\alpha + \kappa \x)}} \ll 1\text{ given that $\Ne \gg 1$ and $\kappa \dx \simeq 1$}, \\
-\Longrightarrow \e^{\beta(\alpha + \kappa \x)} \ll 1. \label{eq:approx_exp}
+\iff \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)} } \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq.
+\ref{eq:df_dx}}, \\
+\iff \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \beta \gamma}, \\
+\Longrightarrow \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \ll 1\text{ given that $\Ne \gg 1$ and $ \gamma  \simeq 1$}, \\
+\Longrightarrow \e^{\beta(\deltaGMin + \Nsite \gamma \x)} \ll 1. \label{eq:approx_exp}
 \end{gather}
@@
 \frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:approx_df_dx}}, \\
-\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)\left( 1 + \e^{\beta(\alpha + \kappa \x)} \right)}{4 \Ne \dx \beta \kappa }\text{ from eq.
+\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx } \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}\text{ from eq.
 \ref{eq:df2_dx2}}, \\
-\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx \beta \kappa }\text{ from eq. \ref{eq:approx_exp}}, \\
-\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \lambda_{\Nstate}(\x\eq) \frac{(1 - \x\eq)\x\eq }{\beta \kappa} \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
-\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \lambda_{\Nstate}(\x\eq) \dx \frac{(1 - \x\eq)\x\eq }{\beta } \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}\text{ given that $\kappa \dx \simeq 1$}, \\
-\Longrightarrow \left| \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \right| \gg \left| \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} \right|\text{ given that $\dx \ll 1$}. \label{eq:approx_df2_df} \\
+\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)\beta \Nsite \gamma}{4 \Ne \dx  }\text{ from eq. \ref{eq:approx_exp}}, \\
+\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \lambda_{\Nstate}(\x\eq) (1 - \x\eq)\x\eq \beta \Nsite \gamma \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
+\Longrightarrow \left| \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \right| \gg \left| \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} \right|\text{ given that $\Nsite \gamma \gg 1$}.
+\label{eq:approx_df2_df} \\
 \end{gather}
@@
 & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\text{ from eq. \ref{eq:approx_df2_df}}, \\
-& \simeq - \frac{1 + \e^{\beta(\alpha + \kappa \x\eq)}}{\beta \kappa}\text{ from eq. \ref{eq:df2_dx2}}, \\
-& \simeq - \frac{1}{\beta \kappa}\text{ from eq. \ref{eq:approx_exp}}. \label{eq:approx_dx_dlnNe}
+& \simeq - \frac{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x\eq)}}{\beta \Nsite \gamma}\text{ from eq. \ref{eq:df2_dx2}}, \\
+& \simeq - \frac{1}{\beta \Nsite \gamma}\text{ from eq. \ref{eq:approx_exp}}. \label{eq:approx_dx_dlnNe}
 \end{align}
@@
 \frac{ \der \dnds\eq}{\der \ln (\Ne)}  & = \frac{ \der \dnds\eq}{\der \x\eq}  \frac{ \der \x\eq}{\der \ln (\Ne)}, \\
-  & \simeq - \frac{1}{\beta \kappa} \text{from eq. \ref{eq:approx_dx_dlnNe} and \ref{eq:approx_domega_dx}}.
+  & \simeq - \frac{1}{\beta \Nsite \gamma} \text{from eq. \ref{eq:approx_dx_dlnNe} and \ref{eq:approx_domega_dx}}.
 \end{align}
-\section*{Numerical results}
+
+\section*{Probability of folding as in Goldstein \& Pollock}
+We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
+From a DNA sequence $\ci$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}\left(\ci\right)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
+\begin{equation}
+G_{\mathrm{F}}\left(\ci\right) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\ci(\site), \ci(r) \right),
+\end{equation}
+where $I(a,b)$ is the pair-wise contact energies between amino-acid $a$ and $b$, using contact potentials estimated by Miya-zawa and Jernigan, and $\mathcal{N}(\site)$ are the neighbor residues of site $\site$ (closer than $7\angstrom$) in the $3$D structure.\\
+
+The free energy of unfolded states $G_{\mathrm{U}}\left(\ci\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
+\begin{equation}
+G_{\mathrm{U}}\left(\ci\right) = \langle G\left(\ci\right) \rangle - kT \ln (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G\left(\ci\right)^2 \rangle - \langle G\left(\ci\right) \rangle^2\right] }{kT}
+\end{equation}
+where the average $\langle . \rangle$ runs other the $55$ decoy $3$D structures, and $k$ is the Boltzmann constant and $T$ the temperature in Kelvin.\\
+
+From the energy of folded and unfolded states, we can compute the difference in free energy between the states:
+\begin{equation}
+\deltaG\left(\ci\right) = G_{\mathrm{F}}\left(\ci\right) - G_{\mathrm{U}}\left(\ci\right)
+\end{equation}
+
 \begin{center}
-\begin{tikzpicture}
-	\begin{axis}[
-	xlabel={$\x$},
-	ymin=0.0, ymax=1.0,
-	samples=100,
-	legend entries={$\ln(\frac{1 - \x}{\x})$,
-		$\alpha = -118\text{, }\kappa = 300\text{, }\Ne = 10^{4}$,
-		$\alpha = -118\text{, }\kappa = 300\text{, }\Ne = 10^{5}$,
-		$\alpha = -16.7\text{, }\kappa = 30\text{, }\Ne = 10^{4}$,
-		$\alpha = -16.7\text{, }\kappa = 30\text{, }\Ne = 10^{5}$},
-	legend cell align=left,
-	minor tick num=2,
-	axis x line=bottom,
-	axis y line=left,
-	legend style={at={(0.5,0.95)},anchor=north west}
-	]
-	\addplot[domain=0.3:0.6,line width=2.0pt, color=BLUE]{ ln((1-x)/x) };		\addplot[domain=0.3:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
-	\addplot[dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
-	\addplot[domain=0.3:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-16.71+30*x))};
-	\addplot[dashed,domain=0.3:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-16.71+30*x))};
-	\end{axis}
-\end{tikzpicture}
+	\includegraphics[width=165mm] {artworks/ModelSimuFold.pdf}
 \end{center}
+
+\begin{mdframed}
+\begin{center}
+	\includegraphics[width=0.6\textwidth] {artworks/GoldsteinPollock/sub-DeltaG-mean.pdf}\\
+	\textbf{$\bm{\deltaG}$ response to change in $\bm{\Ne}$}.
+	$\deltaG$ at equilibrium (blue solid line) is linearly (blue dashed line) dependent on log-$\Ne$, as predicted in the model of Goldstein \& Pollock, with a slope equal to $\beta^{-1}$.
+	\label{fig:EqdndsNe}
+\end{center}
+\end{mdframed}
+
+\section*{Equimutability}
+We define $\x$ as the current phenotype and $\dx$ as the phenotypic effect of a mutation.
+$\s \left( \x, \dx \right)$ is the selection coefficient of a phenotypic effect $\dx$ if the current phenotype is $\x$:
+\begin{equation*}
+\s \left( \x, \dx \right) = \dfrac{w \left( \x + \dx \right) - w \left( \x \right)}{w \left( \x \right)} \simeq w'\left( \x \right) \dx, 
+\end{equation*}
+where $w \left( \x \right)$ is the Wrightian fitness of phenotype $x$.
+We also define $\rho_{\x}(\dx)$ as the distribution of phenotypic effects, at phenotype $\x$.
+At equilibrium, the expected selection coefficient $ 4 \Ne \s \left( \x, \dx \right)$ of mutations that reached fixation must be $0$:
+\begin{align*}
+0 & = \E_{\dx} \left[ 4 \Ne \s\left( \x, \dx \right) \dfrac{ 4 \Ne \s\left( \x, \dx \right) }{1 - \e^{4\Ne\s\left( \x, \dx \right)}} \right] \\ & 
+\simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \x \right)  \right]^2}{1 - \e^{4\Ne f'\left( \x \right) \dx} }  \dx^2  \rho_{\x}(\dx) \der \dx 
+\end{align*}
+And the substitution rate is :
+\begin{align*}
+\omega(\Ne) & = \E_{\dx} \left[ \dfrac{ 4 \Ne \s \left( \x, \dx \right)}{1 - \e^{4\Ne \s \left( \x, \dx \right)}} \right] \simeq \E_{\dx} \left[ \dfrac{ 4 \Ne f'\left( \x \right) \dx}{1 - \e^{4\Ne f'\left( \x \right) \dx}} \right] \\
+& \simeq \int_{-\infty}^{\infty} \dfrac{4 \Ne f'\left( \x \right) \dx}{1 - \e^{4\Ne f'\left( \x \right) \dx } } \rho_{\x}(\dx) \der \dx \\
+\end{align*}
+If $\rho_{\x}(\dx) = \rho(\dx)$ is independent of $\x$:
+\begin{align*}
+0 & \simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \x \right)  \right]^2}{1 - \e^{4\Ne f'\left( \x \right) \dx} } \dx^2 \rho(\dx)  \der \dx \\
+\Longrightarrow & \ 4 \Ne f'\left( \x \right) = K \in \mathbb{R}^{*} \text{ and } \s\left( \x, \dx \right) \simeq \dfrac{K \dx}{4\Ne} \\
+\Longrightarrow & \ \omega( \Ne ) \simeq \int_{-\infty}^{\infty} \dfrac{ K  }{1 - \e^{K \dx } } \dx \rho(\dx)  \der \dx  \text{ is independent of } \Ne
+\end{align*}
+
+\section*{Model of non-specific interactions}
+
+The proteome is assumed to be composed of $m$ protein species, all with same abundance $C$.
+Each macromolecule may either be in free form or engaged in a non-specific interaction.
+Only pairwise interactions are considered, and higher-order interactions are ignored.
+The equilibrium is characterised by:
+\begin{eqnarray}
+[ij] &=& \frac{[i][j]}{C_0} \, e^{\beta E_{ij}}
+\end{eqnarray}
+where $[i]$ and $[j]$ are the concentrations of protein species $i$ and $j$, and $[ij]$ is the concentration of their (non-specific) dimer.
+Here, $E_{ij}$ is the interaction free energy, which can itself be decomposed as a sum of three terms:
+\begin{eqnarray}
+E_{ij} &=& \alpha + E_i + E_j 
+\\
+&=& \alpha + \gamma n (x_i + x_j)
+\end{eqnarray}
+where we assume that each protein has $n=100$ residues at its surface, $x_i$ stands for the fraction of hydrophobic residues at the surface of protein $i$, and each hydrophobic residue makes an additive contribution of $\gamma$ to the total.
+
+By conservation of the total number of molecules:
+\begin{eqnarray}
+C &=& [i] + \sum_{j \neq i} [ij] \\
+&=& [i] + \sum_{j \neq i} \frac{[i][j]}{C_0} \, e^{\beta E_{ij}}
+\end{eqnarray}
+and we note:
+\begin{eqnarray}
+\epsilon_i = \sum_j [ij]
+\end{eqnarray}
+the fraction of protein $i$ sequestered in non-specific interactions.
+We assume that the log fitness is proportional to the total amount of protein sequestered in non-specific interactions:
+\begin{eqnarray}
+\ln f &=& -b \sum_i \epsilon_i
+\end{eqnarray}
+where $b>0$ is a parameter determining the overall stringency of selection against non-specific interactions.
+
+\subsection*{Mean field, weak-interaction limit}
+
+To make the model tractable and compact, we assume that non-specific interactions are weak, i.e.
+$\epsilon_i << 1$ for all $i$.
+We then make a first-order approximation in the $\epsilon_i$'s.
+In addition, we use a mean-field approximation, such that, when considering a specific protein species $i$, we assume that all other proteins have the same fraction $\bar x$ of hydrophobic residues at their surface.
+The value of $\bar x$ could in principle be found using a self-consistent argument, essentially by (1) explicitly calculating the net substitution flux for protein $i$ with fraction $x_i$, under mean-field $\bar x$, and (2) expressing the constraint that this substitution process for protein $i$ is stationary at $x_i = \bar x$.
+This derivation is not conducted here, as it is not needed.
+
+Using these approximations, we can re-express the conservation of total mass as:
+\begin{eqnarray}
+C &=& [i] + (m-1) [i] \frac{C}{C_0} \, e^{\beta (\alpha + \gamma n (\bar x + x_i))}
+\end{eqnarray}
+Here, we have used the fact that $[j] = C(1 - \epsilon_j)$ can be approximated as $[j] \simeq C$ since it is involved in a term already of the order of $\epsilon_i$.
+As a result, all $m-1$ terms of the sum over $j\neq i$ are identical.
+Next, solving for $[i]$ gives:
+\begin{eqnarray}
+[i] &=& \frac{C} {1 + (m-1) \frac{C}{C_0} \, e^{\beta (\alpha + \gamma n (\bar x + x_i))}}
+\\ &\simeq& C \left( 1 - m \frac{C}{C_0} \, e^{\beta (\alpha + \gamma n (\bar x + x_i))} \right)
+\\ &=&
+C (1 - \epsilon_i)
+\end{eqnarray}
+and thus $\epsilon_i$ can be identified with:
+\begin{eqnarray}
+\epsilon_i  &=& m \frac{C}{C_0} \, e^{\beta (\alpha +\gamma n (\bar x + x_i))}
+\end{eqnarray}
+
+Now, assume that the system is at equilibirum (thus $x_i = \bar x$).
+The strength of selection acting on mutations occuring at the surface of protein $i$, of effect size $\delta x = \pm 1/n$, is given by $s = \kappa \delta x$ where:
+\begin{eqnarray}
+\kappa_i &=& b \frac{d \epsilon_i} {d x_i} 
+\\ &=&
+b \beta \gamma n m \frac{C}{C_0} \, e^{\beta (\alpha +\gamma n (\bar x + x_i))}
+\end{eqnarray}
+and thus:
+\begin{eqnarray}
+\ln \kappa_i &=& \ln \left( b \beta \gamma n m \frac{C}{C_0}  \right) \, + \, \beta (\alpha +\gamma n \bar x) \, + \, \beta \gamma n x_i
+\end{eqnarray}
+where only the last term depends on $x_i$.
+Finally, applying the main result of this work to the present case allows us to express the elasticity of $\omega$ as a function of $N_e$ as:
+\begin{eqnarray}
+\chi &=& \frac{d \omega} {d \ln N_e} 
+\\ &=&  2(\lambda - 1) \, \frac{d \ln \kappa_i}{d x_i} 
+\\ &=&  2 (\lambda - 1) \, \frac{1}{\beta \gamma n}
+\end{eqnarray}
+Note that, here, we have used $K=2$ (hydrophobic and polar residues are roughly equally likely to occur by mutation), and assumed $x^* << 1$.
+A more accurate formula could be used without this latter assumption.
+In any case, $\chi$ is now dependent on $x^*$, through $\lambda$.
+
+\subsection*{Empirical calibration}
+
+Based on empirical estimates obtained from Zhang et al.
+The mean fraction of hydrophobic residues at the surface of proteins is $0.22 \pm 0.06$.
+With $n=100$ residues, this makes $22 \pm 6$.
+The mean value for $E_{ij}$ is 7 kT, with a standard deviation of $\sigma = 1.8$ kT.
+Assuming that this standard deviation of $\pm 1.8$ kT is contributed by $\pm 6$ mutations gives $\gamma = 1.8 / 6 = 0.3$ kT or $0.18$ kcal per mole.
+Also, with $x=0.22$, $\lambda \simeq 4$, and thus $\chi = 6  / 30 = 0.2$, thus a much stronger response than under the model based on conformational stability.
+
+\section*{Alternative models for the log-fitness function}
+
+Alternative models of the log-fitness function $\ln f(x)$ are proposed. All of them depend on the folded fraction of the protein of interest, which is given by the Fermi-Dirac distribution:
+\begin{eqnarray}
+p_f &=& \frac{1}{1 + e^{\beta(\alpha + \gamma n x)}}
+\end{eqnarray}
+where $x$ is the fraction of destabilizing mutations, each contributing a $\Delta \Delta G = \gamma$, and $\beta = 1 / kT$.
+Thus, $\alpha = \Delta G_0$ is the free energy of the most stable sequence.
+The misfolded fraction is  $p_m = 1-p_f$.
+In addition, $p_f$ is typically close to 1 (or $p_m << 1$), so that we can use a first-order approximation:
+\begin{eqnarray}
+p_f &=& 1 - p_m
+\\
+&\simeq& 1 - e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+or equivalently
+\begin{eqnarray}
+p_m &\simeq& e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+
+For all models, the selective cost $s$ associated with a point mutation with effect size $\delta x$ is given by:
+\begin{eqnarray}
+s \simeq - \kappa \delta x
+\end{eqnarray}
+where
+\begin{eqnarray}
+\kappa &=& - \frac{d \ln f}{d x} 
+\end{eqnarray}
+Thus, $\kappa$ captures the strength of the selective response.
+All fitness functions considered below are log-concave, and as a result, $\kappa$ is an increasing function of $x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
+The scaled selection coefficient is:
+\begin{eqnarray}
+S = 4N_e s  &=& 4 N_e \kappa \delta x
+\end{eqnarray}
+which depends on the scaled strength of selection $4 N_e \kappa$, reflecting both the intrinsic strength of selection and the effect of population size.
+
+\subsection*{Model 1: fitness equal to folded fraction}
+
+A first model is to assume that the fitness is equal to the folded fraction (Goldstein):
+\begin{eqnarray}
+f(x) &=& p_f
+\end{eqnarray}
+or:
+\begin{eqnarray}
+\ln f(x) &=& \ln (1 - p_m)
+\\
+& \simeq & - p_m
+\\
+& \simeq & - e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+
+The strength of selection is then given by:
+\begin{eqnarray}
+\kappa &=& \beta \gamma n e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+
+This model, however, does not express the fact that selection is typically stronger for proteins characterized by higher levels of expression.
+
+\subsection*{Model 2: selective cost proportional to amount of misfolded protein}
+
+A slight variation on Model 1 is to assume that the selective cost itself is proportional to the total amount of misfolded protein. For a given protein with expression level $y$:
+\begin{eqnarray}
+\ln f = - A y p_m
+\end{eqnarray}
+where $A$ is the cost per misfolded macromolecule. Then, 
+\begin{eqnarray}
+\kappa &=& A \beta \gamma n y  e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+or equivalently
+\begin{eqnarray}
+4 N_e \kappa &\propto& 4 N_e y e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+Thus, under this model, an increase in the effective population size $N_e$ or the expression level $y$ both have the same effect on the molecular evolutionary process undergone by the protein.
+
+\subsection*{Model 3: translational errors}
+
+Model 3 is a variant of model 1, accounting for translational errors.
+Translational errors occur at a rate $\rho$ per residue.
+These errors contribute additional destabilizing mutations, each with effect size $\delta x = 1/n$.
+The total number of translational errors per macromolecule is approximately Poisson distributed:
+\begin{eqnarray}
+\pi_k &=& e^{-\rho n} \frac{(\rho n)^k}{k!}
+\end{eqnarray}
+and the total selective cost is now an average over all possible values of $k$:
+\begin{eqnarray}
+\ln f &=& - A y \sum_k \pi_k e^{\beta(\alpha + \gamma n x + \gamma k)}
+\\
+&=& - A y  e^{\beta(\alpha + \gamma n x)} \sum_k e^{-\rho n} \frac{(\rho n)^k}{k!} e^{\beta \gamma k}
+\\
+&=& - A y  e^{\beta(\alpha + \gamma n x) + \rho n (e^{\beta \gamma} -1 )}
+\\
+& \simeq & - A y  e^{\beta(\alpha + \gamma n x) + \rho \beta \gamma n}
+\\
+& =& - A y  e^{\beta(\alpha + \gamma n (x + \rho))}
+\end{eqnarray}
+In words, the fitness function is the same as for Model 2, except that the trait $x$ (fraction of destabilizing mutations) is shifted by $\rho$, the mean fraction of additional mutations contributed by translation errors.
+This additional factor is independent of $x$, and as a result, the scaled selection strength is essentially the same as for Model 2, up to a proportionality constant (contributed by the shift):
+\begin{eqnarray}
+4 N_e \kappa &\propto& 4 N_e y e^{\beta(\alpha + \gamma n (x + \rho))}
+\\ &\propto& 4 N_e y e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+
+\subsection*{Model 4: cost-benefit argument}
+
+This models assumes that the (1) the expression level is regulated so that the total amount of \emph{functional} macromolecules is maintained at a target level $y$; and (2) the log-fitness is proportional to the ratio of the \emph{total} cost of expression over the benefit contributed by the protein (Beaulieu et al).
+Specifically, the protein is assumed to be regulated so as to reach a level of expression of functional proteins of $y$, and contributes a total benefit $B$ (which depends on its specific function).
+Given that only a fraction $p_f = 1-p_m$ of the total amount of protein expressed by the cell is functional, the total cost of expression $C$ is then equal to:
+\begin{eqnarray}
+C &=& \frac{y}{p_f}
+\\ &\simeq& y (1 + p_m)
+\end{eqnarray}
+Then, the log-fitness is given by:
+\begin{eqnarray}
+\ln f & = & -A \frac{y}{B} \left( 1 + e^{\beta(\alpha + \gamma n x)} \right)
+\\
+&=& -by (1 + e^{\beta(\alpha + \gamma n x)})
+\end{eqnarray}
+where $b = A /B$.
+Compared to model 2 and 3, the log-fitness now has an additional term that depends on the target expression level $y$, but not on trait $x$.
+The scaled strength of selection on mutations affecting $x$ has thus the same functional form as for the two previous models:
+\begin{eqnarray}
+4 N_e \kappa &\propto& 4 N_e y e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+Alternative cost-exp models could also be used, allowing for a non-linear cost function for expression or for some elasticity of the realized equilibrium expression level, as a function of the number of mutations (Gout et al).
+Under these models, the strength of selection is still expected to be an increasing function of $y$, although not linear, e.g.
+\begin{eqnarray}
+4 N_e \kappa &\propto& 4 N_e g(y) e^{\beta(\alpha + \gamma n x)}
+\end{eqnarray}
+where $g$ is some function of $y$.
+
+\bibliographystyle{apalike}
+\bibliography{refs-codons,refs-cds}
 \end{document}

### 2020-03-31 scripts/relaxation_analysis.py

@@ -1,36 +1,0 @@
-#!python3
-import os
-import argparse
-import pandas as pd
-import matplotlib
-
-matplotlib.rcParams['font.family'] = 'monospace'
-matplotlib.use('Agg')
-import matplotlib.pyplot as plt
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
-    parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
-    args = parser.parse_args()
-    print(args.input, args.output)
-
-    for param in ["<dN>/<dN0>", "<dN/dN0>"]:
-        my_dpi = 128
-        fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
-        max_y = 0.0
-        for filepath in args.input:
-            if not os.path.isfile(filepath + ".substitutions.tsv"):
-                continue
-
-            df = pd.read_csv(filepath + ".substitutions.tsv", sep='\t')
-            max_y = max((max(df[param]), max_y))
-            plt.plot(df["AbsoluteStartTime"], df[param], linewidth=3)
-
-        plt.ylim((0, max_y))
-        plt.xlabel(r'$t$')
-        plt.ylabel(param)
-        plt.tight_layout()
-        plt.savefig("{0}/{1}.svg".format(args.output, param.replace("/", "")), format="svg")
-        plt.clf()
-        plt.close('all')

### 2020-04-02 scripts/heatmap_analysis.py
Relaxation can merge all paths, scaling draw the confidence interval and the points

@@ -1,19 +1,22 @@
 #!python3
 import os
 import argparse
-import numpy as np
 import pandas as pd
-import itertools
+import numpy as np
 import statsmodels.api as sm
 import matplotlib
 
-label_size = 16
+label_size = 24
 my_dpi = 128
 matplotlib.rcParams['ytick.labelsize'] = label_size
 matplotlib.rcParams['xtick.labelsize'] = label_size
 matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+
+
+def print_b(text):
+    print('\033[34m' + '\033[1m' + text + '\033[0m')
 
 
 def tex_float(x):
@@ -31,79 +34,66 @@ if __name__ == '__main__':
     parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
     parser.add_argument('-x', '--x_param', required=True, type=str, dest="x_param")
     parser.add_argument('-y', '--y_param', required=False, default="", type=str, dest="y_param")
-    parser.add_argument("--y_param_key", action='append', type=lambda kv: kv.split(":"), dest='y_param_dict')
+    parser.add_argument("--y_param_key", required=False, action='append', type=lambda kv: kv.split(":"),
+                        dest='y_param_dict')
     parser.add_argument('-f', '--node', required=False, default=False, type=bool, dest="node")
     args = parser.parse_args()
-    args.y_param_dict = dict(args.y_param_dict) if (args.y_param_dict is not None) else dict()
+    args.y_param_dict = dict([(float(k), v) for k, v in args.y_param_dict]) if (
+                args.y_param_dict is not None) else dict()
     array_values = dict()
-    len_values = set()
     for filepath in args.input:
-        if not os.path.isfile(filepath):
-            continue
+        if not os.path.isfile(filepath): continue
         for param, vals in pd.read_csv(filepath, sep='\t').items():
-            if args.node and "Node" not in param:
-                continue
+            if args.node and "Node" not in param: continue
             if param not in array_values:
                 array_values[param] = dict()
             x, y = float(filepath.split("_")[-3]), float(filepath.split("_")[-2])
             array_values[param][(x, y)] = vals
-            len_values.add(len(vals))
 
-    assert (len(len_values) == 1)
-    nbr_simu = len_values.pop()
[…]

### 2020-04-02 scripts/relaxation_merge.py

@@ -3,141 +3,50 @@ import os
 import argparse
 import pandas as pd
 import numpy as np
-from scipy.optimize import curve_fit
-import statsmodels.api as sm
+from glob import glob
 import matplotlib
 
-label_size = 16
+label_size = 24
 my_dpi = 128
-nbr_points = 1000
+nbr_points = 100
 matplotlib.rcParams['ytick.labelsize'] = label_size
 matplotlib.rcParams['xtick.labelsize'] = label_size
 matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
 
-
-def exponential(_t, _r, _a, _b):
-    return _a * np.exp(-_r * (_t - t_min)) + _b
-
-
-def tex_float(x):
-    s = "{0:.3g}".format(x)
-    if "e" in s:
-        mantissa, exp = s.split('e')
-        return mantissa + '\\times 10^{' + exp + '}'
-    else:
-        return s
-
-
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument('-f', '--fitting', required=False, default=False, type=bool, dest="fitting")
+    parser.add_argument('-a', '--age', required=True, type=float, dest="age")
+    parser.add_argument('-b', '--branches', required=True, type=int, dest="branches")
     parser.add_argument('-o', '--output', required=True, type=str, dest="output")
+    parser.add_argument("--y_param_key", required=False, action='append', type=lambda kv: kv.split(":"), dest='y_param_dict')
     parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
     args = parser.parse_args()
-    dict_df = {filepath: {k: v[0] for k, v in pd.read_csv(filepath + '.tsv', sep='\t').items()} for filepath in
-               args.input if os.path.isfile(filepath + '.tsv')}
+    args.y_param_dict = dict(args.y_param_dict) if (args.y_param_dict is not None) else dict()
     fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
-    col = "<dN/dN0>"
-    dico_files, dico_node = dict(), dict()
-    for filepath in dict_df.keys():
-        prefix = filepath.split("/")[-2]
-        if prefix not in dico_files:
-            dico_files[prefix] = list()
-        dico_files[prefix].append(filepath)
[…]

### 2020-04-02 scripts/scaling_merge.py

@@ -2,12 +2,14 @@
 import os
 import argparse
 import pandas as pd
+from glob import glob
 import numpy as np
+from collections import defaultdict
 from scipy.optimize import curve_fit
 import statsmodels.api as sm
 import matplotlib
 
-label_size = 16
+label_size = 24
 my_dpi = 128
 nbr_points = 1000
 matplotlib.rcParams['ytick.labelsize'] = label_size
@@ -15,6 +17,14 @@ matplotlib.rcParams['xtick.labelsize'] = label_size
 matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+
+
+def is_float(x):
+    try:
+        float(x)
+        return True
+    except ValueError:
+        return False
 
 
 def exponential(_t, _r, _a, _b):
@@ -34,29 +44,46 @@ if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
     parser.add_argument('-f', '--fitting', required=False, default=False, type=bool, dest="fitting")
     parser.add_argument('-o', '--output', required=True, type=str, dest="output")
-    parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
+    parser.add_argument('-i', '--input', required=True, type=str, dest="input")
+    parser.add_argument('-d', '--distrib', required=False, default=False, type=bool, dest="distrib")
     args = parser.parse_args()
-    dict_df = {filepath: {k: v[0] for k, v in pd.read_csv(filepath + '.tsv', sep='\t').items()} for filepath in
-               args.input if os.path.isfile(filepath + '.tsv')}
-    cols = ["<dN/dN0>"]
+    dict_df = {filepath: {k: v[0] for k, v in pd.read_csv(filepath + ".tsv", sep='\t').items()} for filepath in
+               sorted(glob("{0}/*_exp".format(args.input)))}
     fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
-    for col in cols:
-        dico_node = dict()
+    for col in ["DFE", "mut-ΔΔG"]:
+        if not args.distrib: continue
+        hist = defaultdict(float)
         for filepath in dict_df.keys():
-            if not os.path.isfile(filepath + ".substitutions.tsv"):
-                continue
+            distrib_file = filepath + col + "distrib.tsv"
+            if not os.path.isfile(distrib_file): continue
[…]

### 2020-04-02 scripts/theoretical_equilibrium.py

@@ -78,7 +78,7 @@ if __name__ == '__main__':
     plt.plot(x_range, [10 * scaled_sel_coeff(i, args.alpha) for i in x_range],
              linestyle="--", color=line.get_color(), linewidth=3,
              label="S: " + label.format(args.alpha, args.gamma, args.n, 10 * args.population_size))
-    plt.legend(fontsize=label_size)
+    plt.legend(fontsize=15)
     plt.xlim((x_min, x_max))
     plt.ylim((y_min, y_max))
     plt.tight_layout()

### 2020-04-02 manuscript/manuscript.tex
Results in pdf. Manuscript with supp. mat.

@@
 
-\include{new-commands}
+\include{notations}
 \pgfplotsset{every axis/.append style={line width=1pt}}
@@
 \abstract{
-The substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) had been theoretically expected to decrease with effective population size ($\Ne$).
-Mechanistically, populations with high $\Ne$ would have stronger purifying selection, hence lower $\dnds$, due to the decrease of random drift.
+The quasi-neutral theory of evolution assert that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
+One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
+In molecular sequence, this effect translate in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
 Such theoretical prediction had been observed in empirical data across many clades.
@@
 Together, assumptions about the mapping of sequence to fitness can display a variety of behaviours in the $\dnds$ responses to changes in $\Ne$.
-Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
-We provide a simple mechanistic model that can display a wide range of behaviours depending on the parametrization.
-Our result show that models assuming site-specific fitness landscapes (without epistasis) is at risk of over-estimating the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$, but paradoxically also over-estimating the equilibrium $\dnds$ response to changes in $\Ne$.
+Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
+We apply our framework in the case of fitness proportional to the probability of protein folding.
+Our compact theoretical results match more complex simulations using 3d structure of proteins.
+We assert that models based on the probability of folding are at odd with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
+However, fitness function based on non-specific interactions between proteins match the empirical data.
+We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
 }
 
-\section{Introduction}
+\section*{Introduction}
 Differences in DNA sequences of a gene in separate species are due to the particular history of DNA substitutions along each specie's respective lineage.
@@
 This question is important if we want to extract changes in effective population size from substitution history, as well as to determine the fitness landscape under which protein-coding DNA sequences are evolving.
+We provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
+We apply our framework in the case of fitness proportional to the probability of protein folding.
+Our compact theoretical results match more complex simulations using 3d structure of proteins.
+We assert that models based on the probability of folding are at odd with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
+However, fitness function based on non-specific interactions between proteins match the empirical data.
+We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
 
-\section{Results}
-\subsection{$\bm{\dnds}$ at equilibrium as a function of $\bm{\Ne}$}
+\section*{Results}
+\subsection*{Theoretical approximations}
 
-We provide a toy model of protein evolution under mutation-selection-drift equilibrium, reproducing the observation of more complex models, where we derive mathematical relationship using population-genetic equations.
-This model map the amino-acid sequence to the difference in free energy ($\deltaG$ in kcal/mol) between folded and unfolded state of the whole protein.
-Each amino-acid site in the protein can be in the unstable or stable state.
+We seek to determine the change in $\dnds$ at equilibrium after a change in $\Ne$.
+This model map the amino-acid sequence to a phenotype $\x$, in this case the proportion of unstable sites in the protein.
+Each site can be in one of $\Nstate \geq 2$ states, where only $1$ state is defined the stable state, and $\Nstate - 1$ states are unstable.
+After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$, where $\Nsite$ is the number of sites.
+The Wrightian fitness of phenotype $\x$ is called $f(\x)$
+We demonstrate (See Supplementary Materials) that: 
+\begin{align}
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}},
+\end{align}
+where $\x\eq$ is the phenotype value at equilibrium.
+Thus, the slope of the $\dnds$ elasticity to changes in $\Ne$ is the ratio of first to second derivative of the log-fitness function.
+Moreover, the equilibrium value of $\x\eq$ is given by the equation:
+\begin{align}
+\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}
+\end{align}
+As an example, Wrightian fitness is defined as the probability of our protein to be in the folded state, given by fermi dirac distribution: 
 \begin{equation}
-\deltaG = \deltaGMin + \Nsite * \gamma * \x, 
-\end{equation}
-where $\x$ is the proportion of unstable amino-acid sites in the sequence, $\deltaGMin$ and $\gamma$ are in kcal/mol.
-An optimal protein has a $\deltaG = \deltaGMin$, and pessimal protein has a $\deltaG = \deltaGMin + \Nsite \gamma$.
-Wrightian fitness is defined as the probability of our protein to be in the folded state, given by Boltzmann equation: 
-\begin{equation}
-f(\deltaG) = \dfrac{1}{1 + e^{\beta \deltaG }}, 
-\end{equation}
-where the fixed parameter $\beta=1.686$ mol/kcal at room temperature.\\
-
-We compare our model to the original studies \cite{Williams2006, Goldstein2011, Pollock2012}, where free energy of the folded state is computed using the $3$D structure of the folded state and pair-wise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
-The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
-
-\begin{figure*}[htb!]
-\begin{mdframed}
-\centering
-	\includegraphics[width=0.6\textwidth] {artworks/GoldsteinPollock/dnds_event_tot.pdf}\\
-	\includegraphics[width=0.6\textwidth] {artworks/N300/dnds_event_tot.pdf}\\
-	\includegraphics[width=0.6\textwidth] {artworks/N300Variance/dnds_event_tot.pdf}\\
-	\caption{
-		\textbf{$\bm{\dnds}$ response to change in $\bm{\Ne}$}.
-		Top panel, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ in the model of Goldstein \& Pollock, but is not independent as claimed in \cite{Goldstein2013}.
-		For each population size, $200$ simulation were performed and the average (blue solid line) and standard deviation (blue filling) are shown.
-		This weak correlation matches the prediction of our toy model that the linear relation (blue dashed line) has a slope equal to $(\beta  \Nsite \gamma)^{-1} \simeq 0.0019$.
-		Middle panel, $\alpha=-118$ and $\Nsite \gamma=300\E[\gamma]=300$.
-		Bottom panel, $\alpha=-118$ and each site has it's own normally distributed $\gamma$ with mean $1$ and standard deviation of $1$.
-		The original $3$D model of Goldstein \& Pollock and our simple toy model provide qualitatively and quantitatively the same linear relationship between $\dnds$ and log-$\Ne$, whenever the parameters $\alpha$ and $\Nsite \gamma$ are taken to match the average values observed in the original $3$D model.
-		\label{fig:GoldsteinVsToy}
-	}
-\end{mdframed}
-\end{figure*}
-
-\begin{figure*}[htb!]
-\begin{mdframed}
-	\centering
-	\includegraphics[width=0.8\textwidth] {artworks/ScaleAlphaGammaStdN300/SimuStab/dnds_event_tot.pdf}\\
-	\caption{
-		\textbf{$\bm{\dnds}$ linear response to change in $\bm{\Ne}$}.
-		The parameter $\Nsite \gamma$ is controlling the $\dnds$-$\Ne$ relationship, decreasing $\Nsite \gamma$ implies a stronger negative linear relationship between $\dnds$ and log-$\Ne$.
-The intersect is a function of both $\alpha$ and $\Nsite \gamma$.
-	}
-	\label{fig:EqdndsNe}
-\end{mdframed}
-\end{figure*}
-
-Protein sequence evolution are simulated under an origin-fixation model \cite{McCandlish2014}, where the rate of substitution of a new mutation with selection coefficient $s$ is:
-\begin{equation}
-p_{\text{fix}}(s) = \mu \dfrac{4 \Ne s}{1 + e^{- 4 \Ne s }}, 
-\end{equation}
-where $\mu$ is the mutation rate.\\
-
-Simulations of protein evolution with both models display the same $\dnds$-$\Ne$ relationship, whenever the parameters $\alpha$ and $\Nsite \gamma$ of our model are set to match the average values of the $3$D model (see Figure \ref{fig:EqdndsNe}, panel A).
-In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
-The parameter $\Nsite \gamma=300.0$ kcal/mol, since the average observed increase in $\deltaG$ per destabilizing mutation is $\deltadeltaG = 1.0$ kcal/mol per site and our protein is $300$ sites \cite{Zeldovich2007}.
-The original works claimed that under the $3$D model, $\dnds$ is independent of $\Ne$ because $\deltaG$ is an equimutable parameter.
-We argue instead that $\dnds$ is approximatively linear with log-$\Ne$, and that $\deltaG$ is not equimutable (see Figure \ref{fig:EqdndsNe}, panel B).
-More precisely, this mathematical equation can be derived (see Supplementary Materials) and the slope is inversely proportional to $\Nsite \gamma$ :
-\begin{equation}
-\dnds \propto -\dfrac{1}{\beta \Nsite \gamma} \ln(\Ne).
-\end{equation}
-Meaning that under a value of $\Nsite \gamma=300.0$ kcal/mol, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is $0.002$.
-Thus if we consider $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations, the slope is determined solely by the number of sites in the protein. 
-However, decreasing both $\alpha$ and $\gamma$ together implies a strong negative relationship between $\Ne$ and $\dnds$, as predicted under the nearly-neutral theory of evolution.
-Moreover, by reducing the number of sites of the protein, thus by effectively reducing the potential number of sites that can compensate for the change in $\Ne$, the $\dnds$ at equilibrium after changes in $\Ne$ is stronger \textit{Necessary mechanistic explanation for this fact} (see Figure \ref{fig:EqdndsNe}).
+f(\x) = \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite * \gamma * \x)}}, 
+\end{equation} .
+where $\deltaGMin$ and $\gamma$ are in kcal/mol, and the fixed parameter $\beta=1.686$ mol/kcal at room temperature.
+Thus, $\deltaGMin < 0$ is the difference in free energy between folded and unfolded state when all sites are stable.
+$\Nsite \gamma > 0$ is the expected change in free energy (between folded and unfolded states) when all sites are unstable.
 
@@
 		\begin{axis}[
+		width=0.65\textwidth,
+		height=\axisdefaultheight,
 		xlabel={$\x$},
@@
 		legend entries={$\ln(\frac{1 - \x}{\x})$,
-		$S_{\deltaGMin = -118,\Nsite \gamma = 300,\Ne = 10^{4}}(\x)$,
-		$S_{\deltaGMin = -118,\Nsite \gamma = 300,\Ne = 10^{5}}(\x)$,
-		$S_{\deltaGMin = -16.7,\Nsite \gamma = 30,\Ne = 10^{4}}(\x)$,
-		$S_{\deltaGMin = -16.7,\Nsite \gamma = 30,\Ne = 10^{5}}(\x)$,
+			$S_{\deltaGMin = -118, \gamma = 1,\Ne = 10^{4}}(\x)$,
+			$S_{\deltaGMin = -118, \gamma = 1,\Ne = 10^{5}}(\x)$,
+			$S_{\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{4}}(\x)$,
+			$S_{\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{5}}(\x)$,
 		},
@@
 \end{figure*}
+With such fitness function, we have (See Supplementary Materials) :
+\begin{equation}
+\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma}.
+\end{equation}
+which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
+Meaning that under a value of $\Nsite=300$ sites, $\gamma=1.0$ kcal/mol, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is $0.002$.
 
-\subsection{Epistasis determines the time to relaxation}
+\subsection*{Simulations confirmation}
+We compare our model to the original studies \cite{Williams2006, Goldstein2011, Pollock2012}, where free energy of the folded state is computed using the $3$D structure of the folded state and pair-wise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
+The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
+
+\begin{figure*}[htb!]
+\begin{mdframed}
+\centering
+\includegraphics[width=0.8\textwidth] {artworks/SimuFold-Elasticity.pdf}
+\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Elasticity.pdf}
+	\caption{
+		\textbf{$\bm{\dnds}$ elasticity to change in $\bm{\Ne}$}.
+		For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+		In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally \cite{Goldstein2013}.
+		This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta  \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.
+		Bottom panel, The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+		Moreover, with the Grantham model, the $\dnds$ matches the empirical 3D model of Golstein \& Pollock and the theoretical prediction.
+		\label{fig:GoldsteinVsToy}
+	}
+\end{mdframed}
+\end{figure*}
+Simulations of protein evolution with both models display the same $\dnds$-$\Ne$ relationship, whenever the parameters $\alpha$ and $\gamma$ of our model are set to match the average values of the $3$D model (see Figure \ref{fig:EqdndsNe}).
+In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
+The parameter is set $\gamma=1.0$ kcal/mol, the average observed increase in $\deltaG$ per destabilizing mutation \cite{Zeldovich2007}.
+The original works claimed that under the $3$D model, $\dnds$ is independent of $\Ne$ because $\deltaG$ is an equimutable parameter.
+We argue instead that $\dnds$ is approximatively linear with log-$\Ne$, and that $\deltaG$ is not equimutable (see Figure \ref{fig:EqdndsNe}).
+Thus if we consider $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations, the slope is determined solely by the number of sites in the protein.
+
+\subsection*{Epistasis determines the time to relaxation}
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{artworks/RelaxationScaleAlphaGammaN100/14_1_dNdN0.png}
-	\includegraphics[width=\textwidth]{artworks/RelaxationScaleAlphaGammaN100/100_1_dNdN0.png}\\
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Alpha-Gamma.pdf}
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-DFE-Profile-N50.pdf}
 	\caption{
 		\textbf{$\bm{\dnds}$ relaxation time to changes in $\bm{\Ne}$}.
-		Panel A.
-Our result show that with low epistasis, the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
-		Panel B. With high epistasis, the relaxation time is short.
+		Top panel. $\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+		Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+		The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+		$\beta=1.686$, $\gamma=-10$ for all simulations.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+		Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
+		Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+		Increasing $\Nsite$ implies a higher rate of relaxation.
+		Panel B.
+In context of a fixed fitness landscape, where each amino-acid have different fitnesses (site-specific profile), the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
+In the context of a distribution of fitness effects, the relaxation time is non-existent.
 	}
@@
 
-\begin{figure*}[htb!]
-\begin{mdframed}
-	\centering
-	\includegraphics[width=\textwidth]{artworks/SimuExponentialDecay.png}\\
-	\caption{
-		\textbf{$\bm{\dnds}$ relaxation time under different fitness models.}.
-		Panel A.
-In context of a fixed fitness landscape, where each amino-acid have different fitnesses (site-specific profile), the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
-		Panel B. In the context of a distribution of fitness effects, the relaxation time is non-existent.
-	}
-	\label{fig:relaxProfileDFE}
-\end{mdframed}
-\end{figure*}
-
-\section{Discussion}
+\section*{Discussion}
 
@@
 Our model does not display site variation in $\dnds$, which is a strong limitation, as it has observed in most protein \cite{Echave2017}.
-\section{Materials \& Methods}
+\section*{Materials \& Methods}
 
-\subsection{Simulation with protein folding probability}
+Protein sequence evolution are simulated under an origin-fixation model \cite{McCandlish2014}., where the rate of substitution of a new mutation with selection coefficient $s$ is:
+\begin{equation}
+p_{\text{fix}}(s) = \mu \dfrac{4 \Ne s}{1 + e^{- 4 \Ne s }}, 
+\end{equation}
+where $\mu$ is the mutation rate.
+
+\subsection*{Simulation of protein folding with an additive model of free energy}
 \label{MatMet:folding}
@@
 
-\subsection{Simulation with independent fitness profiles}
+\subsection*{Simulation with independent fitness profiles}
 A fitness profile give a fitness for each amino-acid (vector of size $20$).
@@
 where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \cite{Bloom2017}.
-\begin{center}
-	\includegraphics[width=165mm] {artworks/ModelSimuDiv.pdf}
-\end{center}
-
 The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
 
-\subsection{Simulation with distribution of fitness effects}
+\subsection*{Simulation with distribution of fitness effects (DFE)}
 The selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
@@
 
-\subsection{$\bm{\dnds}$ along the simulation}
+\subsection*{$\bm{\dnds}$ along the simulation}
 From the set of mutant $\setNeighbors$ that are one nucleotide away from $\ci$, we define the subsets $\setNonSynNeighbors$ and $\setSynNeighbors$ that are respectively the set of non-synonymous and synonymoys mutants, where  $\setNonSynNeighbors \cup \setSynNeighbors = \setNeighbors$.
@@
 \end{align}
-\subsection{Reproducibility}
+\subsection*{Reproducibility}
 The simulators written in C++ are publicly available under MIT license at \url{https://github.com/ThibaultLatrille/SimuEvol}.

### 2020-04-02 manuscript/supp-mat.tex
Results in pdf. Manuscript with supp. mat.

@@
 
-\include{new-commands}
+\include{notations}
 
@@
 \part*{Supplementary materials}
-
-\section*{Genotype to phenotype map}
+\tableofcontents
+\section{$\bm{\omega}$ elasticity after a change in $\bm{\Ne}$}
+\subsection{Genotype to phenotype map}
 Define $\Nsite$ as the number of sites in the genotype sequence.
@@
 
-\section*{Selection coefficient}
+\subsection{Selection coefficient}
 $s(\x, \dx)$ is the selection coefficient of an effect $\dx$ if the current phenotype is $\x$:
@@
 s(\x, \dx) &= \frac{f(\x + \dx) - f(\x)}{f( \x)}, \\
- &\simeq \frac{1}{f(\x)}\frac{\der f( \x)}{\der \x} \dx, \\
- &\simeq \frac{\der \ln (f(\x))}{\der \x} \dx, \label{eq:s_from_fitness}
+ &\simeq \frac{1}{f(\x)}\frac{ \partial f(\x) }{\partial \x} \dx, \\
+ &\simeq \frac{ \partial \ln f(\x) }{\partial \x} \dx, \label{eq:s_from_fitness}
 \end{align}
@@
 
-\section*{Probability of fixation}
+\subsection{Probability of fixation}
 The probability of fixation of a mutation with effect $\dx$, for a resident phenotype $\x$ is :
@@
 \end{align}
-\section*{Equilibrium phenotype}
+\subsection{Equilibrium phenotype}
 At equilibrium phenotype $\x\eq$, the expected selection coefficient of mutation that reached fixation must be $0$:
@@
 where  $\lambda_{K}(\x\eq) = \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1)$.
-\section*{Substitution rate ($\bm{\dnds}$) at equilibrium}
+
+\subsection{Substitution rate bias ($\bm{\dnds}$) at equilibrium}
 The substitution rate of selected mutations relative to neutral mutations ($\dnds$) is by definition:
@@
 	\begin{axis}[
+	width=0.65\textwidth,
+	height=\axisdefaultheight,
 	ylabel={$\dnds\eq$},
@@
 	axis y line=left,
-	legend style={at={(0.6,0.34)},anchor=north west}
+	legend style={at={(0.5,0.4)},anchor=north west}
 	]
@@
 	\begin{axis}[
+	width=0.65\textwidth,
+	height=\axisdefaultheight,
 	ylabel={$\frac{\der \dnds\eq}{\der \x\eq}$},
@@
 	axis y line=left,
-	legend style={at={(0.5,0.95)},anchor=north west}
+	legend style={at={(0.5,0.8)},anchor=north west}
 	]
@@
 \end{center}
-
-\section*{Response of equilibrium after a change in $\bm{\Ne}$}
+Moreover, given that the number of state if large enough $\Nstate \gg 1$, the response in equilibrium $\dnds$ due to change in phenotype can be approximated as:
+\begin{align}
+\frac{ \der \dnds\eq}{\der \x\eq}  & = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}\text{ from eq.
+\ref{eq:approx_exp}}, \\
+& \simeq \frac{2 \lambda_{K}(\x\eq)}{\Nstate} + 1, \\
+& \simeq 1. \label{eq:approx_domega_dx}
+\end{align}
+\subsection{$\bm{\omega}$ elasticity after a change in $\bm{\Ne}$}
 Define the function $G(\x, \Ne )$ as:
 \begin{gather}
-G(\x, \Ne ) \equiv \ln \left( \frac{1 - \x}{\x} \right) + \ln (\Nstate-1) + 4 \Ne s(\x, \dx), 
+G(\x, \Ne ) \equiv \lambda_K (\x\eq) + 4 \Ne s(\x, \dx), 
 \end{gather}
@@
 \Longrightarrow \frac{\partial G(\x\eq, \Ne ) }{\partial \x\eq }\frac{ \der \x\eq}{\der \Ne} + \frac{ \partial G(\x\eq, \Ne )}{\partial \Ne} = 0, \\
-\iff \left[ \frac{\partial \ln (1-\x\eq) }{\partial \x\eq } - \frac{\partial \ln (\x\eq) }{\partial \x\eq}  + 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right]\frac{ \der \x\eq}{\der \Ne} + 4 s(\x\eq, \dx) = 0, \\
-\iff \left[ \frac{1}{(1 - \x\eq) \x\eq} + 4 \Ne \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}\dx \right]\frac{ \der \x\eq}{\der \Ne}  = - 4 \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}\dx \text{ from eq.
+\iff \left[  \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq }  + 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right]\frac{ \der \x\eq}{\der \Ne} + 4 s(\x\eq, \dx) = 0, \\
+\iff \left[  \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + 4 \Ne \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}\dx \right]\frac{ \der \x\eq}{\der \Ne}  = - 4 \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}\dx \text{ from eq.
 \ref{eq:s_from_fitness}}, \\
-\iff 4 \dx \left[ \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2} \right] \Ne \frac{ \der \x\eq}{\der \Ne}  = - 4 \dx \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
-\iff \frac{ \der \x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}  \label{eq:dx_dlnNe}.
+\iff 4 \dx \left[ \frac{1}{4 \dx \Ne } \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2} \right] \Ne \frac{ \der \x\eq}{\der \Ne}  = - 4 \dx \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
+\iff \frac{ \der \x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne} \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}  \label{eq:dx_dlnNe}.
 \end{gather}
@@
 \frac{\der \dnds\eq}{\der \ln (\Ne)} & = \frac{\der \dnds\eq}{\der \x\eq} \frac{ \der \x\eq}{\der \ln (\Ne)}, \\
- &= - \frac{\der \dnds\eq}{\der \x\eq} \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}} \text{ from eq.
+ &= - \frac{\der \dnds\eq}{\der \x\eq} \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne } \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}} \text{ from eq.
 \ref{eq:dx_dlnNe}} \label{eq:dw_dlnNe}.
 \end{align}
-\section*{Phenotype to fitness mapping}
-From a given phenotype $\x$, define the fitness as :
+
+Moreover, with the approximation that $\left| 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right| \gg \left| \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } \right|$, meaning that a change in phenotype causes a higher change in scaled selection coefficient than mutational bias, we have:
 \begin{gather}
- f(\x) = \frac{1}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}, \label{eq:fitness}
+\frac{ \der \x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne} \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}},\\
+\Longrightarrow \frac{ \der \x\eq}{\der \ln (\Ne)}  \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}. \label{eq:approx_dx_dlnNe}
 \end{gather}
-where $\deltaGMin = \Delta G_0 < 0$ is the difference in free energy between stable and unstable protein when all sites are stable.
+Together, these approximations leads to the following elasticity in equilibrium $\dnds$ after change in $\Ne$ as:
+\begin{align}
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\label{eq:approx_dw_dlnNe}
+\end{align}
+
+\section{Models for the log-fitness function}
+
+\subsection{Folded fraction}
+All phenotype-fitness function considered below are log-concave, and as a result, $\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}$ is an decreasing function of $\x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
+More precisely, fitness functions depends on the folded fraction of the protein of interest, which is given by the Fermi-Dirac distribution:
+\begin{eqnarray}
+P_{F} (\x) &=& \frac{1}{1 + e^{\beta(\alpha + \gamma \Nsite \x)}}
+\end{eqnarray}
+where $\x$ is the fraction of destabilizing mutations, each contributing a $\Delta \Delta G = \gamma$, and $\beta = 1 / kT$.
+Thus, $\alpha = \Delta G_0 < 0$ is the difference in free energy between folded and unfolded state when all sites are stable.
 $\Nsite \gamma$ is the expected change in $\Delta G$ when all sites are unstable.
-By definition, $\Nsite \gamma \dx = \sfrac{\Nsite \gamma}{\Nsite} = \Delta \Delta G $\\
+The misfolded fraction is  $P_{U} = 1-P_{F}$.
+In addition, $P_{F}$ is typically close to 1 (or $P_{U} \ll 1$), so that we can use a first-order approximation:
+\begin{eqnarray}
+P_{F}(\x) &=& 1 - P_{U}(\x) \\
+&\simeq& 1 - e^{\beta(\alpha + \gamma \Nsite \x)}
+\end{eqnarray}
+or equivalently
+\begin{eqnarray}
+P_{U}(\x) &\simeq& e^{\beta(\alpha + \gamma \Nsite \x)}
+\end{eqnarray}
+
+\subsection{Fitness equal to folded fraction}
+A first model is to assume that the fitness is equal to the folded fraction \cite{Goldstein2013}:
+\begin{gather}
+f(\x) = \frac{1}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}. \label{eq:fitness}
+\end{gather}
 The derivative of fitness w.r.t to phenotype is:
@@
 \frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
- & = - \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \label{eq:df_dx}.
-\end{align}
-And the second derivative of fitness w.r.t to phenotype is:
-\begin{align}
-\frac{\partial^2 \ln (f(\x))}{\partial {\x}^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \right) \text{ from eq.
-\ref{eq:df_dx}}, \\
- & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{\left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}\right)^2}, \\
- & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \frac{ \partial \ln f(\x) }{\partial {\x}} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}.
+& = - \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \label{eq:df_dx},\\
+& \simeq - \beta \Nsite \gamma \e^{\beta(\deltaGMin + \Nsite \gamma \x)}.
 \end{align}
@@
 \end{gather}
-Using $\Ne=10^4$, $\beta=1.686$, $\deltaGMin = -118$, $\Nsite=300$, $\gamma=1$, we have the following:
+
+Using $\Ne=10^4$, $\beta=1.686$, $\deltaGMin = -118$, $\Nsite=300$, $\gamma=1$, we have the following :
 \begin{center}
@@
 		lambda(\x,\k)= ln(\k-1) + ln((1-\x)/\x);
-	},]
+	}]
 	\begin{axis}[
+	width=0.65\textwidth,
+	height=\axisdefaultheight,
 	xlabel={$\x$},
@@
 	axis y line=left,
-	legend style={at={(0.5,0.95)},anchor=north west}
+	legend style={at={(0.7,0.95)},anchor=north west}
 	]
@@
 \end{center}
-\section*{Approximations}
-We assume that $\Ne \gg 1$, $\deltaGMin + \Nsite \gamma \gg 1$ and $ \gamma \simeq 1$.
-\begin{gather}
--S(\x\eq, \dx) \simeq \lambda_{\Nstate}(\x\eq)\text{ from eq. \ref{eq:equilibrium_lambda}}, \\
-\iff s(\x\eq, \dx) \simeq -\frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne}, \\
-\iff \frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:s_from_fitness}}, \label{eq:approx_df_dx}\\
-\iff \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)} } \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq.
+Where in this example we can visually appreciate that the a change in phenotype causes a higher change in scaled selection coefficient than mutational bias (eq. \ref{eq:approx_dx_dlnNe}).
+And the second derivative of fitness w.r.t to phenotype is:
+\begin{align}
+\frac{\partial^2 \ln (f(\x))}{\partial \x^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \right) \text{ from eq.
 \ref{eq:df_dx}}, \\
-\iff \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \simeq \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \beta \gamma}, \\
-\Longrightarrow \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \ll 1\text{ given that $\Ne \gg 1$ and $ \gamma  \simeq 1$}, \\
-\Longrightarrow \e^{\beta(\deltaGMin + \Nsite \gamma \x)} \ll 1. \label{eq:approx_exp}
-\end{gather}
-Moreover, we can get the following:
-\begin{gather}
-\frac{\der \ln (f(\x))}{\der \x} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx}\text{ from eq. \ref{eq:approx_df_dx}}, \\
-\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)}{4 \Ne \dx } \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}\text{ from eq.
-\ref{eq:df2_dx2}}, \\
-\Longrightarrow \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \frac{\lambda_{\Nstate}(\x\eq)\beta \Nsite \gamma}{4 \Ne \dx  }\text{ from eq. \ref{eq:approx_exp}}, \\
-\iff \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \simeq - \lambda_{\Nstate}(\x\eq) (1 - \x\eq)\x\eq \beta \Nsite \gamma \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq}, \\
-\Longrightarrow \left| \frac{ \partial^2 \ln f(\x) }{\partial {\x}^2} \right| \gg \left| \frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} \right|\text{ given that $\Nsite \gamma \gg 1$}.
-\label{eq:approx_df2_df} \\
-\end{gather}
-Together, these approximations leads to the following response in equilibrium phenotype at change in $\Ne$ as:
+ & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{\left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}\right)^2}, \\
+ & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \frac{ \partial \ln f(\x) }{\partial \x} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}, \\
+ & \simeq \beta \Nsite \gamma \frac{ \partial \ln f(\x) }{\partial \x} 
+\end{align}
+Finally, $\omega$ elasticity after a change in $\Ne$ is simply:
 \begin{align}
-\frac{ \der \x\eq}{\der \ln (\Ne)} & = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne  (1 - \x\eq) \x\eq} + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\text{ from eq.
-\ref{eq:dw_dlnNe}}, \\
-& \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\text{ from eq. \ref{eq:approx_df2_df}}, \\
-& \simeq - \frac{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x\eq)}}{\beta \Nsite \gamma}\text{ from eq. \ref{eq:df2_dx2}}, \\
-& \simeq - \frac{1}{\beta \Nsite \gamma}\text{ from eq. \ref{eq:approx_exp}}. \label{eq:approx_dx_dlnNe}
+\frac{ \der \dnds\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma} \text{from eq. \ref{eq:df2_dx2} and \ref{eq:approx_domega_dx}},
 \end{align}
-Moreover, given that the number of state if large enough $\Nstate \gg 1$, the response in equilibrium $\dnds$ due to change in phenotype can be approximated as:
+which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
+This model, however, does not express the fact that selection is typically stronger for proteins characterized by higher levels of expression.
+
+\subsection{Selective cost proportional to amount of misfolded protein}
+
+A slight variation is to assume that the selective cost itself is proportional to the total amount of misfolded protein. For a given protein with expression level $y$:
+\begin{eqnarray}
+\ln f(\x) = - A y P_{U} (\x)
+\end{eqnarray}
+where $A$ is the cost per misfolded macromolecule. Then, 
+\begin{eqnarray}
+\frac{\partial \ln (f(\x))}{\partial \x} &\simeq& -A \beta \gamma \Nsite y  e^{\beta(\alpha + \gamma \Nsite \x)}
+\end{eqnarray}
+or equivalently
+\begin{eqnarray}
+4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite \x)}
+\end{eqnarray}
+Thus, under this model, an increase in the effective population size $\Ne$ or the expression level $y$ both have the same effect on the molecular evolutionary process undergone by the protein.
+Moreover, $\omega$ elasticity after a change in $\Ne$ is the same as before:
 \begin{align}
-\frac{ \der \dnds\eq}{\der \x\eq}  & = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}\text{ from eq.
-\ref{eq:approx_exp}}, \\
-& \simeq \frac{2 \lambda_{K}(\x\eq)}{\Nstate} + 1, \\
-& \simeq 1. \label{eq:approx_domega_dx}
+\frac{ \der \dnds\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma}.
 \end{align}
-Finally,
+\subsection{Translational errors}
+
+An other variant account for translational errors.
+Translational errors occur at a rate $\rho$ per residue.
+These errors contribute additional destabilizing mutations, each with effect size $\delta x = 1/n$.
+The total number of translational errors per macromolecule is approximately Poisson distributed:
+\begin{eqnarray}
+\pi_k &=& e^{-\rho \Nsite} \frac{(\rho \Nsite)^k}{k!}
+\end{eqnarray}
+and the total selective cost is now an average over all possible values of $k$:
+\begin{eqnarray}
+\ln f(\x) &=& - A y \sum_k \pi_k e^{\beta(\alpha + \gamma\Nsite \x + \gamma k)}
+\\
+&=& - A y  e^{\beta(\alpha + \gamma\Nsite \x)} \sum_k e^{-\rho n} \frac{(\rho n)^k}{k!} e^{\beta \gamma k}
+\\
+&=& - A y  e^{\beta(\alpha + \gamma\Nsite \x) + \rho \Nsite (e^{\beta \gamma} -1 )}
+\\
+& \simeq & - A y  e^{\beta(\alpha + \gamma\Nsite \x) + \rho \beta \gamma \Nsite}
+\\
+& =& - A y  e^{\beta(\alpha + \gamma \Nsite (x + \rho))}
+\end{eqnarray}
+In words, the fitness function is the same the previous model, except that the trait $x$ (fraction of destabilizing mutations) is shifted by $\rho$, the mean fraction of additional mutations contributed by translation errors.
+This additional factor is independent of $x$, and as a result, the scaled selection strength is essentially the same, up to a proportionality constant (contributed by the shift):
+\begin{eqnarray}
+4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite (x + \rho))}
+\\ &\propto& -4 \Ne y e^{\beta(\alpha + \gamma\Nsite \x)}
+\end{eqnarray}
+Moreover, $\omega$ elasticity after a change in $\Ne$ is again the same as before:
 \begin{align}
-\frac{ \der \dnds\eq}{\der \ln (\Ne)}  & = \frac{ \der \dnds\eq}{\der \x\eq}  \frac{ \der \x\eq}{\der \ln (\Ne)}, \\
-  & \simeq - \frac{1}{\beta \Nsite \gamma} \text{from eq. \ref{eq:approx_dx_dlnNe} and \ref{eq:approx_domega_dx}}.
+\frac{ \der \dnds\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma}.
 \end{align}
 
-\section*{Probability of folding as in Goldstein \& Pollock}
-We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
-From a DNA sequence $\ci$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}\left(\ci\right)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
-\begin{equation}
-G_{\mathrm{F}}\left(\ci\right) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\ci(\site), \ci(r) \right),
-\end{equation}
-where $I(a,b)$ is the pair-wise contact energies between amino-acid $a$ and $b$, using contact potentials estimated by Miya-zawa and Jernigan, and $\mathcal{N}(\site)$ are the neighbor residues of site $\site$ (closer than $7\angstrom$) in the $3$D structure.\\
+\subsection{Cost-benefit argument}
 
-The free energy of unfolded states $G_{\mathrm{U}}\left(\ci\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
-\begin{equation}
-G_{\mathrm{U}}\left(\ci\right) = \langle G\left(\ci\right) \rangle - kT \ln (1.0\mathrm{E}^{160}) - \dfrac{2 \left[ \langle G\left(\ci\right)^2 \rangle - \langle G\left(\ci\right) \rangle^2\right] }{kT}
-\end{equation}
-where the average $\langle . \rangle$ runs other the $55$ decoy $3$D structures, and $k$ is the Boltzmann constant and $T$ the temperature in Kelvin.\\
-
-From the energy of folded and unfolded states, we can compute the difference in free energy between the states:
-\begin{equation}
-\deltaG\left(\ci\right) = G_{\mathrm{F}}\left(\ci\right) - G_{\mathrm{U}}\left(\ci\right)
-\end{equation}
-
-\begin{center}
-	\includegraphics[width=165mm] {artworks/ModelSimuFold.pdf}
-\end{center}
-
-\begin{mdframed}
-\begin{center}
-	\includegraphics[width=0.6\textwidth] {artworks/GoldsteinPollock/sub-DeltaG-mean.pdf}\\
-	\textbf{$\bm{\deltaG}$ response to change in $\bm{\Ne}$}.
-	$\deltaG$ at equilibrium (blue solid line) is linearly (blue dashed line) dependent on log-$\Ne$, as predicted in the model of Goldstein \& Pollock, with a slope equal to $\beta^{-1}$.
-	\label{fig:EqdndsNe}
-\end{center}
-\end{mdframed}
-
-\section*{Equimutability}
-We define $\x$ as the current phenotype and $\dx$ as the phenotypic effect of a mutation.
-$\s \left( \x, \dx \right)$ is the selection coefficient of a phenotypic effect $\dx$ if the current phenotype is $\x$:
-\begin{equation*}
-\s \left( \x, \dx \right) = \dfrac{w \left( \x + \dx \right) - w \left( \x \right)}{w \left( \x \right)} \simeq w'\left( \x \right) \dx, 
-\end{equation*}
-where $w \left( \x \right)$ is the Wrightian fitness of phenotype $x$.
-We also define $\rho_{\x}(\dx)$ as the distribution of phenotypic effects, at phenotype $\x$.
-At equilibrium, the expected selection coefficient $ 4 \Ne \s \left( \x, \dx \right)$ of mutations that reached fixation must be $0$:
-\begin{align*}
-0 & = \E_{\dx} \left[ 4 \Ne \s\left( \x, \dx \right) \dfrac{ 4 \Ne \s\left( \x, \dx \right) }{1 - \e^{4\Ne\s\left( \x, \dx \right)}} \right] \\ & 
-\simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \x \right)  \right]^2}{1 - \e^{4\Ne f'\left( \x \right) \dx} }  \dx^2  \rho_{\x}(\dx) \der \dx 
-\end{align*}
-And the substitution rate is :
-\begin{align*}
-\omega(\Ne) & = \E_{\dx} \left[ \dfrac{ 4 \Ne \s \left( \x, \dx \right)}{1 - \e^{4\Ne \s \left( \x, \dx \right)}} \right] \simeq \E_{\dx} \left[ \dfrac{ 4 \Ne f'\left( \x \right) \dx}{1 - \e^{4\Ne f'\left( \x \right) \dx}} \right] \\
-& \simeq \int_{-\infty}^{\infty} \dfrac{4 \Ne f'\left( \x \right) \dx}{1 - \e^{4\Ne f'\left( \x \right) \dx } } \rho_{\x}(\dx) \der \dx \\
-\end{align*}
-If $\rho_{\x}(\dx) = \rho(\dx)$ is independent of $\x$:
-\begin{align*}
-0 & \simeq \int_{-\infty}^{\infty} \dfrac{ \left[ 4 \Ne f'\left( \x \right)  \right]^2}{1 - \e^{4\Ne f'\left( \x \right) \dx} } \dx^2 \rho(\dx)  \der \dx \\
-\Longrightarrow & \ 4 \Ne f'\left( \x \right) = K \in \mathbb{R}^{*} \text{ and } \s\left( \x, \dx \right) \simeq \dfrac{K \dx}{4\Ne} \\
-\Longrightarrow & \ \omega( \Ne ) \simeq \int_{-\infty}^{\infty} \dfrac{ K  }{1 - \e^{K \dx } } \dx \rho(\dx)  \der \dx  \text{ is independent of } \Ne
-\end{align*}
-
-\section*{Model of non-specific interactions}
+This models assumes that the :
+\begin{enumerate}
+	\item the expression level is regulated so that the total amount of \emph{functional} macromolecules is maintained at a target level $y$;
+	\item the log-fitness is proportional to the ratio of the \emph{total} cost of expression over the benefit contributed by the protein (Beaulieu et al).
+\end{enumerate}
+Specifically, the protein is assumed to be regulated so as to reach a level of expression of functional proteins of $y$, and contributes a total benefit $B$ (which depends on its specific function).
+Given that only a fraction $P_{F}(\x) = 1-P_{U}(\x)$ of the total amount of protein expressed by the cell is functional, the total cost of expression $C$ is then equal to:
+\begin{eqnarray}
+C(\x) &=& \frac{y}{P_{F}(\x)}
+\\ &\simeq& y (1 + P_{U}(\x))
+\end{eqnarray}
+Then, the log-fitness is given by:
+\begin{eqnarray}
+\ln f (\x) & = & -A \frac{y}{B} \left( 1 + e^{\beta(\alpha + \gamma\Nsite \x)} \right)
+\\
+&=& -by (1 + e^{\beta(\alpha + \gamma\Nsite \x)})
+\end{eqnarray}
+where $b = A /B$.
+Compared to model 2 and 3, the log-fitness now has an additional term that depends on the target expression level $y$, but not on trait $x$.
+The scaled strength of selection on mutations affecting $x$ has thus the same functional form as for the two previous models:
+\begin{eqnarray}
+4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma\Nsite \x)}
+\end{eqnarray}
+Alternative cost-exp models could also be used, allowing for a non-linear cost function for expression or for some elasticity of the realized equilibrium expression level, as a function of the number of mutations (Gout et al).
+Under these models, the strength of selection is still expected to be an increasing function of $y$, although not linear, e.g.
+\begin{eqnarray}
+4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne g(y) e^{\beta(\alpha + \gamma\Nsite \x)}
+\end{eqnarray}
+where $g$ is some function of $y$.
+Moreover, $\omega$ elasticity after a change in $\Ne$ is again the same as before:
+\begin{align}
+\frac{ \der \dnds\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma}.
+\end{align}
+\section{Model of protein-protein interactions}
 
@@
 \\
-&=& \alpha + \gamma n (x_i + x_j)
+&=& \alpha + \gamma \Nsite (x_i + x_j)
 \end{eqnarray}
@@
 \begin{eqnarray}
-\ln f &=& -b \sum_i \epsilon_i
+\ln f(\x) &=& -b \sum_i \epsilon_i
 \end{eqnarray}
@@
 
-\subsection*{Mean field, weak-interaction limit}
+\subsection{Mean field, weak-interaction limit}
 
@@
 \begin{eqnarray}
-C &=& [i] + (m-1) [i] \frac{C}{C_0} \, e^{\beta (\alpha + \gamma n (\bar x + x_i))}
+C &=& [i] + (m-1) [i] \frac{C}{C_0} \, e^{\beta (\alpha + \gamma \Nsite (\bar x + x_i))}
 \end{eqnarray}
@@
 \begin{eqnarray}
-[i] &=& \frac{C} {1 + (m-1) \frac{C}{C_0} \, e^{\beta (\alpha + \gamma n (\bar x + x_i))}}
-\\ &\simeq& C \left( 1 - m \frac{C}{C_0} \, e^{\beta (\alpha + \gamma n (\bar x + x_i))} \right)
+[i] &=& \frac{C} {1 + (m-1) \frac{C}{C_0} \, e^{\beta (\alpha + \gamma \Nsite (\bar x + x_i))}}
+\\ &\simeq& C \left( 1 - m \frac{C}{C_0} \, e^{\beta (\alpha + \gamma \Nsite (\bar x + x_i))} \right)
 \\ &=&
@@
 \begin{eqnarray}
-\epsilon_i  &=& m \frac{C}{C_0} \, e^{\beta (\alpha +\gamma n (\bar x + x_i))}
+\epsilon_i  &=& m \frac{C}{C_0} \, e^{\beta (\alpha +\gamma \Nsite (\bar x + x_i))}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\kappa_i &=& b \frac{d \epsilon_i} {d x_i} 
+\kappa_i &=& b \frac{\der\epsilon_i} {\der x_i} 
 \\ &=&
-b \beta \gamma n m \frac{C}{C_0} \, e^{\beta (\alpha +\gamma n (\bar x + x_i))}
+b \beta \gamma \Nsite m \frac{C}{C_0} \, e^{\beta (\alpha +\gamma \Nsite (\bar x + x_i))}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\ln \kappa_i &=& \ln \left( b \beta \gamma n m \frac{C}{C_0}  \right) \, + \, \beta (\alpha +\gamma n \bar x) \, + \, \beta \gamma n x_i
+\ln \kappa_i &=& \ln \left( b \beta \gamma \Nsite m \frac{C}{C_0}  \right) \, + \, \beta (\alpha +\gamma \Nsite \bar x) \, + \, \beta \gamma\Nsite \x_i
 \end{eqnarray}
 where only the last term depends on $x_i$.
-Finally, applying the main result of this work to the present case allows us to express the elasticity of $\omega$ as a function of $N_e$ as:
+Finally, applying the main result of this work to the present case allows us to express the elasticity of $\omega$ as a function of $\Ne$ as:
 \begin{eqnarray}
-\chi &=& \frac{d \omega} {d \ln N_e} 
-\\ &=&  2(\lambda - 1) \, \frac{d \ln \kappa_i}{d x_i} 
+\chi &=& \frac{\der\omega} {\der \ln \Ne} 
+\\ &=&  2(\lambda - 1) \, \frac{\der\ln \kappa_i}{\der x_i} 
 \\ &=&  2 (\lambda - 1) \, \frac{1}{\beta \gamma n}
@@
 
-\subsection*{Empirical calibration}
+\subsection{Empirical calibration}
 
@@
 
-\section*{Alternative models for the log-fitness function}
+\section{Simulation of protein-coding DNA sequences evolution}
 
-Alternative models of the log-fitness function $\ln f(x)$ are proposed. All of them depend on the folded fraction of the protein of interest, which is given by the Fermi-Dirac distribution:
-\begin{eqnarray}
-p_f &=& \frac{1}{1 + e^{\beta(\alpha + \gamma n x)}}
-\end{eqnarray}
-where $x$ is the fraction of destabilizing mutations, each contributing a $\Delta \Delta G = \gamma$, and $\beta = 1 / kT$.
-Thus, $\alpha = \Delta G_0$ is the free energy of the most stable sequence.
-The misfolded fraction is  $p_m = 1-p_f$.
-In addition, $p_f$ is typically close to 1 (or $p_m << 1$), so that we can use a first-order approximation:
-\begin{eqnarray}
-p_f &=& 1 - p_m
-\\
-&\simeq& 1 - e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
-or equivalently
-\begin{eqnarray}
-p_m &\simeq& e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
+\subsection{Probability of folding using the 3d structure of protein}
 
-For all models, the selective cost $s$ associated with a point mutation with effect size $\delta x$ is given by:
-\begin{eqnarray}
-s \simeq - \kappa \delta x
-\end{eqnarray}
-where
-\begin{eqnarray}
-\kappa &=& - \frac{d \ln f}{d x} 
-\end{eqnarray}
-Thus, $\kappa$ captures the strength of the selective response.
-All fitness functions considered below are log-concave, and as a result, $\kappa$ is an increasing function of $x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
-The scaled selection coefficient is:
-\begin{eqnarray}
-S = 4N_e s  &=& 4 N_e \kappa \delta x
-\end{eqnarray}
-which depends on the scaled strength of selection $4 N_e \kappa$, reflecting both the intrinsic strength of selection and the effect of population size.
+We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
+From a DNA sequence $\ci$ after $t$ substitutions, we compute the free energy of the folded state $G_{\mathrm{F}}\left(\ci\right)$, using the $3$-dimensional structure of the folded state and pair-wise contact energies between neighboring amino-acid residues:
+\begin{equation}
+G_{\mathrm{F}}\left(\ci\right) = \sum_{1 \leq \site \leq \Nsite} \sum_{r \in \mathcal{N}(\site)} I \left(\ci(\site), \ci(r) \right),
+\end{equation}
+where $I(a,b)$ is the pair-wise contact energies between amino-acid $a$ and $b$, using contact potentials estimated by Miya-zawa and Jernigan, and $\mathcal{N}(\site)$ are the neighbor residues of site $\site$ (closer than $7\angstrom$) in the $3$D structure.\\
 
-\subsection*{Model 1: fitness equal to folded fraction}
+The free energy of unfolded states $G_{\mathrm{U}}\left(\ci\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
+\begin{equation}
+G_{\mathrm{U}}\left(\ci\right) = \langle G\left(\ci\right) \rangle - kT \ln (1.0\mathrm{E}^{160}) - \frac{2 \left[ \langle G\left(\ci\right)^2 \rangle - \langle G\left(\ci\right) \rangle^2\right] }{kT}
+\end{equation}
+where the average $\langle . \rangle$ runs other the $55$ decoy $3$D structures, and $k$ is the Boltzmann constant and $T$ the temperature in Kelvin.\\
 
-A first model is to assume that the fitness is equal to the folded fraction (Goldstein):
-\begin{eqnarray}
-f(x) &=& p_f
-\end{eqnarray}
-or:
-\begin{eqnarray}
-\ln f(x) &=& \ln (1 - p_m)
-\\
-& \simeq & - p_m
-\\
-& \simeq & - e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
+From the energy of folded and unfolded states, we can compute the difference in free energy between the states:
+\begin{equation}
+\deltaG\left(\ci\right) = G_{\mathrm{F}}\left(\ci\right) - G_{\mathrm{U}}\left(\ci\right)
+\end{equation}
 
-The strength of selection is then given by:
-\begin{eqnarray}
-\kappa &=& \beta \gamma n e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
+\begin{center}
+	\includegraphics[width=165mm] {artworks/ModelSimuFold.pdf}
+\end{center}
 
-This model, however, does not express the fact that selection is typically stronger for proteins characterized by higher levels of expression.
+\subsection{Simulated $\bm{\dnds}$ elasticity to changes in $\bm{\Ne}$}
 
-\subsection*{Model 2: selective cost proportional to amount of misfolded protein}
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuDfe-Elasticity.pdf}
+\end{center}
+\textbf{$\bm{\dnds}$ elasticity with gamma distributed selection coefficient}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+In the model of gamma distributed fitness effect, $\dnds$ at equilibrium is strongly dependent on log-$\Ne$ where the slope correlation is proportional to the inverse of the shape parameter of the gamma distribution \cite{Welch2008}.
 
-A slight variation on Model 1 is to assume that the selective cost itself is proportional to the total amount of misfolded protein. For a given protein with expression level $y$:
-\begin{eqnarray}
-\ln f = - A y p_m
-\end{eqnarray}
-where $A$ is the cost per misfolded macromolecule. Then, 
-\begin{eqnarray}
-\kappa &=& A \beta \gamma n y  e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
-or equivalently
-\begin{eqnarray}
-4 N_e \kappa &\propto& 4 N_e y e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
-Thus, under this model, an increase in the effective population size $N_e$ or the expression level $y$ both have the same effect on the molecular evolutionary process undergone by the protein.
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuProfile-Elasticity.pdf}
+\end{center}
+\textbf{$\bm{\dnds}$ elasticity with amino-acid fitness profiles}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (relative).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+In the model of site-wise amino-acid fitness profiles taken from \cite{Bloom2017}, $\dnds$ at equilibrium is strongly dependent on log-$\Ne$.
 
-\subsection*{Model 3: translational errors}
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuFold-Elasticity.pdf}
+\end{center}
+\textbf{$\bm{\dnds}$ elasticity with 3D free energy of folding}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed in the original work \cite{Goldstein2013}.
+This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta  \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.
 
-Model 3 is a variant of model 1, accounting for translational errors.
-Translational errors occur at a rate $\rho$ per residue.
-These errors contribute additional destabilizing mutations, each with effect size $\delta x = 1/n$.
-The total number of translational errors per macromolecule is approximately Poisson distributed:
-\begin{eqnarray}
-\pi_k &=& e^{-\rho n} \frac{(\rho n)^k}{k!}
-\end{eqnarray}
-and the total selective cost is now an average over all possible values of $k$:
-\begin{eqnarray}
-\ln f &=& - A y \sum_k \pi_k e^{\beta(\alpha + \gamma n x + \gamma k)}
-\\
-&=& - A y  e^{\beta(\alpha + \gamma n x)} \sum_k e^{-\rho n} \frac{(\rho n)^k}{k!} e^{\beta \gamma k}
-\\
-&=& - A y  e^{\beta(\alpha + \gamma n x) + \rho n (e^{\beta \gamma} -1 )}
-\\
-& \simeq & - A y  e^{\beta(\alpha + \gamma n x) + \rho \beta \gamma n}
-\\
-& =& - A y  e^{\beta(\alpha + \gamma n (x + \rho))}
-\end{eqnarray}
-In words, the fitness function is the same as for Model 2, except that the trait $x$ (fraction of destabilizing mutations) is shifted by $\rho$, the mean fraction of additional mutations contributed by translation errors.
-This additional factor is independent of $x$, and as a result, the scaled selection strength is essentially the same as for Model 2, up to a proportionality constant (contributed by the shift):
-\begin{eqnarray}
-4 N_e \kappa &\propto& 4 N_e y e^{\beta(\alpha + \gamma n (x + \rho))}
-\\ &\propto& 4 N_e y e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Elasticity.pdf}
+\end{center}
+\textbf{$\bm{\dnds}$ elasticity with additive free energy of folding}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$.
+The simulations of our additive free energy model match the theoretical prediction that the slope of the linear relation (dashed line) is equal to $\beta \Nsite \gamma)^{-1} = 0.00198 \simeq = 0.00199$.
+The non-monotony is suspected to be due to the discrete number of sites and states, such that the changes in $\deltaG$ after a mutation is either $-1$, $0$ or $1$.
 
-\subsection*{Model 4: cost-benefit argument}
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Elasticity.pdf}
+\end{center}
+\textbf{Amino-acid distances effect on the elasticity}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+Compared to the simpler model, the elasticity of $\dnds$ to changes in $\Ne$ is monotonous (and smooth) due the realized $\deltaG$ is more continuous.
+Moreover, with the Grantham model, the $\dnds$ is lower and the slope of elasticity lower, closer to the empirical 3D model of Golstein \& Pollock.
 
-This models assumes that the (1) the expression level is regulated so that the total amount of \emph{functional} macromolecules is maintained at a target level $y$; and (2) the log-fitness is proportional to the ratio of the \emph{total} cost of expression over the benefit contributed by the protein (Beaulieu et al).
-Specifically, the protein is assumed to be regulated so as to reach a level of expression of functional proteins of $y$, and contributes a total benefit $B$ (which depends on its specific function).
-Given that only a fraction $p_f = 1-p_m$ of the total amount of protein expressed by the cell is functional, the total cost of expression $C$ is then equal to:
-\begin{eqnarray}
-C &=& \frac{y}{p_f}
-\\ &\simeq& y (1 + p_m)
-\end{eqnarray}
-Then, the log-fitness is given by:
-\begin{eqnarray}
-\ln f & = & -A \frac{y}{B} \left( 1 + e^{\beta(\alpha + \gamma n x)} \right)
-\\
-&=& -by (1 + e^{\beta(\alpha + \gamma n x)})
-\end{eqnarray}
-where $b = A /B$.
-Compared to model 2 and 3, the log-fitness now has an additional term that depends on the target expression level $y$, but not on trait $x$.
-The scaled strength of selection on mutations affecting $x$ has thus the same functional form as for the two previous models:
-\begin{eqnarray}
-4 N_e \kappa &\propto& 4 N_e y e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
-Alternative cost-exp models could also be used, allowing for a non-linear cost function for expression or for some elasticity of the realized equilibrium expression level, as a function of the number of mutations (Gout et al).
-Under these models, the strength of selection is still expected to be an increasing function of $y$, although not linear, e.g.
-\begin{eqnarray}
-4 N_e \kappa &\propto& 4 N_e g(y) e^{\beta(\alpha + \gamma n x)}
-\end{eqnarray}
-where $g$ is some function of $y$.
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Alpha-Elasticity.pdf}
+\end{center}
+\textbf{Effect of $\bm{\alpha}$ on the elasticity}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+The fixed parameters are $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+$\alpha=-118$ are given in the legend.
+Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
+Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
+However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Beta-Elasticity.pdf}
+\end{center}
+\textbf{Effect of $\bm{\beta}$ on the elasticity}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+$\beta=1.686$ are given in the legend.
+Increasing $\beta$ increases the slope of the $\dnds$-$\Ne$ relationship, as predicted in our theoretical model.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Alpha-Gamma-Elasticity.pdf}
+\end{center}
+\textbf{Effect of $\bm{\gamma}$ on the elasticity}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+The fixed parameters are $\Nsite=300$ and $\beta=1.686$. Parameters $\alpha$ and $\gamma$ are given in the legend.
+$\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+The slope of $\dnds$-$\Ne$ relationship is increased with $\gamma$, as predicted by our theoretical model.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-GammaStd-Elasticity.pdf}
+\end{center}
+\textbf{$\bm{\dnds}$ elasticity with additive free energy of folding}.
+$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+The parameters are $\alpha=-118$, $\Nsite=300$, $\beta=1.686$ and each site has it's own gamma distributed $\gamma$ with mean $1$ and standard deviation given in the legend.
+Increasing the variance of $\gamma$ increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites with low $\gamma$ are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
+Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
+However the slope of the $\dnds$-$\Ne$ relationship is not sensibly changed.
+
+\subsection{Simulated relaxation time of $\bm{\dnds}$}
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability.pdf}
+\end{center}
+\textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$}.
+$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line correpond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+$\beta=1.686$, $\gamma=0.2$ and $\alpha=-10$ for all simulations.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+Increasing $\Nsite$ implies a higher $\dnds$ at equilibrium, a lower elasticity of the $\dnds$ to changes in $\Ne$ and a higher rate of relaxation.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Gamma.pdf}
+\end{center}
+\textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$, while correction for $\gamma$}.
+$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+$\beta=1.686$, $\alpha=-10$ for all simulations.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
+Increasing $\Nsite$ implies a higher $\dnds$ at equilibrium, and a higher rate of relaxation.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Alpha.pdf}
+\end{center}
+\textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$, while correction for $\alpha$}.
+$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+$\beta=1.686$, $\gamma=-10$ for all simulations.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+Moreover, $\alpha$ is changed according to $\Nsite$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+Increasing $\Nsite$ implies a lower elasticity of the $\dnds$ to changes in $\Ne$ and a higher rate of relaxation.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Alpha-Gamma.pdf}
+\end{center}
+\textbf{Relaxation time of $\bm{\dnds}$, while correction for $\alpha$ and $\gamma$}.
+$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+$\beta=1.686$, $\gamma=-10$ for all simulations.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
+Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+Increasing $\Nsite$ implies a higher rate of relaxation.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Grantham-Alpha-Gamma.pdf}
+\end{center}
+\textbf{Relaxation time of $\bm{\dnds}$ for the Grantham model}.
+$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+$\beta=1.686$, $\gamma=-10$ for all simulations.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
+Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+Increasing $\Nsite$ implies a higher rate of relaxation.
 

### 2020-04-14 manuscript/figures.tex
Maj manuscript, need more pedagogical writting

\documentclass{beamer}

\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{lmodern}
\usepackage{xfrac, nicefrac}
\usepackage{pgfplots, pgf,tikz}
\usepgfplotslibrary{fillbetween}
\usebackgroundtemplate{\tikz\node[opacity=0]{};}
\setbeamertemplate{footline}[frame number]{}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}{}
\usefonttheme{serif}

\include{notations}
\pgfplotsset{every axis/.append style={line width=1pt}}
\pgfplotscreateplotcyclelist{colors}{LIGHTGREEN\\YELLOW\\RED\\GREEN\\BLUE\\}

\begin{document}
\begin{frame}
\centering
	\begin{tikzpicture}
		\begin{axis}[
		width=\textwidth,
		height=\axisdefaultheight,
		xlabel={$\x$},
		ymin=0.0, ymax=1.0,
		samples=200,
		legend entries={$\ln(\frac{1 - \x}{\x})$,
			${\deltaGMin = -118, \gamma = 1,\Ne = 10^{4}}$,
			${\deltaGMin = -118, \gamma = 1,\Ne = 10^{5}}$,
			${\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{4}}$,
			${\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{5}}$,
		},
		legend cell align=left,
		minor tick num=2,
		axis x line=bottom,
		axis y line=left,
		legend style={at={(0.5,0.95)},anchor=north west}
		]
		\addplot[domain=0.3:0.6,line width=2.0pt, color=BLUE]{ ln((1-x)/x) };
		\addplot[name path=A, domain=0.3:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
		\addplot[name path=B, dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
		\addplot[name path=C,domain=0.3:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-16.71+30*x))};
		\addplot[name path=D,dashed,domain=0.3:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-16.71+30*x))};
		\addplot[fill=RED, opacity=0.2] fill between[ of = A and B];
		\addplot[fill=GREEN, opacity=0.2] fill between[ of = C and D];
		\end{axis}
	\end{tikzpicture}
\end{frame}

\begin{frame}
\centering
	\begin{tikzpicture}[
	declare function={
		omega(\x,\k)= \x*(2*(\x-1)*(ln(\k-1)+ln((1-\x)/\x))/(\k*(\x-1)+1)+ (\k-2)/(\k-1));
	},]
	\begin{axis}[
	width=\textwidth,
	height=\axisdefaultheight,
	ylabel={$\dnds\eq$},
	xlabel={$\x\eq$},
	domain=0:1.0,
	ymin=0.0, ymax=1.1,
	samples=200,
	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.5,0.4)},anchor=north west}
	]
	\addplot[line width=2.0pt, color=BLUE]{omega(x, 2)};
	\addplot[line width=2.0pt, color=YELLOW]{omega(x, 4)};
	\addplot[line width=2.0pt, color=GREEN]{omega(x, 8)};
	\addplot[line width=2.0pt, color=RED]{omega(x, 16)};
	\end{axis}
	\end{tikzpicture}
\end{frame}

\begin{frame}
\centering
	\begin{tikzpicture}[
	declare function={
		domega(\x,\k)= 2*(\k*(\x-1)+1 + (\k*(\x-1)^2+2*\x-1)*(ln(\k-1) + ln((1-\x)/\x)))/(\k*(\x - 1) + 1)^2 + (\k-2) / (\k-1);
	},]
	\begin{axis}[
	width=\textwidth,
	height=\axisdefaultheight,
	ylabel={$\frac{\der \dnds\eq}{\der \x\eq}$},
	xlabel={$\x\eq$},
	domain=0.0:1.0,
	ymin=0, ymax=6,
	samples=200,
	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.5,0.8)},anchor=north west}
	]
	\addplot[line width=2.0pt, color=BLUE]{domega(x, 2)};
	\addplot[line width=2.0pt, color=YELLOW]{domega(x, 4)};
	\addplot[line width=2.0pt, color=GREEN]{domega(x, 8)};
	\addplot[line width=2.0pt, color=RED]{domega(x, 16)};
	\end{axis}
	\end{tikzpicture}
\end{frame}

\begin{frame}
\centering
	\begin{tikzpicture}[
	declare function={
		lambda(\x,\k)= ln(\k-1) + ln((1-\x)/\x);
	}]
	\begin{axis}[
	width=\textwidth,
	height=\axisdefaultheight,
	xlabel={$\x$},
	ymin=0.0, ymax=5.0,
	samples=200,
	legend entries={
		$4\Ne \beta \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}$,
		$\lambda_{2}(\x)$,
		$\lambda_{4}(\x)$,
		$\lambda_{8}(\x)$,
		$\lambda_{16}(\x)$},
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend style={at={(0.7,0.95)},anchor=north west}
	]
	\addplot[domain=0.0:0.38, line width=2.0pt, color=BLUE]{ 4 * 1000 * 1.686 *exp(1.686 * (-118 + 300 *x))};
	\addplot[domain=0.0:0.5,line width=2.0pt, color=YELLOW]{ lambda(x,2) };
	\addplot[domain=0.0:0.8,line width=2.0pt, color=GREEN]{ lambda(x,4) };
	\addplot[domain=0.0:0.9,line width=2.0pt, color=RED]{ lambda(x,8) };
	\addplot[domain=0.0:1.0,line width=2.0pt, color=LIGHTGREEN]{ lambda(x,16) };
	\end{axis}
	\end{tikzpicture}
\end{frame}

\begin{frame}
\begin{figure}[H]
	\centering
	\begin{tikzpicture}[scale = 1.0]
	\begin{axis}[
	width=\textwidth,
	height=\axisdefaultheight,
	ylabel={Fitness $f(\x)$},
	xlabel={Phenotype $\x$},
	cycle list name=colors,
	domain=0:1,
	samples=200,
	legend cell align=left,
	minor tick num=2,
	axis x line=bottom,
	axis y line=left,
	legend entries={$f(\x) = \frac{1}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}$},
	legend style={at={(0.4,0.5)},anchor=south west}
	]
	\addplot[line width=2.0pt, color=BLUE]{ 1 / (1+exp(1.686*(-12 + 30 * x)))};
	\addplot[black]{1.0};
	\end{axis}
	\end{tikzpicture}
\end{figure}
\end{frame}

\end{document}

### 2020-04-14 manuscript/manuscript.tex
Maj manuscript, need more pedagogical writting

@@
 \documentclass{article}
-
 \usepackage{url}
-
 \usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
@@
 \usepackage{xfrac, nicefrac}
-\usepackage{xcolor}
 \usepackage{lmodern}
 \usepackage{enumitem}
-\usepackage{pgfplots, pgf,tikz}
 \usepackage[margin=60pt]{geometry}
@@
 \captionsetup{width=0.85\textwidth}
-
 \include{notations}
-\pgfplotsset{every axis/.append style={line width=1pt}}
-\pgfplotscreateplotcyclelist{colors}{LIGHTGREEN\\YELLOW\\RED\\GREEN\\BLUE\\}
-
 \begin{document}
-
 \title{Substitution rate responses to changes in effective population size.}
-
 \author{T. Latrille, N. Lartillot}
-
 \maketitle
@@
 \abstract{
-The quasi-neutral theory of evolution assert that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
+The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
 One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
-In molecular sequence, this effect translate in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
+In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
 Such theoretical prediction had been observed in empirical data across many clades.
 However, several studies failed to observe such response of $\dnds$ to changes in $\Ne$, or with weak strength or direction.
-Computational model of protein folding have shown that $\dnds$ can be independent of $\Ne$, which can mathematical be proven under certain assumptions.
-Moreover, non-equilibrium properties can implies that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
-Together, assumptions about the mapping of sequence to fitness can display a variety of behaviours in the $\dnds$ responses to changes in $\Ne$.
+Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
+Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
+Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
 Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
-We apply our framework in the case of fitness proportional to the probability of protein folding.
-Our compact theoretical results match more complex simulations using 3d structure of proteins.
-We assert that models based on the probability of folding are at odd with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
-However, fitness function based on non-specific interactions between proteins match the empirical data.
+We apply our framework in the special case of fitness proportional to the probability of protein folding.
+Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
+We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
+However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
 We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
 }
-
 \section*{Introduction}
@@
 In protein-coding DNA sequences, non-synonymous mutations are changing the amino-acid sequence, and by comparing the non-synonymous substitutions rate relative to the synonymous substitution rate (supposedly neutral), a statistic called $\dnds$, we can estimate the global strength of selection and random drift.\\
-
 For a given gene, it has been empirically observed changes of $\dnds$ along lineages of a phylogeny \cite{Yang2001, Zhang2004}.
 More importantly, changes in substitution rate correlate with changes in life-history-traits (longevity, body mass, ...) such as shown by the molecular comparative framework \cite{Lartillot2011,Weber2014}.
-Such covariation are compatible with the theoretical prediction of the nearly-neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
+Such covariations are compatible with the theoretical prediction of the nearly neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
 Indeed, populations with low $\Ne$ would have both a large body size and a high $\dnds$ (weaker selection) due to the increase of random drift.
 If $\dnds$ is empirically correlated to life-history traits, such as body mass and longevity, the direction of correlation could not be replicated across all experiments \cite{Figuet2016}.\\
-
-Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects \cite{Ohta1972, Welch2008}, and also been shown in context of a fixed fitness landscape, where each amino-acid have different fitnesses \cite{Spielman2015a, DosReis2015}.
-However, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly-neutral theory of evolution \cite{Lanfear2014}.
-In the example of a bio-chemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ is independent of $\Ne$ \cite{Goldstein2013}.
+Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects of novel mutations \cite{Ohta1972, Welch2008}.
+Similarly, it has been shown in context of a fixed fitness landscape, where each amino-acid have different fitness \cite{Spielman2015a, DosReis2015}, although sites are considered independently evolving, without epistasis.
+Unexpectedly, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly neutral theory of evolution \cite{Lanfear2014}.
+In the example of a biochemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ as been observed to be independent of $\Ne$ \cite{Goldstein2013}.
 Theoretically, a more general case is obtained whenever the fitness is a concave function of a trait, and such trait is equimutable \cite{Cherry1998}.
 Equimutabibilty meaning a mutation has the same effect on the trait whatever the current trait of the individual.
-Moreover, even in the case of a $\dnds$ response to $\Ne$, this response is far from being $1$-to-$1$ due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$, and subsequentially a decrease in $\dnds$ \cite{Jones2016}.\\
-
+In addition to this complexity, $\dnds$ response to $\Ne$ can be non-monotonic due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$ (adaptive selection), and subsequently a decrease in $\dnds$ (purifying selection) \cite{Jones2016}.\\
 Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
@@
 We provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
-We apply our framework in the case of fitness proportional to the probability of protein folding.
-Our compact theoretical results match more complex simulations using 3d structure of proteins.
-We assert that models based on the probability of folding are at odd with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
-However, fitness function based on non-specific interactions between proteins match the empirical data.
+We apply our framework in the special case of fitness proportional to the probability of protein folding.
+Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
+We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
+However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
 We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
-
 \section*{Results}
 \subsection*{Theoretical approximations}
-
-We seek to determine the change in $\dnds$ at equilibrium after a change in $\Ne$.
-This model map the amino-acid sequence to a phenotype $\x$, in this case the proportion of unstable sites in the protein.
-Each site can be in one of $\Nstate \geq 2$ states, where only $1$ state is defined the stable state, and $\Nstate - 1$ states are unstable.
-After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$, where $\Nsite$ is the number of sites.
-The Wrightian fitness of phenotype $\x$ is called $f(\x)$
-We demonstrate (See Supplementary Materials) that: 
+In this section we theoretically approximate the change in $\dnds$ at equilibrium after a change in $\Ne$, under a model in which the fitness depends on the genotype of the whole sequence.
+Firstly, we decompose the map from genotype to fitness through an intermediary bounded phenotype $0 \leq \x \leq 1$, the proportion of unstable sites in the sequence.
+Each site of the sequence ($\Nsite$ sites) can be in one of $\Nstate \geq 2$ states, where only $1$ of those states is assumed to be a stable state, and $\Nstate - 1$ states are unstable.
+The optimal phenotype is obtained when all of the $\Nsite$ sites of the sequence are stable ($\x = 0$).\\
+To complete the model, the phenotype-fitness map is first considered generally, where the Wrightian fitness of phenotype $\x$ is $f(\x)$.
+After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
+Most importantly the probabilities to increase or decrease $\x$ are different, and depends on the current phenotype $\x$.
+Mutation increasing $\x$ are selected against, while mutations decreasing $\x$ are positively selected. 
+Mutation and selection cancel each other at equilibrium, where the expectation of fitness effects of mutations reaching fixation in the population is $0$.
+Altogether, the equilibrium phenotype obtained for such mutation-selection balance is denoted $\x\eq$, and more importantly is given by the equation (See Supplementary Materials):
 \begin{align}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}},
+\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
 \end{align}
-where $\x\eq$ is the phenotype value at equilibrium.
-Thus, the slope of the $\dnds$ elasticity to changes in $\Ne$ is the ratio of first to second derivative of the log-fitness function.
-Moreover, the equilibrium value of $\x\eq$ is given by the equation:
+This equation of equilibrium phenotype can be further studied to derive the response of $\x\eq$ after a change in $\Ne$, using differential calculus.
+Furthemore, assuming that the number of states is large (typically $20$ for amino-acid), we obtain the change in $\omega\eq$ after a change in $\Ne$ as: 
 \begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
 \end{align}
-As an example, Wrightian fitness is defined as the probability of our protein to be in the folded state, given by fermi dirac distribution: 
+In other words, the slope of the $\dnds$ elasticity to changes in $\Ne$ (in log scale) is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype.
+To note, this elasticity is strictly negative for decreasing log-concave fitness functions.\\
+So far, our result did not assume any particular parametrization of the phenotype-fitness map.
+As an example, already proposed in the literature \cite{Goldstein2011}, fitness is proportional to the probability of our protein to be in the folded state, given by Fermi Dirac distribution: 
 \begin{equation}
-f(\x) = \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite * \gamma * \x)}}, 
-\end{equation} .
-where $\deltaGMin$ and $\gamma$ are in kcal/mol, and the fixed parameter $\beta=1.686$ mol/kcal at room temperature.
-Thus, $\deltaGMin < 0$ is the difference in free energy between folded and unfolded state when all sites are stable.
-$\Nsite \gamma > 0$ is the expected change in free energy (between folded and unfolded states) when all sites are unstable.
-
+f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}, 
+\end{equation}
+where $\deltaGMin < 0$ and $\gamma > 0$ are in kcal/mol, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
+In other words, $\deltaGMin$ is the difference in free energy between folded and unfolded state when all sites are stable.
+$\gamma$ is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
+With such fitness function, the equilibrium phenotype $\x\eq$ is given by the equation: 
+\begin{align}
+\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
+\label{eq:equilibrium}
+\end{align}
+The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
+\begin{equation}
+\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma};
+\end{equation}
+which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
+Moreover, only the free compound parameter $\gamma \Nsite$ has an impact on the slope, an intuition for such result is given in figure \ref{fig:NeChangeInfluence}.
+Meaning that under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is approximately $0.002$.
 \begin{figure*}[htb!]
@@
   \centering
-		\begin{tikzpicture}
-		\begin{axis}[
-		width=0.65\textwidth,
-		height=\axisdefaultheight,
-		xlabel={$\x$},
-		ymin=0.0, ymax=1.0,
-		samples=100,
-		legend entries={$\ln(\frac{1 - \x}{\x})$,
-			$S_{\deltaGMin = -118, \gamma = 1,\Ne = 10^{4}}(\x)$,
-			$S_{\deltaGMin = -118, \gamma = 1,\Ne = 10^{5}}(\x)$,
-			$S_{\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{4}}(\x)$,
-			$S_{\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{5}}(\x)$,
-		},
-		legend cell align=left,
-		minor tick num=2,
-		axis x line=bottom,
-		axis y line=left,
-		legend style={at={(0.5,0.95)},anchor=north west}
-		]
-		\addplot[domain=0.3:0.6,line width=2.0pt, color=BLUE]{ ln((1-x)/x) };
-		\addplot[name path=A, domain=0.3:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
-		\addplot[name path=B, dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
-		\addplot[name path=C,domain=0.3:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-16.71+30*x))};
-		\addplot[name path=D,dashed,domain=0.3:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-16.71+30*x))};
-		\addplot[fill=RED, opacity=0.2] fill between[ of = A and B];
-		\addplot[fill=GREEN, opacity=0.2] fill between[ of = C and D];
-		\end{axis}
-		\end{tikzpicture}
+  \includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
   \caption{
-			\textbf{$\bm{\dnds}$ response to change in $\bm{\Ne}$}.
-			The equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=S_{\deltaGMin,\Nsite \gamma,\Ne}(\x\eq)$ (see Supplementary Materials).
-$S$ is proportional to $\Ne$, but increase exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate.
+   \textbf{Theoretical model.}
+   Panel A.
+Illustration of the Genotype-phenotype-fitness map.
+The phenotype ($\x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of unstable sites in the sequence.
+The fitness is a decreasing log-concave function of the phenotype.
+   Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$.
+   For the special case $K=2$, the equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$.
+The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate.
+Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$ (x-axis of the crossing with the blue solid line).
+On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably.
+Moreover, changes in $\x\eq$ reflects the changes in $\dnds\eq$ since both are related by a monotonous function (See Supplementary Materials).
+The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
   }
@@
 \end{figure*}
-With such fitness function, we have (See Supplementary Materials) :
-\begin{equation}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma}.
-\end{equation}
-which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
-Meaning that under a value of $\Nsite=300$ sites, $\gamma=1.0$ kcal/mol, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is $0.002$.
 
 \subsection*{Simulations confirmation}
-We compare our model to the original studies \cite{Williams2006, Goldstein2011, Pollock2012}, where free energy of the folded state is computed using the $3$D structure of the folded state and pair-wise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
+Our compact theoretical results are based on assumptions that are broken in practice, for example assuming that 
+\begin{enumerate}
+ \item $\Nsite$ is large;
+ \item selection coefficient is well approximated by the fitness derivative;
+ \item genetic code is not taken into account;
+ \item all unstable states are equivalently unstable.
+\end{enumerate}
+We thus first tested the soundness of our theoretical results when such assumptions are relaxed, by simulating DNA sequences evolution (See Methods). 
+Firstly, simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}).
+Secondly the parameter $\deltaGMin$ hardly changes the slope of the linear regression, as expected also theoretically.
+Decreasing $\deltaGMin$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
+Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
+However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.\\
+
+Finally, we compare our results to more complex models of sequence evolution leveraging the $3$D structure of protein \cite{Williams2006, Goldstein2011, Pollock2012}.
+In such models, the free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
 The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
-
+The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$.
+Using extensive simulations, we argue instead that $\dnds$ is approximately linear with log-$\Ne$ (see Figure \ref{fig:GoldsteinVsToy}, panel C).
+Moreover, the observed slope match the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
+In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
 \begin{figure*}[htb!]
@@
 \centering
-\includegraphics[width=0.8\textwidth] {artworks/SimuFold-Elasticity.pdf}
-\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Elasticity.pdf}
+\includegraphics[width=0.9\textwidth] {artworks/Elasticity.png}
  \caption{
   \textbf{$\bm{\dnds}$ elasticity to change in $\bm{\Ne}$}.
-		For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-		In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally \cite{Goldstein2013}.
+  For each population size, $100$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+  Panel A.
+The fixed parameters are $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+$\alpha$ are given in the legend.
+Decreasing $\alpha$ (to more negative values) increases $\dnds$ but changes hardly the slope of the linear regression, as expected theoretically.
+  $\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+  Panel B. The fixed parameters are $\Nsite=300$ and $\beta=1.686$. Parameters $\alpha$ and $\gamma$ are given in the legend.
+  $\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+  The slope of $\dnds$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
+  Panel C. In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally \cite{Goldstein2013}.
   This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.
-		Bottom panel, The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+  Bottom D, the fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
   Moreover, with the Grantham model, the $\dnds$ matches the empirical 3D model of Golstein \& Pollock and the theoretical prediction.
@@
 \end{figure*}
-Simulations of protein evolution with both models display the same $\dnds$-$\Ne$ relationship, whenever the parameters $\alpha$ and $\gamma$ of our model are set to match the average values of the $3$D model (see Figure \ref{fig:EqdndsNe}).
-In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
-The parameter is set $\gamma=1.0$ kcal/mol, the average observed increase in $\deltaG$ per destabilizing mutation \cite{Zeldovich2007}.
-The original works claimed that under the $3$D model, $\dnds$ is independent of $\Ne$ because $\deltaG$ is an equimutable parameter.
-We argue instead that $\dnds$ is approximatively linear with log-$\Ne$, and that $\deltaG$ is not equimutable (see Figure \ref{fig:EqdndsNe}).
-Thus if we consider $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations, the slope is determined solely by the number of sites in the protein.
-
 \subsection*{Epistasis determines the time to relaxation}
-
-Although the equilibrium value of $\dnds$ after changes in $\Ne$ is an important feature of the $\dnds$-$\Ne$ relationship, an other parameter that is often neglected is the relaxation time to reach the new equilibrium $\dnds$ \cite{Jones2016}.
+Although the equilibrium value of $\dnds$ after changes in $\Ne$ is an important feature of the $\dnds$-$\Ne$ relationship, another parameter that is scarcely studied is the relaxation time to reach the new equilibrium $\dnds$ \cite{Jones2016}.
+We observed in our simulations that the determining factor of the relaxation time is the number of sites $\Nsite$ (See Figure \ref{fig:relaxStability}).
+Such observations match the theoretical prediction that high epistasis leads to faster return to equilibrium, since more opportunities are available.
+In the case of a fitness landscape with epistasis, after a change in $\Ne$, any site that is used to climb either up or down the fitness landscape will have diminishing return in other sites of the sequences.
+Not taking into account epistasis have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$ .
+Using a site-specific fitness landscape, where each amino-acid have different fitness, thus with no epistasis, the relaxation time is long since every site has to adapt to the new change in $\Ne$.
+However, using a fixed distribution of fitness effect (DFE) does not suffer from underestimating the relaxation rate, since the effect is instantaneous (see Figure \ref{fig:relaxStability}).
 \begin{figure*}[htb!]
@@
  \centering
-	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Alpha-Gamma.pdf}
-	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-DFE-Profile-N50.pdf}
+ \includegraphics[width=0.9\textwidth] {artworks/Relaxation.pdf}
  \caption{
   \textbf{$\bm{\dnds}$ relaxation time to changes in $\bm{\Ne}$}.
-		Top panel. $\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
-		Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
-		The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
+  $\dnds$ relaxation after a brutal change in $\Ne$.
+  Solid line corresponds to the average over replicates and the shaded area correspond to the $90\%$ interval among replicates. 
+  The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $700$ million years.
+  Panel A.
 $\beta=1.686$, $\gamma=-10$ for all simulations.
-The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
+The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates ($r$) is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
   Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
   Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-		Increasing $\Nsite$ implies a higher rate of relaxation.
+  Increasing $\Nsite$ implies a reduced time to reach the new equilibrium.
   Panel B.
-In context of a fixed fitness landscape, where each amino-acid have different fitnesses (site-specific profile), the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
+In context of a fixed fitness landscape, where each amino-acid has different fitness (site-specific profile), the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
 In the context of a distribution of fitness effects, the relaxation time is non-existent.
@@
 \end{figure*}
-In the case of a fitness landscape with epistasis, after a change in $\Ne$, any site that is used to climb either up or down the fitness landscape will have diminishing return in other site of the sequences.
-Not taking into account epistasis have the consequence of over-estimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$ \ref{fig:relaxStability}).
 
-Using a site-specific fitness landscape, where each amino-acid have different fitnesses, thus with no epistasis, the relaxation time is long since every site has to adapt to the new change in $\Ne$.
-However, using a distribution of fitness effect (DFE) does not suffer from under-estimating the relaxation rate, since the effect is instantaneous (see Figure \ref{fig:relaxProfileDFE}).
-
+\subsection*{Distribution of fitness effects}
+DNA mutations changing a genotype can result in a change of phenotype, and ultimately a change in fitness.
+From a specific genotype, all the possible mutations thus result in a distribution a phenotypic effect (DPE) and fitness effects (DFE).
+The DPE and DFE are not known a priori, but are the resulting consequence of the mutation-selection-drift balance (see Figure \ref{fig:Distribution}).
+Empirically, these distributions are of particular importance since they can be obtained experimentally or inferred with other data.
+As an example, DFE can be inferred from polymorphism dataset \cite{Eyre-walker2007, Galtier2016}. 
+Moreover, the distribution of $\deltadeltaG=\gamma \Nsite \dx$ of novel mutations can be obtained experimentally.
+\begin{figure*}[htb!]
+ \begin{mdframed}
+  \centering
+  \includegraphics[width=0.9\textwidth] {artworks/DPE-DFE.pdf}
+  \caption{
+   \textbf{Distribution of phenotypic and fitness effects.}
+   Distribution of fitness effects and phenotypic effect for novels non-synonymous mutations observed along a simulation at the mutation-selection balance.
+$\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+   Each side of the distribution is fitted to a gamma distribution, shown in dotted line. 
+   Panel A. Distribution of observed $\deltadeltaG$, which fit adequately the gamma distribution for negative $\deltadeltaG$ (stabilizing mutations). 
+   Panel A.
+Distribution of observed selection coefficient, which fit adequately the gamma distribution for both positive and negative selection coefficient.
+However the shape parameter estimated is not the same for positive and negative selection coefficients.
+  }
+  \label{fig:Distribution}
+ \end{mdframed}
+\end{figure*}
 \section*{Discussion}
-
 Goldstein \& Pollock argued that model of protein evolution should consider how the acceptance of mutations depends on the sequence context in which they arise (epistasis), as to correctly represent resultant rate heterogeneity along the sequence, to predict the role of compensatory substitutions in protein evolution, predict which of the 10\% of deleterious mutations in humans are harmless in other species, or to accurately represent the rate and time dependence of convergence and homoplasy \cite{Goldstein2017}.
-We argue that not taking into account epistasis also have the consequence of over-estimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$.
-Paradoxically, not taking into account epistasis means over-estimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.\\
-
-If using a distribution of fitness effect (DFE) is mathematically convenient and does not suffer from under-estimating the relaxation rate, it is also over-estimating the response of $\dnds$ to changes in $\Ne$ at equilibrium (\textit{should show proof and mechanistic explanation that the selection coefficient shall not be independent of $\Ne$}).
+We argue that not taking into account epistasis also have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$.
+Paradoxically, not taking into account epistasis means overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.\\
+If using a distribution of fitness effect (DFE) is mathematically convenient and does not suffer from underestimating the relaxation rate, it is also overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.
 As such, developing inference of fitness landscape will require to assume site-interdependent fitness landscapes.\\
-
-Our model does not display site variation in $\dnds$, which is a strong limitation, as it has observed in most protein \cite{Echave2017}.
 \section*{Materials \& Methods}
-
-Protein sequence evolution are simulated under an origin-fixation model \cite{McCandlish2014}., where the rate of substitution of a new mutation with selection coefficient $s$ is:
+Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}., where the rate of substitution of a new mutation with selection coefficient $s$ is:
 \begin{equation}
@@
 where $\mu$ is the mutation rate.
-
 \subsection*{Simulation of protein folding with an additive model of free energy}
@@
 where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
-
 From the resident sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
-For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codon that are one mutation away and that are not stop codon.
+For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codons that are one mutation away and that are not stop codon.
 For each mutant sequence $\cj \in \setNeighbors$, we compute $\deltaG^{t+1}$ from the updated sequence $\cj$, and subsequently the selection coefficient of the mutant:
@@
 \end{equation}
-The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
+The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
 \begin{equation}
@@
 where $\mu_{\itoj}$ is the mutation rate between $\ci$ and $\cj$, determined by the underlying $4x4$ nucleotide mutation rate matrix, and ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
-
 \subsection*{Simulation with independent fitness profiles}
@@
 where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \cite{Bloom2017}.
-The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
-
+The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
 \subsection*{Simulation with distribution of fitness effects (DFE)}
@@
 \end{equation}
-The next change in the protein coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
-
+The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
 \subsection*{$\bm{\dnds}$ along the simulation}
-From the set of mutant $\setNeighbors$ that are one nucleotide away from $\ci$, we define the subsets $\setNonSynNeighbors$ and $\setSynNeighbors$ that are respectively the set of non-synonymous and synonymoys mutants, where  $\setNonSynNeighbors \cup \setSynNeighbors = \setNeighbors$.
+From the set of mutants $\setNeighbors$ that is one nucleotide away from $\ci$, we define the subsets $\setNonSynNeighbors$ and $\setSynNeighbors$ that are respectively the set of non-synonymous and synonymous mutants, where $\setNonSynNeighbors \cup \setSynNeighbors = \setNeighbors$.
 As in previous works \cite{Spielman2015a, DosReis2015, Jones2016}, the ratio of non-synonymous over synonymous substitution rates of the sequence is defined as :
@@
 The scripts and instructions necessary to reproduce the experiments are available at \url{https://github.com/ThibaultLatrille/GenotypePhenotypeFitness}.
-
 \bibliographystyle{apalike}

### 2020-04-14 manuscript/presentation.tex
Maj manuscript, need more pedagogical writting

\documentclass{article}
\usepackage{url}
\usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
\usepackage{adjustbox}
\usepackage{float}
\usepackage{caption}
\usepackage{mdframed}
\usepackage{lmodern}
\usepackage{bm,bbold}
\usepackage{xfrac, nicefrac}
\usepackage{lmodern}
\usepackage{enumitem}
\usepackage[margin=90pt]{geometry}
\include{notations}

\begin{document}
~\\
The equilibrium phenotype ($\x\eq$)	obtained at mutation-selection balance solves the equation:
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
\end{align}
Furthemore, assuming that the number of states ($\Nstate$) is large, typically $20$ for amino-acid, the response in $\omega\eq$ after a change in $\Ne$ is: 
\begin{align}
\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \dfrac{\dfrac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\dfrac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
\end{align}

\newpage
~\\
Fitness is proportional to the probability of our protein to be in the folded state, given by Fermi Dirac distribution: 
\begin{equation}
f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}.
\end{equation}
\begin{itemize}
	\setlength\itemsep{-0.1em}
	\item $\beta=1.686$ mol/kcal is a fixed parameter at room temperature.
	\item $\deltaGMin < 0$ (in kcal/mol) is the difference in free energy between folded and unfolded state when all sites are stable.
	\item $\gamma > 0$ (in kcal/mol) is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
\end{itemize}
The equilibrium phenotype $\x\eq$ solves the equation: 
\begin{align}
\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
\label{eq:equilibrium}
\end{align}
The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
\begin{equation}
\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma}.
\end{equation}

\newpage
~\\
The equilibrium $\x\eq$ is obtained when the exponentially increasing function intersect the blue line. 
\begin{itemize}
	\setlength\itemsep{-0.1em}
	\item $\beta \Nsite \gamma$ is the exponential growth rate. $\beta$ and $ \Nsite$
	are considered fixed.
	\item When $\gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$.
	\item When $\gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably.
\end{itemize}
The value of $\alpha$ has been chosen given all other parameters such that $\x\eq$ is the same for $\Ne=10^4$.\\

$\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$
~\\
\begin{enumerate}
	\item $\Nsite$ is large;
	\item Selection coefficient is well approximated by the fitness derivative;
	\item Genetic code is not taken into account;
	\item All unstable states are equivalently unstable.
\end{enumerate}
\newpage
~\\
\begin{itemize}
	\item $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
	\item In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
	\item Such theoretical prediction had been observed in empirical data.
	\item Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
	\item Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
	\item Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
\end{itemize}

\newpage
~\\
\begin{itemize}
	\item Provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
	\item Apply our framework in the special case of fitness proportional to the probability of protein folding.
	\item Tested the soundness of our theoretical results when some assumptions are relaxed.
	\item Interpret our theoretical results in the light of empirical data.
	\item If the sequence is totally stable, only deleterious mutations can appear, moreover they would have a low selection coefficient.
	\item Deleterious mutations will reach fixation, increasing the phenotype, until the sequence is "marginally stable".
	\item What is the equilibrium phenotype at mutation-selection balance?
\end{itemize}

\newpage
\begin{itemize}
	\item Fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
	\item Theoretical slope is -0.00198 and observed is -0.00126
	\item Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before the protein became "marginally stable".  
	\item With increased $\x\eq$, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
	\item The slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.
	\item The slope of $\dnds$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
	\item Free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues.
	\item Free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states
	\item Left panel. Using the 3D structure of protein, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally.
	\item Right panel. Parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
	\item Empirical estimates are $\gamma \simeq 1$ for $\Nsite=300$ and $\beta=1.686$, leading to $\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -0.002$.
	\item The x-axis can be either $\Ne$ or protein abundance.
	\item Estimation using polymorphism and divergence data observed a slope of $\simeq -0.04$, approximately $20$ times greater than our prediction. 
	\item Models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship.
	\item However, theoretical approximations apply more broadly to protein-protein interactions, where stable state are hydrophilic amino-acid and unstable states are hydrophobic amino-acid at the surface of the protein.
	\item From empirical estimates on protein-protein interactions, $\frac{ \der \omega\eq}{\der \ln (\Ne)} \sim -0.2$, a much stronger response than under the model based on conformational stability.
\end{itemize}
\newpage
~\\
\begin{itemize}
	\item Importance of epistatic interactions in the $\Ne$-$\dnds$ relationship
	\item Epistasis determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
\end{itemize}

\end{document}

### 2020-04-14 manuscript/supp-mat.tex
Maj manuscript, need more pedagogical writting

@@
 \usepackage{caption}
-\usepackage{mdframed}
 \usepackage{lmodern}
@@
 \usepackage[margin=60pt]{geometry}
-
+\pdfinclusioncopyfonts=1
+\captionsetup{width=0.85\textwidth}
 \include{notations}
-
 \begin{document}
@@
 \end{gather}
-
 \subsection{Selection coefficient}
@@
 where $S(\x\eq, \dx) = 4\Ne s (\x\eq, \dx)$ is the scaled selection coefficient.
-
 \subsection{Probability of fixation}
@@
 where  $\lambda_{K}(\x\eq) = \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1)$.
-
 \subsection{Substitution rate bias ($\bm{\dnds}$) at equilibrium}
@@
 \begin{center}
-	\begin{tikzpicture}[
-	declare function={
-		omega(\x,\k)= \x*(2*(\x-1)*(ln(\k-1)+ln((1-\x)/\x))/(\k*(\x-1)+1)+ (\k-2)/(\k-1));
-	},]
-	\begin{axis}[
-	width=0.65\textwidth,
-	height=\axisdefaultheight,
-	ylabel={$\dnds\eq$},
-	xlabel={$\x\eq$},
-	domain=0:1.0,
-	ymin=0.0, ymax=1.1,
-	samples=100,
-	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
-	legend cell align=left,
-	minor tick num=2,
-	axis x line=bottom,
-	axis y line=left,
-	legend style={at={(0.5,0.4)},anchor=north west}
-	]
-	\addplot[line width=2.0pt, color=BLUE]{omega(x, 2)};
-	\addplot[line width=2.0pt, color=YELLOW]{omega(x, 4)};
-	\addplot[line width=2.0pt, color=GREEN]{omega(x, 8)};
-	\addplot[line width=2.0pt, color=RED]{omega(x, 16)};
-	\end{axis}
-	\end{tikzpicture}
+ \includegraphics[width=0.8\textwidth, page=2] {figures.pdf}
 \end{center}
@@
 \begin{center}
-	\begin{tikzpicture}[
-	declare function={
-		domega(\x,\k)= 2*(\k*(\x-1)+1 + (\k*(\x-1)^2+2*\x-1)*(ln(\k-1) + ln((1-\x)/\x)))/(\k*(\x - 1) + 1)^2 + (\k-2) / (\k-1);
-	},]
-	\begin{axis}[
-	width=0.65\textwidth,
-	height=\axisdefaultheight,
-	ylabel={$\frac{\der \dnds\eq}{\der \x\eq}$},
-	xlabel={$\x\eq$},
-	domain=0.0:1.0,
-	ymin=0, ymax=6,
-	samples=50,
-	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
-	legend cell align=left,
-	minor tick num=2,
-	axis x line=bottom,
-	axis y line=left,
-	legend style={at={(0.5,0.8)},anchor=north west}
-	]
-	\addplot[line width=2.0pt, color=BLUE]{domega(x, 2)};
-	\addplot[line width=2.0pt, color=YELLOW]{domega(x, 4)};
-	\addplot[line width=2.0pt, color=GREEN]{domega(x, 8)};
-	\addplot[line width=2.0pt, color=RED]{domega(x, 16)};
-	\end{axis}
-	\end{tikzpicture}
+ \includegraphics[width=0.8\textwidth, page=3] {figures.pdf}
 \end{center}
@@
 \end{align}
-
 Moreover, with the approximation that $\left| 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right| \gg \left| \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } \right|$, meaning that a change in phenotype causes a higher change in scaled selection coefficient than mutational bias, we have:
@@
 \end{align}
-
 \section{Models for the log-fitness function}
-
 \subsection{Folded fraction}
-All phenotype-fitness function considered below are log-concave, and as a result, $\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}$ is an decreasing function of $\x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
+All phenotype-fitness functions considered below are log-concave, and as a result, $\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}$ is an decreasing function of $\x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
 More precisely, fitness functions depends on the folded fraction of the protein of interest, which is given by the Fermi-Dirac distribution:
@@
 \end{eqnarray}
-
 \subsection{Fitness equal to folded fraction}
@@
 \end{gather}
-
 Using $\Ne=10^4$, $\beta=1.686$, $\deltaGMin = -118$, $\Nsite=300$, $\gamma=1$, we have the following :
 \begin{center}
-	\begin{tikzpicture}[
-	declare function={
-		lambda(\x,\k)= ln(\k-1) + ln((1-\x)/\x);
-	}]
-	\begin{axis}[
-	width=0.65\textwidth,
-	height=\axisdefaultheight,
-	xlabel={$\x$},
-	ymin=0.0, ymax=5.0,
-	samples=100,
-	legend entries={
-		$4\Ne \beta \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}$,
-		$\lambda_{2}(\x)$,
-		$\lambda_{4}(\x)$,
-		$\lambda_{8}(\x)$,
-		$\lambda_{16}(\x)$},
-	legend cell align=left,
-	minor tick num=2,
-	axis x line=bottom,
-	axis y line=left,
-	legend style={at={(0.7,0.95)},anchor=north west}
-	]
-	\addplot[domain=0.0:0.38, line width=2.0pt, color=BLUE]{ 4 * 1000 * 1.686 *exp(1.686 * (-118 + 300 *x))};
-	\addplot[domain=0.0:0.5,line width=2.0pt, color=YELLOW]{ lambda(x,2) };
-	\addplot[domain=0.0:0.8,line width=2.0pt, color=GREEN]{ lambda(x,4) };
-	\addplot[domain=0.0:0.9,line width=2.0pt, color=RED]{ lambda(x,8) };
-	\addplot[domain=0.0:1.0,line width=2.0pt, color=LIGHTGREEN]{ lambda(x,16) };
-	\end{axis}
-	\end{tikzpicture}
+\includegraphics[width=0.8\textwidth, page=4] {figures.pdf}
 \end{center}
@@
 This model, however, does not express the fact that selection is typically stronger for proteins characterized by higher levels of expression. 
-
 \subsection{Selective cost proportional to amount of misfolded protein}
-
 A slight variation is to assume that the selective cost itself is proportional to the total amount of misfolded protein. For a given protein with expression level $y$:
@@
 \subsection{Translational errors}
-
 An other variant account for translational errors.
@@
 \end{align}
-
 \subsection{Cost-benefit argument}
-
 This models assumes that the :
 \begin{enumerate}
-	\item the expression level is regulated so that the total amount of \emph{functional} macromolecules is maintained at a target level $y$;
+ \item the expression level is regulated so that the total number of \emph{functional} macromolecules is maintained at a target level $y$;
  \item the log-fitness is proportional to the ratio of the \emph{total} cost of expression over the benefit contributed by the protein (Beaulieu et al).
@@
 \section{Model of protein-protein interactions}
-
 The proteome is assumed to be composed of $m$ protein species, all with same abundance $C$.
@@
 Only pairwise interactions are considered, and higher-order interactions are ignored.
-The equilibrium is characterised by:
+The equilibrium is characterized by:
 \begin{eqnarray}
@@
 where $b>0$ is a parameter determining the overall stringency of selection against non-specific interactions.
-
 \subsection{Mean field, weak-interaction limit}
-
 To make the model tractable and compact, we assume that non-specific interactions are weak, i.e.
@@
 In addition, we use a mean-field approximation, such that, when considering a specific protein species $i$, we assume that all other proteins have the same fraction $\bar x$ of hydrophobic residues at their surface.
-The value of $\bar x$ could in principle be found using a self-consistent argument, essentially by (1) explicitly calculating the net substitution flux for protein $i$ with fraction $x_i$, under mean-field $\bar x$, and (2) expressing the constraint that this substitution process for protein $i$ is stationary at $x_i = \bar x$.
+The value of $\bar x$ could in principle be found using a self-consistent argument, essentially by (1) explicitly calculating the net substitution flux for protein $i$ with fraction $x_i$, under mean field $\bar x$, and (2) expressing the constraint that this substitution process for protein $i$ is stationary at $x_i = \bar x$.
 This derivation is not conducted here, as it is not needed.
-
 Using these approximations, we can re-express the conservation of total mass as:
@@
 \end{eqnarray}
-
-Now, assume that the system is at equilibirum (thus $x_i = \bar x$).
-The strength of selection acting on mutations occuring at the surface of protein $i$, of effect size $\delta x = \pm 1/n$, is given by $s = \kappa \delta x$ where:
+Now, assume that the system is at equilibrium (thus $x_i = \bar x$).
+The strength of selection acting on mutations occurring at the surface of protein $i$, of effect size $\delta x = \pm 1/n$, is given by $s = \kappa \delta x$ where:
 \begin{eqnarray}
@@
 In any case, $\chi$ is now dependent on $x^*$, through $\lambda$.
-
 \subsection{Empirical calibration}
-
 Based on empirical estimates obtained from Zhang et al.
@@
 \section{Simulation of protein-coding DNA sequences evolution}
-
 \subsection{Probability of folding using the 3d structure of protein}
-
 We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
@@
 where $I(a,b)$ is the pair-wise contact energies between amino-acid $a$ and $b$, using contact potentials estimated by Miya-zawa and Jernigan, and $\mathcal{N}(\site)$ are the neighbor residues of site $\site$ (closer than $7\angstrom$) in the $3$D structure.\\
-
 The free energy of unfolded states $G_{\mathrm{U}}\left(\ci\right)$ is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states:
@@
 where the average $\langle . \rangle$ runs other the $55$ decoy $3$D structures, and $k$ is the Boltzmann constant and $T$ the temperature in Kelvin.\\
-
 From the energy of folded and unfolded states, we can compute the difference in free energy between the states:
@@
 \end{equation}
-
 \begin{center}
@@
 \subsection{Simulated $\bm{\dnds}$ elasticity to changes in $\bm{\Ne}$}
-
 \begin{center}
@@
 In the model of gamma distributed fitness effect, $\dnds$ at equilibrium is strongly dependent on log-$\Ne$ where the slope correlation is proportional to the inverse of the shape parameter of the gamma distribution \cite{Welch2008}.
-
 \begin{center}
@@
 In the model of site-wise amino-acid fitness profiles taken from \cite{Bloom2017}, $\dnds$ at equilibrium is strongly dependent on log-$\Ne$. 
-
-\begin{center}
-	\includegraphics[width=0.8\textwidth] {artworks/SimuFold-Elasticity.pdf}
-\end{center}
-\textbf{$\bm{\dnds}$ elasticity with 3D free energy of folding}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed in the original work \cite{Goldstein2013}.
-This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta  \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.
-
 \begin{center}
@@
 The non-monotony is suspected to be due to the discrete number of sites and states, such that the changes in $\deltaG$ after a mutation is either $-1$, $0$ or $1$.
-
-\begin{center}
-	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Elasticity.pdf}
-\end{center}
-\textbf{Amino-acid distances effect on the elasticity}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-Compared to the simpler model, the elasticity of $\dnds$ to changes in $\Ne$ is monotonous (and smooth) due the realized $\deltaG$ is more continuous.
-Moreover, with the Grantham model, the $\dnds$ is lower and the slope of elasticity lower, closer to the empirical 3D model of Golstein \& Pollock.
-
-\begin{center}
-	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Alpha-Elasticity.pdf}
-\end{center}
-\textbf{Effect of $\bm{\alpha}$ on the elasticity}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-The fixed parameters are $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-$\alpha=-118$ are given in the legend.
-Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
-Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
-However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.
-
+Such non-monotony is not observed with the Grantham model, in which the $\dnds$ is lower and the slope of elasticity lower, closer to the empirical 3D model of Golstein \& Pollock.
 \begin{center}
@@
 Increasing $\beta$ increases the slope of the $\dnds$-$\Ne$ relationship, as predicted in our theoretical model.
-
-\begin{center}
-	\includegraphics[width=0.8\textwidth] {artworks/SimuStab-Alpha-Gamma-Elasticity.pdf}
-\end{center}
-\textbf{Effect of $\bm{\gamma}$ on the elasticity}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-The fixed parameters are $\Nsite=300$ and $\beta=1.686$. Parameters $\alpha$ and $\gamma$ are given in the legend.
-$\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-The slope of $\dnds$-$\Ne$ relationship is increased with $\gamma$, as predicted by our theoretical model.
-
 \begin{center}
@@
 However the slope of the $\dnds$-$\Ne$ relationship is not sensibly changed.
-
 \subsection{Simulated relaxation time of $\bm{\dnds}$}
-
 \begin{center}
@@
 \textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$}.
-$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
-Solid line correpond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+$\dnds$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line corresponds to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
 The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
@@
 Increasing $\Nsite$ implies a higher $\dnds$ at equilibrium, a lower elasticity of the $\dnds$ to changes in $\Ne$ and a higher rate of relaxation.
-
 \begin{center}
@@
 \textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$, while correction for $\gamma$}.
-$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
-Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+$\dnds$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line corresponds to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
 The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
@@
 Increasing $\Nsite$ implies a higher $\dnds$ at equilibrium, and a higher rate of relaxation.
-
-\begin{center}
-	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Alpha.pdf}
-\end{center}
-\textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$, while correction for $\alpha$}.
-$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
-Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
-The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
-$\beta=1.686$, $\gamma=-10$ for all simulations.
-The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Moreover, $\alpha$ is changed according to $\Nsite$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-Increasing $\Nsite$ implies a lower elasticity of the $\dnds$ to changes in $\Ne$ and a higher rate of relaxation.
-
-\begin{center}
-	\includegraphics[width=0.8\textwidth] {artworks/Relaxation-Stability-Alpha-Gamma.pdf}
-\end{center}
-\textbf{Relaxation time of $\bm{\dnds}$, while correction for $\alpha$ and $\gamma$}.
-$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
-Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
-The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.
-$\beta=1.686$, $\gamma=-10$ for all simulations.
-The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
-Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-Increasing $\Nsite$ implies a higher rate of relaxation.
-
 \begin{center}
@@
 \textbf{Relaxation time of $\bm{\dnds}$ for the Grantham model}.
-$\dnds$ relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
-Solid line correspond to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
+$\dnds$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+Solid line corresponds to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
 The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $900$ million years.

### 2020-04-22 manuscript/figures.tex
Refactored manuscript

@@
 		xlabel={$\x$},
-		ymin=0.0, ymax=1.0,
+		ymin=0.0, ymax=5.0,
 		samples=200,
-		legend entries={$\ln(\frac{1 - \x}{\x})$,
-			${\deltaGMin = -118, \gamma = 1,\Ne = 10^{4}}$,
-			${\deltaGMin = -118, \gamma = 1,\Ne = 10^{5}}$,
-			${\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{4}}$,
-			${\deltaGMin = -16.7, \gamma = 0.1,\Ne = 10^{5}}$,
+		legend entries={$\ln(\frac{1 - \x}{\x}) + \ln(19)$,
+			${\alpha = -118, \gamma = 1,\Ne = 10^{4}}$,
+			${\alpha = -118, \gamma = 1,\Ne = 10^{5}}$,
+			${\alpha = -15.7, \gamma = 0.1,\Ne = 10^{4}}$,
+			$\tiny{\alpha = -15.7, \gamma = 0.1,\Ne = 10^{5}}$,
 		},
@@
 		axis y line=left,
-		legend style={at={(0.5,0.95)},anchor=north west}
+		legend style={at={(0.05,0.25)},anchor=south west}
 		]
-		\addplot[domain=0.3:0.6,line width=2.0pt, color=BLUE]{ ln((1-x)/x) };
-		\addplot[name path=A, domain=0.3:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
+		\addplot[domain=0.1:0.39,line width=2.0pt, color=BLUE]{ ln(19*(1-x)/x) };
+		\addplot[name path=A, domain=0.1:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
 		\addplot[name path=B, dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
-		\addplot[name path=C,domain=0.3:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-16.71+30*x))};
-		\addplot[name path=D,dashed,domain=0.3:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-16.71+30*x))};
+		\addplot[name path=C,domain=0.1:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-15.71+30*x))};
+		\addplot[name path=D,dashed,domain=0.1:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-15.71+30*x))};
 		\addplot[fill=RED, opacity=0.2] fill between[ of = A and B];
@@
 	samples=200,
-	legend entries={$\Nstate=2$, $\Nstate=4$, $\Nstate=8$, $\Nstate=16$},
+	legend entries={$\Nstate=2$, $\Nstate=5$, $\Nstate=10$, $\Nstate=20$},
 	legend cell align=left,
@@
 	\addplot[line width=2.0pt, color=BLUE]{omega(x, 2)};
-	\addplot[line width=2.0pt, color=YELLOW]{omega(x, 4)};
-	\addplot[line width=2.0pt, color=GREEN]{omega(x, 8)};
-	\addplot[line width=2.0pt, color=RED]{omega(x, 16)};
+	\addplot[line width=2.0pt, color=YELLOW]{omega(x, 5)};
+	\addplot[line width=2.0pt, color=GREEN]{omega(x, 10)};
+	\addplot[line width=2.0pt, color=RED]{omega(x, 20)};
 	\end{axis}
@@
 	\addplot[line width=2.0pt, color=BLUE]{domega(x, 2)};
-	\addplot[line width=2.0pt, color=YELLOW]{domega(x, 4)};
-	\addplot[line width=2.0pt, color=GREEN]{domega(x, 8)};
-	\addplot[line width=2.0pt, color=RED]{domega(x, 16)};
+	\addplot[line width=2.0pt, color=YELLOW]{domega(x, 5)};
+	\addplot[line width=2.0pt, color=GREEN]{domega(x, 10)};
+	\addplot[line width=2.0pt, color=RED]{domega(x, 20)};
 	\end{axis}
@@
 	legend entries={
-		$4\Ne \beta \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}$,
+		$4\Ne \beta \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}}$,
 		$\lambda_{2}(\x)$,
-		$\lambda_{4}(\x)$,
-		$\lambda_{8}(\x)$,
-		$\lambda_{16}(\x)$},
+		$\lambda_{5}(\x)$,
+		$\lambda_{10}(\x)$,
+		$\lambda_{20}(\x)$},
 	legend cell align=left,
@@
 	\addplot[domain=0.0:0.5,line width=2.0pt, color=YELLOW]{ lambda(x,2) };
-	\addplot[domain=0.0:0.8,line width=2.0pt, color=GREEN]{ lambda(x,4) };
-	\addplot[domain=0.0:0.9,line width=2.0pt, color=RED]{ lambda(x,8) };
-	\addplot[domain=0.0:1.0,line width=2.0pt, color=LIGHTGREEN]{ lambda(x,16) };
+	\addplot[domain=0.0:0.8,line width=2.0pt, color=GREEN]{ lambda(x,5) };
+	\addplot[domain=0.0:0.9,line width=2.0pt, color=RED]{ lambda(x,10) };
+	\addplot[domain=0.0:1.0,line width=2.0pt, color=LIGHTGREEN]{ lambda(x,20) };
 	\end{axis}
@@
 	axis y line=left,
-	legend entries={$f(\x) = \frac{1}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}$},
+	legend entries={$f(\x) = \frac{1}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}}$},
 	legend style={at={(0.4,0.5)},anchor=south west}

### 2020-04-22 manuscript/manuscript.tex
Refactored manuscript

@@
 \section*{Introduction}
-Differences in DNA sequences of a gene in separate species are due to the particular history of DNA substitutions along each specie's respective lineage.
-These substitutions at the level of the population are the result of point mutations at the level of individuals, subject to evolutionary forces such as selection and random drift.
-In protein-coding DNA sequences, non-synonymous mutations are changing the amino-acid sequence, and by comparing the non-synonymous substitutions rate relative to the synonymous substitution rate (supposedly neutral), a statistic called $\dnds$, we can estimate the global strength of selection and random drift.\\
-For a given gene, it has been empirically observed changes of $\dnds$ along lineages of a phylogeny \cite{Yang2001, Zhang2004}.
-More importantly, changes in substitution rate correlate with changes in life-history-traits (longevity, body mass, ...) such as shown by the molecular comparative framework \cite{Lartillot2011,Weber2014}.
-Such covariations are compatible with the theoretical prediction of the nearly neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
-Indeed, populations with low $\Ne$ would have both a large body size and a high $\dnds$ (weaker selection) due to the increase of random drift.
-If $\dnds$ is empirically correlated to life-history traits, such as body mass and longevity, the direction of correlation could not be replicated across all experiments \cite{Figuet2016}.\\
-Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects of novel mutations \cite{Ohta1972, Welch2008}.
-Similarly, it has been shown in context of a fixed fitness landscape, where each amino-acid have different fitness \cite{Spielman2015a, DosReis2015}, although sites are considered independently evolving, without epistasis.
-Unexpectedly, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly neutral theory of evolution \cite{Lanfear2014}.
-In the example of a biochemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ as been observed to be independent of $\Ne$ \cite{Goldstein2013}.
-Theoretically, a more general case is obtained whenever the fitness is a concave function of a trait, and such trait is equimutable \cite{Cherry1998}.
-Equimutabibilty meaning a mutation has the same effect on the trait whatever the current trait of the individual.
-In addition to this complexity, $\dnds$ response to $\Ne$ can be non-monotonic due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$ (adaptive selection), and subsequently a decrease in $\dnds$ (purifying selection) \cite{Jones2016}.\\
-Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
-This question is important if we want to extract changes in effective population size from substitution history, as well as to determine the fitness landscape under which protein-coding DNA sequences are evolving.
-We provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
-We apply our framework in the special case of fitness proportional to the probability of protein folding.
-Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
-We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
-However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
-We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
+Molecular sequences differ across species due to the particular history of DNA substitutions along each specie's respective lineage.
+These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
+These forces have effects at different levels, mutations are carried by molecular sequence, selection is beared by individual's fitness, and random genetic drift is a population effect.
+The strength of random genetic genetic, or stochastic sampling of mutations, is less pronounced in lineage with high effective population size ($\Ne$), and as a consequence the purification by selection of deleterious mutations is more effective.
+Such idea has emerged in the framework of quasi-neutral theory of evolution, whereby predicting a decrease in the substitution rate per unit of time of selected mutations along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.
+This prediction is valid under the assumption that most mutations are deleterious and that their selective effect on individuals are drawn from a fixed distribution \cite{Welch2008}.
+Furthermore, this prediction assumes a constant mutation rate per unit of time, but such assumption can be relaxed by normalizing the substitution rate of selected mutations by the substitution rate of neutral mutations such as to discard this confounding factor \cite{Ohta1972}.
+This relative substitution rate of selected mutations, called $\dnds$ is this study, is a theoretical macroscopic observable in molecular sequence evolution, given we can separate selected and neutral mutations.
+Practically, $\dnds$ can be approximated in protein-coding DNA sequences, where selected mutations are the non-synonymous changes affecting the amino-acid sequence, and the non-synonymous mutations are supposedly neutral, this statistic is called $dNdS$.\\
+% \dnds decrease with \Ne under the assumption of fixed DFE
+
+The context of protein coding sequence fostered another modeling of selection, based on genotype-fitness map instead of distribution of fitness effects.
+In such approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998}.
+For example, if the target amino-acid has an higher fitness, the selective effect of the mutation is positive, and reciprocally negative for lower fitness.
+Such modeling approach account for higher substitutions rate between amino-acid with similar bio-chemical properties.
+Fundamentally, in such framework the fitness depends solely on the actual genotype, not on the trajectory of mutations leading to this sequence.
+Even thought the modeling approach is substantially different, assuming each site of the sequence has its own genotype-fitness map also predicts a negative correlation between $\dnds$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.\\
+% \dnds decrease with \Ne under the assumption site fitness landscape
+
+Empirically, fluctuation of $\dnds$ along lineages of phylogenies have been inferred using protein-coding DNA alignments \cite{Yang2001, Zhang2004}.
+Subsequently, inference methods combining molecular sequences and lineage specific quantitative traits have asserted that $\dnds$ correlates positively with longevity and body mass, traits which have been sampled only in the extant lineages \cite{Lartillot2011,Weber2014}.
+Moreover, longevity and body mass are proxies of $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}.
+As such, empirical evidences suggest a negative correlation between $\dnds$ and $\Ne$, confirming the theoretical prediction of the quasi-neutral theory of evolution.
+However, this correlation could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation could be in the inverse direction or not statistically significant \cite{Figuet2016}.\\
+% \dnds decrease with \Ne empirically, but not always
+
+Mitigated empirical evidence that $\dnds$ decrease with $\Ne$ encouraged alternative explanatory mechanism \cite{Lanfear2014}.
+One striking result proved that $\dnds$ can independent of $\Ne$, when fitness is a function of a phenotype and that such phenotype is equimutable.
+Equimutability states that the probabilities of phenotype changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
+Such argument has been invoked in the \textit{in silico} observation that $\dnds$ is seamingly independent of $\Ne$ if the protein thermodynamic stability is computed using a 3D structural model of the protein, and that the fitness is proportional to the folding probability of the protein \cite{Goldstein2013}.
+The given explanation is that probabilities of changes in free energy of folding ($\deltadeltaG$) is independent of the current free energy ($\deltaG$), which is considered an equimutable phenotype.\\
+% Explanation could be that \dnds is independent of \Ne
+
+However, this assumption is relatively strong, as for example if a protein is completely stable, only destabilizing mutations can occur, and more generally the probabilities of destabilizing mutational events is dependent on the actual composition of the protein.
+Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\dnds$ relationship to $\Ne$ is unknown.
+The goal of this study is to derive the $\dnds$ relationship to $\Ne$ to intrinsic molecular parameters of the model.
+More generally, we develop mathematical tools to derive the $\dnds$ relationship to $\Ne$ in the context of a genotype-phenotype-fitness map.
+With such results, we stress to test the assumption that protein stability is a proxy of fitness in the light of empirical data.\\
+% The assumptions are strong, we want to relax equimutability which is unlikely
+
+Orthogonally to the relationship between $\dnds$ and $\Ne$, independent studies have argued that protein abundance is also a predictor of $\dnds$.
+As such, we also seek to derive the $\dnds$ relationship to protein abundance, and discuss the empirical implication.\\
+% \dnds is relationship to protein abundance
+
+Lastly, the theoretical results discussed so far are all obtained when the balance between mutation, selection and drift is at equilibrium.
+However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ leads first to an increase of $\dnds$ due to adaptive selection, and subsequently a decrease in $\dnds$ due to stronger purifying selection \cite{Jones2016}.
+Studying only equilibrium properties can thus be misleading, and the dynamical response of $\dnds$ to changes in $\Ne$ must also be addressed by our model.\\
+% Result at equilibrium, what about dynamical properties?
+
 \section*{Results}
 \subsection*{Theoretical approximations}
-In this section we theoretically approximate the change in $\dnds$ at equilibrium after a change in $\Ne$, under a model in which the fitness depends on the genotype of the whole sequence.
-Firstly, we decompose the map from genotype to fitness through an intermediary bounded phenotype $0 \leq \x \leq 1$, the proportion of unstable sites in the sequence.
-Each site of the sequence ($\Nsite$ sites) can be in one of $\Nstate \geq 2$ states, where only $1$ of those states is assumed to be a stable state, and $\Nstate - 1$ states are unstable.
-The optimal phenotype is obtained when all of the $\Nsite$ sites of the sequence are stable ($\x = 0$).\\
-To complete the model, the phenotype-fitness map is first considered generally, where the Wrightian fitness of phenotype $\x$ is $f(\x)$.
-After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
-Most importantly the probabilities to increase or decrease $\x$ are different, and depends on the current phenotype $\x$.
-Mutation increasing $\x$ are selected against, while mutations decreasing $\x$ are positively selected. 
-Mutation and selection cancel each other at equilibrium, where the expectation of fitness effects of mutations reaching fixation in the population is $0$.
-Altogether, the equilibrium phenotype obtained for such mutation-selection balance is denoted $\x\eq$, and more importantly is given by the equation (See Supplementary Materials):
+In this section we present an analytical approximate solution for the change in $\dnds$ at equilibrium after a change in $\Ne$.
+The main results are given in a general case of any phenotype-fitness map, and also derived in the specific case of fitness proportional to the total amount of folded protein.
+All development are available in Supplementary Materials in the most general case.
+The first step define the relationship from genotype to fitness, through an intermediary phenotype.
+The second step present the equation of the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
+Finally, an approximate response of $\dnds$ to changes in $\Ne$ is derived from the equilibrium equation. \\
+
+We base our work on numerical model of protein stability \cite{Williams2006, Goldstein2011, Pollock2012}, on which we make simplifying assumptions to derive analytical equations.
+In the original model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
+Technically, the free energy of the folded are unfolded conformations are computed using $3$D conformations of the protein.
+In such model, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.\\
+
+We approximate this model such that the destabilizing effect of an amino-acid does not depend on a $3$D representation of the protein, but each site contribute independently to $\deltaG$.
+With such approximation, at a given site of sequence each amino-acid is either stabilizing or destabilizing the protein, and an unstable amino-acid contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ compare to a stable amino-acid.
+If all amino-acids of the sequence are stable, $\deltaG$ equal to $ \alpha < 0$. 
+In this model, the more compact phenotype summarizing the genotype is the proportion of unstable amino-acid in the sequence, defined as $0 \leq \x \leq 1$, and $\deltaG$ is a linear function of $\x$:
 \begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
+\deltaG (\x) = \alpha + \Nsite \gamma \x,
 \end{align}
-This equation of equilibrium phenotype can be further studied to derive the response of $\x\eq$ after a change in $\Ne$, using differential calculus.
-Furthemore, assuming that the number of states is large (typically $20$ for amino-acid), we obtain the change in $\omega\eq$ after a change in $\Ne$ as: 
+where $\Nsite$ is the number of sites in the sequence. \\
+
+From a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of unstable amino-acid, or do nothing if the mutation is between both unstable (or equivalently stable) amino-acids.
+To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable and only one amino-acid is stable and the $19$ others are unstable.
+Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
+\begin{gather}
+\begin{cases}
+\dx &\text{ with probability } 1-x, \\
+0 &\text{ with probability } \frac{18 \x }{19}, \\
+-\dx &\text{ with probability } \frac{\x}{19}.\\
+\end{cases} \label{eq:proba}
+\end{gather}
+In the extreme case of optimal phenotype ($\x = 0$), only destabilization mutations are proposed.
+Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between unstable amino-acids (effect $0$) is proportional to $\x$. 
+Once obtained the phenotypic effects of mutations and their probabilities, definition of the phenotype-fitness map is necessary derive the selective effect of mutations.\\
+
+Subsequent equations are valid for the more general case where the Wrightian fitness is a log-concave differentiable function $f$ of phenotype $\x$.
+As an example, we develop the argument when the Wrightian fitness of an individual is the proportion of protein in folded conformation.
+This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm. 
+This fitness is then given by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
 \begin{align}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
+f(\x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma \x)}},\\
+\Rightarrow f(\x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma \x)}, 
 \end{align}
-In other words, the slope of the $\dnds$ elasticity to changes in $\Ne$ (in log scale) is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype.
-To note, this elasticity is strictly negative for decreasing log-concave fitness functions.\\
-So far, our result did not assume any particular parametrization of the phenotype-fitness map.
-As an example, already proposed in the literature \cite{Goldstein2011}, fitness is proportional to the probability of our protein to be in the folded state, given by Fermi Dirac distribution: 
-\begin{equation}
-f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}, 
-\end{equation}
-where $\deltaGMin < 0$ and $\gamma > 0$ are in kcal/mol, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
-In other words, $\deltaGMin$ is the difference in free energy between folded and unfolded state when all sites are stable.
-$\gamma$ is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
-With such fitness function, the equilibrium phenotype $\x\eq$ is given by the equation: 
+where $\alpha$ and $\gamma$ are defined as above, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
+Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
 \begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
-\label{eq:equilibrium}
+s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(\x) }{\partial \x} \label{eq:s} \\
+\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x)} \label{eq:s-unfolded}
 \end{align}
-The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
-\begin{equation}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma};
-\end{equation}
-which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
-Moreover, only the free compound parameter $\gamma \Nsite$ has an impact on the slope, an intuition for such result is given in figure \ref{fig:NeChangeInfluence}.
-Meaning that under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is approximately $0.002$.
+
+With the complete genotype-phenotype-fitness map defined (see Figure \ref{fig:NeChangeInfluence}, left panel), one can study the evolutionary processes unfolds.
+Starting from an optimal sequence, destabilizing mutations will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, a point of equilibrium called marginal stability.
+Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
+Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average.
+Formally, and after simplification, the equilibrium phenotype denoted $\x\eq$ is given by the equation:
+\begin{align}
+\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
+\Rightarrow \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (19)  & \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x\eq)}. \label{eq:equilibrium}
+\end{align}
+This equation cannot be solved explicitly for $\x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $\x\eq$ is given in Figure \ref{fig:NeChangeInfluence} (right panel).
+\\
 \begin{figure*}[htb!]
@@
 The fitness is a decreasing log-concave function of the phenotype.
-   Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$.
-   For the special case $K=2$, the equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$.
+			Panel B.
+Illustration of the $\dnds$ elasticity after a change in $\Ne$.
+The equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma \x\eq)}$.
 The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate.
-Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$ (x-axis of the crossing with the blue solid line).
-On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably.
-Moreover, changes in $\x\eq$ reflects the changes in $\dnds\eq$ since both are related by a monotonous function (See Supplementary Materials).
+Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only changes subtly $\x\eq$ (x-axis of the crossing with the blue solid line).
+On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) involve a stronger response of $\x\eq$.
+Moreover, changes in $\x\eq$ reflects the changes in $\dnds$ since both are related by a monotonous function (equation \ref{eq:dnds}).
 The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
@@
 
+Moreover, at equilibrium we can derive the expected substitution rate of mutations changing the sequence, and by normalizying by the substitution rate of neutral mutations, $\dnds$ simply approximates to:
+\begin{gather}
+\dnds \simeq \x\eq \label{eq:dnds}
+\end{gather}
+This simple approximation is due to substitutions at unstable sites (proportion $\x\eq$) to one of the other unstable amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (see equation \ref{eq:proba}).\\
+
+From the equilibrium equation \ref{eq:equilibrium}, differential calculus allow to derive the response of $\x\eq$ and ultimately $\dnds$ after a change in $\Ne$.
+We define the elasticity $\chi$ as slope of the $\dnds$ response to changes in $\Ne$ in log scale, which simplify to a compact equation: 
+\begin{align}
+\chi = \frac{ \der \dnds}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}, \\
+\Rightarrow \chi & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
+\end{align}
+The elasticity is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype.
+To note, this elasticity is strictly negative for decreasing log-concave fitness functions, asserting that $\dnds$ is a decreasing function of $\Ne$.
+In the case of fitness equal to the folded proportion, $\chi$ is independent of $\x\eq$, meaning that $\dnds$ is linearly decreasing with $\Ne$ in log space.
+Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship, where only $\gamma$ and $\Nsite$ are free parameters. \\
+
+Other parametrization of the fitness function shed light on the robustness of the theoretical results.
+If the selective coefficient itself is not equal (equation \ref{eq:s-unfolded}) but proportional to the total amount of misfolded protein, and the expression level of the protein is denoted $y$, then the response of $\dnds$ to changes in $\Ne$ is the same as the response to changes in $y$ (See Supplementary Materials):
+\begin{align}
+s & \propto - \beta \gamma y  e^{\beta(\alpha + \gamma \Nsite \x)} \\
+\Rightarrow \chi = \frac{ \der \dnds}{\der \ln (y)} & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
+\end{align}
+Meaning we should observe the same relationship between $\dnds$ and $\Ne$ that between $\dnds$ and expression level.\\
+
+Under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the elasticity $\chi \simeq -0.002$.
+In other words, for a increase in $\Ne$ of $4$ orders of magnitude, $\dnds$ decrease approximately of $0.01$, a subtle relationship that requires laborious  effort to be detected in simulated data, and large amount of empirical data with few bias and noise to be detectable.
+
 \subsection*{Simulations confirmation}
-Our compact theoretical results are based on assumptions that are broken in practice, for example assuming that 
-\begin{enumerate}
- \item $\Nsite$ is large;
- \item selection coefficient is well approximated by the fitness derivative;
- \item genetic code is not taken into account;
- \item all unstable states are equivalently unstable.
-\end{enumerate}
-We thus first tested the soundness of our theoretical results when such assumptions are relaxed, by simulating DNA sequences evolution (See Methods). 
-Firstly, simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}).
-Secondly the parameter $\deltaGMin$ hardly changes the slope of the linear regression, as expected also theoretically.
-Decreasing $\deltaGMin$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
-Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
-However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.\\
+In order to test the soundness of our compact theoretical results, we relaxed several assumptions by simulating DNA sequences evolution (See Methods), under various range of $\Ne$ and reporting the average $\dnds$ observed along the computation.
+With respect to mutation, we assumed that all amino-acids transition are equiprobable, in other word the complexity of the genetic code is not taken into account.
+Simulating evolution of DNA sequence, and taking into account a matrix a mutation rate between nucleotide allow to test directly the robustness of the results to this assumption.
+Furthermore, with regards to fitness, we assumed that all unstable amino-acids are equivalently unstable.
+This assumption is relaxed by choosing for every site of the sequence one stable amino-acid, and the impact on the phenotype ($\x$) of other amino-acids is not binary, but scaled by the Grantham amino-acid distance \cite{Grantham1974}.
+Also, the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
+The robustness of such approximation is test in the simulations with sequences of a finite number of sites ($\Nsite=300$).\\
 
-Finally, we compare our results to more complex models of sequence evolution leveraging the $3$D structure of protein \cite{Williams2006, Goldstein2011, Pollock2012}.
-In such models, the free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
-The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
-The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$.
+Simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}, panel B).
+Secondly we observe that effect of the parameter $\alpha$ is subtle on the slope of the linear regression, as expected also theoretically (See Figure \ref{fig:GoldsteinVsToy}, panel A).
+Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching the point of marginal stability.\\
+
+Finally, we relaxed our assumption that each site of the sequence contribute independently to $\deltaG$, by taking into account the $3$D structure of protein.
+We implemented the original model \cite{Williams2006, Goldstein2011, Pollock2012}, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985} (see Supplementary Materials).
+The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$ \cite{Goldstein2013}.
 Using extensive simulations, we argue instead that $\dnds$ is approximately linear with log-$\Ne$ (see Figure \ref{fig:GoldsteinVsToy}, panel C).
 Moreover, the observed slope match the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
-In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
+In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the optimal sequence of $300$ sites \cite{Goldstein2011}.
 \begin{figure*}[htb!]
@@
 \end{figure*}
-\subsection*{Epistasis determines the time to relaxation}
+\subsection*{Epistasis and time to relaxation}
 Although the equilibrium value of $\dnds$ after changes in $\Ne$ is an important feature of the $\dnds$-$\Ne$ relationship, another parameter that is scarcely studied is the relaxation time to reach the new equilibrium $\dnds$ \cite{Jones2016}.
@@
 
-\subsection*{Distribution of fitness effects}
-DNA mutations changing a genotype can result in a change of phenotype, and ultimately a change in fitness.
-From a specific genotype, all the possible mutations thus result in a distribution a phenotypic effect (DPE) and fitness effects (DFE).
-The DPE and DFE are not known a priori, but are the resulting consequence of the mutation-selection-drift balance (see Figure \ref{fig:Distribution}).
-Empirically, these distributions are of particular importance since they can be obtained experimentally or inferred with other data.
-As an example, DFE can be inferred from polymorphism dataset \cite{Eyre-walker2007, Galtier2016}. 
-Moreover, the distribution of $\deltadeltaG=\gamma \Nsite \dx$ of novel mutations can be obtained experimentally.
-\begin{figure*}[htb!]
- \begin{mdframed}
-  \centering
-  \includegraphics[width=0.9\textwidth] {artworks/DPE-DFE.pdf}
-  \caption{
-   \textbf{Distribution of phenotypic and fitness effects.}
-   Distribution of fitness effects and phenotypic effect for novels non-synonymous mutations observed along a simulation at the mutation-selection balance.
-$\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-   Each side of the distribution is fitted to a gamma distribution, shown in dotted line. 
-   Panel A. Distribution of observed $\deltadeltaG$, which fit adequately the gamma distribution for negative $\deltadeltaG$ (stabilizing mutations). 
-   Panel A.
-Distribution of observed selection coefficient, which fit adequately the gamma distribution for both positive and negative selection coefficient.
-However the shape parameter estimated is not the same for positive and negative selection coefficients.
-  }
-  \label{fig:Distribution}
- \end{mdframed}
-\end{figure*}
+Previous studies have argued that model of protein evolution should consider epistasis for various empirical reasons such as rate heterogeneity along the sequence, rate and time dependence of convergence, role of compensatory substitutions, prediction of deleterious mutations \cite{Goldstein2017}.
+We argue that epistasis also have an important role the elasticity of $\dnds$ to changes in $\Ne$.
+If epistasis is not taken into account, the elasticity of $\dnds$ is spread along time incompatible with biological evolution. \\
+% Epistasis influence relaxation time
+
 \section*{Discussion}
-Goldstein \& Pollock argued that model of protein evolution should consider how the acceptance of mutations depends on the sequence context in which they arise (epistasis), as to correctly represent resultant rate heterogeneity along the sequence, to predict the role of compensatory substitutions in protein evolution, predict which of the 10\% of deleterious mutations in humans are harmless in other species, or to accurately represent the rate and time dependence of convergence and homoplasy \cite{Goldstein2017}.
-We argue that not taking into account epistasis also have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$.
-Paradoxically, not taking into account epistasis means overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.\\
-If using a distribution of fitness effect (DFE) is mathematically convenient and does not suffer from underestimating the relaxation rate, it is also overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.
-As such, developing inference of fitness landscape will require to assume site-interdependent fitness landscapes.\\
+The negative relationship between the relative substitution rate of selected mutations ($\dnds$) and $\Ne$ has been predicted theoretically under the quasi-neutral theory of evolution.
+Empirical data showed mitigated confirmation of this prediction, encouraging alternative explanatory mechanism.
+One such explanation, developed in the context of genotype-phenotype-fitness map, demonstrated the absence of $\dnds$ elasticity to changes in $\Ne$ if the probabilities of phenotype changes are invariant to the current phenotype.
+We argue that this assumption can be relaxed, and provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in this context.
+We apply our framework in the special case of fitness proportional to the proportion of folded proteins. \\
+% Goal and theoretical results
+
+Our theoretical results demonstrate that the relationship between $\dnds$ and $\Ne$ (in log space) is linear with a negative slope.
+This slope is a function of molecular parameters of the model, and is inversely proportional to the product of the sequence size and the average change in conformational energy of destabilizing mutations.
+Based on empirical estimate of the molecular parameters, the elasticity of $\dnds$ to changes in $\Ne$ would be $\chi \simeq 0.002$.
+Our compact theoretical results are supported by more complex simulations of protein evolution relaxing several assumptions.
+% Our theoretical results are robsut to approximations
+
+Particularly, our theoretical prediction match a numerical model of protein evolution, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations.
+Previous studies using this model presented an apparent lack of $\dnds$ elasticity to changes in $\Ne$ \cite{Goldstein2013}.
+We argue the apparent lack of elasticity is in fact due to the subtle relation, and require extensive computation to be detected.\\
+% Goldstein \& Pollock model don't have a null elasticity
+
+Empirically, inference of the $\dnds$ elasticity in Primates, using polymorphism and divergence data estimated $\chi \simeq $, approximately $X$ times greater than our theoretical prediction \cite{Brevet2019}.
+More empirical data across clades would be required to infirm such model, but in the light of empirical data estimating the relationship between $\Ne$ and $\dnds$, models based on the probability of folding are not reasonably fitted.\\
+% Our result show the fitness proportional to proportion of folded folding is at odd with empirical data
+
+Furthermore, our theoretical results manifest that the relationship between $\Ne$ and $\dnds$ is the same as the relationship between protein abundance and $\Ne$, whenever protein fitness is determined by the deleterious effect of unfolded proteins.
+As such, dataset relating $\dnds$ to $\Ne$ and protein can be used confirm or infirm this model.\\
+% Stress the importance that elasticity is valid for Ne and protein abondance
+
+Notably, theoretical approximations apply more broadly to protein-protein interactions (See Supplementary Materials), where protein may either be in free form or engaged in a non-specific interaction.
+In non-specific interactions, stable site are hydrophilic amino-acid and unstable site are hydrophobic amino-acid at the protein surface.
+Fitting this model with empirical estimates \cite{Zhang2008}, we obtain an elasticity of $\chi = 0.2$ (See Supplementary Materials) thus a much stronger response than under the model based on conformational stability.
+This effect is due to less sites in the protein being involved in protein-protein interaction than for conformational stability, in addition to the lower energy engaged in the contacts.
+As such, in light of empirical data available, the assumption that protein fitness is determined such as to avoid non-specific interaction is more likely than the assumption that protein fitness is determined by the deleterious effect of unfolded proteins.
+\\
+% Protein-protein interactions
+
+% DFE can also be used to fit to empirical data - Shall we present this ?
+
 \section*{Materials \& Methods}
-Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}., where the rate of substitution of a new mutation with selection coefficient $s$ is:
+Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}, where one sequence represent the whole population.
+From the resident DNA sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
+For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codons that are one mutation away and that are not stop codon.
+For each mutant sequence $\cj \in \setNeighbors$, we compute its fitness and subsequently the selection coefficient of the mutant:
 \begin{equation}
-p_{\text{fix}}(s) = \mu \dfrac{4 \Ne s}{1 + e^{- 4 \Ne s }}, 
+s \left( \ci,\cj\right) = \dfrac{ f \left(\cj \right) - f \left(\ci\right) }{f\left( \ci \right)}.
 \end{equation}
-where $\mu$ is the mutation rate.
-\subsection*{Simulation of protein folding with an additive model of free energy}
-\label{MatMet:folding}
-From a DNA sequence $\ci$ after $t$ substitutions, the protein's difference in free energy between folded and unfolded state is given by:
-\begin{equation*}
-\deltaG\left(\ci\right) = \deltaGMin + \Nsite \gamma * \x\left(\ci\right), 
-\end{equation*}
-where $\x\left(\ci\right)$ is the proportion of non-optimal amino-acid sites in the sequence.
-Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Boltzmann equation: 
-\begin{equation}
-f(\deltaG\left(\ci\right)) = \dfrac{P_{\mathrm{F}}\left(\ci\right)}{P_{\mathrm{F}}\left(\ci\right) + P_{\mathrm{U}}\left(\ci\right)} = \dfrac{e^{-\beta G_{\mathrm{F}}\left(\ci\right) }}{e^{-\beta G_{\mathrm{F}} \left(\ci\right) } + e^{-\beta G_{\mathrm{U}}\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }},
-\end{equation}
-where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
-From the resident sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
-For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codons that are one mutation away and that are not stop codon.
-For each mutant sequence $\cj \in \setNeighbors$, we compute $\deltaG^{t+1}$ from the updated sequence $\cj$, and subsequently the selection coefficient of the mutant:
-\begin{equation}
-s \left( \ci,\cj\right) = \dfrac{ f\left( \deltaG \left(\cj\right) \right) - f\left( \deltaG \left(\ci\right) \right)}{f\left( \deltaG \left(\ci\right) \right)}.
-\end{equation}
-The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
+The next event of mutant invading the population, and the time to reach such event is chosen using Gillespie algorithm, according to the rates of substitution between sequences:
 \begin{equation}
@@
 where $\mu_{\itoj}$ is the mutation rate between $\ci$ and $\cj$, determined by the underlying $4x4$ nucleotide mutation rate matrix, and ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
-\subsection*{Simulation with independent fitness profiles}
-A fitness profile give a fitness for each amino-acid (vector of size $20$).
+Various optimization are implemented to reduce the computation time of mutant's fitnesses.
+The simulation starts with a burn-in period to reach mutation-selection-drift equilibrium.
+\subsection*{Models of fitness function}
+\label{MatMet:folding}
+Under a simulation of protein folding with an additive model of free energy, the protein's difference in free energy between folded and unfolded state is given by:
+\begin{equation*}
+\deltaG\left(\ci\right) = \alpha + \Nsite \gamma * \x\left(\ci\right), 
+\end{equation*}
+where $0 \leq \x\left(\ci\right) \leq 1$ is the distance of $\ci$ to the optimal sequence.
+For each site of sequence, the optimal amino-acids are chosen randomly at initialization, and the distance between the current amino-acid and the optimal is scaled by the Grantham amino-acid distance \cite{Grantham1974}.
+Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Fermi-Dirac distribution: 
+\begin{equation}
+f(\deltaG\left(\ci\right)) = \dfrac{e^{-\beta \deltaG\left(\ci\right) }}{1 + e^{-\beta \deltaG\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }}, 
+\end{equation}
+where $\beta$ is the inverse of the temperature ($\beta=1/kT$).\\
+
+Under a simulation with site independent fitness profiles, a fitness profile give a fitness for each amino-acid (vector of size $20$).
 Each site of the protein has a specific amino-acid fitness profile.
@@
 where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \cite{Bloom2017}.
-The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
-\subsection*{Simulation with distribution of fitness effects (DFE)}
-The selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
+
+Under simulation with a fixed distribution of fitness effects (DFE), the selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
 \begin{equation}
@@
 \end{equation}
-The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
 \subsection*{$\bm{\dnds}$ along the simulation}
@@
 \end{align}
+And $\dnds$ is taken as the average of the time-dependent $\dnds(t)$ along the simulation.
 \subsection*{Reproducibility}

### 2020-04-22 manuscript/supp-mat.tex
Refactored manuscript

@@
 \end{center}
+Moreover, given that the number of state if large enough $\Nstate \gg 1$, the equilibrium $\dnds$ can be approximated as:
+\begin{align}
+\dnds\eq  & = \x\eq \left[ \frac{2 (\x\eq-1)  \lambda_{K}(\x\eq) }{\Nstate (\x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right], \\
+& \simeq \x\eq
+\end{align}
 And the derivative of $\dnds\eq$ w.r.t to $\x\eq$ is: 
@@
 \begin{align}
-\frac{ \der \dnds\eq}{\der \x\eq}  & = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}\text{ from eq.
-\ref{eq:approx_exp}}, \\
+\frac{ \der \dnds\eq}{\der \x\eq}  & = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}, \\
 & \simeq \frac{2 \lambda_{K}(\x\eq)}{\Nstate} + 1, \\
@@
 \begin{gather}
-f(\x) = \frac{1}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}}. \label{eq:fitness}
+f(\x) = \frac{1}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}}. \label{eq:fitness}
 \end{gather}
@@
 \begin{align}
-\frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
-& = - \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \label{eq:df_dx},\\
-& \simeq - \beta \Nsite \gamma \e^{\beta(\deltaGMin + \Nsite \gamma \x)}.
+\frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\alpha + \Nsite \gamma \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
+& = - \beta \Nsite \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}} \label{eq:df_dx},\\
+& \simeq - \beta \Nsite \gamma \e^{\beta(\alpha + \Nsite \gamma \x)}.
 \end{align}
@@
 \begin{gather}
-\lambda_{K}(\x\eq) = 4\Ne \beta \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x\eq)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x\eq)}}  \text{ from eq.
-\ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
+\lambda_{K}(\x\eq) = 4\Ne \beta \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x\eq)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x\eq)}}  \text{ from eq. \ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
 \end{gather}
-Using $\Ne=10^4$, $\beta=1.686$, $\deltaGMin = -118$, $\Nsite=300$, $\gamma=1$, we have the following :
+Using $\Ne=10^4$, $\beta=1.686$, $\alpha = -118$, $\Nsite=300$, $\gamma=1$, we have the following :
 \begin{center}
@@
 \begin{align}
-\frac{\partial^2 \ln (f(\x))}{\partial \x^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \right) \text{ from eq.
+\frac{\partial^2 \ln (f(\x))}{\partial \x^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}} \right) \text{ from eq.
 \ref{eq:df_dx}}, \\
- & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\deltaGMin + \Nsite \gamma \x)}}{\left( 1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}\right)^2}, \\
- & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\deltaGMin + \Nsite \gamma \x)}} \frac{ \partial \ln f(\x) }{\partial \x} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}, \\
+ & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{\left( 1 + \e^{\beta(\alpha + \Nsite \gamma \x)}\right)^2}, \\
+ & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}} \frac{ \partial \ln f(\x) }{\partial \x} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}, \\
  & \simeq \beta \Nsite \gamma \frac{ \partial \ln f(\x) }{\partial \x} 
@@
 
+\subsection*{Distribution of fitness effects}
+DNA mutations changing a genotype can result in a change of phenotype, and ultimately a change in fitness.
+From a specific genotype, all the possible mutations thus result in a distribution a phenotypic effect (DPE) and fitness effects (DFE).
+The DPE and DFE are not known a priori, but are the resulting consequence of the mutation-selection-drift balance.
+Empirically, these distributions are of particular importance since they can be obtained experimentally or inferred with other data.
+As an example, DFE can be inferred from polymorphism dataset \cite{Eyre-walker2007, Galtier2016}. 
+Moreover, the distribution of $\deltadeltaG=\gamma \Nsite \dx$ of novel mutations can be obtained experimentally.
+
+\begin{center}
+	\includegraphics[width=0.8\textwidth] {artworks/DPE-DFE.pdf}
+\end{center}
+
+Distribution of fitness effects and phenotypic effect for novels non-synonymous mutations observed along a simulation at the mutation-selection balance.
+$\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+Each side of the distribution is fitted to a gamma distribution, shown in dotted line. 
+Panel A. Distribution of observed $\deltadeltaG$, which fit adequately the gamma distribution for negative $\deltadeltaG$ (stabilizing mutations). 
+Panel A.
+Distribution of observed selection coefficient, which fit adequately the gamma distribution for both positive and negative selection coefficient.
+However the shape parameter estimated is not the same for positive and negative selection coefficients.
+
 \bibliographystyle{apalike}

### 2020-04-22 manuscript/manuscript.tex
Nicolas corrections

@@
 \captionsetup{width=0.85\textwidth}
+\renewcommand{\baselinestretch}{1.5}
+
 \include{notations}
@@
 	}
+
 	\section*{Introduction}
-Molecular sequences differ across species due to the particular history of DNA substitutions along each specie's respective lineage.
+
+	[Concernant l'introduction, j'ai pris la liberté de lisser directement le texte, de rajouter quelques petites phrases ici et là, pour compléter et ajuster le fil de l'argument.
+Il me semble pour autant que je n'ai pas trop trahi ton fil directeur, du moins dans la première partie.
+Sur la fin de l'intro, par contre, je pense qu'il faut remanier un peu plus.
+En particulier, (1) la question de l'articulation avec la question de la relation entre dnds et expression, et (2) le fait de bien terminer toute l'analyse de l'état des lieux avant d'en venir à la contribution spécifique de ce manuscrit; dans la version que tu avais proposé, ces deux aspects avaient tendance à se mélanger un peu sur la fin.]
+
+	[je propose de scinder ton premier paragraphe en deux: un premier de perspective un peu plus générale sur l'enjeu de démêler les forces évolutives à partir des patterns de substitution, un deuxième qui en vient plus spécifiquement au rôle de la dérive génétique, et qui s'articule du coup plus directement avec la question de la théorie quasi-neutre (en passant, en anglais, c'est nearly-neutral, et non quasi-neutral).
+Du coup, concernant ce premier paragraphe-ci, il faut aller jusqu'au bout de l'argument, et j'ai rajouté deux phrases en ce sens.]
+	Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
 	These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
-These forces have effects at different levels, mutations are carried by molecular sequence, selection is beared by individual's fitness, and random genetic drift is a population effect.
-The strength of random genetic genetic, or stochastic sampling of mutations, is less pronounced in lineage with high effective population size ($\Ne$), and as a consequence the purification by selection of deleterious mutations is more effective.
-Such idea has emerged in the framework of quasi-neutral theory of evolution, whereby predicting a decrease in the substitution rate per unit of time of selected mutations along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.
-This prediction is valid under the assumption that most mutations are deleterious and that their selective effect on individuals are drawn from a fixed distribution \cite{Welch2008}.
-Furthermore, this prediction assumes a constant mutation rate per unit of time, but such assumption can be relaxed by normalizing the substitution rate of selected mutations by the substitution rate of neutral mutations such as to discard this confounding factor \cite{Ohta1972}.
-This relative substitution rate of selected mutations, called $\dnds$ is this study, is a theoretical macroscopic observable in molecular sequence evolution, given we can separate selected and neutral mutations.
-Practically, $\dnds$ can be approximated in protein-coding DNA sequences, where selected mutations are the non-synonymous changes affecting the amino-acid sequence, and the non-synonymous mutations are supposedly neutral, this statistic is called $dNdS$.\\
-% \dnds decrease with \Ne under the assumption of fixed DFE
+	These forces have effects at different levels: mutations are carried by molecular sequence, selection is mediated at the level of individuals, while random genetic drift is a population effect.
+	Yet, they jointly contribute to the long-term molecular evolutionary process.
+	Thus, the challenge of molecular evolution is to tease out their respective contributions, based on comparative analyses.
 
-The context of protein coding sequence fostered another modeling of selection, based on genotype-fitness map instead of distribution of fitness effects.
-In such approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998}.
-For example, if the target amino-acid has an higher fitness, the selective effect of the mutation is positive, and reciprocally negative for lower fitness.
-Such modeling approach account for higher substitutions rate between amino-acid with similar bio-chemical properties.
-Fundamentally, in such framework the fitness depends solely on the actual genotype, not on the trajectory of mutations leading to this sequence.
-Even thought the modeling approach is substantially different, assuming each site of the sequence has its own genotype-fitness map also predicts a negative correlation between $\dnds$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.\\
-% \dnds decrease with \Ne under the assumption site fitness landscape
+	[The specific role of random drift $\to$ the nearly-neutral argument.]
+	One main aspect of this challenge is to correctly evaluate the role of random drift in the long term evolutionary process.
+	Population genetics theory implies that the strength of drift, due to the stochastic sampling of mutations, is less pronounced in lineage with large effective population size ($\Ne$), and as a consequence, the purification by selection of weakly deleterious mutations is more effective.
+	This fundamental idea is at the core of the nearly-neutral theory of evolution.
+	The nearly-neutral theory posits that a substantial fraction of mutations are weakly deleterious [d'abord expliciter cette composante importante de la théorie quasi-neutre].
+	As a result, the theory predicts a decrease in the substitution rate per unit of time of selected mutations along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}, relative to the neutral substitution rate.
+	[In its original version, the nearly-neutral theory also assumes a constant mutation rate per unit of time, but such assumption can be relaxed by normalizing the substitution rate of selected mutations by the substitution rate of neutral mutations such as to discard this confounding factor \cite{Ohta1972}.
+{\it pas tout à fait vrai: en fait, la théorie invoquait un effet temps de génération, qui était compensé par un effet Ne, résultant en une horloge moléculaire au niveau non-synonyme.
+Mais cela me parait un peu compliqué.
+Peut-être zapper ce point?}]
 
-Empirically, fluctuation of $\dnds$ along lineages of phylogenies have been inferred using protein-coding DNA alignments \cite{Yang2001, Zhang2004}.
-Subsequently, inference methods combining molecular sequences and lineage specific quantitative traits have asserted that $\dnds$ correlates positively with longevity and body mass, traits which have been sampled only in the extant lineages \cite{Lartillot2011,Weber2014}.
-Moreover, longevity and body mass are proxies of $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}.
-As such, empirical evidences suggest a negative correlation between $\dnds$ and $\Ne$, confirming the theoretical prediction of the quasi-neutral theory of evolution.
-However, this correlation could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation could be in the inverse direction or not statistically significant \cite{Figuet2016}.\\
-% \dnds decrease with \Ne empirically, but not always
+	The relative substitution rate of selected, relative to neutral, mutations (called $\dnds$ is this study) is a theoretical macroscopic observable in molecular sequence evolution [formulation compliquée..
+is an observable quantity of the molecular evolutionary process?], at least if we can separately estimate the rate of neutral mutations in a phylogenetic or comparative context.
+Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while those selected mutations are the non-synonymous changes affecting the amino-acid sequence, $\dnds$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
+Thus, in practice, the nearly-neutral argument translates into a predicted decrease in $\dnds =  dN/dS$ as a function of long-term changes in $N_e$.
 
-Mitigated empirical evidence that $\dnds$ decrease with $\Ne$ encouraged alternative explanatory mechanism \cite{Lanfear2014}.
-One striking result proved that $\dnds$ can independent of $\Ne$, when fitness is a function of a phenotype and that such phenotype is equimutable.
-Equimutability states that the probabilities of phenotype changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
-Such argument has been invoked in the \textit{in silico} observation that $\dnds$ is seamingly independent of $\Ne$ if the protein thermodynamic stability is computed using a 3D structural model of the protein, and that the fitness is proportional to the folding probability of the protein \cite{Goldstein2013}.
-The given explanation is that probabilities of changes in free energy of folding ($\deltadeltaG$) is independent of the current free energy ($\deltaG$), which is considered an equimutable phenotype.\\
-% Explanation could be that \dnds is independent of \Ne
+	This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects \cite{Welch2008} {\it citer également Kimura, 1979}.
+Be more specific here: Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $dnds$ as a function of $N_e$ (i.e.
+$\dnds \sim N_e^{-\beta}$), where $\beta$ is the shape parameter of the DFE.
+In practice, DFEs are strongly leptokurtic: thus, predicts a weak negative relation between $\dnds$ and $N_e$.
 
-However, this assumption is relatively strong, as for example if a protein is completely stable, only destabilizing mutations can occur, and more generally the probabilities of destabilizing mutational events is dependent on the actual composition of the protein.
-Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\dnds$ relationship to $\Ne$ is unknown.
-The goal of this study is to derive the $\dnds$ relationship to $\Ne$ to intrinsic molecular parameters of the model.
-More generally, we develop mathematical tools to derive the $\dnds$ relationship to $\Ne$ in the context of a genotype-phenotype-fitness map.
-With such results, we stress to test the assumption that protein stability is a proxy of fitness in the light of empirical data.\\
+	The context of protein coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
+	In this alternative approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998}.
+	For example, if the target amino-acid has a higher fitness than the current amino acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
+	Such a modeling approach predicts higher substitution rates between amino-acid with similar biochemical properties.
+	More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
+	Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\dnds$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.
+
+	Empirically, variation in [fluctuation is more in the short term, I would say] $\dnds$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \cite{Yang2001, Zhang2004}.
+[Ici, donner un compte-rendu un peu plus exhaustif des analyses empiriques publiées au sujet de la relation dnds / Ne.
+Possiblement en deux temps: les analyses non intégratives: Popadin, et d'autres; puis, les approches de type coevol.]
+	Subsequently, inference methods combining molecular sequences and lineage specific quantitative traits have found that $\dnds$ correlates positively with life-history traits such as longevity and body mass \cite{Lartillot2011,Weber2014}.
+	Since longevity and body mass typically correlate negatively with $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}, these empirical correlations suggest a negative correlation between $\dnds$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
+	However, this correlation between $\dnds$ and life-history traits could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation was found to be either not statistically significant (ref?), or even in the opposite direction (i.e.
+negative between $dnds$ and LHT) \cite{Figuet2016} (other refs?).
+
+	Mitigated empirical evidence about the response of $\dnds$ to changes in $\Ne$ encouraged the search for alternative explanatory mechanisms \cite{Lanfear2014}.
+	In this direction, one striking result was the proof that $\dnds$ is in fact predicted to be independent of $\Ne$ under very general cirumstances, namely [je propose d'être plus exhaustif sur l'ensemble des conditions des résultats de Cherry], whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
+Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
+	This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \cite{Goldstein2013}.
+The thermodynamic stability is itself computed using a 3D structural model of the protein.
+These simulation experiments have led to the observation that $\dnds$ is essentially independent of $\Ne$.
+The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.
+
+	However, the equimutability assumption is a relatively strong one, which also conflicts with simple combinatorial considerations about the relation between sequence and phenotype.
+For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur.
+More generally, assuming that the stability of a protein sequence reflects an underlying fraction of positions having already accepted destabilizing amino-acids, then the probability of destabilizing mutational events is in turn expected to directly depend on the current stability the protein.
+
+	[Ici, il faut élaborer beaucoup plus en détail la question de l'articulation avec niveaux d'expression. Voici ce que je propose.]
+	In addition, if empirical evidence for a negative correlation of $\dnds$ with $N_e$ is still not totally convincing, another empirical correlation is known to be much more robust, namely, that of $\dnds$ with expression level or protein abundance.
+Indeed, expression level is one of the best predictors of $\dnds$, with highly expressed proteins typically having lower $\dnds$ values.
+Of note, the slope of the response of $\dnds$ to changes in expression level is relatively weak (ref Drummond), although clearly significant.
+Importantly, many of the theoretical models of protein-coding sequence evolution that have been invoked to explain this correlation between $\dnds$ and expression level are such that the response of $\dnds$ to changes in expression level or in $N_e$ should be very similar.
+Conversely, under strict equimutability, most of these models would predict that $\dnds$ should be essentially independent, not just of $N_e$, but also of expression level, thus at odds with empirical observations.
+
+	[La aussi, il faut ressaisir l'ensemble de la question.
+Il y a plusieurs enjeux: il faut se positionner de façon claire, mais diplomatique, par rapport aux résultats et interprétations déjà proposés, par Cherry, Goldstein et d'autres: concéder que la relation omega Ne est faible, mais pas inexistante, et devant être quantifiée.
+D'autre part, il faut commencer à dessiner une perspective qui justifie le travail que tu vas présenter, et bien souligner comment ce travail théorique pourrait s'articuler avec les questionnements empiriques.]
+	Overall, there is no doubt that the impact of changes in $N_e$ (or in expression level) on $\dnds$ is at best weak, making empirical estimation more difficult.
+Even if weak, however, an exact theoretical understanding and quantification of how the slope of this relation depends on the underlying map between genotype, phenotype and fitness, would be useful.
+[Ultimately, it could be confronted with more refined empirical estimates of the relation between $\dnds$ and both $N_e$ and expression level - à développer ici: rcomment la pente dépend de paramètres structuraux, et du coup, est potentiellement informative au sujet de ce qui détermine ultimement le paysage sélectif].
+
+	Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium.
+However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\dnds$ due to adaptive selection, and subsequently a decrease in $\dnds$ due to stronger purifying selection in the long term \cite{Jones2016}.
+	Studying only equilibrium properties can thus be misleading.
+For that reason, the dynamical response of $\dnds$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized [là aussi, justifier l'articulation entre compréhension théorique et impact potentiel sur l'analyse empirique].
+
+	%Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\dnds$ relationship to $\Ne$ is unknown.
+	[Et maintenant seulement, tu peux décliner le contenu du manuscrit.] The goal of this study is to characterize how the response of $\dnds$ to changes in $\Ne$ quantitatively depends on the intrinsic [structural?] molecular parameters of the model.
+To this end, we develop a general mathematical approach for deriving the $\dnds$ relationship to $\Ne$ in the context of a given genotype-phenotype-fitness map.
+[Based on these results, we stress to test the assumption that protein stability is a proxy of fitness in the light of empirical data -- Me paraît être dit de façon un peu trop brutale, et trop naive aussi: aucun doute que la contrainte conformationnelle est une composante essentielle de la sélection purificatrice sur les protéines.
+La question est plutôt de savoir dans quelle mesure les éventuelles modulations du dnds en fonction de Ne dépendent de cette composante-là de la fitness, ou alors d'autres enjeux biophysiques, ou autres.
+Pourrais dire quelque chose d'un peu plus indirect, par exemple: we briefly discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the slope of the relation between $\dnds$ and $N_e$.
+Quoi qu'il en soit, la fin de ce paragraphe est encore incomplète et pourra être revue une fois que la discussion aura été complétée.
+]
 	% The assumptions are strong, we want to relax equimutability which is unlikely
 
-Orthogonally to the relationship between $\dnds$ and $\Ne$, independent studies have argued that protein abundance is also a predictor of $\dnds$.
-As such, we also seek to derive the $\dnds$ relationship to protein abundance, and discuss the empirical implication.\\
-% \dnds is relationship to protein abundance
+	\section*{Results}
 
-Lastly, the theoretical results discussed so far are all obtained when the balance between mutation, selection and drift is at equilibrium.
-However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ leads first to an increase of $\dnds$ due to adaptive selection, and subsequently a decrease in $\dnds$ due to stronger purifying selection \cite{Jones2016}.
-Studying only equilibrium properties can thus be misleading, and the dynamical response of $\dnds$ to changes in $\Ne$ must also be addressed by our model.\\
-% Result at equilibrium, what about dynamical properties?
+	\subsection*{Models of evolution}
 
-\section*{Results}
-\subsection*{Theoretical approximations}
-In this section we present an analytical approximate solution for the change in $\dnds$ at equilibrium after a change in $\Ne$.
-The main results are given in a general case of any phenotype-fitness map, and also derived in the specific case of fitness proportional to the total amount of folded protein.
-All development are available in Supplementary Materials in the most general case.
-The first step define the relationship from genotype to fitness, through an intermediary phenotype.
-The second step present the equation of the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
-Finally, an approximate response of $\dnds$ to changes in $\Ne$ is derived from the equilibrium equation. \\
+	[ Il faut présenter le cadre général: deux niveaux de modèles: modèle abstrait, modèle thermo, et justifier leur articulation].
+	The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$ (i.e.
+such that the coding positions of the sequence contribute additively to the trait), under directional selection specified by a decreasing and log-concave fitness function $f(x)$.
+As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability.
+This model is inspired from previous work \cite{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations.
+	%Throughout our derivation, we present the results both for the general case and for the more specific example.
 
-We base our work on numerical model of protein stability \cite{Williams2006, Goldstein2011, Pollock2012}, on which we make simplifying assumptions to derive analytical equations.
-In the original model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
-Technically, the free energy of the folded are unfolded conformations are computed using $3$D conformations of the protein.
-In such model, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.\\
+	In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
+	Technically, the free energy of the folded are unfolded conformations are computed based on the $3$D conformations of the protein, and using a statistical potential.
+	In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.
 
-We approximate this model such that the destabilizing effect of an amino-acid does not depend on a $3$D representation of the protein, but each site contribute independently to $\deltaG$.
+	We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighbouring residues. Instead, each site contributes independently to $\deltaG$. 
 	With such approximation, at a given site of sequence each amino-acid is either stabilizing or destabilizing the protein, and an unstable amino-acid contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ compare to a stable amino-acid.
-If all amino-acids of the sequence are stable, $\deltaG$ equal to $ \alpha < 0$. 
-In this model, the more compact phenotype summarizing the genotype is the proportion of unstable amino-acid in the sequence, defined as $0 \leq \x \leq 1$, and $\deltaG$ is a linear function of $\x$:
+[dire plus explicitement: en chaque position, un acide aminé optimal, tous les autres considérés comme également destabilisateurs].
+	If all amino-acids of the sequence are stable [je dirais plutôt: an amino-acid is stabilizing / destabilizing at a given position, plutôt que stable / unstable: the protein is stable / unstable, not the amino acid], $\deltaG$ is at its minimum, equal to $ \alpha < 0$.
+	In this model, the phenotype of a given genotype (i.e.
+sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq \x \leq 1$.
+Thus, $\deltaG$ is a linear function of $\x$:
 	\begin{align}
@@
 
-From a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of unstable amino-acid, or do nothing if the mutation is between both unstable (or equivalently stable) amino-acids.
+	[Some words about the phenotype - fitness map].
+The fitness is assumed to be equal to the proportion of protein in folded conformation.
+This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm (supp information).
+	This fitness is then given by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
+	\begin{align}
+	f(\x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma \x)}},\\
+	\Rightarrow f(\x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma \x)}, 
+	\end{align}
+	where $\alpha$ and $\gamma$ are defined as above, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
+
+	\subsection*{Elasticity of $\dnds$ to changes in $N_e$. Analytical approximation}
+
+	[Vu que c'est le résultat essentiel, je propose de mettre la notion d'élasticité beaucoup plus clairement en avant.]
+	In this section we present an analytical approximate solution for the change in [response of equilibrium] $\dnds$ at equilibrium after a change in $\Ne$.
+By analogy with thermodynamics, we call this reponse the \emph{elasticity} of $\dnds$ to changes in $N_e$, and denote it as $\chi$:
+	\begin{align}
+	\chi = \frac{ \der \dnds}{\der \ln (\Ne)}
+	\end{align}
+	The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case where the fitness is proportional to the total amount of folded protein [in the specific case of the biophysical model?].
+	All development are available in Supplementary Materials in the most general case.
+
+	[Paragraphe à revoir en fonction du changement de contexte opéré par le fait que le modèle est maintenant décrit dans la section précédente.]
+	The first step is to define the relationship from genotype to fitness, through an intermediary phenotype
+	The second step present the equation of the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
+	Finally, an approximate response of $\dnds$ to changes in $\Ne$ is derived from the equilibrium equation.
+
+	First, from a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of unstable amino-acid, or do nothing if the mutation is between both unstable (or equivalently stable) amino-acids.
 	To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable and only one amino-acid is stable and the $19$ others are unstable.
@@
 	\end{gather}
-In the extreme case of optimal phenotype ($\x = 0$), only destabilization mutations are proposed.
+	In the extreme case of an optimal phenotype ($\x = 0$), only destabilization mutations are proposed.
 	Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between unstable amino-acids (effect $0$) is proportional to $\x$. 
-Once obtained the phenotypic effects of mutations and their probabilities, definition of the phenotype-fitness map is necessary derive the selective effect of mutations.\\
+	[Here, could point out that the mutation bias is proportional to $(1-x)/x$?
+And explain that this is fundamentally a combinatorial effect, due to the number of mutational opportunities available in either direction.]
 
-Subsequent equations are valid for the more general case where the Wrightian fitness is a log-concave differentiable function $f$ of phenotype $\x$.
-As an example, we develop the argument when the Wrightian fitness of an individual is the proportion of protein in folded conformation.
-This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm. 
-This fitness is then given by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
-\begin{align}
-f(\x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma \x)}},\\
-\Rightarrow f(\x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma \x)}, 
-\end{align}
-where $\alpha$ and $\gamma$ are defined as above, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
-Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
+	%Once obtained the phenotypic effects of mutations and their probabilities, definition of the phenotype-fitness map is necessary derive the selective effect of mutations.\\
+
+	Second, we need to determine the strength of selection acting on mutations. Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
 	\begin{align}
@@
 	\end{align}
+	[conversely, stabilizing mutations will be under positive selection, $+s$ - relever le fait que $s$ dépend de $x$, et qu'à cause de la log-concavité, en valeur absolue, la force de la sélection croit avec $x$.]
 
 	With the complete genotype-phenotype-fitness map defined (see Figure \ref{fig:NeChangeInfluence}, left panel), one can study the evolutionary processes unfolds.
-Starting from an optimal sequence, destabilizing mutations will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, a point of equilibrium called marginal stability.
+[à reformuler, vu que la description du map gen/phen/fit est passée plus haut.
+Par exemple: Based on these expressions for mutational and selective pressure, one can study..., fig 1 ?].
+	Starting from an optimal sequence, mostly destabilizing mutations will occur, which will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, at which point the protein will reach a point of equilibrium called marginal stability.
 	Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
@@
 	\end{align}
+	[intercaler entre les deux équations: EQ... in the general case, and EQ... in the more specific case of the biophysical model]
+	[Rajouter une interprétation: this equation essentially expresses the mutation-selection equilibrium: the left-hand side of the equation is the log of the mutation bias at $x$, while the right-hand side is simply $4 N_e s$].
+
 	This equation cannot be solved explicitly for $\x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $\x\eq$ is given in Figure \ref{fig:NeChangeInfluence} (right panel).
-\\
+[Ici, développer un peu l'explication: mutation bias decreases weakly with $x$ (blue curve on figure) (of note: not equimutable, but nearly so..).
+while the strength of selection increases sharply with $x$.
+Equilibrium at their intersection.
+Increasing Ne: shifting the selective response upward, which then results in a leftward shift of the equilbrium -- the amount of change on $x$, however, is very small, if the $s$ curve is very steep around the equilibrium set point, etc.]
+
 	\begin{figure*}[htb!]
@@
 
-Moreover, at equilibrium we can derive the expected substitution rate of mutations changing the sequence, and by normalizying by the substitution rate of neutral mutations, $\dnds$ simply approximates to:
+	[Ici, manque un lien logique: the argument thus far explains how $x*$ varies with $N_e$.
+To capture how $\dnds$ varies with $N_e$, we also need to obtain an expression for $\dnds$ as a function of $x*$ ?
+quelque chose du genre?] Moreover, at equilibrium we can derive the expected substitution rate of mutations changing the sequence, and by normalizying by the substitution rate of neutral mutations, $\dnds$ simply approximates to:
 	\begin{gather}
@@
 
-From the equilibrium equation \ref{eq:equilibrium}, differential calculus allow to derive the response of $\x\eq$ and ultimately $\dnds$ after a change in $\Ne$.
-We define the elasticity $\chi$ as slope of the $\dnds$ response to changes in $\Ne$ in log scale, which simplify to a compact equation: 
+	Finally, from the equilibrium equation \ref{eq:equilibrium}, differential calculus allow to derive the response of $\x\eq$ and ultimately $\dnds$ after a change in $\Ne$ (see supp info).
+	%We define the elasticity $\chi$ as slope of the $\dnds$ response to changes in $\Ne$ in log scale, which simplify to a compact equation: 
+	In the end, the elasticity of $\dnds$ to changes in $\Ne$ can be expressed in terms of a very compact equation: 
 	\begin{align}
@@
 	\end{align}
-The elasticity is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype.
-To note, this elasticity is strictly negative for decreasing log-concave fitness functions, asserting that $\dnds$ is a decreasing function of $\Ne$.
-In the case of fitness equal to the folded proportion, $\chi$ is independent of $\x\eq$, meaning that $\dnds$ is linearly decreasing with $\Ne$ in log space.
-Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship, where only $\gamma$ and $\Nsite$ are free parameters. \\
+	[Ici aussi, intercaler un petit morceau de texte entre les deux équations.
+Par ailleurs, je me disais que ce serait utile d'interpréter un peu plus directement la version générale en termes de courbure.
+Je propose l'expression courbure relative pour désigner le rapport de la seconde sur la première dérivée (relative à la pente, finalement..).]
+	The elasticity is thus equal to the inverse of the \emph{relative curvature}, i.e.
+the ratio of the second to the first derivatives, of the log-fitness function, taken at the equilibrium phenotype.
+Of note, this elasticity is strictly negative for decreasing log-concave fitness functions, asserting that $\dnds$ is a decreasing function of $\Ne$.
+In addition, the elasticity itself is low (i.e.
+$\dnds$ responds more weakly) for strongly concave log-fitness functions -- thus formally capturing the intuitive argument developed above (figure 1B).
+
+	In the case of the biophysical model, $\chi$ is essentially independent of $\x\eq$, meaning that $\dnds$ is linearly decreasing with $\Ne$ in log scale.
+	Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship, where only $\gamma$ and $\Nsite$ are free parameters.
+[Bon, ici, il faudrait qualifier les choses: pourquoi dire que beta n'est pas un paramètre?
+il y a tout de même des organismes hyper-thermophiles.
+Par ailleurs, c'est intéressant de voir que, bien que l'équation générale semble suggérer en apparence que l'élasticité ne dépend que de la fonction de fitness, en fait, ce que montre le modèle biophysique, c'est que l'élasticité peut en réalité être modulé de deux manières: en jouant sur la relation génotype / phénotype (ici, gamma), ou en jouant sur la relation phenotype fitness (ici, beta).
+Raison de plus pour considérer beta comme un paramètre potentiellement libre lui aussi.]
 
@@
 	In other words, for a increase in $\Ne$ of $4$ orders of magnitude, $\dnds$ decrease approximately of $0.01$, a subtle relationship that requires laborious  effort to be detected in simulated data, and large amount of empirical data with few bias and noise to be detectable.
+In addition, it would imply an equally weak relation between $\dnds$ and expression level.
 
-\subsection*{Simulations confirmation}
+	\subsection*{Simulation experiments}
+
 	In order to test the soundness of our compact theoretical results, we relaxed several assumptions by simulating DNA sequences evolution (See Methods), under various range of $\Ne$ and reporting the average $\dnds$ observed along the computation.
+[Là, ta phrase dit un peu tout d'un coup.
+Peut-être dire en plusieurs temps: Our theoretical derivation of the elasticity of $\dnds$ to changes in $N_e$ is based on several simplifying assumptions about the evolutionary model and makes multiple approximations.
+In order to test the robustness of our main result, we therefore conducted systematic simulation experiments, relaxing several of these assumptions.
+In each case, simulations were conducted under a broad range of values of $\Ne$, monitoring the average $\dnds$ observed at equilibrium and plotting the scaling of these measured equilibrium $\dnds$ as a function of $N_e$.
+Puis un nouveau paragraphe sur les différentes modalités qui sont relaxées.]
+
+	[Specifically, with respect to mutations, our derivation assumes that... ]
 	With respect to mutation, we assumed that all amino-acids transition are equiprobable, in other word the complexity of the genetic code is not taken into account.
 	Simulating evolution of DNA sequence, and taking into account a matrix a mutation rate between nucleotide allow to test directly the robustness of the results to this assumption.
-Furthermore, with regards to fitness, we assumed that all unstable amino-acids are equivalently unstable.
+	Furthermore, with regards to fitness [euh non...
+with regards to the phenotypic effects of amino-acid changes], we assumed that all unstable [destabilizing] amino-acids are equivalently unstable [have an identical impact on protein stability.
+In reality, one would expect conservative amino-acid replacements to be less destabilizing than radical changes].
 	This assumption is relaxed by choosing for every site of the sequence one stable amino-acid, and the impact on the phenotype ($\x$) of other amino-acids is not binary, but scaled by the Grantham amino-acid distance \cite{Grantham1974}.
-Also, the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
+[Ici, ca me paraîtrait plus logique d'avoir déjà posé, même dans le modèle le plus simple, qu'en chaque position, il y a de toute manière un acide aminé optimal -- même si en pratique, sous le modèle uniforme, on n'a finalement pas besoin de préciser lequel, néanmoins, conceptuellement, le modèle ne fait de sens que parce qu'on imagine qu'il y en a un, donc ca mérite d'être déjà présent dès le début.
+Jj'ai proposé quelque chose en ce sens tout au début de la section résultats.
+Du coup, une fois que ce point est posé plus haut, alors il suffit de dire que les effets destabilisateurs en chaque position, au lieu d'être identiques pour tous les acides aminés alternatifs, sont maintenant proportionnels à la distance de Grantham entre l'acide aminé optimal en cette position et l'acide aminé proposé par la mutation non-synonyme.]
+	Also [Finally], the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
 	The robustness of such approximation is test in the simulations with sequences of a finite number of sites ($\Nsite=300$).\\
 
-Simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}, panel B).
-Secondly we observe that effect of the parameter $\alpha$ is subtle on the slope of the linear regression, as expected also theoretically (See Figure \ref{fig:GoldsteinVsToy}, panel A).
-Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching the point of marginal stability.\\
+	Simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is indeed linear, at least in the range explored here, and that the slope of the linear regression matches the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}, panel B [devrait toujours commencer par panel A, puis B..
+quitte à inverser les panels sur ta figure]).
+	Secondly we observe that the parameter $\alpha$ has virtually no effect on the slope of the linear regression, as expected also theoretically (See Figure \ref{fig:GoldsteinVsToy}, panel A).
+Instead, decreasing $\alpha$ (to more negative values) merely results in an overall increase in $\dnds$ over the whole range of $N_e$ (i.e.
+has an impact on the intercept, not on the slope).
+This is due to the fact that decreasing $\alpha$ shifts the equilibrium to higher $\x\eq$, since more unstable sites can then reach fixation before reaching the point of marginal stability.\\
 
-Finally, we relaxed our assumption that each site of the sequence contribute independently to $\deltaG$, by taking into account the $3$D structure of protein.
+	Finally, we relaxed our assumption that each site of the sequence contribute independently to $\deltaG$, by taking into account the $3$D structure of protein [and using a statistical potential to estimate $\deltaG$].
 	We implemented the original model \cite{Williams2006, Goldstein2011, Pollock2012}, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985} (see Supplementary Materials).
@@
 	Using extensive simulations, we argue instead that $\dnds$ is approximately linear with log-$\Ne$ (see Figure \ref{fig:GoldsteinVsToy}, panel C).
-Moreover, the observed slope match the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
+	[Ici, je dirais les choses plus diplomatiquement:
+	The original works showed that under such $3$D model $\dnds$ is approximately independent of $\Ne$ \cite{Goldstein2013}.
+	Using extensive simulations in order to obtain sufficient resolution, we observe that $\dnds$ is in fact weakly dependent on $N_e$, being approximately linear with log-$\Ne$ (see Figure \ref{fig:GoldsteinVsToy}, panel C).].
+	Moreover, the observed slope matches the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
 	In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the optimal sequence of $300$ sites \cite{Goldstein2011}.
@@
 	We observed in our simulations that the determining factor of the relaxation time is the number of sites $\Nsite$ (See Figure \ref{fig:relaxStability}).
-Such observations match the theoretical prediction that high epistasis leads to faster return to equilibrium, since more opportunities are available.
-In the case of a fitness landscape with epistasis, after a change in $\Ne$, any site that is used to climb either up or down the fitness landscape will have diminishing return in other sites of the sequences.
-Not taking into account epistasis have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$ .
-Using a site-specific fitness landscape, where each amino-acid have different fitness, thus with no epistasis, the relaxation time is long since every site has to adapt to the new change in $\Ne$.
-However, using a fixed distribution of fitness effect (DFE) does not suffer from underestimating the relaxation rate, since the effect is instantaneous (see Figure \ref{fig:relaxStability}).
+	These observations match the theoretical prediction that high epistasis leads to faster return to equilibrium, since more mutational opportunities are available for driving the trait closer to equilibrium.
+	More specifically, in the case of a fitness landscape with epistasis, after a change in $\Ne$, any site that contributes to pushing the trait is used to climb either up or down the fitness landscape will have diminishing return in other sites of the sequences.
+[Formulation pas très claire..
+je me demande si la phrase juste avant ne suffit pas..]
+
+	[Not taking into account epistasis have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$ .
+'overestimating' dans le contexte présent ne veut pas dire grand chose: on n'estime rien à ce stade.
+Par ailleurs, sauter d'un seul coup à des modèles de type mutsel site-indépendents ou DFE est un peu abrupt.
+Mettre en contexte.
+Par exemple:]
+	It may be useful to compare the relaxation pattern observed here with the predictions under two alternative models of sequence evolution, representing two extreme scenarios.
+On one side, under a site-independent fitness landscape, such as those contemplated by many current mutation-selection models (Halpern and Bruno, Rodrigue), thus with no epistasis, every site has to adapt on its own to the new change in $\Ne$.
+As a result, the relaxation time is vey long, of the order of the inverse of the per-site substitution rate.
+On the other side, assuming a fixed distribution of fitness effect (DFE) [does not suffer from underestimating the relaxation rate - de nouveau, erreur de catégorie ici, on n'estime rien], the response of $\dnds$ is instantaneous (see Figure \ref{fig:relaxStability}).
 	\begin{figure*}[htb!]
@@
 
+	[L'ensemble du paragraphe ci-dessous me semble plus être de la discussion - par ailleurs, il faudrait le retravailler un peu]
 	Previous studies have argued that model of protein evolution should consider epistasis for various empirical reasons such as rate heterogeneity along the sequence, rate and time dependence of convergence, role of compensatory substitutions, prediction of deleterious mutations \cite{Goldstein2017}.
-We argue that epistasis also have an important role the elasticity of $\dnds$ to changes in $\Ne$.
-If epistasis is not taken into account, the elasticity of $\dnds$ is spread along time incompatible with biological evolution. \\
-% Epistasis influence relaxation time
+	We argue that epistasis also has an important role in the non-stationary response of $\dnds$ to changes in $\Ne$ [attention: on a précédemment défini l'elasticité comme la réponse du omega d'équilibre à des changements de log Ne, on ne peut donc pas parler d'élasticité hors équilibre comme cela.
+Je n'utiliserais pas ce terme dans ce contexte].
+	If epistasis is not taken into account, the relaxation time of $\dnds$ is incompatible with biological evolution [phrase un peu abrupte, telle quelle: incompatble avec quoi exactement?].
 

### 2020-05-09 manuscript/figures.tex
Added Figures and updated manuscript

@@
 	height=\axisdefaultheight,
-		xlabel={$\x$},
+	xlabel={$x$},
 	ymin=0.0, ymax=5.0,
 	samples=200,
-		legend entries={$\ln(\frac{1 - \x}{\x}) + \ln(19)$,
+	legend entries={$\ln(\frac{1 - x}{x}) + \ln(19)$,
 		${\alpha = -118, \gamma = 1,\Ne = 10^{4}}$,
@@
 height=\axisdefaultheight,
-	ylabel={$\dnds\eq$},
-	xlabel={$\x\eq$},
+ylabel={$\omega\eq$},
+xlabel={$x\eq$},
 domain=0:1.0,
@@
 height=\axisdefaultheight,
-	ylabel={$\frac{\der \dnds\eq}{\der \x\eq}$},
-	xlabel={$\x\eq$},
+ylabel={$\frac{\der \omega\eq}{\der x\eq}$},
+xlabel={$x\eq$},
 domain=0.0:1.0,
@@
 height=\axisdefaultheight,
-	xlabel={$\x$},
+xlabel={$x$},
 ymin=0.0, ymax=5.0,
@@
 legend entries={
-		$4\Ne \beta \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}}$,
-		$\lambda_{2}(\x)$,
-		$\lambda_{5}(\x)$,
-		$\lambda_{10}(\x)$,
-		$\lambda_{20}(\x)$},
+$4\Ne \beta \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma x)}}$,
+$\lambda_{2}(x)$,
+$\lambda_{5}(x)$,
+$\lambda_{10}(x)$,
+$\lambda_{20}(x)$},
 legend cell align=left,
@@
 height=\axisdefaultheight,
-	ylabel={Fitness $f(\x)$},
-	xlabel={Phenotype $\x$},
+ylabel={Fitness $f(x)$},
+xlabel={Phenotype $x$},
 cycle list name=colors,
@@
 axis y line=left,
-	legend entries={$f(\x) = \frac{1}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}}$},
+legend entries={$f(x) = \frac{1}{1 + \e^{\beta(\alpha + \Nsite \gamma x)}}$},
 legend style={at={(0.4,0.5)},anchor=south west}

### 2020-05-09 manuscript/manuscript.tex
Added Figures and updated manuscript

--- 2020-04-14 manuscript/manuscript.tex
@@
 		One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
-In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
+		In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$) with respect to $\Ne$.
 		Such theoretical prediction had been observed in empirical data across many clades.
-However, several studies failed to observe such response of $\dnds$ to changes in $\Ne$, or with weak strength or direction.
-Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
-Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
-Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
-Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
-We apply our framework in the special case of fitness proportional to the probability of protein folding.
+		However, several studies failed to observe such response of $\omega$ to changes in $\Ne$, or with weak strength or direction.
+		Computational models of protein folding have observed that $\omega$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
+		Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\omega$ and then a decrease.
+		Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\omega$ responses to changes in $\Ne$.
+		Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\omega$ in the context of a genotype-phenotype-fitness map.
+		We derive general formulas, and specify an application in the case of fitness proportional to the probability of protein folding.
 		Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
-We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
+		We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\omega$ relationship in population genetic dataset.
 		However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
-We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
+		We also stress the importance of epistatic interactions in the $\Ne$-$\omega$ relationship, such that it determines to time to reach a new equilibrium.
 	}
+
 	\section*{Introduction}
-Differences in DNA sequences of a gene in separate species are due to the particular history of DNA substitutions along each specie's respective lineage.
-These substitutions at the level of the population are the result of point mutations at the level of individuals, subject to evolutionary forces such as selection and random drift.
-In protein-coding DNA sequences, non-synonymous mutations are changing the amino-acid sequence, and by comparing the non-synonymous substitutions rate relative to the synonymous substitution rate (supposedly neutral), a statistic called $\dnds$, we can estimate the global strength of selection and random drift.\\
-For a given gene, it has been empirically observed changes of $\dnds$ along lineages of a phylogeny \cite{Yang2001, Zhang2004}.
-More importantly, changes in substitution rate correlate with changes in life-history-traits (longevity, body mass, ...) such as shown by the molecular comparative framework \cite{Lartillot2011,Weber2014}.
-Such covariations are compatible with the theoretical prediction of the nearly neutral theory of evolution \cite{Ohta1972, Ohta1992}, where the underlying mechanism is changes in effective population size ($\Ne$).
-Indeed, populations with low $\Ne$ would have both a large body size and a high $\dnds$ (weaker selection) due to the increase of random drift.
-If $\dnds$ is empirically correlated to life-history traits, such as body mass and longevity, the direction of correlation could not be replicated across all experiments \cite{Figuet2016}.\\
-Theoretical prediction of $\dnds$ decreasing as a function of $\Ne$ had been developed in the context of a fixed distribution of fitness effects of novel mutations \cite{Ohta1972, Welch2008}.
-Similarly, it has been shown in context of a fixed fitness landscape, where each amino-acid have different fitness \cite{Spielman2015a, DosReis2015}, although sites are considered independently evolving, without epistasis.
-Unexpectedly, it has also been argued that the existence of a $\dnds$ response to $\Ne$ is not straightforward, even in the context of nearly neutral theory of evolution \cite{Lanfear2014}.
-In the example of a biochemical mechanistic computational model, where the fitness is determined by the probability of folding, $\dnds$ as been observed to be independent of $\Ne$ \cite{Goldstein2013}.
-Theoretically, a more general case is obtained whenever the fitness is a concave function of a trait, and such trait is equimutable \cite{Cherry1998}.
-Equimutabibilty meaning a mutation has the same effect on the trait whatever the current trait of the individual.
-In addition to this complexity, $\dnds$ response to $\Ne$ can be non-monotonic due to non-equilibrium properties, such that an increase in $\Ne$ implies firstly a burst of $\dnds$ (adaptive selection), and subsequently a decrease in $\dnds$ (purifying selection) \cite{Jones2016}.\\
-Our goal in this present work to reconcile empirical observation with theoretical results, such as to delimit which models and assumptions are reasonable.
-This question is important if we want to extract changes in effective population size from substitution history, as well as to determine the fitness landscape under which protein-coding DNA sequences are evolving.
-We provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
-We apply our framework in the special case of fitness proportional to the probability of protein folding.
-Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
-We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship in population genetic dataset.
-However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
-We also stress the importance of epistatic interactions in the $\Ne$-$\dnds$ relationship, such that it determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
+
+	Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
+	These substitutions in molecular sequences are the result of the interplay between evolutionary forces such as mutation, selection and random genetic drift.
+	These forces have effects at different levels: mutations are carried by molecular sequence, selection is mediated at the level of individuals, while random genetic drift is a population effect.
+	Yet, they jointly contribute to the long-term molecular evolutionary process.
+	Thus, the challenge of molecular evolution is to tease out their respective contributions, based on comparative analyses.
+
+	One main aspect of this challenge is to correctly evaluate the role of random drift in the long term evolutionary process.
+	Population genetics theory implies that the strength of drift, due to the stochastic sampling of mutations, is less pronounced in lineage with large effective population size ($\Ne$), and as a consequence, the purification by selection of weakly deleterious mutations is more effective.
+	This fundamental idea is at the core of the nearly-neutral theory of evolution.
+	The nearly-neutral theory posits that a substantial fraction of mutations are weakly deleterious.
+	As a result, the theory predicts that the substitution rate of selected mutations relative to the neutral substitution rate, called $\omega$, decreases along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.
+
+	This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects (DFE) \cite{Kimura1979, Welch2008}.
+	Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $\omega$ as a function of $\Ne$ (i.e.
+$\omega \sim \Ne^{-\beta}$), where $\beta$ is the shape parameter of the DFE.
+	In practice, DFEs are strongly leptokurtic, thus predicts a weak negative relation between $\omega$ and $\Ne$.
+
+	The substitution rate of selected mutations relative to neutral substitution rate ($\omega$) is an observable quantity of the molecular evolutionary process, at least if we can separately estimate the rate of neutral and selected mutations in a phylogenetic or comparative context.
+	Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while those selected mutations are the non-synonymous changes affecting the amino-acid sequence, $\omega$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
+	Thus, in practice, the nearly-neutral argument translates into a predicted decrease in $\omega = dN/dS$ as a function of long-term changes in $\Ne$.
+
+	The context of protein coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
+	In this alternative approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998, Rodrigue2010, Tamuri2012}.
+	For example, if the target amino-acid has a higher fitness than the current amino-acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
+	Such a modeling approach predicts higher substitution rates between amino-acid with similar biochemical properties.
+	More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
+	Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\omega$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.
+
+	Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \cite{Yang2001, Zhang2004}. 
+	Combining estimations of $\omega$ along branches to life-history traits such as body mass or generation time revealed a positive correlation \cite{Popadin2007, Nikolaev2007}
+	Subsequently, integrative inference methods combining molecular sequences and lineage specific quantitative traits have also found that $\omega$ correlates positively with traits such as longevity and body mass \cite{Lartillot2011, Figuet2017}.
+	Since longevity and body mass typically correlate negatively with $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}, these empirical correlations suggest a negative correlation between $\omega$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
+	However, this correlation between $\omega$ and life-history traits could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation was found to be either not statistically significant (\cite{Lartillot2012}), or even in the opposite direction \cite{Lanfear2010, Nabholz2013, Weber2014, Figuet2016}.
+
+	Mitigated empirical evidence about the response of $\omega$ to changes in $\Ne$ encouraged the search for alternative explanatory mechanisms \cite{Lanfear2014}.
+	In this direction, one striking result was the proof that $\omega$ is in fact predicted to be independent of $\Ne$ under very general circumstances, namely, whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
+	Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
+	This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \cite{Goldstein2013}.
+	The thermodynamic stability is itself computed using a 3D structural model of the protein. These simulation experiments have led to the observation that $\omega$ is essentially independent of $\Ne$.
+	The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.
+
+	However, the equimutability assumption is a relatively strong one, which also conflicts with simple combinatorial considerations about the relation between sequence and phenotype.
+For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur.
+More generally, assuming that the stability of a protein sequence reflects an underlying fraction of positions having already accepted destabilizing amino-acids, then the probability of destabilizing mutational events is in turn expected to directly depend on the current stability of the protein.
+
+	In addition, if empirical evidence for a negative correlation of $\omega$ with $\Ne$ is still not totally convincing, another empirical correlation is known to be much more robust, namely, that of $\omega$ with expression level or protein abundance \cite{Duret2000, Rocha2004, Wang2011, Song2017}.
+Indeed, expression level is one of the best predictors of $\omega$, with highly expressed proteins typically having lower $\omega$ values.
+Of note, the slope of the response of $\omega$ to changes in expression level is relatively weak \cite{Drummond2005}, although clearly significant.
+Importantly, many of the theoretical models of protein-coding sequence evolution that have been invoked to explain this correlation between $\omega$ and expression level are such that the response of $\omega$ to changes in expression level or in $\Ne$ should be very similar.
+Conversely, under strict equimutability, most of these models would predict that $\omega$ should be essentially independent, not just of $\Ne$, but also of expression level, thus at odds with empirical observations.
+
+	Overall, there is no doubt that the impact of changes in $\Ne$ (or in expression level) on $\omega$ is at best weak, making empirical estimation more difficult.
+	Even if weak, however, an exact theoretical understanding and quantification of how the slope of this relation depends on the underlying map between genotype, phenotype and fitness, would be useful.
+	Ultimately, relating proteins structural parameters to the $\omega$-$\Ne$ relationship reveal an intersection between disparate empirical dataset,  coming from protein thermodynamic on one side and comparative genomic on the other side.
+
+	Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium.
+However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\omega$ due to adaptive selection, and subsequently a decrease in $\omega$ due to stronger purifying selection in the long term \cite{Jones2016}.
+	Studying only equilibrium properties can thus be misleading.
+For that reason, the dynamical response of $\omega$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized.
+	Dynamical properties of the $\omega$ response to changes of $\Ne$ are of theoretical interest, but are also empirically relevant such that if overlook they could thwart the relation between theoretical expectation and empirical estimates.
+
+	% Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\omega$ relationship to $\Ne$ is unknown.
+	The goal of this study is to characterize the dynamical and equilibrium response of $\omega$ to changes in $\Ne$, and relate this response to structural parameters of the model.
+	To this end, we develop a general mathematical approach for deriving the $\omega$ relationship to $\Ne$ (and expression level) in the context of a given genotype-phenotype-fitness map.
+	Based on these results, we discuss the consequences of the assumption that protein stability is a proxy of fitness, and the signatures persisting in DNA sequences observable under such assumption.
+	In the light of previously published empirical estimates from protein thermodynamic, and from comparative genomics, we discuss the articulation between empirical data and our mechanistic model.
+	We also discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the slope of the relation between $\omega$ and $\Ne$.
+
 	\section*{Results}
-\subsection*{Theoretical approximations}
-In this section we theoretically approximate the change in $\dnds$ at equilibrium after a change in $\Ne$, under a model in which the fitness depends on the genotype of the whole sequence.
-Firstly, we decompose the map from genotype to fitness through an intermediary bounded phenotype $0 \leq \x \leq 1$, the proportion of unstable sites in the sequence.
-Each site of the sequence ($\Nsite$ sites) can be in one of $\Nstate \geq 2$ states, where only $1$ of those states is assumed to be a stable state, and $\Nstate - 1$ states are unstable.
-The optimal phenotype is obtained when all of the $\Nsite$ sites of the sequence are stable ($\x = 0$).\\
-To complete the model, the phenotype-fitness map is first considered generally, where the Wrightian fitness of phenotype $\x$ is $f(\x)$.
-After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
-Most importantly the probabilities to increase or decrease $\x$ are different, and depends on the current phenotype $\x$.
-Mutation increasing $\x$ are selected against, while mutations decreasing $\x$ are positively selected. 
-Mutation and selection cancel each other at equilibrium, where the expectation of fitness effects of mutations reaching fixation in the population is $0$.
-Altogether, the equilibrium phenotype obtained for such mutation-selection balance is denoted $\x\eq$, and more importantly is given by the equation (See Supplementary Materials):
+
+	\subsection*{Models of evolution}
+
+	The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$ (i.e.
+such that the coding positions of the sequence contribute additively to the trait), under directional selection specified by a decreasing and log-concave fitness function $f(x)$.
+As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability (See figure \ref{fig:NeChangeInfluence}).
+This model is inspired from previous work \cite{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations.
+	%Throughout our derivation, we present the results both for the general case and for the more specific example.
+
+	In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
+	Technically, the free energy of the folded are unfolded conformations are computed based on the $3$D conformations of the protein, and using a statistical potential.
+	In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.
+
+	We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighbouring residues.
+	Instead, each site contributes independently to $\deltaG$. 
+	With such approximation, for each site of the sequence only one amino-acid is stabilizing the protein. 
+	All $19$ other amino-acids are equally destabilizing, and each site contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ per destabilizing amino-acids.
+	If all amino-acids of the sequence are stabilizing, $\deltaG$ is optimal and equal to $ \alpha < 0$. 
+	In this model, the most succinct phenotype of a given genotype (i.e.
+sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq x \leq 1$.
+Thus, $\deltaG$ is a linear function of $x$:
 	\begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
+		\deltaG (x) = \alpha + \Nsite \gamma x,
 	\end{align}
-This equation of equilibrium phenotype can be further studied to derive the response of $\x\eq$ after a change in $\Ne$, using differential calculus.
-Furthemore, assuming that the number of states is large (typically $20$ for amino-acid), we obtain the change in $\omega\eq$ after a change in $\Ne$ as: 
+	where $\Nsite$ is the number of sites in the sequence.
+
+	For a given $\deltaG$, thermodynamic equations allows to derive the proportion of protein in folded conformation in the cytoplasm, which is assumed to be a proxy for fitness.
+	This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm (supp information).
+	Analytically, this fitness by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
 	\begin{align}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
+		f(x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma x)}},\\
+		\Rightarrow f(x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma x)}, 
 	\end{align}
-In other words, the slope of the $\dnds$ elasticity to changes in $\Ne$ (in log scale) is the ratio of first to the second derivative of the log-fitness function, taken at the equilibrium phenotype.
-To note, this elasticity is strictly negative for decreasing log-concave fitness functions.\\
-So far, our result did not assume any particular parametrization of the phenotype-fitness map.
-As an example, already proposed in the literature \cite{Goldstein2011}, fitness is proportional to the probability of our protein to be in the folded state, given by Fermi Dirac distribution: 
-\begin{equation}
-f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}, 
-\end{equation}
-where $\deltaGMin < 0$ and $\gamma > 0$ are in kcal/mol, and the fixed parameter $\beta$ is $1.686$ mol/kcal at room temperature.
-In other words, $\deltaGMin$ is the difference in free energy between folded and unfolded state when all sites are stable.
-$\gamma$ is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
-With such fitness function, the equilibrium phenotype $\x\eq$ is given by the equation: 
+	where $\alpha$ and $\gamma$ are defined as above, and the parameter $\beta$ is $1.686$ mol/kcal at room temperature.
+
+	\subsection*{Susceptibility of $\omega$ to changes in $\Ne$. Analytical approximation}
+
+	In this section we present an analytical approximate solution for the response of equilibrium $\omega$ after a change in $\Ne$ (in log space).
+We call this reponse the susceptibility of $\omega$ to changes in $\Ne$, and denote it as $\chi$:
 	\begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
-\label{eq:equilibrium}
+		\chi = \frac{ \der \omega}{\der \ln (\Ne)}
 	\end{align}
-The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
-\begin{equation}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq -\dfrac{1}{\beta \Nsite \gamma};
-\end{equation}
-which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
-Moreover, only the free compound parameter $\gamma \Nsite$ has an impact on the slope, an intuition for such result is given in figure \ref{fig:NeChangeInfluence}.
-Meaning that under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the slope of the linear relationship between $\dnds$ and log-$\Ne$ is approximately $0.002$.
+	Broadly, deriving $\chi$ is done in two steps.
+	The first step is to determine the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
+	Subsequently, differential calculus on the response of the equilibrium phenotype to a change in $\Ne$ allows to ultimately derive an equation for $\chi$.
+	The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case where the fitness is proportional to the total amount of folded protein in the specific case of the biophysical model.
+	All development are available in Supplementary Materials in the most general case.
+
+	First, from a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of destabilizing amino-acid, or do nothing if the mutation is between both destabilizing amino-acids.
+	To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable.
+	Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
+	\begin{gather}
+		\begin{cases}
+		\dx &\text{ with probability } 1-x, \\
+		0 &\text{ with probability } \frac{18 x }{19}, \\
+		-\dx &\text{ with probability } \frac{x}{19}.\\
+		\end{cases} \label{eq:proba}
+	\end{gather}
+	In the extreme case of an optimal phenotype ($x = 0$), only destabilization mutations are proposed.
+	Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between destabilizing amino-acids (effect $0$) is proportional to $x$. 
+	Or in other words, the mutation bias is proportional to $(1-x)/x$, which this is fundamentally a combinatorial effect, due to the number of mutational opportunities available in either direction.
+
+	Second, we need to determine the strength of selection acting on mutations.
+	Destabilizing mutations are selected against with a negative selection coefficient which can be approximated by:
+	\begin{align}
+		s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(x) }{\partial x} \label{eq:s} \\
+		\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma x)}. \label{eq:s-unfolded}
+	\end{align}
+	Conversely, stabilizing mutations will be under positive selection with opposite sign but same absolute value.
+	It is important to realize that the selective effect is dependent on $x$, and that because fitnesses are log-concave, the absolute value of $s$ increases with regards to $x$.
+
+	Based on these expressions for mutational and selective pressure, one can study the evolutionary process unfurls.
+	Starting from an optimal sequence, mostly destabilizing mutations will occur, which will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, at which point the protein will reach a point of equilibrium called marginal stability.
+	Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
+	Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average \cite{Goldstein2013}.
+	Formally, and after simplification, the equilibrium phenotype denoted $x\eq$ is given by the equation:
+	\begin{align}
+		\ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(x\eq) }{\partial {x\eq}} \text{ in the general case,} \\
+		\Rightarrow \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)} \text{,} \label{eq:equilibrium}
+	\end{align}
+	in the more specific case of the biophysical model.
+This equation essentially expresses the mutation-selection equilibrium: the left-hand side of the equation is the log of the mutation bias at $x$, while the right-hand side is simply $4 \Ne s$.
+
+	This equation cannot be solved explicitly for $x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $x\eq$ is given in figure \ref{fig:NeChangeInfluence}B.
+	Mutation bias decreases weakly with $x$ (blue curve on figure) while the strength of selection increases sharply with $x$.
+	The equilibrium phenotype is obtain at their intersection (eq \ref{eq:equilibrium}).
+	An increase in $\Ne$ leads to shifting the selective response upward, which then results in a leftward shift of the equilbrium phenotype. 
+	The amount of change on $x\eq$, however, is very small if the selection coefficient curve is very steep around the equilibrium set point.
+	Of note, the phenotype is not equimutable, meaning the mutational bias is not constant, but nearly so since the derivative of the mutational bias is greatly lower than the derivative of the selection coefficient around the equilibrium point.
+
 	\begin{figure*}[htb!]
@@
 Illustration of the Genotype-phenotype-fitness map.
-The phenotype ($\x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of unstable sites in the sequence.
+The phenotype ($x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of destabilizing sites in the sequence.
 The fitness is a decreasing log-concave function of the phenotype.
-   Panel B. Illustration of the $\dnds$ elasticity after a change in $\Ne$.
-   For the special case $K=2$, the equilibrium $\x\eq$ is determined by the equation $\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$.
-The right-hand side of the equation increases exponentially with $\x$ where $\Nsite \gamma$ is the exponential growth rate.
-Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$ (x-axis of the crossing with the blue solid line).
-On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably.
-Moreover, changes in $\x\eq$ reflects the changes in $\dnds\eq$ since both are related by a monotonous function (See Supplementary Materials).
+				Panel B.
+Illustration of the $\omega$ susceptibility after a change in $\Ne$.
+The equilibrium $x\eq$ is determined by the equation $\ln(\frac{1 - x\eq}{x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)}$.
+The right-hand side of the equation increases exponentially with $x$ where $\Nsite \gamma$ is the exponential growth rate.
+Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only changes subtly $x\eq$ (x-axis of the crossing with the blue solid line).
+On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) involve a stronger response of $x\eq$.
+Moreover, changes in $x\eq$ reflects the changes in $\omega$ since both are related by a monotonous function (equation \ref{eq:dnds}).
 The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
@@
 
-\subsection*{Simulations confirmation}
-Our compact theoretical results are based on assumptions that are broken in practice, for example assuming that 
-\begin{enumerate}
- \item $\Nsite$ is large;
- \item selection coefficient is well approximated by the fitness derivative;
- \item genetic code is not taken into account;
- \item all unstable states are equivalently unstable.
-\end{enumerate}
-We thus first tested the soundness of our theoretical results when such assumptions are relaxed, by simulating DNA sequences evolution (See Methods). 
-Firstly, simulations demonstrate that the relation between $\dnds$ and log-$\Ne$ is linear, and that the slope of the linear regression match the expected theoretical value (See Figure \ref{fig:GoldsteinVsToy}).
-Secondly the parameter $\deltaGMin$ hardly changes the slope of the linear regression, as expected also theoretically.
-Decreasing $\deltaGMin$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
-Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
-However the slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.\\
+	The argument thus far explains how $x\eq$ varies with $\Ne$.
+	To capture how $\omega$ varies with $\Ne$, we also need to obtain an expression for $\omega$ as a function of $x\eq$. 
+	At equilibrium we can derive the expected substitution rate of mutations changing the amino-acid sequence, and by normalizying by the substitution rate of neutral mutations, $\omega$ simply approximates to:
+	\begin{gather}
+		\omega \simeq x\eq \label{eq:dnds}
+	\end{gather}
+	This simple approximation is due to substitutions at destabilizing sites (proportion $x\eq$) to one of the other destabilizing amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (see equation \ref{eq:proba}).
 
-Finally, we compare our results to more complex models of sequence evolution leveraging the $3$D structure of protein \cite{Williams2006, Goldstein2011, Pollock2012}.
-In such models, the free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985}.
-The free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states (see Supplementary Materials).
-The original works claimed that under such $3$D model $\dnds$ is independent of $\Ne$.
-Using extensive simulations, we argue instead that $\dnds$ is approximately linear with log-$\Ne$ (see Figure \ref{fig:GoldsteinVsToy}, panel C).
-Moreover, the observed slope match the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
-In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the most stable sequence of $300$ sites \cite{Goldstein2011}.
+	Finally, $x\eq$ at equilibrium (eq \ref{eq:equilibrium}) is implicitly a function of $\Ne$ and the other parameters.
+	Implicit derivation of the equilibrium with regards to $\Ne$ (see supp info) allows the derive the response of $x\eq$ and ultimately $\omega$ after a change in $\Ne$.
+	After simplifications, the susceptibility of $\omega$ to changes in $\Ne$ can be approximated in terms of a very compact equation for the general case: 
+	\begin{gather}
+		\chi = \frac{ \der \omega}{\der \ln (\Ne)} \simeq - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}
+	\end{gather}
+	The susceptibility is thus equal to the inverse of the relative curvature, i.e.
+the ratio of the second to the first derivatives, of the log-fitness function, taken at the equilibrium phenotype.
+Of note, this susceptibility is strictly negative for decreasing log-concave fitness functions, asserting that $\omega$ is a decreasing function of $\Ne$.
+In addition, the susceptibility itself is low (i.e.
+$\omega$ responds more weakly) for strongly concave log-fitness functions, thus formally capturing the intuitive argument developed above (figure \ref{fig:NeChangeInfluence}B).
+
+	In the specific case of the biophysical model, the susceptibility ($\chi$) simplify to:
+	\begin{gather}
+		\chi \simeq -\dfrac{1}{\beta \Nsite \gamma}, 
+	\end{gather}
+	Meaning that $\omega$ is linearly decreasing with $\Ne$ in log scale, since $\chi$ is independent of $x\eq$ and thus the exact value of the equilibrium phenotype has no impact on the slope.
+	Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship.
+	In practice, $\gamma = \Delta \Delta G$ determine the slope of the linear relationship between $\omega$ and $\Ne$, while $\alpha = \Delta G_{\text{min}}$ can never obtained solely from this slope.
+	Importantly, only empirical relative values of $\Ne$ (multiplicative) and $\omega$ (additive) are sufficient to estimate $\chi$, absolute values are not necessary to estimate the slope.
+	Conversely, if one seek to determines $\Delta G_{\text{min}}$, one need the intercept of the linear relationship, and thus absolute value of $\omega$ and $\Ne$ are necessary.
+	% Although the general equation seems to suggest in appearance that elasticity only depends on the fitness function, in fact, what the biophysical model shows is that the elasticity can actually be modulated in two ways: by playing on the genotype / phenotype relationship (here, $\gamma$), or by playing on the fitness phenotype relationship (here, beta).
+All the more reason to consider beta as a potentially free parameter too.
+
+	Other parametrization of the fitness function shed light on the robustness of the theoretical results.
+	If the selective coefficient itself is not equal (equation \ref{eq:s-unfolded}) but proportional to the total amount of misfolded protein, and the expression level of the protein is denoted $A$, then the response of $\omega$ to changes in $\Ne$ is the same as the response to changes in $A$ (supplementary materials):
+	\begin{align}
+	s & \propto - \beta \gamma y e^{\beta(\alpha + \gamma \Nsite x)} \\
+	\Rightarrow \chi \frac{ \der \omega}{\der \ln (\Ne)} = \frac{ \der \omega}{\der \ln (A)} & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
+	\end{align}
+	Meaning we should observe the same relationship between $\omega$ and either effective population size ($\Ne$) or expression level ($A$).
+
+	Under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the susceptibility $\chi \simeq -0.002$.
+	In other words, for a increase in $\Ne$ of $6$ orders of magnitude, $\omega$ decrease approximately of $0.01$, a subtle relationship that requires laborious effort to be detected in simulated data, and large amount of empirical data with few bias and noise to be detectable.
+In addition, it would imply an equally weak relation between $\omega$ and expression level.
+
+	\subsection*{Simulation experiments}
+
+	Our theoretical derivation of the susceptibility of $\omega$ to changes in $\Ne$ is based on several simplifying assumptions about the evolutionary model and makes multiple approximations.
+In order to test the robustness of our main result, we therefore conducted systematic simulation experiments, relaxing several of these assumptions.
+In each case, simulations were conducted under a broad range of values of $\Ne$, monitoring the average $\omega$ observed at equilibrium and plotting the scaling of these measured equilibrium $\omega$ as a function of $\Ne$.
+
+	Specifically, with respect to mutations, our derivation assumes that all amino-acids transition are equiprobable, in other word the complexity of the genetic code is not taken into account.
+	Simulating evolution of DNA sequence, and taking into account a matrix a mutation rate between nucleotide allow to test directly the robustness of the results to this assumption.
+	Furthermore, with regards to the phenotypic effects of amino-acid changes, we assumed that all destabilizing amino-acids are have an identical impact on protein stability.
+	In reality, one would expect conservative amino-acid replacements to be less destabilizing than radical changes.
+	This assumption is relaxed in our simulation, indeed destabilizing effects in each position, instead of being identical for all alternative amino-acids, are now proportional to the Grantham distance \cite{Grantham1974} between the optimal amino-acid in this position and the amino-acid proposed by the non-synonymous mutation.
+	Finally, the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
+	The robustness of such approximation is test in the simulations with sequences of a finite number of sites ($\Nsite=300$).
+
+	Simulations demonstrate that the relation between $\omega$ and log-$\Ne$ is indeed linear, at least in the range explored here, and that the slope of the linear regression matches the expected theoretical value (figure \ref{fig:GoldsteinVsToy}A).
+	Secondly we observe that the parameter $\alpha$ has virtually no effect on the slope of the linear regression, as expected also theoretically (figure \ref{fig:GoldsteinVsToy}B).
+Instead, decreasing $\alpha$ (to more negative values) merely results in an overall increase in $\omega$ over the whole range of $\Ne$ (i.e.
+has an impact on the intercept, not on the slope).
+This is due to the fact that decreasing $\alpha$ shifts the equilibrium to higher $x\eq$, since more destabilizing sites can then reach fixation before reaching the point of marginal stability.
+
+	Finally, we relaxed our assumption that each site of the sequence contribute independently to $\deltaG$, by taking into account the $3$D structure of protein and using a statistical potential to estimate $\deltaG$.
+	We implemented the original model \cite{Williams2006, Goldstein2011, Pollock2012}, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985} (see Supplementary Materials).
+	The original works showed that under such $3$D model $\omega$ is approximately independent of $\Ne$ \cite{Goldstein2013}.
+	Using extensive simulations in order to obtain sufficient resolution, we observe that $\omega$ is in fact weakly dependent on $\Ne$, being approximately linear with log-$\Ne$ (figure \ref{fig:GoldsteinVsToy}C).
+	Moreover, the observed slope matches the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
+	In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the optimal sequence of $300$ sites \cite{Goldstein2011} (figure \ref{fig:GoldsteinVsToy}D).
 	\begin{figure*}[htb!]
@@
 			\caption{
-  \textbf{$\bm{\dnds}$ elasticity to change in $\bm{\Ne}$}.
+				\textbf{$\bm{\omega}$ susceptibility to change in $\bm{\Ne}$}.
 				For each population size, $100$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
@@
 $\alpha$ are given in the legend.
-Decreasing $\alpha$ (to more negative values) increases $\dnds$ but changes hardly the slope of the linear regression, as expected theoretically.
-  $\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+Decreasing $\alpha$ (to more negative values) increases $\omega$ but changes hardly the slope of the linear regression, as expected theoretically.
+				$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
 				Panel B. The fixed parameters are $\Nsite=300$ and $\beta=1.686$. Parameters $\alpha$ and $\gamma$ are given in the legend.
-  $\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-  The slope of $\dnds$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
-  Panel C. In the model of 3D free energy of folding, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally \cite{Goldstein2013}.
+				$\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+				The slope of $\omega$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
+				Panel C. In the model of 3D free energy of folding, $\omega$ at equilibrium is weakly dependent on log-$\Ne$, but not completely as originally claimed \cite{Goldstein2013}.
 				This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.
 				Bottom D, the fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-  Moreover, with the Grantham model, the $\dnds$ matches the empirical 3D model of Golstein \& Pollock and the theoretical prediction.
+				Moreover, with the Grantham model, the $\omega$ matches the empirical 3D model of Golstein \& Pollock and the theoretical prediction.
 				\label{fig:GoldsteinVsToy}
@@
 	\end{figure*}
-\subsection*{Epistasis determines the time to relaxation}
-Although the equilibrium value of $\dnds$ after changes in $\Ne$ is an important feature of the $\dnds$-$\Ne$ relationship, another parameter that is scarcely studied is the relaxation time to reach the new equilibrium $\dnds$ \cite{Jones2016}.
-We observed in our simulations that the determining factor of the relaxation time is the number of sites $\Nsite$ (See Figure \ref{fig:relaxStability}).
-Such observations match the theoretical prediction that high epistasis leads to faster return to equilibrium, since more opportunities are available.
-In the case of a fitness landscape with epistasis, after a change in $\Ne$, any site that is used to climb either up or down the fitness landscape will have diminishing return in other sites of the sequences.
-Not taking into account epistasis have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$ .
-Using a site-specific fitness landscape, where each amino-acid have different fitness, thus with no epistasis, the relaxation time is long since every site has to adapt to the new change in $\Ne$.
-However, using a fixed distribution of fitness effect (DFE) does not suffer from underestimating the relaxation rate, since the effect is instantaneous (see Figure \ref{fig:relaxStability}).
+	\subsection*{Epistasis and time to relaxation}
+	Although the equilibrium value of $\omega$ after changes in $\Ne$ is an important feature of the $\omega$-$\Ne$ relationship, another characteristic that is scarcely studied is the dynamic aspect, particularly the relaxation time to reach the new equilibrium $\omega$ \cite{Jones2016}.
+	We observed in our simulations that the determining factor of the relaxation time is the number of sites $\Nsite$ (figure \ref{fig:relaxStability}A).
+	These observations match the theoretical prediction that high epistasis leads to faster return to equilibrium, since more mutational opportunities are available for driving the trait closer to equilibrium.
+
+	It may be useful to compare the relaxation pattern observed here with the predictions under two alternative models of sequence evolution, representing two extreme scenarios.
+On one side, under a site-independent fitness landscape, such as those contemplated by many current mutation-selection models \cite{Halpern1998, Rodrigue2010}], thus with no epistasis, every site has to adapt on its own to the new change in $\Ne$.
+As a result, the relaxation time is vey long, of the order of the inverse of the per-site substitution rate.
+	On the other side, assuming a fixed distribution of fitness effect (DFE), the response of $\omega$ is instantaneous (figure \ref{fig:relaxStability}B).
+
+	Disregarding simulations under a fixed DFE, another characteristic observed is the non-continuity of $\omega$ after a change in $\Ne$.
+	Both an increase and decrease in $\Ne$ lead to a discontinuity in $\omega$.
+	Most importantly, $\omega$ is suddenly higher regardless of the change's direction in $\Ne$, then smoothly reaching the equilibrium value (figure \ref{fig:relaxStability}A \& \ref{fig:relaxStability}B).
+	The similarities in these non-equilibrium properties can both be explained mechanistically.
+	Under low $\Ne$ (between $0$ and $100$ My in the simulation), the phenotype is far away from the optimal phenotype and loosely constrained because the strength of selection is weaker.
+	The reaction to an increase in $\Ne$ (at $100$ My in the simulation) is first a higher constrain and a traction toward a more optimal phenotype, which translate into suddenly higher $\omega$.
+	Conversely, under high $\Ne$ (between $100$ and $400$ My in the simulation) the phenotype is closer to optimal and tightly constrained.
+	The reaction to a decrease in $\Ne$ is first a lower constrain and thus a $\omega$ closer to the neutral case, which translate into higher $\omega$.
+	To note, an increase in $\Ne$ can theoretically and possibly lead to a temporarily $\omega > 1$ due to adaptive evolution \cite{Jones2016}, while an decrease in $\Ne$ always imply $\omega < 1$ due to at most a neutral regime of relaxed selection.
+
 	\begin{figure*}[htb!]
@@
 			\caption{
-  \textbf{$\bm{\dnds}$ relaxation time to changes in $\bm{\Ne}$}.
-  $\dnds$ relaxation after a brutal change in $\Ne$.
-  Solid line corresponds to the average over replicates and the shaded area correspond to the $90\%$ interval among replicates. 
+				\textbf{$\bm{\omega}$ relaxation after a change in $\bm{\Ne}$}.
+				Solid line corresponds to the average over $1000$ replicates and the shaded area correspond to the $90\%$ interval among replicates. 
 				The mutation rate ($\mu$) is $1e{-8}$ per year per site, and the total time of the computation is $700$ million years.
 				Panel A. 
-$\beta=1.686$, $\gamma=-10$ for all simulations.
-The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates ($r$) is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-  Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
-  Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-  Increasing $\Nsite$ implies a reduced time to reach the new equilibrium.
+				$\beta=1.686$ for all simulations.
+				The DNA sequence of $500$ sites is divided into exons of equal size.
+				However the number of sites per exon changes between simulations from $\Nsite=5$ to $\Nsite=500$.
+				Moreover, $\gamma$ is changed according to the exon size such that the product is kept constant, thus the susceptibility of the $\omega$ to changes in $\Ne$ is kept constant.
+				Finally, $\alpha$ is changed accordingly such that the equilibrium value $x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+				As such, whatever the exon size, $x\eq$ and $\chi$ are kept constant and thus the observed effect is due to the number of sites in the exon.
+				We observe that increasing the number of sites implies a reduced time to reach the new equilibrium.
 				Panel B.
-In context of a fixed fitness landscape, where each amino-acid has different fitness (site-specific profile), the time taken to reach the new equilibrium value of $\dnds$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
-In the context of a distribution of fitness effects, the relaxation time is non-existent.
+In context of a fixed fitness landscape, where each amino-acid has different fitness (site-specific profile), the time taken to reach the new equilibrium value of $\omega$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
+In the context of a distribution of fitness effects (DFE), the relaxation time is non-existent and the new equilibrium value of $\omega$ is reached instantaneously.
 			}
@@
 
-\subsection*{Distribution of fitness effects}
-DNA mutations changing a genotype can result in a change of phenotype, and ultimately a change in fitness.
-From a specific genotype, all the possible mutations thus result in a distribution a phenotypic effect (DPE) and fitness effects (DFE).
-The DPE and DFE are not known a priori, but are the resulting consequence of the mutation-selection-drift balance (see Figure \ref{fig:Distribution}).
-Empirically, these distributions are of particular importance since they can be obtained experimentally or inferred with other data.
-As an example, DFE can be inferred from polymorphism dataset \cite{Eyre-walker2007, Galtier2016}. 
-Moreover, the distribution of $\deltadeltaG=\gamma \Nsite \dx$ of novel mutations can be obtained experimentally.
-\begin{figure*}[htb!]
- \begin{mdframed}
-  \centering
-  \includegraphics[width=0.9\textwidth] {artworks/DPE-DFE.pdf}
-  \caption{
-   \textbf{Distribution of phenotypic and fitness effects.}
-   Distribution of fitness effects and phenotypic effect for novels non-synonymous mutations observed along a simulation at the mutation-selection balance.
-$\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-   Each side of the distribution is fitted to a gamma distribution, shown in dotted line. 
-   Panel A. Distribution of observed $\deltadeltaG$, which fit adequately the gamma distribution for negative $\deltadeltaG$ (stabilizing mutations). 
-   Panel A.
-Distribution of observed selection coefficient, which fit adequately the gamma distribution for both positive and negative selection coefficient.
-However the shape parameter estimated is not the same for positive and negative selection coefficients.
-  }
-  \label{fig:Distribution}
- \end{mdframed}
-\end{figure*}
 	\section*{Discussion}
-Goldstein \& Pollock argued that model of protein evolution should consider how the acceptance of mutations depends on the sequence context in which they arise (epistasis), as to correctly represent resultant rate heterogeneity along the sequence, to predict the role of compensatory substitutions in protein evolution, predict which of the 10\% of deleterious mutations in humans are harmless in other species, or to accurately represent the rate and time dependence of convergence and homoplasy \cite{Goldstein2017}.
-We argue that not taking into account epistasis also have the consequence of overestimating the relaxation time to return to equilibrium of $\dnds$ after changes in $\Ne$.
-Paradoxically, not taking into account epistasis means overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.\\
-If using a distribution of fitness effect (DFE) is mathematically convenient and does not suffer from underestimating the relaxation rate, it is also overestimating the response of $\dnds$ to changes in $\Ne$ at equilibrium.
-As such, developing inference of fitness landscape will require to assume site-interdependent fitness landscapes.\\
+	% Goal and theoretical results
+	The negative relationship between the relative substitution rate of selected mutations ($\omega$) and the effective population size ($\Ne$) has been predicted theoretically under the quasi-neutral theory of evolution.
+	Empirical data showed mitigated confirmation of this prediction, encouraging alternative explanatory mechanism.
+	One such explanation, developed in the context of genotype-phenotype-fitness map, demonstrated the absence of $\omega$ susceptibility to changes in $\Ne$ if the probabilities of phenotype changes are invariant to the current phenotype.
+	We argue that this assumption can be relaxed, and provide theoretical tools to derive the relationship between $\omega$ and $\Ne$ at equilibrium in this context.
+	We apply our framework in the special case of fitness proportional to the proportion of folded proteins. 
+	Our theoretical results demonstrate that the relationship between $\omega$ and $\Ne$ (in log space) is linear with a negative slope.
+	This slope is a function of structural parameters of the protein, and is inversely proportional to the product of the sequence size and the average change in conformational energy of destabilizing mutations, leading in practice to a very subtle and weak relationship.
+
+	Our compact theoretical results are supported by more complex simulations of protein evolution relaxing several assumptions.
+	Particularly, our theoretical prediction match a numerical model of protein evolution, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations.
+	Previous studies using this model presented an apparent lack of $\omega$ susceptibility to changes in $\Ne$ \cite{Goldstein2013}.
+	We argue the apparent lack of susceptibility is in fact due to the subtle relation, and require extensive computation to be detected.
+
+	\subsection*{Empirical relevance}
+
+	% Our theoretical results are robsut to approximations
+	% Goldstein \& Pollock model don't have a null susceptibility
+	Based on empirical estimate of the structural parameters ($\beta = 1.686,\ \gamma=1, \Nsite=300$), the slope of the linear relationship between $\omega$ and log-$\Ne$ would be $\chi \simeq -0.002$.
+	Empirically, inference of the $\omega$ susceptibility in Primates, using both polymorphism and divergence data estimated $\chi \simeq X$, approximately $X$ times greater than our theoretical prediction \cite{Brevet2019}.
+	More empirical data across different clades would be required to robustly consolidate such empirical observation.
+	However, in the light of empirical data tending to demonstrate negative relationship between $\omega$ and $\Ne$, models based on the probability of folding are not sufficient to explain the observation of such stastically significant relationship.
+	% Our result show the fitness proportional to proportion of folded folding is at odd with empirical data
+
+	% Stress the importance that susceptibility is valid for Ne and protein abondance
+	Furthermore, our theoretical results demonstrate that the relationship between $\Ne$ and $\omega$ is the qualitatively and quantitatively the same as between protein abundance and $\omega$, whenever protein fitness is determined by the deleterious effect of unfolded proteins.
+	As such, dataset relating $\omega$ to either $\Ne$ or protein expression could in principle be used to determine whether the slope is the same.
+	However such endeavor is easily prone to biases and confounding factor, due to study across different clades and across different genes.
+
+	% Protein-protein interactions
+	Notably, our theoretical approximations apply more broadly to protein-protein interactions (using a mean-field argument, supplementary materials), where protein may either be in free form or engaged in a non-specific interactions with other proteins.
+	In non-specific interactions, stabilizing amino-acids at the protein surface are hydrophilic and destabilizing amino-acid are hydrophobic (sticking to  hydrophobic residues in other proteins).
+	Fitting this model with empirical structural estimates \cite{Zhang2008}, we obtain an susceptibility of $\chi \simeq -0.2$ thus a much stronger response than under the model based on conformational stability.
+	This effect is due to less sites in the protein being involved in protein-protein interaction than for conformational stability, in addition to the lower energy engaged in the contacts.
+	Comprehensively, the assumption that protein fitness is in part determined such as to avoid non-specific interaction leads to a stronger response of $\omega$ to changes in $\Ne$, is more compatible with empirical data obtained in comparative genomics.
+	% Such path might worth be explored !
+
+	Finally, an integrative inference framework of comparative genomics parameterized by structural parameter, expression level and effective population size can in principle allow to robustly test our prediction with empirical data.
+	Orthogonally, this inference framework could allow to estimate unknown variable, for example ancestral effective population size, with enough signal and calibration of the other parameters. 
+	Such inference framework should be careful with some caveats that we will subsequently discuss.
+	\subsection*{Inference models}
+	% Categories of models
+	Models of inference are classified broadly into phenomenological and mechanistics.
+	Mechanistic models dissect the causation chains and construct model for first-principles, while phenomonological models aims to determine the statistical significativity of parameters.
+
+	% Phenomenological models
+	An example of such phenomenological model are the models parametrized directly in $\omega = dN/dS$.
+	They will for example tell us which branch of the tree, or which site of the sequence has a decrease or increase in $\omega$ but not the underlying reason for such changes, which can be either mutation, selection, drift or another evolutionary force.
+	For example, in such models, a mutation and its reverse mutation do not have opposite selection coefficient.
+	As such they can not disentangle the potentially confounding effect of mutation, selection and drift.
+	Neither can they predict the nature of the functional relationship between population genetics or structural physico-chemical parameters to $\omega$.
+	As such, of mechanistic models are necessary to construct an integrative comparative framework.
+
+	% Site specific mechanistic models of evolution 
+	On the other hand, mechanistic models of evolution mapping genotype to fitness have so far assumed site-independent fitness map, mainly for computational rationals.
+	Previous studies have argued that model of protein evolution should consider epistasis for various empirical reasons such as rate heterogeneity along the sequence, rate and time dependence of convergence, role of compensatory substitutions \cite{Goldstein2017}.
+	We argue that epistasis also has an important role in the response of $\omega$ to changes in $\Ne$, both in terms of susceptibility and dynamic of the response.
+	Fundamentally, we argue that any site-specific model implies a slow dynamic and a strong susceptibility, and adding epistasis to the model imply a faster dynamic and a weaker susceptibility.
+	Intuitively, this effect originate that each site is to adapt independently to changes in $\Ne$ leading to overall a slow response (substitutions must  affect all sites), and a strong susceptibility since each site will change its position in the fitness landscape.
+	Taking into account epistasis, the burden of adapting to changes in $\Ne$ is shared by more sites, such that all of them don't have to adapt. 
+	As a conclusion, both mechanistic and phenomenological models of inference are at risk to endure some culprits.
+	In retrospect, these pitfalls have been anticipated thanks to the intuitions developed in the construction of this mechanistic genotype-phenotype-fitness model.
+
+	\subsection*{Theoretical modeling}
+	% Analogy to physics
+	This study elaborate the signature on DNA sequences of a long-term evolutionary process as emergent from population-genetics and structural physico-chemical first principles.
+	Hence, physico-chemical vocabulary and theory are recruited to describe the molecular details.
+	More widely, our modeling is imprinted with vocabulary and analogy to physical systems, with wording in the likes of forces, susceptibility and dynamic.
+	We argue analogy between evolutionary and mechanstic system is not restricted to wording, but allows to visualize and intuit of the causality chains and which approximation can be made such as to derive analytical tractable results.
+	Because it is constructive approach, and not solely intuitive, the robustness of theoretical results to approximations can be asserted by computational implementation. 
+	Computational models offers a means to test the validity and robustness, while the mathematical models offers an intuitive mechanistic mental analogy.
+
+	% Mechanical analogy
+	% Elasticity with springs ? Surface tension with molecule and radius of the 
+	% Elasticity ? Susceptibility (dimensionless) ? Compliance (units of metres per newton) ? Flexibility (units of metres per newton) ? 
+	In the mechanical analogy, the phenotype represent a spatial coordinate of the system, which is determined by external evolutionary forces; namely mutation, selection and drift.
+	These evolutionary forces are homologous to springs attached and stretching the phenotype in various directions and with various intensity, and the aim is to find the equilibrium point.
+	Once the equilibrium point is determined, the deformation of the phenotype (called strain) can be studied whenever applying an differential stress to the system.
+	Elasticity refers to a reversible deformation of a system upon a external force, hence its ability to return to its original state when that force is removed.
+	Intuitively, changing the free energy of the reference phenotype ($\Delta G_{\text{min}} = \alpha$) leads in first approximation to changes in the equilibrium point but not the stiffness of the springs, because only the reference is changed, meaning the elastic coefficient is conserved as found in our theoretical results.
+	Conversely, stiffness ot the strings increases with regards to $\Delta \Delta G = \gamma$, meaning a that change in the stress applied to the system ($\Ne$) result in a weak deformation ($\omega$) at the new equilibrium.
+
+	% Thermodynamic analogy 
+	% Analogy to Clapeyron formula ? TdP/dT = ΔH/ΔV, where T is analogous to Ne
+	In the thermodynamic analogy, the phenotype is a macroscopic resultant of microscopic states of each site of the sequence.
+	Such example could be homologous to the Ising model, and each site of the sequence is a magnetic dipole which can be up or down (destabilizing or stabilizing).
+	The phenotype is an aggregate of the microscopic states, simply the proportion of destabilizing states.
+	Under external forces, the macroscopic proportion is statistically determined by the potential energy of the system, and the susceptibility ()analogous to magnetic susceptibility) of the system is the response of macroscopic observable to change in temperature (inverse of $\Ne$).
+
+	% Fundation to integrate other forces
+	Altogether, this framework can be used as a premise % (groundwork, foundation, starting point,...)
+	to integrate other external evolutionnary forces.
+	Additional molecular processes could be included, for instance GC-biased gene conversion inducing a bias in the nucleic composition of genomes. 
+	The change of equilibrium due to the GC-bias, and the magnitude of the response could be investigated, as to describe the signatures observable on molecular sequence, factoring the effect of mutation, selection and drift.
+	Finally, we hope the result of this work, and more broadly the framework can foster the understanding of observable signature of a long-term evolutionary process as emergent from ecological parameters and molecular physico-chemical first principles.
+
 	\section*{Materials \& Methods}
-Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}., where the rate of substitution of a new mutation with selection coefficient $s$ is:
+	Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}, where one sequence represent the whole population.
+	From the resident DNA sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
+	For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codons that are one mutation away and that are not stop codon.
+	For each mutant sequence $\cj \in \setNeighbors$, we compute its fitness and subsequently the selection coefficient of the mutant:
 	\begin{equation}
-p_{\text{fix}}(s) = \mu \dfrac{4 \Ne s}{1 + e^{- 4 \Ne s }}, 
+	s \left( \ci,\cj\right) = \dfrac{ f \left(\cj \right) - f \left(\ci\right) }{f\left( \ci \right)}.
 	\end{equation}
-where $\mu$ is the mutation rate.
-\subsection*{Simulation of protein folding with an additive model of free energy}
-\label{MatMet:folding}
-From a DNA sequence $\ci$ after $t$ substitutions, the protein's difference in free energy between folded and unfolded state is given by:
-\begin{equation*}
-\deltaG\left(\ci\right) = \deltaGMin + \Nsite \gamma * \x\left(\ci\right), 
-\end{equation*}
-where $\x\left(\ci\right)$ is the proportion of non-optimal amino-acid sites in the sequence.
-Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Boltzmann equation: 
-\begin{equation}
-f(\deltaG\left(\ci\right)) = \dfrac{P_{\mathrm{F}}\left(\ci\right)}{P_{\mathrm{F}}\left(\ci\right) + P_{\mathrm{U}}\left(\ci\right)} = \dfrac{e^{-\beta G_{\mathrm{F}}\left(\ci\right) }}{e^{-\beta G_{\mathrm{F}} \left(\ci\right) } + e^{-\beta G_{\mathrm{U}}\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }},
-\end{equation}
-where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
-From the resident sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
-For a protein of $\Nsite$ amino-acid sites, $\left| \setNeighbors \right| \leq 9 \Nsite$, since each codon has a maximum of $9$ possible mutant codons that are one mutation away and that are not stop codon.
-For each mutant sequence $\cj \in \setNeighbors$, we compute $\deltaG^{t+1}$ from the updated sequence $\cj$, and subsequently the selection coefficient of the mutant:
-\begin{equation}
-s \left( \ci,\cj\right) = \dfrac{ f\left( \deltaG \left(\cj\right) \right) - f\left( \deltaG \left(\ci\right) \right)}{f\left( \deltaG \left(\ci\right) \right)}.
-\end{equation}
-The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, according to the rates of substitution between codons:
+	The next event of mutant invading the population, and the time to reach such event is chosen using Gillespie algorithm, according to the rates of substitution between sequences:
 	\begin{equation}
@@
 	where $\mu_{\itoj}$ is the mutation rate between $\ci$ and $\cj$, determined by the underlying $4x4$ nucleotide mutation rate matrix, and ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
-\subsection*{Simulation with independent fitness profiles}
-A fitness profile give a fitness for each amino-acid (vector of size $20$).
+	Various optimization are implemented to reduce the computation time of mutant's fitnesses.
+	The simulation starts with a burn-in period to reach mutation-selection-drift equilibrium.
+	\subsection*{Models of fitness function}
+	\label{MatMet:folding}
+	Under a simulation of protein folding with an additive model of free energy, the protein's difference in free energy between folded and unfolded state is given by:
+	\begin{equation*}
+	\deltaG\left(\ci\right) = \alpha + \Nsite \gamma * x\left(\ci\right), 
+	\end{equation*}
+	where $0 \leq x\left(\ci\right) \leq 1$ is the distance of $\ci$ to the optimal sequence.
+	For each site of sequence, the optimal amino-acids are chosen randomly at initialization, and the distance between the current amino-acid and the optimal is scaled by the Grantham amino-acid distance \cite{Grantham1974}.
+	Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Fermi-Dirac distribution: 
+	\begin{equation}
+	f(\deltaG\left(\ci\right)) = \dfrac{e^{-\beta \deltaG\left(\ci\right) }}{1 + e^{-\beta \deltaG\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }}, 
+	\end{equation}
+	where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
+
+	Under a simulation with site independent fitness profiles, a fitness profile give a fitness for each amino-acid (vector of size $20$).
 	Each site of the protein has a specific amino-acid fitness profile.
@@
 	where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \cite{Bloom2017}.
-The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
-\subsection*{Simulation with distribution of fitness effects (DFE)}
-The selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
+
+	Under simulation with a fixed distribution of fitness effects (DFE), the selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
 	\begin{equation}
@@
 	\end{equation}
-The next change in the protein-coding DNA and the time to next the event is chosen using Gillespie algorithm, as in \ref{MatMet:folding}.
-\subsection*{$\bm{\dnds}$ along the simulation}
+	\subsection*{$\bm{\omega}$ along the simulation}
 	From the set of mutants $\setNeighbors$ that is one nucleotide away from $\ci$, we define the subsets $\setNonSynNeighbors$ and $\setSynNeighbors$ that are respectively the set of non-synonymous and synonymous mutants, where $\setNonSynNeighbors \cup \setSynNeighbors = \setNeighbors$.
@@
 	\begin{align}
-\dnds(t) &= \dfrac{\sum_{\cj \in \setNonSynNeighbors} \submatrix_{\itoj}}{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj}} \left( \dfrac{\sum_{\cj \in \setSynNeighbors} \submatrix_{\itoj}}{\sum_{\cj \in \setSynNeighbors} \mu_{\itoj}} \right)^{-1}\\
+	\omega(t) &= \dfrac{\sum_{\cj \in \setNonSynNeighbors} \submatrix_{\itoj}}{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj}} \left( \dfrac{\sum_{\cj \in \setSynNeighbors} \submatrix_{\itoj}}{\sum_{\cj \in \setSynNeighbors} \mu_{\itoj}} \right)^{-1}\\
 	&= \dfrac{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj} \dfrac{4 \Ne s \left( \ci,\cj\right)}{{1 - \e^{-4 \Ne \left( \ci,\cj\right)} }}}{\sum_{\cj \in \setNonSynNeighbors} \mu_{\itoj}} 
 	\end{align}
+	And $\omega$ is taken as the average of the time-dependent $\omega(t)$ along the simulation.
 	\subsection*{Reproducibility}

### 2020-05-09 manuscript/supp-mat.tex
Added Figures and updated manuscript

@@
 Each site can be in one of $\Nstate \geq 2$ states, where only $1$ state is defined the stable state, and $\Nstate - 1$ states are unstable.
-For a given genotype sequence, define phenotype $0 \leq \x \leq 1$ as the current proportion of sites in the unstable state.
-After a mutation, given that only one site can change at a time, the absolute change of $\x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
-Define $\rho_{\x}(\dx)$ as the probability to get a change of phenotype equal to $\dx$, if the current phenotype is $\x$:\\
+For a given genotype sequence, define phenotype $0 \leq x \leq 1$ as the current proportion of sites in the unstable state.
+After a mutation, given that only one site can change at a time, the absolute change of $x$ is either $0$ or $\dx=\sfrac{1}{\Nsite}$.
+Define $\rho_{x}(\dx)$ as the probability to get a change of phenotype equal to $\dx$, if the current phenotype is $x$:\\
 \begin{gather}
 \begin{cases}
-\dx &\text{ with probability } \rho_{\x}(\dx) = 1-x, \\
-0 &\text{ with probability } \rho_{\x}(0) = \x \left[1 - \frac{1}{\Nstate - 1}\right], \\
--\dx &\text{ with probability } \rho_{\x}(-\dx) = \frac{\x}{\Nstate - 1}.\\
+\dx &\text{ with probability } \rho_{x}(\dx) = 1-x, \\
+0 &\text{ with probability } \rho_{x}(0) = x \left[1 - \frac{1}{\Nstate - 1}\right], \\
+-\dx &\text{ with probability } \rho_{x}(-\dx) = \frac{x}{\Nstate - 1}.\\
 \end{cases} \label{eq:proba}
@@
 \subsection{Selection coefficient}
-$s(\x, \dx)$ is the selection coefficient of an effect $\dx$ if the current phenotype is $\x$:
+$s(x, \dx)$ is the selection coefficient of an effect $\dx$ if the current phenotype is $x$:
 \begin{align}
-s(\x, \dx) &= \frac{f(\x + \dx) - f(\x)}{f( \x)}, \\
- &\simeq \frac{1}{f(\x)}\frac{ \partial f(\x) }{\partial \x} \dx, \\
- &\simeq \frac{ \partial \ln f(\x) }{\partial \x} \dx, \label{eq:s_from_fitness}
+s(x, \dx) &= \frac{f(x + \dx) - f(x)}{f( x)}, \\
+ &\simeq \frac{1}{f(x)}\frac{ \partial f(x) }{\partial x} \dx, \\
+ &\simeq \frac{ \partial \ln f(x) }{\partial x} \dx, \label{eq:s_from_fitness}
 \end{align}
-where $f( \x)$ is the Wrightian fitness of phenotype $x$. And thus we also have:
+where $f( x)$ is the Wrightian fitness of phenotype $x$. And thus we also have:
 \begin{gather}
-s(\x, -\dx) \simeq - s(\x, \dx) \text{ from eq. \ref{eq:s_from_fitness}} \label{eq:s_minus_deltax}, \\
-\iff S(\x, -\dx) \simeq - S(\x, \dx), \label{eq:S_minus_deltax}
+s(x, -\dx) \simeq - s(x, \dx) \text{ from eq. \ref{eq:s_from_fitness}} \label{eq:s_minus_deltax}, \\
+\iff S(x, -\dx) \simeq - S(x, \dx), \label{eq:S_minus_deltax}
 \end{gather}
-where $S(\x\eq, \dx) = 4\Ne s (\x\eq, \dx)$ is the scaled selection coefficient.
+where $S(x\eq, \dx) = 4\Ne s (x\eq, \dx)$ is the scaled selection coefficient.
 \subsection{Probability of fixation}
-The probability of fixation of a mutation with effect $\dx$, for a resident phenotype $\x$ is :
+The probability of fixation of a mutation with effect $\dx$, for a resident phenotype $x$ is :
 \begin{align}
-p_{\text{fix}}(\x, \dx) &= \frac{ 2 s(\x, \dx)}{1 - \e^{-4\Ne s(\x, \dx)}}, \\
- &= \frac{ 2 s(\x, \dx)}{1 - \e^{-S(\x, \dx)}}. \label{eq:pfix}
+p_{\text{fix}}(x, \dx) &= \frac{ 2 s(x, \dx)}{1 - \e^{-4\Ne s(x, \dx)}}, \\
+ &= \frac{ 2 s(x, \dx)}{1 - \e^{-S(x, \dx)}}. \label{eq:pfix}
 \end{align}
@@
 \begin{gather}
-p_{\text{fix}}(\x, 0) = \frac{ 1}{2 \Ne} \label{eq:pfix0}.
+p_{\text{fix}}(x, 0) = \frac{ 1}{2 \Ne} \label{eq:pfix0}.
 \end{gather}
@@
 \begin{align}
-\frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} & = \frac{ 2 \Ne 2 s(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \text{ from eq. \ref{eq:pfix} and \ref{eq:pfix0}},\\
-& = \frac{S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} \label{eq:pfixratio}.
+\frac{p_{\text{fix}}(x, \dx)}{p_{\text{fix}}(x, 0)} & = \frac{ 2 \Ne 2 s(x, \dx)}{1 - \e^{-S(x, \dx)}} \text{ from eq. \ref{eq:pfix} and \ref{eq:pfix0}},\\
+& = \frac{S(x, \dx)}{1 - \e^{-S(x, \dx)}} \label{eq:pfixratio}.
 \end{align}
 \subsection{Equilibrium phenotype}
-At equilibrium phenotype $\x\eq$, the expected selection coefficient of mutation that reached fixation must be $0$:
+At equilibrium phenotype $x\eq$, the expected selection coefficient of mutation that reached fixation must be $0$:
 \begin{gather}
- 0 = \E_{\dx} \left[ s(\x\eq, \dx) p_{\text{fix}}(\x\eq, \dx) \right], \\
-\iff 0 = \frac{ 2 s(\x\eq, \dx)^2}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) + s(\x\eq, 0) \frac{ \rho_{\x\eq}(0)}{2\Ne} + \frac{ 2s(\x\eq, -\dx)^2}{1 - \e^{-S(\x\eq, -\dx)}} \rho_{\x\eq}(-\dx)\text{ from eq.
+ 0 = \E_{\dx} \left[ s(x\eq, \dx) p_{\text{fix}}(x\eq, \dx) \right], \\
+\iff 0 = \frac{ 2 s(x\eq, \dx)^2}{1 - \e^{-S(x\eq, \dx)}}   \rho_{x\eq}(\dx) + s(x\eq, 0) \frac{ \rho_{x\eq}(0)}{2\Ne} + \frac{ 2s(x\eq, -\dx)^2}{1 - \e^{-S(x\eq, -\dx)}} \rho_{x\eq}(-\dx)\text{ from eq.
 \ref{eq:pfix} and \ref{eq:pfix0}}, \\
-\Longrightarrow \frac{ 2s(\x\eq, \dx)^2}{1 - \e^{-S(\x\eq, \dx)}}   \rho_{\x\eq}(\dx) \simeq \frac{ - 2s(\x\eq, \dx)^2}{1 - \e^{S(\x\eq, \dx)}}   \rho_{\x\eq}(-\dx) \text{ from eq.
+\Longrightarrow \frac{ 2s(x\eq, \dx)^2}{1 - \e^{-S(x\eq, \dx)}}   \rho_{x\eq}(\dx) \simeq \frac{ - 2s(x\eq, \dx)^2}{1 - \e^{S(x\eq, \dx)}}   \rho_{x\eq}(-\dx) \text{ from eq.
 \ref{eq:S_minus_deltax}}, \\
-\iff \frac{ \rho_{\x\eq}(\dx)}{\rho_{\x\eq}(-\dx)} \simeq \e^{-S(\x\eq, \dx)} \frac{ \e^{-S(\x\eq, \dx)} - 1}{ \e^{-S(\x\eq, \dx)} \left( 1 - \e^{S(\x\eq, \dx)} \right)}, \\
-\iff \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq -S(\x\eq, \dx) \text{ from eq. \ref{eq:proba}} \label{eq:equilibrium}, \\
-\iff \lambda_{K}(\x\eq) \simeq -S(\x\eq, \dx) \label{eq:equilibrium_lambda},
+\iff \frac{ \rho_{x\eq}(\dx)}{\rho_{x\eq}(-\dx)} \simeq \e^{-S(x\eq, \dx)} \frac{ \e^{-S(x\eq, \dx)} - 1}{ \e^{-S(x\eq, \dx)} \left( 1 - \e^{S(x\eq, \dx)} \right)}, \\
+\iff \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (\Nstate-1) \simeq -S(x\eq, \dx) \text{ from eq. \ref{eq:proba}} \label{eq:equilibrium}, \\
+\iff \lambda_{K}(x\eq) \simeq -S(x\eq, \dx) \label{eq:equilibrium_lambda},
 \end{gather}
-where  $\lambda_{K}(\x\eq) = \ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1)$.
-\subsection{Substitution rate bias ($\bm{\dnds}$) at equilibrium}
-The substitution rate of selected mutations relative to neutral mutations ($\dnds$) is by definition:
+where  $\lambda_{K}(x\eq) = \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (\Nstate-1)$.
+\subsection{Substitution rate bias ($\bm{\omega}$) at equilibrium}
+The substitution rate of selected mutations relative to neutral mutations ($\omega$) is by definition:
 \begin{align}
-\dnds & = \E_{\dx} \left[ \frac{p_{\text{fix}}(\x, \dx)}{p_{\text{fix}}(\x, 0)} \right], \\
- & = (1 - \x) \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} + \x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) + \frac{\x}{\Nstate-1} \frac{ S(\x, -\dx)}{1 - \e^{-S(\x, -\dx)}} \text{ from eq.
+\omega & = \E_{\dx} \left[ \frac{p_{\text{fix}}(x, \dx)}{p_{\text{fix}}(x, 0)} \right], \\
+ & = (1 - x) \frac{ S(x, \dx)}{1 - \e^{-S(x, \dx)}} + x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) + \frac{x}{\Nstate-1} \frac{ S(x, -\dx)}{1 - \e^{-S(x, -\dx)}} \text{ from eq.
 \ref{eq:proba}, \ref{eq:pfix} and \ref{eq:pfix0}}, \\
- & = (1 - \x) \frac{ S(\x, \dx)}{1 - \e^{-S(\x, \dx)}} - \frac{\x}{\Nstate-1}  \frac{ S(\x, \dx)}{1 - \e^{S(\x, \dx)}} +  \x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) \text{ from eq.
+ & = (1 - x) \frac{ S(x, \dx)}{1 - \e^{-S(x, \dx)}} - \frac{x}{\Nstate-1}  \frac{ S(x, \dx)}{1 - \e^{S(x, \dx)}} +  x \left(\frac{\Nstate - 2}{\Nstate - 1}\right) \text{ from eq.
 \ref{eq:S_minus_deltax}}.
 \end{align}
-$\dnds\eq$ at equilibrium is then determined by the phenotype at equilibrium $\x\eq$:
+$\omega\eq$ at equilibrium is then determined by the phenotype at equilibrium $x\eq$:
 \begin{align}
-\dnds\eq &= (1 - \x\eq) \frac{ S(\x\eq, \dx)}{1 - \e^{-S(\x\eq, \dx)}} - \frac{\x\eq}{\Nstate-1} \frac{ S(\x\eq, \dx)}{1 - \e^{S(\x\eq, \dx)}} + \x\eq \left(\frac{\Nstate - 2}{\Nstate - 1}\right) , \\
- &= \x\eq \left[ \frac{2 (\x\eq-1)  \lambda_{K}(\x\eq) }{\Nstate (\x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right] \text{ from eq. \ref{eq:equilibrium}}.
+\omega\eq &= (1 - x\eq) \frac{ S(x\eq, \dx)}{1 - \e^{-S(x\eq, \dx)}} - \frac{x\eq}{\Nstate-1} \frac{ S(x\eq, \dx)}{1 - \e^{S(x\eq, \dx)}} + x\eq \left(\frac{\Nstate - 2}{\Nstate - 1}\right) , \\
+ &= x\eq \left[ \frac{2 (x\eq-1)  \lambda_{K}(x\eq) }{\Nstate (x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right] \text{ from eq. \ref{eq:equilibrium}}.
 \end{align}
@@
 \end{center}
-Moreover, given that the number of state if large enough $\Nstate \gg 1$, the equilibrium $\dnds$ can be approximated as:
+Moreover, given that the number of state if large enough $\Nstate \gg 1$, the equilibrium $\omega$ can be approximated as:
 \begin{align}
-\dnds\eq  & = \x\eq \left[ \frac{2 (\x\eq-1)  \lambda_{K}(\x\eq) }{\Nstate (\x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right], \\
-& \simeq \x\eq
+\omega\eq  & = x\eq \left[ \frac{2 (x\eq-1)  \lambda_{K}(x\eq) }{\Nstate (x\eq-1)+1} + \frac{\Nstate - 2}{\Nstate - 1}\right], \\
+& \simeq x\eq
 \end{align}
-And the derivative of $\dnds\eq$ w.r.t to $\x\eq$ is: 
+And the derivative of $\omega\eq$ w.r.t to $x\eq$ is: 
 \begin{gather}
-\frac{\der \dnds\eq}{\der \x\eq} = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1} \label{eq:dw_dx}.
+\frac{\der \omega\eq}{\der x\eq} = 2 \left[ \frac{\Nstate (x\eq- 1) + 1+\left[\Nstate (x\eq-1)^2+2 x\eq-1\right] \lambda_{K}(x\eq) }{(\Nstate (x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1} \label{eq:dw_dx}.
 \end{gather}
@@
 \end{center}
-Moreover, given that the number of state if large enough $\Nstate \gg 1$, the response in equilibrium $\dnds$ due to change in phenotype can be approximated as:
+Moreover, given that the number of state if large enough $\Nstate \gg 1$, the response in equilibrium $\omega$ due to change in phenotype can be approximated as:
 \begin{align}
-\frac{ \der \dnds\eq}{\der \x\eq}  & = 2 \left[ \frac{\Nstate (\x\eq- 1) + 1+\left[\Nstate (\x\eq-1)^2+2 \x\eq-1\right] \lambda_{K}(\x\eq) }{(\Nstate (\x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}, \\
-& \simeq \frac{2 \lambda_{K}(\x\eq)}{\Nstate} + 1, \\
+\frac{ \der \omega\eq}{\der x\eq}  & = 2 \left[ \frac{\Nstate (x\eq- 1) + 1+\left[\Nstate (x\eq-1)^2+2 x\eq-1\right] \lambda_{K}(x\eq) }{(\Nstate (x\eq-1)+1)^2}\right] + \frac{\Nstate - 2}{\Nstate - 1}, \\
+& \simeq \frac{2 \lambda_{K}(x\eq)}{\Nstate} + 1, \\
 & \simeq 1. \label{eq:approx_domega_dx}
@@
 \subsection{$\bm{\omega}$ elasticity after a change in $\bm{\Ne}$}
-Define the function $G(\x, \Ne )$ as:
+Define the function $G(x, \Ne )$ as:
 \begin{gather}
-G(\x, \Ne ) \equiv \lambda_K (\x\eq) + 4 \Ne s(\x, \dx), 
+G(x, \Ne ) \equiv \lambda_K (x\eq) + 4 \Ne s(x, \dx), 
 \end{gather}
-The equilibrium equation (eq. \ref{eq:equilibrium}) states that $G(\x\eq, \Ne )=0$, meaning that $\x\eq$ is implicitly a function of $\Ne$:
+The equilibrium equation (eq. \ref{eq:equilibrium}) states that $G(x\eq, \Ne )=0$, meaning that $x\eq$ is implicitly a function of $\Ne$:
 \begin{gather}
-G(\x\eq(\Ne), \Ne ) = 0, \\
-\Longrightarrow \frac{\partial G(\x\eq, \Ne ) }{\partial \x\eq }\frac{ \der \x\eq}{\der \Ne} + \frac{ \partial G(\x\eq, \Ne )}{\partial \Ne} = 0, \\
-\iff \left[  \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq }  + 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right]\frac{ \der \x\eq}{\der \Ne} + 4 s(\x\eq, \dx) = 0, \\
-\iff \left[  \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + 4 \Ne \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}\dx \right]\frac{ \der \x\eq}{\der \Ne}  = - 4 \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}\dx \text{ from eq.
+G(x\eq(\Ne), \Ne ) = 0, \\
+\Longrightarrow \frac{\partial G(x\eq, \Ne ) }{\partial x\eq }\frac{ \der x\eq}{\der \Ne} + \frac{ \partial G(x\eq, \Ne )}{\partial \Ne} = 0, \\
+\iff \left[  \frac{\partial \lambda_K (x\eq) }{\partial x\eq }  + 4 \Ne \frac{ \partial s(x\eq, \dx) }{\partial x\eq } \right]\frac{ \der x\eq}{\der \Ne} + 4 s(x\eq, \dx) = 0, \\
+\iff \left[  \frac{\partial \lambda_K (x\eq) }{\partial x\eq } + 4 \Ne \frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}\dx \right]\frac{ \der x\eq}{\der \Ne}  = - 4 \frac{ \partial \ln f(x\eq) }{\partial {x\eq}}\dx \text{ from eq.
 \ref{eq:s_from_fitness}}, \\
-\iff 4 \dx \left[ \frac{1}{4 \dx \Ne } \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2} \right] \Ne \frac{ \der \x\eq}{\der \Ne}  = - 4 \dx \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}, \\
-\iff \frac{ \der \x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne} \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}  \label{eq:dx_dlnNe}.
+\iff 4 \dx \left[ \frac{1}{4 \dx \Ne } \frac{\partial \lambda_K (x\eq) }{\partial x\eq } + \frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2} \right] \Ne \frac{ \der x\eq}{\der \Ne}  = - 4 \dx \frac{ \partial \ln f(x\eq) }{\partial {x\eq}}, \\
+\iff \frac{ \der x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{1}{4 \dx \Ne} \frac{\partial \lambda_K (x\eq) }{\partial x\eq } + \frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}  \label{eq:dx_dlnNe}.
 \end{gather}
@@
 \begin{align}
-\frac{\der \dnds\eq}{\der \ln (\Ne)} & = \frac{\der \dnds\eq}{\der \x\eq} \frac{ \der \x\eq}{\der \ln (\Ne)}, \\
- &= - \frac{\der \dnds\eq}{\der \x\eq} \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne } \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}} \text{ from eq.
+\frac{\der \omega\eq}{\der \ln (\Ne)} & = \frac{\der \omega\eq}{\der x\eq} \frac{ \der x\eq}{\der \ln (\Ne)}, \\
+ &= - \frac{\der \omega\eq}{\der x\eq} \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{1}{4 \dx \Ne } \frac{\partial \lambda_K (x\eq) }{\partial x\eq } + \frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}} \text{ from eq.
 \ref{eq:dx_dlnNe}} \label{eq:dw_dlnNe}.
 \end{align}
-Moreover, with the approximation that $\left| 4 \Ne \frac{ \partial s(\x\eq, \dx) }{\partial \x\eq } \right| \gg \left| \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } \right|$, meaning that a change in phenotype causes a higher change in scaled selection coefficient than mutational bias, we have:
+Moreover, with the approximation that $\left| 4 \Ne \frac{ \partial s(x\eq, \dx) }{\partial x\eq } \right| \gg \left| \frac{\partial \lambda_K (x\eq) }{\partial x\eq } \right|$, meaning that a change in phenotype causes a higher change in scaled selection coefficient than mutational bias, we have:
 \begin{gather}
-\frac{ \der \x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{1}{4 \dx \Ne} \frac{\partial \lambda_K (\x\eq) }{\partial \x\eq } + \frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}},\\
-\Longrightarrow \frac{ \der \x\eq}{\der \ln (\Ne)}  \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}. \label{eq:approx_dx_dlnNe}
+\frac{ \der x\eq}{\der \ln (\Ne)}  = - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{1}{4 \dx \Ne} \frac{\partial \lambda_K (x\eq) }{\partial x\eq } + \frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}},\\
+\Longrightarrow \frac{ \der x\eq}{\der \ln (\Ne)}  \simeq - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}. \label{eq:approx_dx_dlnNe}
 \end{gather}
-Together, these approximations leads to the following elasticity in equilibrium $\dnds$ after change in $\Ne$ as:
+Together, these approximations leads to the following elasticity in equilibrium $\omega$ after change in $\Ne$ as:
 \begin{align}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\frac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}\label{eq:approx_dw_dlnNe}
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}\label{eq:approx_dw_dlnNe}
 \end{align}
@@
 \subsection{Folded fraction}
-All phenotype-fitness functions considered below are log-concave, and as a result, $\frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}$ is an decreasing function of $\x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
+All phenotype-fitness functions considered below are log-concave, and as a result, $\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}$ is an decreasing function of $x$; the less stable the protein already is, the stronger the purifying selection against additional destabilizing mutations.
 More precisely, fitness functions depends on the folded fraction of the protein of interest, which is given by the Fermi-Dirac distribution:
 \begin{eqnarray}
-P_{F} (\x) &=& \frac{1}{1 + e^{\beta(\alpha + \gamma \Nsite \x)}}
+P_{F} (x) &=& \frac{1}{1 + e^{\beta(\alpha + \gamma \Nsite x)}}
 \end{eqnarray}
-where $\x$ is the fraction of destabilizing mutations, each contributing a $\Delta \Delta G = \gamma$, and $\beta = 1 / kT$.
+where $x$ is the fraction of destabilizing mutations, each contributing a $\Delta \Delta G = \gamma$, and $\beta = 1 / kT$.
 Thus, $\alpha = \Delta G_0 < 0$ is the difference in free energy between folded and unfolded state when all sites are stable.
@@
 \begin{eqnarray}
-P_{F}(\x) &=& 1 - P_{U}(\x) \\
-&\simeq& 1 - e^{\beta(\alpha + \gamma \Nsite \x)}
+P_{F}(x) &=& 1 - P_{U}(x) \\
+&\simeq& 1 - e^{\beta(\alpha + \gamma \Nsite x)}
 \end{eqnarray}
@@
 \begin{eqnarray}
-P_{U}(\x) &\simeq& e^{\beta(\alpha + \gamma \Nsite \x)}
+P_{U}(x) &\simeq& e^{\beta(\alpha + \gamma \Nsite x)}
 \end{eqnarray}
@@
 \begin{gather}
-f(\x) = \frac{1}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}}. \label{eq:fitness}
+f(x) = \frac{1}{1 + \e^{\beta(\alpha + \Nsite \gamma x)}}. \label{eq:fitness}
 \end{gather}
@@
 \begin{align}
-\frac{\partial \ln (f(\x))}{\partial \x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\alpha + \Nsite \gamma \x)} \right)}{\partial \x} \text{ from eq. \ref{eq:fitness}}, \\
-& = - \beta \Nsite \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}} \label{eq:df_dx},\\
-& \simeq - \beta \Nsite \gamma \e^{\beta(\alpha + \Nsite \gamma \x)}.
+\frac{\partial \ln (f(x))}{\partial x}  & = - \frac{\partial \ln \left( 1 + \e^{\beta(\alpha + \Nsite \gamma x)} \right)}{\partial x} \text{ from eq. \ref{eq:fitness}}, \\
+& = - \beta \Nsite \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma x)}} \label{eq:df_dx},\\
+& \simeq - \beta \Nsite \gamma \e^{\beta(\alpha + \Nsite \gamma x)}.
 \end{align}
-The equilibrium phenotype ($\x\eq$) is :
+The equilibrium phenotype ($x\eq$) is :
 \begin{gather}
-\lambda_{K}(\x\eq) = 4\Ne \beta \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x\eq)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x\eq)}}  \text{ from eq. \ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
+\lambda_{K}(x\eq) = 4\Ne \beta \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma x\eq)}}{1 + \e^{\beta(\alpha + \Nsite \gamma x\eq)}}  \text{ from eq. \ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
 \end{gather}
@@
 \begin{align}
-\frac{\partial^2 \ln (f(\x))}{\partial \x^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial \x} \left( \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}} \right) \text{ from eq.
+\frac{\partial^2 \ln (f(x))}{\partial x^2} & = - \beta \Nsite \gamma \frac{\partial}{\partial x} \left( \frac{\e^{\beta(\alpha + \Nsite \gamma x)}}{1 + \e^{\beta(\alpha + \Nsite \gamma x)}} \right) \text{ from eq.
 \ref{eq:df_dx}}, \\
- & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma \x)}}{\left( 1 + \e^{\beta(\alpha + \Nsite \gamma \x)}\right)^2}, \\
- & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\alpha + \Nsite \gamma \x)}} \frac{ \partial \ln f(\x) }{\partial \x} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}, \\
- & \simeq \beta \Nsite \gamma \frac{ \partial \ln f(\x) }{\partial \x} 
+ & = - \beta \Nsite \gamma  \beta \Nsite \gamma \frac{\e^{\beta(\alpha + \Nsite \gamma x)}}{\left( 1 + \e^{\beta(\alpha + \Nsite \gamma x)}\right)^2}, \\
+ & = \frac{\beta \Nsite \gamma}{1 + \e^{\beta(\alpha + \Nsite \gamma x)}} \frac{ \partial \ln f(x) }{\partial x} \text{ from eq. \ref{eq:df_dx}} \label{eq:df2_dx2}, \\
+ & \simeq \beta \Nsite \gamma \frac{ \partial \ln f(x) }{\partial x} 
 \end{align}
@@
 \begin{align}
-\frac{ \der \dnds\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma} \text{from eq. \ref{eq:df2_dx2} and \ref{eq:approx_domega_dx}},
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma} \text{from eq. \ref{eq:df2_dx2} and \ref{eq:approx_domega_dx}},
 \end{align}
-which is independent of $\x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
+which is independent of $x\eq$, meaning $\omega$ is linearly decreasing with $\Ne$ in log space.
 This model, however, does not express the fact that selection is typically stronger for proteins characterized by higher levels of expression. 
@@
 \begin{eqnarray}
-\ln f(\x) = - A y P_{U} (\x)
+\ln f(x) = - A y P_{U} (x)
 \end{eqnarray}
@@
 \begin{eqnarray}
-\frac{\partial \ln (f(\x))}{\partial \x} &\simeq& -A \beta \gamma \Nsite y  e^{\beta(\alpha + \gamma \Nsite \x)}
+\frac{\partial \ln (f(x))}{\partial x} &\simeq& -A \beta \gamma \Nsite y  e^{\beta(\alpha + \gamma \Nsite x)}
 \end{eqnarray}
@@
 \begin{eqnarray}
-4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite \x)}
+4 \Ne \frac{\partial \ln (f(x))}{\partial x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite x)}
 \end{eqnarray}
@@
 \begin{align}
-\frac{ \der \dnds\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma}.
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \frac{1}{\beta \Nsite \gamma}.
 \end{align}
@@
 \begin{eqnarray}
-\ln f(\x) &=& - A y \sum_k \pi_k e^{\beta(\alpha + \gamma\Nsite \x + \gamma k)}
+\ln f(x) &=& - A y \sum_k \pi_k e^{\beta(\alpha + \gamma\Nsite x + \gamma k)}
 \\
-&=& - A y  e^{\beta(\alpha + \gamma\Nsite \x)} \sum_k e^{-\rho n} \frac{(\rho n)^k}{k!} e^{\beta \gamma k}
+&=& - A y  e^{\beta(\alpha + \gamma\Nsite x)} \sum_k e^{-\rho n} \frac{(\rho n)^k}{k!} e^{\beta \gamma k}
 \\
-&=& - A y  e^{\beta(\alpha + \gamma\Nsite \x) + \rho \Nsite (e^{\beta \gamma} -1 )}
+&=& - A y  e^{\beta(\alpha + \gamma\Nsite x) + \rho \Nsite (e^{\beta \gamma} -1 )}
 \\
-& \simeq & - A y  e^{\beta(\alpha + \gamma\Nsite \x) + \rho \beta \gamma \Nsite}
+& \simeq & - A y  e^{\beta(\alpha + \gamma\Nsite x) + \rho \beta \gamma \Nsite}
 \\
@@
 \begin{eqnarray}
-4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite (x + \rho))}
-\\ &\propto& -4 \Ne y e^{\beta(\alpha + \gamma\Nsite \x)}
+4 \Ne \frac{\partial \ln (f(x))}{\partial x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite (x + \rho))}
+\\ &\propto& -4 \Ne y e^{\beta(\alpha + \gamma\Nsite x)}
 \end{eqnarray}
@@
 Specifically, the protein is assumed to be regulated so as to reach a level of expression of functional proteins of $y$, and contributes a total benefit $B$ (which depends on its specific function).
-Given that only a fraction $P_{F}(\x) = 1-P_{U}(\x)$ of the total amount of protein expressed by the cell is functional, the total cost of expression $C$ is then equal to:
+Given that only a fraction $P_{F}(x) = 1-P_{U}(x)$ of the total amount of protein expressed by the cell is functional, the total cost of expression $C$ is then equal to:
 \begin{eqnarray}
-C(\x) &=& \frac{y}{P_{F}(\x)}
-\\ &\simeq& y (1 + P_{U}(\x))
+C(x) &=& \frac{y}{P_{F}(x)}
+\\ &\simeq& y (1 + P_{U}(x))
 \end{eqnarray}
@@
 \begin{eqnarray}
-\ln f (\x) & = & -A \frac{y}{B} \left( 1 + e^{\beta(\alpha + \gamma\Nsite \x)} \right)
+\ln f (x) & = & -A \frac{y}{B} \left( 1 + e^{\beta(\alpha + \gamma\Nsite x)} \right)
 \\
-&=& -by (1 + e^{\beta(\alpha + \gamma\Nsite \x)})
+&=& -by (1 + e^{\beta(\alpha + \gamma\Nsite x)})
 \end{eqnarray}
@@
 \begin{eqnarray}
-4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma\Nsite \x)}
+4 \Ne \frac{\partial \ln (f(x))}{\partial x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma\Nsite x)}
 \end{eqnarray}
@@
 \begin{eqnarray}
-4 \Ne \frac{\partial \ln (f(\x))}{\partial \x} &\propto& -4 \Ne g(y) e^{\beta(\alpha + \gamma\Nsite \x)}
+4 \Ne \frac{\partial \ln (f(x))}{\partial x} &\propto& -4 \Ne g(y) e^{\beta(\alpha + \gamma\Nsite x)}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\ln f(\x) &=& -b \sum_i \epsilon_i
+\ln f(x) &=& -b \sum_i \epsilon_i
 \end{eqnarray}
@@
 \begin{eqnarray}
-\ln \kappa_i &=& \ln \left( b \beta \gamma \Nsite m \frac{C}{C_0}  \right) \, + \, \beta (\alpha +\gamma \Nsite \bar x) \, + \, \beta \gamma\Nsite \x_i
+\ln \kappa_i &=& \ln \left( b \beta \gamma \Nsite m \frac{C}{C_0}  \right) \, + \, \beta (\alpha +\gamma \Nsite \bar x) \, + \, \beta \gamma\Nsite x_i
 \end{eqnarray}
@@
 
-\subsection{Simulated $\bm{\dnds}$ elasticity to changes in $\bm{\Ne}$}
+\subsection{Simulated $\bm{\omega}$ elasticity to changes in $\bm{\Ne}$}
 \begin{center}
@@
 \end{center}
-\textbf{$\bm{\dnds}$ elasticity with gamma distributed selection coefficient}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+\textbf{$\bm{\omega}$ elasticity with gamma distributed selection coefficient}.
+$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
 For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-In the model of gamma distributed fitness effect, $\dnds$ at equilibrium is strongly dependent on log-$\Ne$ where the slope correlation is proportional to the inverse of the shape parameter of the gamma distribution \cite{Welch2008}.
+In the model of gamma distributed fitness effect, $\omega$ at equilibrium is strongly dependent on log-$\Ne$ where the slope correlation is proportional to the inverse of the shape parameter of the gamma distribution \cite{Welch2008}.
 \begin{center}
@@
 \end{center}
-\textbf{$\bm{\dnds}$ elasticity with amino-acid fitness profiles}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (relative).
+\textbf{$\bm{\omega}$ elasticity with amino-acid fitness profiles}.
+$\omega$ at equilibrium as a function of $\bm{\Ne}$ (relative).
 For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-In the model of site-wise amino-acid fitness profiles taken from \cite{Bloom2017}, $\dnds$ at equilibrium is strongly dependent on log-$\Ne$. 
+In the model of site-wise amino-acid fitness profiles taken from \cite{Bloom2017}, $\omega$ at equilibrium is strongly dependent on log-$\Ne$. 
 \begin{center}
@@
 \end{center}
-\textbf{$\bm{\dnds}$ elasticity with additive free energy of folding}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+\textbf{$\bm{\omega}$ elasticity with additive free energy of folding}.
+$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
 For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
@@
 The simulations of our additive free energy model match the theoretical prediction that the slope of the linear relation (dashed line) is equal to $\beta \Nsite \gamma)^{-1} = 0.00198 \simeq = 0.00199$.
-The non-monotony is suspected to be due to the discrete number of sites and states, such that the changes in $\deltaG$ after a mutation is either $-1$, $0$ or $1$.
-Such non-monotony is not observed with the Grantham model, in which the $\dnds$ is lower and the slope of elasticity lower, closer to the empirical 3D model of Golstein \& Pollock.
-\begin{center}
- \includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Beta-Elasticity.pdf}
-\end{center}
-\textbf{Effect of $\bm{\beta}$ on the elasticity}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+%The non-monotony is suspected to be due to the discrete number of sites and states, such that the changes in $\deltaG$ after a mutation is either $-1$, $0$ or $1$.
+Such non-monotony is not observed with the Grantham model, in which the $\omega$ is lower and the slope of elasticity lower, closer to the empirical 3D model of Golstein \& Pollock.
+%\begin{center}
+% \includegraphics[width=0.8\textwidth] {artworks/SimuStab-Grantham-Beta-Elasticity.pdf}
+%\end{center}
+%\textbf{Effect of $\bm{\beta}$ on the elasticity}.
+%$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+%For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+%The fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
 $\beta=1.686$ are given in the legend.
-Increasing $\beta$ increases the slope of the $\dnds$-$\Ne$ relationship, as predicted in our theoretical model.
-\begin{center}
- \includegraphics[width=0.8\textwidth] {artworks/SimuStab-GammaStd-Elasticity.pdf}
-\end{center}
-\textbf{$\bm{\dnds}$ elasticity with additive free energy of folding}.
-$\dnds$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-The parameters are $\alpha=-118$, $\Nsite=300$, $\beta=1.686$ and each site has it's own gamma distributed $\gamma$ with mean $1$ and standard deviation given in the legend.
-Increasing the variance of $\gamma$ increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites with low $\gamma$ are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
-Once many sites are unstable, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
-However the slope of the $\dnds$-$\Ne$ relationship is not sensibly changed.
-\subsection{Simulated relaxation time of $\bm{\dnds}$}
+%Increasing $\beta$ increases the slope of the $\omega$-$\Ne$ relationship, as predicted in our theoretical model.
+%\begin{center}
+% \includegraphics[width=0.8\textwidth] {artworks/SimuStab-GammaStd-Elasticity.pdf}
+%\end{center}
+%\textbf{$\bm{\omega}$ elasticity with additive free energy of folding}.
+%$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
+%For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
+%The parameters are $\alpha=-118$, $\Nsite=300$, $\beta=1.686$ and each site has it's own gamma distributed $\gamma$ with mean $1$ and standard deviation given in the legend.
+%Increasing the variance of $\gamma$ increases $\omega$, by shifting the equilibrium to higher $x\eq$ since more unstable sites with low $\gamma$ are fixed before reaching sensible deleterious selection coefficient against unstable mutations.
+Once many sites are unstable, the $\omega$ is higher since non-synonymous mutations between unstable states are effectively neutral.
+However the slope of the $\omega$-$\Ne$ relationship is not sensibly changed.
+\subsection{Simulated relaxation time of $\bm{\omega}$}
 \begin{center}
@@
 \end{center}
-\textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$}.
-$\dnds$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+\textbf{Relaxation time of $\bm{\omega}$ dependence on $\Nsites$}.
+$\omega$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
 Solid line corresponds to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
@@
 The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Increasing $\Nsite$ implies a higher $\dnds$ at equilibrium, a lower elasticity of the $\dnds$ to changes in $\Ne$ and a higher rate of relaxation.
+Increasing $\Nsite$ implies a higher $\omega$ at equilibrium, a lower elasticity of the $\omega$ to changes in $\Ne$ and a higher rate of relaxation.
 \begin{center}
@@
 \end{center}
-\textbf{Relaxation time of $\bm{\dnds}$ dependence on $\Nsites$, while correction for $\gamma$}.
-$\dnds$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+\textbf{Relaxation time of $\bm{\omega}$ dependence on $\Nsites$, while correction for $\gamma$}.
+$\omega$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
 Solid line corresponds to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
@@
 The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
-Increasing $\Nsite$ implies a higher $\dnds$ at equilibrium, and a higher rate of relaxation.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\omega$ to changes in $\Ne$ is kept constant.
+Increasing $\Nsite$ implies a higher $\omega$ at equilibrium, and a higher rate of relaxation.
 \begin{center}
@@
 \end{center}
-\textbf{Relaxation time of $\bm{\dnds}$ for the Grantham model}.
-$\dnds$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
+\textbf{Relaxation time of $\bm{\omega}$ for the Grantham model}.
+$\omega$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
 Solid line corresponds to the average over replicates ($r$) and the shaded area correspond to the $90\%$ interval among replicates. 
@@
 The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\dnds$ to changes in $\Ne$ is kept constant.
-Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $\x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\omega$ to changes in $\Ne$ is kept constant.
+Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
 Increasing $\Nsite$ implies a higher rate of relaxation.

### 2020-06-16 Scaling/simulated_experiment.py
Scripts and config files to replicate the figures

@@ -1,50 +1,0 @@
-#!python3
-import argparse
-import os
-from subprocess import run
-
-
-def create_experiment(experiment, screen, sbatch, nbr_cpu):
-    exp_path = os.getcwd() + '/Experiments/' + experiment
-
-    os.makedirs(exp_path, exist_ok=True)
-    os.system('cp config.yaml {0}'.format(exp_path))
-    os.remove(exp_path + "/Snakefile") if os.path.exists(exp_path + "/Snakefile") else None
-    os.symlink(os.getcwd() + "/Snakefile", exp_path + "/Snakefile")
-    run_file = exp_path + "/snakeslurm.sh"
-    with open(run_file, 'w') as w:
-        w.write("#!/usr/bin/env bash\n")
-        run_str = 'snakemake '
-        if sbatch:
-            run_str += '-j 99 --cluster "sbatch -J {0} -p long -N 1 ' \
-                       '-o {1}/slurm.%x.%j.out -e {1}/slurm.%x.%j.err '.format(experiment, exp_path)
-            run_str += '--cpus-per-task={params.threads} --mem={params.mem} -t {params.time}"\n'
-        else:
-            run_str += "-j {0} --printshellcmds".format(nbr_cpu)
-        w.write(run_str)
-    os.system("chmod 755 " + run_file)
-    cmd = 'cd ' + exp_path + ' && ./snakeslurm.sh'
-    screen_cmd = 'screen -dmS ' + experiment + ' bash -c "' + cmd + '"'
-    with open(exp_path + "/screen.sh", 'w') as w:
-        w.write("#!/usr/bin/env bash\n")
-        w.write(screen_cmd)
-    if screen:
-        print(screen_cmd)
-        run(screen_cmd, shell=True)
-    else:
-        print(cmd)
-        run(cmd, shell=True)
-
-
-if __name__ == '__main__':
-    parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument('-e', '--experiment', required=False, type=str, default="selcoeff", dest="experiment")
-    parser.add_argument('-s', '--screen', required=False, type=bool, default=False, dest="screen")
-    parser.add_argument('-b', '--sbatch', required=False, type=bool, default=False, dest="sbatch")
-    parser.add_argument('-j', '--nbr_cpu', required=False, type=int, default=6, dest="nbr_cpu")
-    args = parser.parse_args()
[…]

### 2020-06-16 copy_artworks_after_run_all.sh

@@ -1,0 +1,18 @@
+#!/usr/bin/env bash
+\cp -f ./Relaxation/Experiments/figure-4A/merge_all/Stability.pdf ./manuscript/artworks/Relaxation-Stability-Alpha-Gamma.pdf
+\cp -f ./Relaxation/Experiments/Supp-7-1/merge_all/Stability.pdf ./manuscript/artworks/Relaxation-Stability.pdf
+\cp -f ./Relaxation/Experiments/Supp-7-2/merge_all/Stability.pdf ./manuscript/artworks/Relaxation-Stability-Gamma.pdf
+\cp -f ./Relaxation/Experiments/Supp-7-3/merge_all/Stability.pdf ./manuscript/artworks/Relaxation-Stability-Grantham-Alpha-Gamma.pdf
+\cp -f ./Scaling/Experiments/figure-3A/heatmap_SimuStab_plot/dnd0_event_tot.pdf ./manuscript/artworks/SimuStab-Alpha-Gamma-Elasticity.pdf
+\cp -f ./Heatmap/Experiments/figure-3B/heatmap_SimuStab_plot/dnd0_event_tot.pdf ./manuscript/artworks/SimuStab-Grantham-Alpha-Elasticity.pdf
+\cp -f ./Heatmap/Experiments/figure-3C/heatmap_SimuFold_plot/dnd0_event_tot.pdf ./manuscript/artworks/SimuFold-Elasticity.pdf
+\cp -f ./Heatmap/Experiments/figure-3C/heatmap_SimuFold_plot/sub-ΔG-mean.pdf ./manuscript/artworks/SimuFold-Elasticity-DG.pdf
+\cp -f ./Heatmap/Experiments/figure-3D/heatmap_SimuStab_plot/dnd0_event_tot.pdf ./manuscript/artworks/SimuStab-Grantham-Elasticity.pdf
+\cp -f ./Heatmap/Experiments/figure-4B/merge_all/Stability.pdf ./manuscript/artworks/Relaxation-DFE-Profile-N500.pdf
+\cp -f ./Heatmap/Experiments/Supp-6-1/heatmap_SimuDfe_plot/dnd0_event_tot.pdf ./manuscript/artworks/SimuDfe-Elasticity.pdf
[…]

### 2020-06-16 run_all.sh

@@ -1,10 +1,7 @@
 #!/usr/bin/env bash
-for EXPERIMENT in ./Heatmap/Experiments/*; do
-  NAME=$(basename "${EXPERIMENT}")
-  echo "${NAME}"
-  cd ${EXPERIMENT}
-  # sed -i 's#X_PARAM_MAGNITUDE: 1#X_PARAM_MAGNITUDE: 2#g' config.yaml
-  snakemake --unlock
-  snakemake --printshellcmds --rerun-incomplete -j 8
-  cd ../../..
+CPU=2
+for FOLDER in "Heatmap" "Scaling" "Relaxation"; do
+    for EXPERIMENT in ${FOLDER}/*-6-4.yaml; do
+      python3 ./scripts/simulated_experiment.py -f ${FOLDER} -c $(basename "${EXPERIMENT}") -j ${CPU}
+    done
 done

### 2020-06-16 scripts/heatmap_analysis.py

@@ -13,17 +13,24 @@ matplotlib.rcParams['xtick.labelsize'] = label_size
 matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+from cycler import cycler
+
+plt.rc('axes', prop_cycle=cycler(color=["#5D80B4", "#E29D26", "#8FB03E", "#EB6231", "#857BA1"]))
 
 
 def print_b(text):
     print('\033[34m' + '\033[1m' + text + '\033[0m')
 
 
-def tex_float(x):
-    s = "{0:.3g}".format(x)
-    if "e" in s:
-        mantissa, exp = s.split('e')
-        return mantissa + '\\times 10^{' + exp + '}'
+def grec_letter(s):
+    if s == "gamma" or s == "alpha" or s == "beta" or s == "chi" or s == "omega":
+        return '\\' + s
+    elif s == "pop_size":
+        return "N_{\\mathrm{e}}"
+    elif s == "gamma_std":
+        return "\\sigma ( \\gamma )"
+    elif s == "gamma_distribution_shape":
+        return "k"
     else:
         return s
 
@@ -32,68 +39,80 @@ if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
     parser.add_argument('-o', '--output', required=True, type=str, dest="output")
     parser.add_argument('-i', '--input', required=True, type=str, nargs='+', dest="input")
-    parser.add_argument('-x', '--x_param', required=True, type=str, dest="x_param")
-    parser.add_argument('-y', '--y_param', required=False, default="", type=str, dest="y_param")
-    parser.add_argument("--y_param_key", required=False, action='append', type=lambda kv: kv.split(":"),
-                        dest='y_param_dict')
     parser.add_argument('-f', '--node', required=False, default=False, type=bool, dest="node")
     args = parser.parse_args()
-    args.y_param_dict = dict([(float(k), v) for k, v in args.y_param_dict]) if (
-                args.y_param_dict is not None) else dict()
     array_values = dict()
+    li = []
     for filepath in args.input:
         if not os.path.isfile(filepath): continue
+        x, y = float(filepath.split("_")[-3]), float(filepath.split("_")[-2])
[…]

### 2020-06-16 scripts/relaxation_merge.py

@@ -14,7 +14,9 @@ matplotlib.rcParams['xtick.labelsize'] = label_size
 matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+from cycler import cycler
 
+plt.rc('axes', prop_cycle=cycler(color=["#5D80B4", "#E29D26", "#8FB03E", "#EB6231", "#857BA1"]))
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
     parser.add_argument('-a', '--age', required=True, type=float, dest="age")
@@ -41,7 +43,7 @@ if __name__ == '__main__':
         plt.fill_between(t_range, np.percentile(array, 5, axis=0), np.percentile(array, 95, axis=0), alpha=0.2)
 
     plt.xlabel(r'$t$', fontsize=label_size)
-    plt.ylabel("\\omega", fontsize=label_size)
+    plt.ylabel("$\\omega$", fontsize=label_size)
     for s in range(1, args.branches):
         plt.axvline(x=s * args.age, linewidth=3, color='black')
     plt.xlim((0, args.branches * args.age))

### 2020-06-16 scripts/replicate_analysis.py

@@ -1,0 +1,23 @@
+import os
+from glob import glob
+import pandas as pd
+
+os.chdir("../")
+os.makedirs("Analysis", exist_ok=True)
+li = list()
+
+for exp in ["Heatmap", "Scaling"]:
+    for exp_replicate in sorted(glob(exp + "/Experiments/*")):
+        for heatmap_folder in sorted(glob(exp_replicate + "/heatmap_*_plot/")):
+            if not os.path.isfile(heatmap_folder + "results.tsv"): continue
+            li.append(pd.read_csv(heatmap_folder + "results.tsv", sep='\t').assign(filename=heatmap_folder))
+
+df = pd.concat(li, axis=0, ignore_index=True)
+print(df)
+
+for name in set(df["name"]):
+    if "dnd" not in name: continue
+    for mean in set(df["mean"]):
+        for log in set(df["log"]):
+            sub = df[(df["mean"] == mean) & (df["name"] == name) & (df["log"] == log)]
+            sub.to_csv("Analysis/{0}_mean{1}_log{2}.tsv".format(name, mean, log), index=False, sep="\t")

### 2020-06-16 scripts/scaling_merge.py

@@ -6,6 +6,7 @@ from glob import glob
 import numpy as np
 from collections import defaultdict
 from scipy.optimize import curve_fit
+from scipy.stats import gamma
 import statsmodels.api as sm
 import matplotlib
 
@@ -14,9 +15,11 @@ my_dpi = 128
 nbr_points = 1000
 matplotlib.rcParams['ytick.labelsize'] = label_size
 matplotlib.rcParams['xtick.labelsize'] = label_size
-matplotlib.rcParams["font.family"] = ["Latin Modern Mono"]
 matplotlib.use('Agg')
 import matplotlib.pyplot as plt
+from cycler import cycler
+
+plt.rc('axes', prop_cycle=cycler(color=["#5D80B4", "#E29D26", "#8FB03E", "#EB6231", "#857BA1"]))
 
 
 def is_float(x):
@@ -49,24 +52,59 @@ if __name__ == '__main__':
     args = parser.parse_args()
     dict_df = {filepath: {k: v[0] for k, v in pd.read_csv(filepath + ".tsv", sep='\t').items()} for filepath in
                sorted(glob("{0}/*_exp".format(args.input)))}
-    fig = plt.figure(figsize=(1920 / my_dpi, 1080 / my_dpi), dpi=my_dpi)
-    for col in ["DFE", "mut-ΔΔG"]:
+    params = pd.concat([pd.read_csv(f + ".parameters.tsv", sep='\t') for f in dict_df], axis=0, ignore_index=True)
+    nunique = params.apply(pd.Series.nunique)
+    params = params.drop(nunique[nunique != 1].index, axis=1).drop_duplicates()
+    if ("exon_size" in params) and ("gamma" in params):
+        params["chi"] = - 1 / (1.686 * params["exon_size"] * params["gamma"])
+    params.to_csv(args.output.replace(".tsv", ".parameters.tsv"), index=False, sep="\t")
+    fig = plt.figure(figsize=(1920 / (2 * my_dpi), 1080 / my_dpi), dpi=my_dpi)
+    for col, label in [("DFE", "S"), ("mut-ΔΔG", "$\\Delta \\Delta$G")]:
         if not args.distrib: continue
-        hist = defaultdict(float)
+        li = list()
         for filepath in dict_df.keys():
             distrib_file = filepath + col + "distrib.tsv"
             if not os.path.isfile(distrib_file): continue
-            for head, val in pd.read_csv(distrib_file, sep='\t').items():
-                try: head_f = float(head)
[…]

### 2020-06-16 scripts/simulated_experiment.py

@@ -4,11 +4,12 @@ import os
 from subprocess import run
 
 
-def create_experiment(experiment, screen, sbatch, nbr_cpu):
-    exp_path = os.getcwd() + '/Experiments/' + experiment
+def create_experiment(folder, experiment, screen, sbatch, nbr_cpu):
+    os.chdir(folder)
+    exp_path = os.getcwd() + '/Experiments/' + experiment.replace(".yaml", "")
 
     os.makedirs(exp_path, exist_ok=True)
-    os.system('cp config.yaml {0}'.format(exp_path))
+    os.system('cp {0} {1}/config.yaml'.format(experiment, exp_path))
     os.remove(exp_path + "/Snakefile") if os.path.exists(exp_path + "/Snakefile") else None
     os.symlink(os.getcwd() + "/Snakefile", exp_path + "/Snakefile")
     run_file = exp_path + "/snakeslurm.sh"
@@ -38,7 +39,8 @@ def create_experiment(experiment, screen, sbatch, nbr_cpu):
 
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument('-e', '--experiment', required=False, type=str, default="selcoeff", dest="experiment")
+    parser.add_argument('-f', '--folder', required=False, type=str, default="Heatmap", dest="folder")
+    parser.add_argument('-c', '--config', required=False, type=str, default="figure-4A.yaml", dest="experiment")
     parser.add_argument('-s', '--screen', required=False, type=bool, default=False, dest="screen")
     parser.add_argument('-b', '--sbatch', required=False, type=bool, default=False, dest="sbatch")
     parser.add_argument('-j', '--nbr_cpu', required=False, type=int, default=6, dest="nbr_cpu")
@@ -47,4 +49,4 @@ if __name__ == '__main__':
         import uuid
 
         args.experiment = str(uuid.uuid4())[:8]
-    create_experiment(args.experiment, args.screen, args.sbatch, args.nbr_cpu)
+    create_experiment(args.folder, args.experiment, args.screen, args.sbatch, args.nbr_cpu)

### 2020-06-16 scripts/theoretical_equilibrium.py

@@ -15,12 +15,13 @@ import matplotlib.pyplot as plt
 
 if __name__ == '__main__':
     parser = argparse.ArgumentParser(formatter_class=argparse.ArgumentDefaultsHelpFormatter)
-    parser.add_argument('-o', '--output', required=True, type=str, dest="output")
-    parser.add_argument('-g', '--exon_size', required=True, type=int, dest="n")
-    parser.add_argument('-p', '--population_size', required=True, type=float, dest="population_size")
-    parser.add_argument('-a', '--alpha', required=True, type=float, dest="alpha")
-    parser.add_argument('-k', '--gamma', required=True, type=float, dest="gamma")
-    parser.add_argument('-b', '--beta', required=True, type=float, dest="beta")
+    parser.add_argument('-o', '--output', default="theoretical_eq", type=str, dest="output")
+    parser.add_argument('--exon_size', default=300, type=int, dest="n")
+    parser.add_argument('--population_size', default=10000, type=float, dest="population_size")
+    parser.add_argument('--alpha', default=-118, type=float, dest="alpha")
+    parser.add_argument('--gamma', default=1.0, type=float, dest="gamma")
+    parser.add_argument('--beta', default=1.686, type=float, dest="beta")
+    parser.add_argument('--nbr_states', default=20, type=int, dest="nbr_states")
     args, unknown = parser.parse_known_args()
     dict_df = dict()
 
@@ -43,7 +44,7 @@ if __name__ == '__main__':
             return float("inf")
         elif x == 1.0:
             return -float("inf")
-        return np.log((1 - x) / x)
+        return np.log((1 - x) / x) + np.log(args.nbr_states - 1)
 
 
     def self_consistent_eq(x, alpha):
@@ -83,4 +84,4 @@ if __name__ == '__main__':
     plt.ylim((y_min, y_max))
     plt.tight_layout()
     plt.savefig("{0}.pdf".format(args.output), format="pdf", dpi=my_dpi)
-    pd.DataFrame(dict_df).to_csv(args.output, index=False, sep="\t")
+    pd.DataFrame(dict_df).to_csv("{0}.tsv".format(args.output), index=False, sep="\t")

### 2020-06-16 manuscript/figures.tex
Manuscript formatted to MBE guidelines

@@
 	width=\textwidth,
-	height=\axisdefaultheight,
+	height=\axisdefaultheight*0.75,
 	xlabel={$x$},
@@
 	]
-	\addplot[domain=0.1:0.39,line width=2.0pt, color=BLUE]{ ln(19*(1-x)/x) };
-	\addplot[name path=A, domain=0.1:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
+	\addplot[domain=0.2:0.50,line width=2.0pt, color=BLUE]{ ln(19*(1-x)/x) };
+	\addplot[name path=A, domain=0.2:0.38, line width=2.0pt, color=RED]{4*10000*1.686*exp(1.686*(-118+300*x))};
 	\addplot[name path=B, dashed,domain=0.3:0.37, line width=2.0pt, color=RED]{4*100000*1.686*exp(1.686*(-118+300*x))};
-	\addplot[name path=C,domain=0.1:0.39, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-15.71+30*x))};
-	\addplot[name path=D,dashed,domain=0.1:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-15.71+30*x))};
+	\addplot[name path=C,domain=0.2:0.383, line width=2.0pt, color=GREEN]{4*10000*0.1*1.686*exp(1.686*(-15.71+30*x))};
+	\addplot[name path=D,dashed,domain=0.2:0.35, line width=2.0pt, color=GREEN]{4*100000*0.1*1.686*exp(1.686*(-15.71+30*x))};
 	\addplot[fill=RED, opacity=0.2] fill between[ of = A and B];
@@
 height=\axisdefaultheight,
-ylabel={Fitness $f(x)$},
-xlabel={Phenotype $x$},
+ylabel={Fitness},
+xlabel={Phenotype},
 cycle list name=colors,
-domain=0:1,
+domain=0:0.4,
 samples=200,
@@
 ]
-\addplot[line width=2.0pt, color=BLUE]{ 1 / (1+exp(1.686*(-12 + 30 * x)))};
+\addplot[line width=2.0pt, color=BLUE]{ 1 / (1+exp(1.686*(-5 + 10 * x)))};
 \addplot[black]{1.0};

### 2020-06-16 manuscript/manuscript.tex
Manuscript formatted to MBE guidelines

@@
-\documentclass{article}
-\usepackage{url}
+\documentclass{MBE}%
+\usepackage[bookmarks]{hyperref}
+\hypersetup{
+	colorlinks,
+	citecolor=black,
+	filecolor=black,
+	linkcolor=black,
+	urlcolor=black
+}
 \usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
-\usepackage{adjustbox}
-\usepackage{float}
-\usepackage{caption}
-\usepackage{mdframed}
 \usepackage{lmodern}
@@
 \usepackage{xfrac, nicefrac}
-\usepackage{lmodern}
-\usepackage{enumitem}
-\usepackage[margin=60pt]{geometry}
 \pdfinclusioncopyfonts=1
-\captionsetup{width=0.85\textwidth}
-\renewcommand{\baselinestretch}{1.5}
+\include{notations}
 
-\include{notations}
+\jshort{mbe}
+\volname{}
+\jvolume{}
+\jvol{}
+\jissue{}
+\pubyear{}
+
+\mstype{Article}
+
 \begin{document}
-	\title{Substitution rate responses to changes in effective population size.}
-	\author{T. Latrille, N. Lartillot}
+
+\title[Substitution rate susceptibility] {Substitution rate response to changes in effective population size and protein expression level.}
+
+\author[Latrille \& Lartillot] {
+	T. \surname{Latrille}$^{\ast,1,2}$,
+	N. \surname{Lartillot}$^{1}$
+}
+
+\address{
+	$^{1}$Université de Lyon, Université Lyon 1, CNRS, Laboratoire de Biométrie et Biologie Évolutive UMR 5558, 69622 Villeurbanne, France.\\
+	$^{2}$École normale supérieure de Lyon, Université de Lyon, Université Lyon 1, Lyon, France}
+
+\coresp{thibault.latrille@ens-lyon.org or nicolas.lartillot@univ-lyon1.fr}
+
+\abstract{
+Molecular sequences are shaped by selection, where the strength of selection relative to drift is arbitrated by effective population size ($\Ne$).
+Populations with high $\Ne$ are expected to undergo stronger purifying selection, and consequently a decrease in substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$=dN/dS).
+However, computational models based on biophysics of protein stability have suggested that $\omega$ can also be independent of $\Ne$, proven under general conditions.
+Together, the response of $\omega$ to changes in $\Ne$ depends on the specific mapping from sequence to fitness.
+Importantly,  increase in protein expression level has been found empirically to result in decrease of $\omega$, predicted by theoretical results recruiting the selection for protein stability.
+In this present work we derive a theoretical approximation for the response of $\omega$ to changes in $\Ne$ and expression level, under an explicit genotype-phenotype-fitness map.
+The method is generally valid for additive traits and log-concave fitness functions.
+We applied the results to protein undergoing selection for their conformational stability,  corroborated by simulations under more complex models.
+We predict a weak response of $\omega$ to changes in either $\Ne$ or expression level, which are interchangeable.
+Based on empirical data, we propose that fitness based on the conformational stability is not a sufficient mechanism to explain the amplitude of $\omega$ variations observed.
+However, other aspects of protein biophysics might be explored such as protein-protein interactions, which can lead to stronger response of $\omega$ to changes in $\Ne$.
+}
+
+\keyword{protein stability, substitution rate, population-genetics, drift, expression level.}
+
 \maketitle
 
-	\abstract{
-		The quasi-neutral theory of evolution asserts that the effective population size ($\Ne$) plays an important role in shaping the evolution of molecular sequences.
-		One consequence is that $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
-		In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$) with respect to $\Ne$.
-		Such theoretical prediction had been observed in empirical data across many clades.
-		However, several studies failed to observe such response of $\omega$ to changes in $\Ne$, or with weak strength or direction.
-		Computational models of protein folding have observed that $\omega$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
-		Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\omega$ and then a decrease.
-		Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\omega$ responses to changes in $\Ne$.
-		Our goal in this present work is to provide theoretical tools to derive the relationship between $\Ne$ and $\omega$ in the context of a genotype-phenotype-fitness map.
-		We derive general formulas, and specify an application in the case of fitness proportional to the probability of protein folding.
-		Our compact theoretical results are supported by more complex simulations using 3d structure of proteins.
-		We assert that models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\omega$ relationship in population genetic dataset.
-		However, our framework applied to the case of non-specific interactions between proteins suits more the empirical data.
-		We also stress the importance of epistatic interactions in the $\Ne$-$\omega$ relationship, such that it determines to time to reach a new equilibrium.
-	}
-
 \section*{Introduction}
-
 Molecular sequences differ across species due to the particular history of DNA substitutions along their respective lineages.
@@
 This fundamental idea is at the core of the nearly-neutral theory of evolution.
-	The nearly-neutral theory posits that a substantial fraction of mutations are weakly deleterious.
-	As a result, the theory predicts that the substitution rate of selected mutations relative to the neutral substitution rate, called $\omega$, decreases along a lineage with higher $\Ne$ \cite{Ohta1972, Ohta1992}.
+This theory posits that a substantial fraction of substitutions are weakly deleterious, and as result predicts that the substitution rate of selected sites relative to the neutral substitution rate, called $\omega$, decreases along a lineage with higher $\Ne$ \citep{Ohta1972, Ohta1992}.
 
-	This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects (DFE) \cite{Kimura1979, Welch2008}.
+This prediction has been more quantitatively examined under the assumption that the selective effects of mutations are drawn from a fixed distribution of fitness effects (DFE) \citep{Kimura1979, Welch2008}.
 Assuming a gamma DFE, a key result obtained in this context is an approximate allometric scaling of $\omega$ as a function of $\Ne$ (i.e.
-$\omega \sim \Ne^{-\beta}$), where $\beta$ is the shape parameter of the DFE.
+$\omega \sim \Ne^{-k}$), where $k$ is the shape parameter of the DFE.
 In practice, DFEs are strongly leptokurtic, thus predicts a weak negative relation between $\omega$ and $\Ne$.
@@
 The substitution rate of selected mutations relative to neutral substitution rate ($\omega$) is an observable quantity of the molecular evolutionary process, at least if we can separately estimate the rate of neutral and selected mutations in a phylogenetic or comparative context.
-	Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while those selected mutations are the non-synonymous changes affecting the amino-acid sequence, $\omega$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
-	Thus, in practice, the nearly-neutral argument translates into a predicted decrease in $\omega = dN/dS$ as a function of long-term changes in $\Ne$.
+Practically, in the case of protein-coding DNA sequences, and assuming that synonymous mutations are neutral, while non-synonymous changes affecting the amino-acid sequence are under selection, $\omega$ can be identified with the ratio of non-synonymous over synonymous substitution rates ($dN/dS$).
+Thus the nearly-neutral argument translates into a predicted decrease in $\omega = dN/dS$ as a function of long-term changes in $\Ne$.
 
-	The context of protein coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
-	In this alternative approach, the selective effect of a mutation depends on the fitnesses of both source and target amino-acids involved on the mutation \cite{Halpern1998, Rodrigue2010, Tamuri2012}.
+The context of protein-coding sequences fostered another modeling approach, based on genotype-fitness map instead of distribution of fitness effects.
+In this alternative approach, the selective effect of a mutation depends on the fitness of both source and target amino-acids involved on the mutation \citep{Halpern1998, Rodrigue2010, Tamuri2012}.
 For example, if the target amino-acid has a higher fitness than the current amino-acid, the selective effect of the mutation is positive, and reciprocally negative for a lower fitness.
@@
 More fundamentally, the fitness depends solely on the current genotype, not on the trajectory of mutations leading to this sequence.
-	Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\omega$ and $\Ne$ \cite{Spielman2015a, DosReis2015}.
+Even though this modeling approach differs substantially from the one assuming a fixed DFE, it also predicts a negative correlation between $\omega$ and $\Ne$ \citep{Spielman2015a, DosReis2015}.
 
-	Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \cite{Yang2001, Zhang2004}. 
-	Combining estimations of $\omega$ along branches to life-history traits such as body mass or generation time revealed a positive correlation \cite{Popadin2007, Nikolaev2007}
-	Subsequently, integrative inference methods combining molecular sequences and lineage specific quantitative traits have also found that $\omega$ correlates positively with traits such as longevity and body mass \cite{Lartillot2011, Figuet2017}.
-	Since longevity and body mass typically correlate negatively with $\Ne$, with low $\Ne$ corresponding to lineage with a large body size and extended longevity \cite{Romiguier2014}, these empirical correlations suggest a negative correlation between $\omega$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
-	However, this correlation between $\omega$ and life-history traits could not be replicated across all experiments, and depending on the clade or the potential biases taken into account, the correlation was found to be either not statistically significant (\cite{Lartillot2012}), or even in the opposite direction \cite{Lanfear2010, Nabholz2013, Weber2014, Figuet2016}.
-
-	Mitigated empirical evidence about the response of $\omega$ to changes in $\Ne$ encouraged the search for alternative explanatory mechanisms \cite{Lanfear2014}.
-	In this direction, one striking result was the proof that $\omega$ is in fact predicted to be independent of $\Ne$ under very general circumstances, namely, whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
-	Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \cite{Cherry1998}.
-	This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \cite{Goldstein2013}.
-	The thermodynamic stability is itself computed using a 3D structural model of the protein. These simulation experiments have led to the observation that $\omega$ is essentially independent of $\Ne$.
+Conversely, one striking theoretical result was the proof that $\omega$ is in fact predicted to be independent of $\Ne$ under very general circumstances, namely, whenever (i) the fitness is a log-concave function of a phenotype and (ii) the phenotype itself is equimutable.
+Equimutability states that the distribution of phenotypic changes due to mutations is independent the current phenotype of the individual \citep{Cherry1998}.
+This general theoretical argument has been invoked in the context of \textit{in silico} experiments of protein sequence evolution, assuming that proteins are under selection for their thermodynamic stability, the fitness being proportional to the folding probability of the protein \citep{Goldstein2013}.
+The thermodynamic stability is itself computed using a 3D structural model of the protein.
+These computational experiments have led to the observation that $\omega$ is essentially independent of $\Ne$.
 The explanation proposed for this result is that the distribution of changes in free energy of folding ($\deltadeltaG$) due to mutations is approximately independent of the current free energy ($\deltaG$), thus making the free energy of folding essentially equimutable.
 
-	However, the equimutability assumption is a relatively strong one, which also conflicts with simple combinatorial considerations about the relation between sequence and phenotype.
+However, the equimutability assumption is a relatively strong one, which also conflicts with combinatorial considerations about the relation between sequence and phenotype \citep{Serohijos2012}.
 For example, if a protein sequence is already maximally stable, only destabilizing (or neutral) mutations can occur.
@@
 
-	In addition, if empirical evidence for a negative correlation of $\omega$ with $\Ne$ is still not totally convincing, another empirical correlation is known to be much more robust, namely, that of $\omega$ with expression level or protein abundance \cite{Duret2000, Rocha2004, Wang2011, Song2017}.
-Indeed, expression level is one of the best predictors of $\omega$, with highly expressed proteins typically having lower $\omega$ values.
-Of note, the slope of the response of $\omega$ to changes in expression level is relatively weak \cite{Drummond2005}, although clearly significant.
-Importantly, many of the theoretical models of protein-coding sequence evolution that have been invoked to explain this correlation between $\omega$ and expression level are such that the response of $\omega$ to changes in expression level or in $\Ne$ should be very similar.
-Conversely, under strict equimutability, most of these models would predict that $\omega$ should be essentially independent, not just of $\Ne$, but also of expression level, thus at odds with empirical observations.
+Together, depending on the theoretical model mapping sequence to fitness, $\omega$ can be either independent or negatively correlated to $\Ne$, or even positively if considering adaptive evolution and environmental changes \citep{Lanfear2014}.
+Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred using phylogenetic codon models applied to empirical sequences \citep{Yang2001, Zhang2004}. 
+Combining estimations of $\omega$ along branches to life-history traits such as body mass or generation time uncovered a positive correlation \citep{Popadin2007, Nikolaev2007}.
+Subsequently, integrative inference methods combining molecular sequences and lineage specific quantitative traits have also found that $\omega$ correlates positively with traits such as longevity and body mass \citep{Lartillot2011, Figuet2017}.
+Since lineage with a large body size and extended longevity typically correspond to low $\Ne$ \citep{Romiguier2014}, these empirical correlations suggest a negative correlation between $\omega$ and $\Ne$, thus confirming the theoretical prediction of the nearly-neutral theory of evolution.
+However, the universality and robustness of the correlation between $\omega$ and life-history traits is still debated. 
+Results have been not entirely consistent across independent studies, the correlation was found to be either not statistically significant \citep{Lartillot2012}, or even in the opposite direction depending on a specific clade or the potential biases taken into account \citep{Lanfear2010, Nabholz2013, Lanfear2014, Figuet2016}.
 
-	Overall, there is no doubt that the impact of changes in $\Ne$ (or in expression level) on $\omega$ is at best weak, making empirical estimation more difficult.
-	Even if weak, however, an exact theoretical understanding and quantification of how the slope of this relation depends on the underlying map between genotype, phenotype and fitness, would be useful.
-	Ultimately, relating proteins structural parameters to the $\omega$-$\Ne$ relationship reveal an intersection between disparate empirical dataset,  coming from protein thermodynamic on one side and comparative genomic on the other side.
+In addition, if empirical evidence for a negative correlation of $\omega$ with $\Ne$ is still not totally convincing, another empirical correlation is known to be much more robust.
+Indeed, expression level or protein abundance is one of the best predictors of $\omega$, with highly expressed proteins typically having lower $\omega$ values, a correlation clearly significant although relatively weak \citep{Duret2000, Rocha2004, Drummond2005a, Zhang2015, Song2017}.
+Importantly, theoretical models also based on protein stability have been invoked to explain this negative correlation between $\omega$ and expression level \citep{Wilke2006, Drummond2008}, such that selection against protein misfolding -due to toxicity- induces abundant proteins to evolve to greater stability, where the protein is more constrained and evolve slowly \citep{Serohijos2012}.
+
+Indication that expression level and $\Ne$ play a similar roles on the evolution of protein have been noticed.
+Precisely, under models of selection against protein misfolding, the free energy of folding $\deltaG$ varies similarly along a gradient of either $\Ne$ or expression level \citep{Serohijos2013}.
+Additionally, under strict equimutability of $\deltaG$, computational models observed that $\omega$ is also independent of expression levels \citep{Serohijos2012}, similar to the phenomenon observed w.r.t $\Ne$.
+However, the theoretical response of $\omega$ w.r.t to both $\Ne$ and expression level has not been quantified, and most importantly has not been related to the specific map between genotype, phenotype and fitness.
+This analytical development would be useful to confront empirical data, relating $\omega$ to both $\Ne$ and expression level.
+Ultimately, relating proteins structural parameters to response of $\omega$ reveals an articulation between empirical data from protein thermodynamic on one side and comparative genomics on the other side.
 
 Lastly, the theoretical results discussed so far are all valid only when the balance between mutation, selection and drift is at equilibrium.
-However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\omega$ due to adaptive selection, and subsequently a decrease in $\omega$ due to stronger purifying selection in the long term \cite{Jones2016}.
+However, under a model of site-dependent genotype-fitness map, an increase in $\Ne$ first leads to an increase of $\omega$ due to adaptive selection, and subsequently a decrease in $\omega$ due to stronger purifying selection in the long term \citep{Jones2016}.
 Studying only equilibrium properties can thus be misleading.
-For that reason, the dynamical response of $\omega$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized.
-	Dynamical properties of the $\omega$ response to changes of $\Ne$ are of theoretical interest, but are also empirically relevant such that if overlook they could thwart the relation between theoretical expectation and empirical estimates.
+For that reason, the dynamic response of $\omega$ to changes in $\Ne$ must also be addressed, quantified, and its connection with the underlying selective landscape better characterized.
+Dynamic properties of the $\omega$ response to changes of $\Ne$ are of theoretical interest, but are also empirically relevant such that if overlook they could thwart the relation between theoretical expectation and empirical estimates.
+% Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\omega$ relationship to $\Ne$ is unknown.
 
-	% Under the assumption that fitness relates to protein stability, and that such stability if a function of the genotype, the theoretical $\omega$ relationship to $\Ne$ is unknown.
-	The goal of this study is to characterize the dynamical and equilibrium response of $\omega$ to changes in $\Ne$, and relate this response to structural parameters of the model.
-	To this end, we develop a general mathematical approach for deriving the $\omega$ relationship to $\Ne$ (and expression level) in the context of a given genotype-phenotype-fitness map.
-	Based on these results, we discuss the consequences of the assumption that protein stability is a proxy of fitness, and the signatures persisting in DNA sequences observable under such assumption.
-	In the light of previously published empirical estimates from protein thermodynamic, and from comparative genomics, we discuss the articulation between empirical data and our mechanistic model.
-	We also discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the slope of the relation between $\omega$ and $\Ne$.
+This study aims to characterize the dynamic and equilibrium response of $\omega$ to changes in $\Ne$ and expression level, and relate this response to structural parameters of the model.
+We develop a general mathematical approach to derive the $\omega$ response to $\Ne$ and expression level in the context of a given genotype-phenotype-fitness map.
+In the light of previously published empirical estimates from protein thermodynamic and comparative genomics, we discuss the articulation between empirical data and our mechanistic model.
+We also discuss some of the alternative biophysical mechanisms that could determine the selective landscape on protein-coding sequences, and how they would modulate the response of $\omega$ to changes in $\Ne$ and expression level.
 
@@
 
-	The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$ (i.e.
-such that the coding positions of the sequence contribute additively to the trait), under directional selection specified by a decreasing and log-concave fitness function $f(x)$.
-As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability (See figure \ref{fig:NeChangeInfluence}).
-This model is inspired from previous work \cite{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations.
+The results that are presented below are valid for a general category of models of sequence evolution, based on an additive trait $x$, such that the coding positions of the sequence contribute additively to the trait.
+The trait is under directional selection specified by a decreasing and log-concave fitness function $f(x)$.
+As a specific example, we more specifically consider a model of protein evolution under the constraint of thermodynamic stability (See figure \ref{fig:Summary}, left panel).
+This model is inspired from previous work \citep{Williams2006, Goldstein2011, Pollock2012}, except that we make several simplifying assumptions, allowing us to derive analytical equations.
 %Throughout our derivation, we present the results both for the general case and for the more specific example.
 
+\begin{figure*}[htb!]
+\begin{center}
+\includegraphics[width=\textwidth, page=1] {artworks/summary.pdf}
+\end{center}
+ \caption{
+ \textbf{Outline of the theoretical results.}
+ Left panel.
+The complete state of the amino-acid sequence is the genotype.
+The phenotype ($x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of destabilizing sites in the sequence.
+The fitness is a decreasing log-concave function of the phenotype, depending on structural parameters of the model ($\Nsites, \DeltaGmin, \ \gamma, \ \beta$).
+ Middle panel.
+Once defined the relation from genotype to fitness at the individual level, the evolutionary process consists of point substitutions of the amino-acid representative sequence of the population.
+Substitutions are mutation reaching fixation, thus the rate of substitution depends mechanistically on the interplay between mutation, selection and genetic drift.
+ For a given effective population size ($\Ne$), the evolutionary process result in an average value of the phenotype ($\smash{x\eq}$) and an average relative substitution rate of selected mutation ($\omega$).
+Averaging over time is equivalent to determining the statistical equilibrium, by ergodicity of the stochastic process, given we can derive the probabilities of transitions between phenotypes.
+ Right panel. The scaling of equilibrium $\omega$ under a range of $\Ne$ give us the susceptibility ($\chi$), as a function of the structural parameters defined in the phenotype-fitness map.
+ }
+ \label{fig:Summary}
+\end{figure*}
 In the original biophysical model, the protein stability is the difference in free energy between folded and unfolded conformations, called $\deltaG$ in kcal/mol.
-	Technically, the free energy of the folded are unfolded conformations are computed based on the $3$D conformations of the protein, and using a statistical potential.
-	In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation.
+Technically, the free energy of the folded and unfolded conformations are computed based on the $3$D conformations of the protein and statistical potentials.
+In such models, the stabilizing or destabilizing effect of an amino-acid at a particular site depends on the amino-acids present in the vicinity in $3$D conformation, taking into account specific epistasis \citep{Dasmeh2018}.
 
-	We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighbouring residues.
+We approximate this model such that the destabilizing effect of an amino-acid does not depend on other neighboring residues, disregarding specific epistasis \citep{Dasmeh2014}.
 Instead, each site contributes independently to $\deltaG$. 
 With such approximation, for each site of the sequence only one amino-acid is stabilizing the protein. 
-	All $19$ other amino-acids are equally destabilizing, and each site contribute to an excess of $\gamma > 0$ (in kcal/mol) to $\deltaG$ per destabilizing amino-acids.
-	If all amino-acids of the sequence are stabilizing, $\deltaG$ is optimal and equal to $ \alpha < 0$. 
+All $19$ other amino-acids are equally destabilizing, and each site contribute to an excess of $\gamma = \Delta \Delta G > 0$ (in kcal/mol) to $\deltaG$ per destabilizing amino-acids.
+If all amino-acids of the sequence are stabilizing, $\deltaG$ is optimal and equal to $ \DeltaGmin = \Delta G_{\text{min}} < 0$. 
 In this model, the most succinct phenotype of a given genotype (i.e.
 sequence) is just the proportion of destabilizing amino-acids in the sequence, defined as $0 \leq x \leq 1$.
-Thus, $\deltaG$ is a linear function of $x$:
+Thus $\deltaG$ is a linear function of $x$:
 \begin{align}
-		\deltaG (x) = \alpha + \Nsite \gamma x,
+ \deltaG (x) = \DeltaGmin + \Nsite \gamma x,
 \end{align}
@@
 For a given $\deltaG$, thermodynamic equations allows to derive the proportion of protein in folded conformation in the cytoplasm, which is assumed to be a proxy for fitness.
-	This fitness function is motivated in part that a protein must be folded to perform its function, but can also be justified by the toxic effect of misfolded proteins in the cytoplasm (supp information).
-	Analytically, this fitness by the Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation\cite{Goldstein2011}: 
+This fitness function is motivated in part that a protein must be folded to perform its function, however this a strong assumption which will be relaxed subsequently to take into account protein expression level.
+Analytically, this fitness is given by Fermi Dirac distribution and is typically close to $1$, leading to a first-order approximation \citep{Goldstein2011}: 
 \begin{align}
-		f(x) = \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma x)}},\\
-		\Rightarrow f(x) \simeq 1 - e^{\beta (\alpha + \Nsite \gamma x)}, 
+ f(x) = \dfrac{1}{1 + e^{\beta (\DeltaGmin + \Nsite \gamma x)}},\\
+ \Rightarrow f(x) \simeq 1 - e^{\beta (\DeltaGmin + \Nsite \gamma x)}, 
 \end{align}
-	where $\alpha$ and $\gamma$ are defined as above, and the parameter $\beta$ is $1.686$ mol/kcal at room temperature.
+where $\DeltaGmin$ and $\gamma$ are defined as above, and the parameter $\beta$ is $1.686$ mol/kcal at room temperature.
 
@@
 In this section we present an analytical approximate solution for the response of equilibrium $\omega$ after a change in $\Ne$ (in log space).
-We call this reponse the susceptibility of $\omega$ to changes in $\Ne$, and denote it as $\chi$:
+We call this response the susceptibility of $\omega$ to changes in $\Ne$, and denote it as $\chi$:
 \begin{align}
-		\chi = \frac{ \der \omega}{\der \ln (\Ne)}
+ \chi = \frac{ \der \omega}{\der \ln (\Ne)} \label{eq:chi}
 \end{align}
-	Broadly, deriving $\chi$ is done in two steps.
-	The first step is to determine the phenotype at equilibrium, when evolutionnary forces of mutation, selection and genetic drift compensate each others.
+Broadly, deriving $\chi$ is done in two steps (See figure \ref{fig:Summary}).
+First, we determine the phenotype at equilibrium, when evolutionary forces of mutation, selection and genetic drift compensate each other.
 Subsequently, differential calculus on the response of the equilibrium phenotype to a change in $\Ne$ allows to ultimately derive an equation for $\chi$.
-	The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case where the fitness is proportional to the total amount of folded protein in the specific case of the biophysical model.
-	All development are available in Supplementary Materials in the most general case.
+The main results are given in the general case of any phenotype-fitness map, and also derived in the specific case of the biophysical model, where the fitness is proportional to the total amount of folded protein.
+All developments are available in supplementary materials in the most general case.
 
-	First, from a given genotype, mutations changing the sequence have various effect, they can increase or decrease the proportion of destabilizing amino-acid, or do nothing if the mutation is between both destabilizing amino-acids.
-	To derive the probabilities of such events to occur, we also make the simplyfing assumption that the transition between amino-acids are equiprobable.
+From a given genotype, mutations changing the sequence have various effects, they can increase or decrease the proportion of destabilizing amino-acid, or do nothing if the mutation is between two destabilizing amino-acids.
+To derive the probabilities of such events to occur, we also make the simplifying assumption that all transitions between amino-acids are equiprobable.
 Altogether, any mutation in the sequence can then have a phenotypic effect of $0$ or $\dx=\sfrac{1}{\Nsite}$, and the probabilities of transition are:
@@
 \end{gather}
-	In the extreme case of an optimal phenotype ($x = 0$), only destabilization mutations are proposed.
-	Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or between destabilizing amino-acids (effect $0$) is proportional to $x$. 
+In the extreme case of an optimal phenotype ($x = 0$), only destabilizing mutations are proposed.
+Moreover, the probability to propose stabilizing mutation (effect $-\dx$), or in between destabilizing amino-acids (effect $0$) is proportional to $x$. 
 Or in other words, the mutation bias is proportional to $(1-x)/x$, which this is fundamentally a combinatorial effect, due to the number of mutational opportunities available in either direction.
@@
  s & \simeq \frac{1}{\Nsite}\frac{ \partial \ln f(x) }{\partial x} \label{eq:s} \\
-		\Rightarrow s & \simeq - \beta \gamma e^{\beta (\alpha + \Nsite \gamma x)}. \label{eq:s-unfolded}
+ \Rightarrow s & \simeq - \beta \gamma e^{\beta (\DeltaGmin + \Nsite \gamma x)}. \label{eq:s-unfolded}
 \end{align}
 Conversely, stabilizing mutations will be under positive selection with opposite sign but same absolute value.
-	It is important to realize that the selective effect is dependent on $x$, and that because fitnesses are log-concave, the absolute value of $s$ increases with regards to $x$.
+It is important to realize that the selective effect is dependent on $x$, and that because fitness are log-concave, the absolute value of $s$ increases with regards to $x$.
 
-	Based on these expressions for mutational and selective pressure, one can study the evolutionary process unfurls.
-	Starting from an optimal sequence, mostly destabilizing mutations will occur, which will reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, at which point the protein will reach a point of equilibrium called marginal stability.
+Based on these expressions for mutational and selective pressure, one can study the evolutionary process trajectory.
+Starting from an optimal sequence, mostly destabilizing mutations will occur, but they may reach fixation and accumulate until the selection coefficient against new deleterious mutations is too strong, at which point the protein will reach a point of equilibrium called marginal stability \citep{Taverna2002, Bloom2007}.
 Most importantly the probability of fixation of mutations is affected by genetic drift, parameterized by effective population size ($\Ne$).
-	Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average \cite{Goldstein2013}.
-	Formally, and after simplification, the equilibrium phenotype denoted $x\eq$ is given by the equation:
+Altogether, at the equilibrium between mutation, selection and drift, the selection coefficient of new advantageous and deleterious mutations reaching fixations are expected to be null on average \citep{Goldstein2013}.
+Formally, and after simplification, the equilibrium phenotype denoted $x\eq$ is given in the general case by the equation :
 \begin{align}
-		\ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(x\eq) }{\partial {x\eq}} \text{ in the general case,} \\
-		\Rightarrow \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)} \text{,} \label{eq:equilibrium}
+ \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(x\eq) }{\partial {x\eq}} \\
+ \Rightarrow \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) & \simeq 4\Ne \beta \gamma e^{\beta (\DeltaGmin + \Nsite \gamma x\eq)} \text{,} \label{eq:equilibrium}
 \end{align}
 in the more specific case of the biophysical model.
-This equation essentially expresses the mutation-selection equilibrium: the left-hand side of the equation is the log of the mutation bias at $x$, while the right-hand side is simply $4 \Ne s$.
+This equation essentially expresses the mutation-selection equilibrium: the left-hand side of the equation is the log of the mutation bias at $x$, while the right-hand side is simply $4 \Ne s$, the scaled selection coefficient.
 
-	This equation cannot be solved explicitly for $x\eq$, but an intuition on the consequences of change in $\Ne$ to the equilibirum phenotype $x\eq$ is given in figure \ref{fig:NeChangeInfluence}B.
-	Mutation bias decreases weakly with $x$ (blue curve on figure) while the strength of selection increases sharply with $x$.
-	The equilibrium phenotype is obtain at their intersection (eq \ref{eq:equilibrium}).
-	An increase in $\Ne$ leads to shifting the selective response upward, which then results in a leftward shift of the equilbrium phenotype. 
-	The amount of change on $x\eq$, however, is very small if the selection coefficient curve is very steep around the equilibrium set point.
-	Of note, the phenotype is not equimutable, meaning the mutational bias is not constant, but nearly so since the derivative of the mutational bias is greatly lower than the derivative of the selection coefficient around the equilibrium point.
+This equation cannot be solved explicitly for $x\eq$, but a qualitative intuition on the consequences of change in $\Ne$ to the equilibrium phenotype $x\eq$ is given in figure \ref{fig:NeChangeInfluence}.
+Intuitively an increase in $\Ne$ results in a more optimal phenotype, closer to $0$.
+Mutation bias (left-hand side of equation \ref{eq:equilibrium}) decreases with $x$ while the strength of selection (right-hand side of equation \ref{eq:equilibrium}) increases with $x$, and the equilibrium phenotype is obtained at their intersection.
+An increase in $\Ne$ (green dotted curve) leads to shifting the selective response upward, which then results in a leftward shift of the equilibrium phenotype, closer to $0$.
+The leftward shift is smaller for selective strength with steeper curve, resulting in qualitatively weaker susceptibility of the equilibrium phenotype to changes in $\Ne$
 
 \begin{figure*}[htb!]
-		\begin{mdframed}
-			\centering
-			\includegraphics[width=0.9\textwidth, page=1] {artworks/theoretical.pdf}
+ \begin{center}
+ 	 \includegraphics[width=\textwidth, page=1] {artworks/theoretical.pdf}
+ \end{center}
  \caption{
-				\textbf{Theoretical model.}
-				Panel A.
-Illustration of the Genotype-phenotype-fitness map.
-The phenotype ($x$) is a real-valued summary of the genotype, and is defined in our model as the fraction of destabilizing sites in the sequence.
-The fitness is a decreasing log-concave function of the phenotype.
-				Panel B.
-Illustration of the $\omega$ susceptibility after a change in $\Ne$.
-The equilibrium $x\eq$ is determined by the equation $\ln(\frac{1 - x\eq}{x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)}$.
-The right-hand side of the equation increases exponentially with $x$ where $\Nsite \gamma$ is the exponential growth rate.
-Thus when $\Nsite \gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only changes subtly $x\eq$ (x-axis of the crossing with the blue solid line).
-On the other hand, when $\Nsite \gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) involve a stronger response of $x\eq$.
-Moreover, changes in $x\eq$ reflects the changes in $\omega$ since both are related by a monotonous function (equation \ref{eq:dnds}).
-The value of $\alpha$ has been chosen given all other parameters such that the solid lines all cross at the same point.
+ \textbf{Phenotype susceptibility after a change in $\bm{\Ne}$}
+ The equilibrium phenotype $x\eq$ is determined by the equation $\smash{\ln(\frac{1 - x\eq}{x\eq}) + \ln(19)=4\Ne \beta \gamma e^{\beta (\DeltaGmin + \Nsite \gamma x\eq)}}$.
+The right-hand side of the equation increases exponentially with $x$ where $\Nsite \gamma$ is the exponential growth rate ({\color{RED}red} and {\color{GREEN}green} curves).
+ The parameter $\gamma$ is chosen such that $\gamma$ is increased by an order of magnitude in the red compared to the green curve, moreover, the value of $\DeltaGmin$ is chosen such that the equilibrium value $\smash{x\eq}$ is kept constant by solving numerically equation \ref{eq:equilibrium}.
+ Whenever $\Nsite \gamma$ is large ({\color{RED}red solid} curve), the exponential growth is steep and increasing $\Ne$ (dotted curve) only changes subtly the equilibrium phenotype by a factor $\smash{\color{RED}\Delta x\eq}$.
+On the other hand, whenever $\Nsite \gamma$ is low ({\color{GREEN}green} solid curve), the exponential growth is steadier and increasing $\Ne$ (dotted curve) involve a stronger susceptibility of the equilibrium phenotype by a factor $\smash{\color{GREEN}\Delta x\eq}$.
+Moreover, changes in $\smash{x\eq}$ reflects the changes in $\omega$ since both are approximately equal (equation \ref{eq:dnds}).
  }
  \label{fig:NeChangeInfluence}
-		\end{mdframed}
 \end{figure*}
 
-	The argument thus far explains how $x\eq$ varies with $\Ne$.
+The results thus far ties only the equilibrium phenotype ($x\eq$) to $\Ne$.
 To capture how $\omega$ varies with $\Ne$, we also need to obtain an expression for $\omega$ as a function of $x\eq$. 
-	At equilibrium we can derive the expected substitution rate of mutations changing the amino-acid sequence, and by normalizying by the substitution rate of neutral mutations, $\omega$ simply approximates to:
+At equilibrium we can derive (supplementary materials) the expected substitution rate of mutations changing the amino-acid sequence, and by normalizing by the substitution rate of neutral mutations, $\omega$ simply approximates to:
 \begin{gather}
@@
 \end{gather}
-	This simple approximation is due to substitutions at destabilizing sites (proportion $x\eq$) to one of the other destabilizing amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (see equation \ref{eq:proba}).
+This simple approximation is due to substitutions at destabilizing sites (proportion $x\eq$) to one of the other destabilizing amino-acids, which are effectively neutral and compose the largest proportion of proposed mutations with substantial probability of fixation (equation \ref{eq:proba}).
 
-	Finally, $x\eq$ at equilibrium (eq \ref{eq:equilibrium}) is implicitly a function of $\Ne$ and the other parameters.
-	Implicit derivation of the equilibrium with regards to $\Ne$ (see supp info) allows the derive the response of $x\eq$ and ultimately $\omega$ after a change in $\Ne$.
-	After simplifications, the susceptibility of $\omega$ to changes in $\Ne$ can be approximated in terms of a very compact equation for the general case: 
+Taken together, implicit derivation of the equilibrium with regards to $\Ne$ yields the susceptibility (equation \ref{eq:chi}) of $\omega$ to a change in $\Ne$:
 \begin{gather}
-		\chi = \frac{ \der \omega}{\der \ln (\Ne)} \simeq - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}
+ \chi = \frac{ \der \omega}{\der \ln (\Ne)} \simeq - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{\Nsite}{4 \Ne} \frac{\partial \ln [ (1 - x\eq) / x\eq ]}{\partial x\eq } + \frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}.
 \end{gather}
-	The susceptibility is thus equal to the inverse of the relative curvature, i.e.
-the ratio of the second to the first derivatives, of the log-fitness function, taken at the equilibrium phenotype.
+The two terms of the denominator correspond to the derivative of respectively the mutational bias and scaled selection coefficient.
+However, mutational bias decreases weakly with $x$ (blue curve on figure \ref{fig:NeChangeInfluence}) while the strength of selection increases sharply with $x$ (red and green curves).
+As a consequence, the phenotype is not equimutable, meaning the mutational bias is not constant, but nearly so since the derivative of the mutational bias is greatly lower than the derivative of the selection coefficient around the equilibrium point.
+This simplification lead to a very compact equation for susceptibility $\chi$ in the general case: 
+\begin{gather}
+\chi \simeq - \frac{\frac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\frac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}
+\end{gather}
+The susceptibility is thus equal to the inverse of the relative curvature, i.e. the ratio of the second to the first derivatives, of the log-fitness function, taken at the equilibrium phenotype.
 Of note, this susceptibility is strictly negative for decreasing log-concave fitness functions, asserting that $\omega$ is a decreasing function of $\Ne$.
 In addition, the susceptibility itself is low (i.e.
-$\omega$ responds more weakly) for strongly concave log-fitness functions, thus formally capturing the intuitive argument developed above (figure \ref{fig:NeChangeInfluence}B).
+$\omega$ responds more weakly) for strongly concave log-fitness functions, thus quantitatively capturing the intuition developed in figure \ref{fig:NeChangeInfluence}, explicitly that the amount of change on $\omega$ is very small if the selection coefficient curve is very steep around the equilibrium set point (red curve compared to green curve).
 
-	In the specific case of the biophysical model, the susceptibility ($\chi$) simplify to:
+In the specific case of the biophysical model, the susceptibility ($\chi$) further simplifies to:
 \begin{gather}
-		\chi \simeq -\dfrac{1}{\beta \Nsite \gamma}, 
+ \chi \simeq -\dfrac{1}{\beta \Nsite \gamma}, \label{eq:chi_application}
 \end{gather}
-	Meaning that $\omega$ is linearly decreasing with $\Ne$ in log scale, since $\chi$ is independent of $x\eq$ and thus the exact value of the equilibrium phenotype has no impact on the slope.
+Meaning that $\omega$ is linearly decreasing with $\Ne$ (in log scale) since $\chi$ is independent of $x\eq$, or, in other words, the exact value of the equilibrium phenotype has no impact on the slope.
 Moreover, only the compound parameter $\beta \gamma \Nsite$ has an impact on the slope of the linear relationship.
-	In practice, $\gamma = \Delta \Delta G$ determine the slope of the linear relationship between $\omega$ and $\Ne$, while $\alpha = \Delta G_{\text{min}}$ can never obtained solely from this slope.
-	Importantly, only empirical relative values of $\Ne$ (multiplicative) and $\omega$ (additive) are sufficient to estimate $\chi$, absolute values are not necessary to estimate the slope.
-	Conversely, if one seek to determines $\Delta G_{\text{min}}$, one need the intercept of the linear relationship, and thus absolute value of $\omega$ and $\Ne$ are necessary.
+In practice, the structural and molecular parameters of the model are $\DeltaGmin = \Delta G_{\text{min}}$ and $\gamma = \Delta \Delta G$ such that the slope of the linear relationship between $\omega$ and $\Ne$ is affected by $\Delta \Delta G$ but not by $\Delta G_{\text{min}}$.
+Importantly, only relative values of $\Ne$ (multiplicative) and $\omega$ (additive) are sufficient to estimate empirical value $\chi$, absolute values are not necessary.
 % Although the general equation seems to suggest in appearance that elasticity only depends on the fitness function, in fact, what the biophysical model shows is that the elasticity can actually be modulated in two ways: by playing on the genotype / phenotype relationship (here, $\gamma$), or by playing on the fitness phenotype relationship (here, beta).
 All the more reason to consider beta as a potentially free parameter too.
+\subsection*{Susceptibility of $\omega$ to changes in protein expression.}
 
-	Other parametrization of the fitness function shed light on the robustness of the theoretical results.
-	If the selective coefficient itself is not equal (equation \ref{eq:s-unfolded}) but proportional to the total amount of misfolded protein, and the expression level of the protein is denoted $A$, then the response of $\omega$ to changes in $\Ne$ is the same as the response to changes in $A$ (supplementary materials):
+\label{sec:expression}
+
+Effective population size is not the sole predictor of $\omega$, and expression level (or protein abundance) is also negatively correlated to $\omega$. 
+The previous model, however, does not express the fact that selection is typically stronger for proteins characterized by higher levels of expression.
+The previous general equations can be directly applied when we assume that each misfolded protein has the same relative effect on fitness \citep{Drummond2005a, Wilke2006, Drummond2008, Serohijos2012}.
+For a given protein with expression level $y$ and a cost $c$ per misfolded macromolecule, the fitness and selective coefficient are defined as follow, leading to the mutation-selection-drift equilibrium:
+\begin{gather}
+\ln (f(x)) \simeq - c y e^{\beta (\DeltaGmin + \Nsite \gamma x)} \\
+\Rightarrow s  \simeq - \beta \gamma c y e^{\beta(\DeltaGmin + \gamma \Nsite x)} \\
+\Rightarrow \ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) = 4\Ne y A \beta \gamma \Nsite e^{\beta(\alpha + \gamma \Nsite x\eq)}
+\end{gather}
+Importantly, $\Ne$ and $y$ are confounded factors in the mutation-selection-drift equilibrium equation, both appears only once and they are multiplied.
+This means that increasing either $\Ne$ or $y$ leads to same change in equilibrium phenotype, hence the same $\omega$, and ultimately the same susceptibility:
 \begin{align}
-	s & \propto - \beta \gamma y e^{\beta(\alpha + \gamma \Nsite x)} \\
-	\Rightarrow \chi \frac{ \der \omega}{\der \ln (\Ne)} = \frac{ \der \omega}{\der \ln (A)} & \simeq -\dfrac{1}{\beta \Nsite \gamma}.
+\chi = \frac{ \der \omega\eq}{\der \ln (\Ne)} = \frac{ \der \omega\eq}{\der \ln (y)} \simeq - \frac{1}{\beta \Nsite \gamma}. \label{eq:chi_application_y}
 \end{align}
-	Meaning we should observe the same relationship between $\omega$ and either effective population size ($\Ne$) or expression level ($A$).
-
-	Under empirically relevant value of $\Nsite=300$ sites and $\gamma=1.0$ kcal/mol \cite{Zeldovich2007}, the susceptibility $\chi \simeq -0.002$.
-	In other words, for a increase in $\Ne$ of $6$ orders of magnitude, $\omega$ decrease approximately of $0.01$, a subtle relationship that requires laborious effort to be detected in simulated data, and large amount of empirical data with few bias and noise to be detectable.
-In addition, it would imply an equally weak relation between $\omega$ and expression level.
+Equation \ref{eq:chi_application_y} can also be proven under more general hypothesis relating phenotype to fitness, for example if the selective cost is due to translational errors.
+If the protein is assumed to be regulated such as to reach a specific level of functional proteins abundance under a general cost-benefit argument \citep{Cherry2010,Gout2010}, a multiplicative factor depending solely on the expression level is prefixed (see supplementary materials).
+Altogether, we theoretically obtain the same linear decrease of $\omega$ w.r.t either effective population size or expression level (in log space) under a broad variety of hypothesis.
 
@@
 
-	Our theoretical derivation of the susceptibility of $\omega$ to changes in $\Ne$ is based on several simplifying assumptions about the evolutionary model and makes multiple approximations.
+Our theoretical derivation of the susceptibility of $\omega$ to changes in $\Ne$ (and expression level) is based on several simplifying assumptions about the evolutionary model and makes multiple approximations.
 In order to test the robustness of our main result, we therefore conducted systematic simulation experiments, relaxing several of these assumptions.
@@
 
-	Specifically, with respect to mutations, our derivation assumes that all amino-acids transition are equiprobable, in other word the complexity of the genetic code is not taken into account.
-	Simulating evolution of DNA sequence, and taking into account a matrix a mutation rate between nucleotide allow to test directly the robustness of the results to this assumption.
-	Furthermore, with regards to the phenotypic effects of amino-acid changes, we assumed that all destabilizing amino-acids are have an identical impact on protein stability.
+Specifically, with respect to mutations, our derivation assumes that all amino-acid transitions are equiprobable, in other words, the complexity of the genetic code is not taken into account.
+Simulating evolution of DNA sequence, and taking into account a matrix a mutation rate between nucleotide allows to test directly the robustness of the results to this assumption.
+Furthermore, with regard to the phenotypic effects of amino-acid changes, we previously assumed that all destabilizing amino-acids are having an identical impact on protein stability.
 In reality, one would expect conservative amino-acid replacements to be less destabilizing than radical changes.
-	This assumption is relaxed in our simulation, indeed destabilizing effects in each position, instead of being identical for all alternative amino-acids, are now proportional to the Grantham distance \cite{Grantham1974} between the optimal amino-acid in this position and the amino-acid proposed by the non-synonymous mutation.
-	Finally, the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
-	The robustness of such approximation is test in the simulations with sequences of a finite number of sites ($\Nsite=300$).
+This assumption is relaxed in our simulation, such that destabilizing mutations in each position are now proportional to the Grantham distance \citep{Grantham1974} between the optimal amino-acid in this position and the amino-acid proposed by the non-synonymous mutation.
+Additionally, the number of sites in the sequence ($\Nsite$) is assumed to be large such that the selection coefficient is well approximated by the fitness derivative (equation \ref{eq:s}).
+The robustness of such approximation is tested in the simulations with finite sequences ($\Nsite=300$).
+% Finally, the substitution rate is known to vary among sites within a protein, and each site does not contribute equally to the stability of the protein \citep{Echave2016}.
+% To mimic this effect, the destabilizing effect of mutations is scaled by a factor specific to each site, drawn from gamma distribution (see supplementary materials).
 
 Simulations demonstrate that the relation between $\omega$ and log-$\Ne$ is indeed linear, at least in the range explored here, and that the slope of the linear regression matches the expected theoretical value (figure \ref{fig:GoldsteinVsToy}A).
-	Secondly we observe that the parameter $\alpha$ has virtually no effect on the slope of the linear regression, as expected also theoretically (figure \ref{fig:GoldsteinVsToy}B).
-Instead, decreasing $\alpha$ (to more negative values) merely results in an overall increase in $\omega$ over the whole range of $\Ne$ (i.e.
+Secondly, we observe that the parameter $\DeltaGmin$ has virtually no effect on the slope of the linear regression, as also expected theoretically (figure \ref{fig:GoldsteinVsToy}B).
+Instead, decreasing $\DeltaGmin$ (to more negative values) merely results in an overall increase in $\omega$ over the whole range of $\Ne$ (i.e.
 has an impact on the intercept, not on the slope).
-This is due to the fact that decreasing $\alpha$ shifts the equilibrium to higher $x\eq$, since more destabilizing sites can then reach fixation before reaching the point of marginal stability.
+This is due to the fact that decreasing $\DeltaGmin$ shifts the equilibrium to higher $x\eq$, since more destabilizing sites can then reach fixation before reaching the point of marginal stability.
 
-	Finally, we relaxed our assumption that each site of the sequence contribute independently to $\deltaG$, by taking into account the $3$D structure of protein and using a statistical potential to estimate $\deltaG$.
-	We implemented the original model \cite{Williams2006, Goldstein2011, Pollock2012}, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations and pairwise contact potential energies between neighboring amino-acid residues \cite{Miyazawa1985} (see Supplementary Materials).
-	The original works showed that under such $3$D model $\omega$ is approximately independent of $\Ne$ \cite{Goldstein2013}.
+Finally, we relaxed our assumption that each site of the sequence contributes independently to $\deltaG$, by taking into account the $3$D structure of protein and using a statistical potential to estimate $\deltaG$.
+We implemented the original model \citep{Williams2006, Goldstein2011, Pollock2012}, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations and pairwise contact potential energies between neighboring amino-acid residues \citep{Miyazawa1985} (supplementary materials).
+The original works showed that under such $3$D model, $\omega$ is approximately independent of $\Ne$ \citep{Goldstein2013}.
 Using extensive simulations in order to obtain sufficient resolution, we observe that $\omega$ is in fact weakly dependent on $\Ne$, being approximately linear with log-$\Ne$ (figure \ref{fig:GoldsteinVsToy}C).
 Moreover, the observed slope matches the theoretical value considering $\deltadeltaG = 1.0$ kcal/mol for destabilizing mutations and $\Nsite=300$. 
-	In such experiment, $\alpha=-118$ kcal/mol as it is the $\deltaG$ of the optimal sequence of $300$ sites \cite{Goldstein2011} (figure \ref{fig:GoldsteinVsToy}D).
+In such experiment, $\DeltaGmin=-118$ kcal/mol as it is the $\deltaG$ of the optimal sequence of $300$ sites \citep{Goldstein2011} (figure \ref{fig:GoldsteinVsToy}D).
 \begin{figure*}[htb!]
-		\begin{mdframed}
-			\centering
-			\includegraphics[width=0.9\textwidth] {artworks/Elasticity.png}
+ \begin{center}
+ 	 \includegraphics[width=\textwidth] {artworks/elasticity-low.pdf}
+ \end{center}
  \caption{
  \textbf{$\bm{\omega}$ susceptibility to change in $\bm{\Ne}$}.
-				For each population size, $100$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-				Panel A.
-The fixed parameters are $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-$\alpha$ are given in the legend.
-Decreasing $\alpha$ (to more negative values) increases $\omega$ but changes hardly the slope of the linear regression, as expected theoretically.
-				$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
-				Panel B. The fixed parameters are $\Nsite=300$ and $\beta=1.686$. Parameters $\alpha$ and $\gamma$ are given in the legend.
-				$\gamma$ is increased and $\alpha$ is changed accordingly such that the equilibrium value $x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-				The slope of $\omega$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
-				Panel C. In the model of 3D free energy of folding, $\omega$ at equilibrium is weakly dependent on log-$\Ne$, but not completely as originally claimed \cite{Goldstein2013}.
-				This weak dependence matches the theoretical prediction of our additive free energy model that the linear relation (dashed line) has a slope equal to $(\beta \Nsite \gamma)^{-1} = 0.00198 \simeq 0.00124$.
-				Bottom D, the fixed parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
-				Moreover, with the Grantham model, the $\omega$ matches the empirical 3D model of Golstein \& Pollock and the theoretical prediction.
+ Scaling experiment simulating sequence evolution and recording the average $\omega$ (y-axis) observed at equilibrium as a function of $\Ne$ (x-axis).
+ Along the x-axis, $200$ replicate simulations are performed for each different $\Ne$, the average (solid lines) and $90\%$ confidence interval (shaded area) of $\omega$ are shown.
+ Linear regression (dashed lines) allows to estimate the susceptibility ($\hat{\chi}$), which is compared to theoretical expectation.
+ The fixed parameters are $\Nsite=300$, $\beta=1.686$.
+ Simulations are conducted under our additive phenotype model (expect panel C), and for each non-optimal amino-acid $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
+ Panel A; $\DeltaGmin$ and $\gamma$ are given in the legend.
+ $\gamma$ is increased and $\DeltaGmin$ is changed accordingly such that the equilibrium value $\smash{x\eq}$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+ The estimated susceptibility ($\hat{\chi}$) decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
+ Panel B; $\gamma=1$ and $\DeltaGmin$ are given in the legend.
+Decreasing $\DeltaGmin$ (to more negative values) increases $\omega$ but changes hardly the slope of the linear regression, as expected theoretically.
+ Panel C; Folding free energy that determines the stability of the folded native state is computed using $3$D structural conformations and pairwise contact potential energies between neighboring amino-acid residues.
+The estimated susceptibility $\hat{\chi}=0.00124$, and $\omega$ at equilibrium is weakly dependent on log-$\Ne$, but not completely as originally claimed \citep{Goldstein2013}.
+ Under the approximation of an additive free energy model and with empirical estimate of $\gamma = \Delta \Delta G \simeq 1$, our theoretical prediction is $\chi = 0.00198$, close to the observation using $3$D structural conformations.
+ Panel D; structural parameters are matched to their empirical estimates of $\DeltaGmin=\Delta G_{\text{min}} = -118$ and $\gamma = \Delta \Delta G = 1$.
+ Under such parametrization, the scaling behavior is similar to panel C, although with less variance.
  \label{fig:GoldsteinVsToy}
  }
-		\end{mdframed}
 \end{figure*}
-	\subsection*{Epistasis and time to relaxation}
-	Although the equilibrium value of $\omega$ after changes in $\Ne$ is an important feature of the $\omega$-$\Ne$ relationship, another characteristic that is scarcely studied is the dynamic aspect, particularly the relaxation time to reach the new equilibrium $\omega$ \cite{Jones2016}.
-	We observed in our simulations that the determining factor of the relaxation time is the number of sites $\Nsite$ (figure \ref{fig:relaxStability}A).
-	These observations match the theoretical prediction that high epistasis leads to faster return to equilibrium, since more mutational opportunities are available for driving the trait closer to equilibrium.
+
+\subsection*{Time to relaxation}
+
+Although the equilibrium value of $\omega$ after changes in $\Ne$ is an important feature of the $\omega$-$\Ne$ relationship, another characteristic that is scarcely studied is the dynamic aspect \citep{Jones2016}, particularly the relaxation time to reach the new equilibrium $\omega$.
+We observed in our simulations that the determining factor of the relaxation time is the number of sites $\Nsite$ (figure \ref{fig:relaxStability}A), such that the return to equilibrium is faster for longer sequences.
+These observations match the theoretical prediction that more mutational opportunities are available for longer sequences, driving the trait close to equilibrium as a faster rate.
 
 It may be useful to compare the relaxation pattern observed here with the predictions under two alternative models of sequence evolution, representing two extreme scenarios.
-On one side, under a site-independent fitness landscape, such as those contemplated by many current mutation-selection models \cite{Halpern1998, Rodrigue2010}], thus with no epistasis, every site has to adapt on its own to the new change in $\Ne$.
-As a result, the relaxation time is vey long, of the order of the inverse of the per-site substitution rate.
+On one side, having fitness modeled at the level of sites, such as those contemplated by many current mutation-selection models \citep{Halpern1998, Rodrigue2010, Tamuri2012}, thus every site has to adapt on its own to the new change in $\Ne$.
+The relaxation time is very long, on the order of the inverse of the per-site substitution rate.
 On the other side, assuming a fixed distribution of fitness effect (DFE), the response of $\omega$ is instantaneous (figure \ref{fig:relaxStability}B).
+Our model is effectively in between these two extreme scenarios.
 
-	Disregarding simulations under a fixed DFE, another characteristic observed is the non-continuity of $\omega$ after a change in $\Ne$.
+Another characteristic observed is the non-continuity of $\omega$ after a change in $\Ne$.
 Both an increase and decrease in $\Ne$ lead to a discontinuity in $\omega$.
-	Most importantly, $\omega$ is suddenly higher regardless of the change's direction in $\Ne$, then smoothly reaching the equilibrium value (figure \ref{fig:relaxStability}A \& \ref{fig:relaxStability}B).
+Most importantly $\omega$ is suddenly higher regardless of the change's direction in $\Ne$, then smoothly reaching the equilibrium value (figure \ref{fig:relaxStability}A \& \ref{fig:relaxStability}B).
 The similarities in these non-equilibrium properties can both be explained mechanistically.
-	Under low $\Ne$ (between $0$ and $100$ My in the simulation), the phenotype is far away from the optimal phenotype and loosely constrained because the strength of selection is weaker.
-	The reaction to an increase in $\Ne$ (at $100$ My in the simulation) is first a higher constrain and a traction toward a more optimal phenotype, which translate into suddenly higher $\omega$.
-	Conversely, under high $\Ne$ (between $100$ and $400$ My in the simulation) the phenotype is closer to optimal and tightly constrained.
-	The reaction to a decrease in $\Ne$ is first a lower constrain and thus a $\omega$ closer to the neutral case, which translate into higher $\omega$.
-	To note, an increase in $\Ne$ can theoretically and possibly lead to a temporarily $\omega > 1$ due to adaptive evolution \cite{Jones2016}, while an decrease in $\Ne$ always imply $\omega < 1$ due to at most a neutral regime of relaxed selection.
-
+Under low $\Ne$, the phenotype is far away from the optimal phenotype because the strength of selection is weaker.
+The reaction to an increase in $\Ne$ is first an short traction toward a more optimal phenotype, which can results into suddenly higher $\omega$, seemingly to an adaptation.
+Conversely, under high $\Ne$ the phenotype is closer to optimal and the purification of deleterious mutations is stronger.
+The reaction to a decrease in $\Ne$ is a relaxation of the purification and thus a $\omega$ closer to the neutral case, which results into higher $\omega$ until the point of marginal stability.
+To note, an increase in $\Ne$ can theoretically and possibly lead to a temporarily $\omega > 1$ due to adaptive evolution \citep{Jones2016}, while an decrease in $\Ne$ always imply $\omega < 1$ due to at most a neutral regime of relaxed selection.
 \begin{figure*}[htb!]
-		\begin{mdframed}
-			\centering
-			\includegraphics[width=0.9\textwidth] {artworks/Relaxation.pdf}
+ \begin{center}
+ 	\includegraphics[width=\textwidth] {artworks/Relaxation.pdf}
+ \end{center}
  \caption{
@@
  However the number of sites per exon changes between simulations from $\Nsite=5$ to $\Nsite=500$.
-				Moreover, $\gamma$ is changed according to the exon size such that the product is kept constant, thus the susceptibility of the $\omega$ to changes in $\Ne$ is kept constant.
-				Finally, $\alpha$ is changed accordingly such that the equilibrium value $x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
-				As such, whatever the exon size, $x\eq$ and $\chi$ are kept constant and thus the observed effect is due to the number of sites in the exon.
+ Moreover, $\gamma$ is changed according to the exon size such that the product is kept constant, and as a consequence the susceptibility of $\omega$ to changes in $\Ne$ is kept constant.
+ Finally, $\DeltaGmin$ is changed accordingly such that the equilibrium value $\smash{x\eq}$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
+ As such, whatever the exon size, $\smash{x\eq}$ and $\chi$ are kept constant and thus the observed effect is due to the number of sites in the exon.
  We observe that increasing the number of sites implies a reduced time to reach the new equilibrium.
@@
 In context of a fixed fitness landscape, where each amino-acid has different fitness (site-specific profile), the time taken to reach the new equilibrium value of $\omega$ after a change in $\Ne$ is long, such that relaxation rate is on the order of the mutation rate.
-In the context of a distribution of fitness effects (DFE), the relaxation time is non-existent and the new equilibrium value of $\omega$ is reached instantaneously.
+In the context of a fixed distribution of fitness effects (DFE), the relaxation time is non-existent and the new equilibrium value of $\omega$ is reached instantaneously.
  }
  \label{fig:relaxStability}
-		\end{mdframed}
 \end{figure*}
@@
 % Goal and theoretical results
-	The negative relationship between the relative substitution rate of selected mutations ($\omega$) and the effective population size ($\Ne$) has been predicted theoretically under the quasi-neutral theory of evolution.
-	Empirical data showed mitigated confirmation of this prediction, encouraging alternative explanatory mechanism.
-	One such explanation, developed in the context of genotype-phenotype-fitness map, demonstrated the absence of $\omega$ susceptibility to changes in $\Ne$ if the probabilities of phenotype changes are invariant to the current phenotype.
-	We argue that this assumption can be relaxed, and provide theoretical tools to derive the relationship between $\omega$ and $\Ne$ at equilibrium in this context.
-	We apply our framework in the special case of fitness proportional to the proportion of folded proteins. 
-	Our theoretical results demonstrate that the relationship between $\omega$ and $\Ne$ (in log space) is linear with a negative slope.
-	This slope is a function of structural parameters of the protein, and is inversely proportional to the product of the sequence size and the average change in conformational energy of destabilizing mutations, leading in practice to a very subtle and weak relationship.
+
+We provide a compact analytical result for the equilibrium response of $\omega$ to changes in $\Ne$, and relate this response to the parametrization of the genotype-phenotype-fitness map.
+An application to selection against protein misfolding prove that the susceptibility of $\omega$ to variation in $\Ne$ (in log space) is linear with a negative slope.
+Furthermore, this application demonstrate that effective population size and protein expression level are interchangeable in respect to their impact on the response of $\omega$.
+Overall, the susceptibility ($\chi$), is a function of structural parameters of the protein, and is inversely proportional to the product of three terms; namely the sequence size, the inverse temperature ($\beta$), and the average change in conformational energy of destabilizing mutations ($\Delta \Delta G$).
+This product can be several orders of magnitude greater than $1$ in practice, such that the susceptibility of $\omega$ is its inverse, very subtle and weak.
+
+%The negative relationship between the relative substitution rate of selected mutations ($\omega$) and the effective population size ($\Ne$) has been predicted theoretically under the quasi-neutral theory of evolution.
+%Empirical data showed mitigated confirmation of this prediction, encouraging alternative explanatory mechanism.
+%One such explanation, developed in the context of genotype-phenotype-fitness map, demonstrated the absence of $\omega$ susceptibility to changes in $\Ne$ if the probabilities of phenotype changes are invariant to the current phenotype.
+%We argue that this assumption can be relaxed, and provide theoretical tools to derive the relationship between $\omega$ and $\Ne$ at equilibrium in this context.
+%We apply our framework in the special case of fitness proportional to the proportion of folded proteins.
 
 Our compact theoretical results are supported by more complex simulations of protein evolution relaxing several assumptions.
-	Particularly, our theoretical prediction match a numerical model of protein evolution, in which the free energy of the folded are unfolded conformations are computed using $3$D structures of the protein conformations.
-	Previous studies using this model presented an apparent lack of $\omega$ susceptibility to changes in $\Ne$ \cite{Goldstein2013}.
-	We argue the apparent lack of susceptibility is in fact due to the subtle relation, and require extensive computation to be detected.
+Particularly, our theoretical prediction match a numerical model of protein evolution, in which the free energy of the folded and unfolded conformations are computed using $3$D structures of the protein conformations.
+Previous studies using this model presented an apparent lack of $\omega$ susceptibility to changes in $\Ne$ \citep{Goldstein2013}.
+We argue the apparent lack of susceptibility is in fact due to a very subtle and weak relation, and require extensive computation to be detected.
+Based on empirical estimate of the structural parameters $\beta = 1.686$, $\Nsite=300$ sites and $\gamma=\Delta \Delta G = 1.0$ kcal/mol \citep{Zeldovich2007}, the susceptibility $\hat{\chi} \simeq -0.002$.
+In other words, for a relative increase in $\Ne$ or expression level of $6$ orders of magnitude, a factor approximately equal $0.01$ is subtracted to $\omega$, a subtle relationship that requires laborious effort to be detected in simulated data.
 
-	\subsection*{Empirical relevance}
-
-	% Our theoretical results are robsut to approximations
-	% Goldstein \& Pollock model don't have a null susceptibility
-	Based on empirical estimate of the structural parameters ($\beta = 1.686,\ \gamma=1, \Nsite=300$), the slope of the linear relationship between $\omega$ and log-$\Ne$ would be $\chi \simeq -0.002$.
-	Empirically, inference of the $\omega$ susceptibility in Primates, using both polymorphism and divergence data estimated $\chi \simeq X$, approximately $X$ times greater than our theoretical prediction \cite{Brevet2019}.
-	More empirical data across different clades would be required to robustly consolidate such empirical observation.
-	However, in the light of empirical data tending to demonstrate negative relationship between $\omega$ and $\Ne$, models based on the probability of folding are not sufficient to explain the observation of such stastically significant relationship.
-	% Our result show the fitness proportional to proportion of folded folding is at odd with empirical data
+\subsection*{Adequacy to empirical data}
+Empirically, variation in $\omega$ along the branches of phylogenetic trees has been inferred and correlated to proxies of $\Ne$, such as body size or other life-history traits.
+These promising results showed mitigated confirmation of a negative $\omega$ susceptibility to changes in $\Ne$ \citep{Lanfear2014}.
+More recently, phylogenetic integrative methods refined the estimate of covariation between$\omega$ and $\Ne$ along lineages by leveraging polymorphism data and genetic diversity of supposedly neutral site, such as to estimate $\Ne$ by taking into account confounding factors \citep{Brevet2019}.
+These methods can give an estimate of $\hat{\chi} \simeq 0.02$ (see supplementary materials) at least one order of magnitude times greater than our theoretical prediction.
+More empirical data across different clades would be required to robustly consolidate such empirical observation, but as of yet are challenging our theoretical predictions.
 
 % Stress the importance that susceptibility is valid for Ne and protein abondance
-	Furthermore, our theoretical results demonstrate that the relationship between $\Ne$ and $\omega$ is the qualitatively and quantitatively the same as between protein abundance and $\omega$, whenever protein fitness is determined by the deleterious effect of unfolded proteins.
-	As such, dataset relating $\omega$ to either $\Ne$ or protein expression could in principle be used to determine whether the slope is the same.
-	However such endeavor is easily prone to biases and confounding factor, due to study across different clades and across different genes.
+Conjointly, the theoretical prediction of a weak negative $\omega$ susceptibility to changes in protein expression level is in stark contrast with various empirical studies \citep{Duret2000, Rocha2004, Wang2011, Song2017}.
+Furthermore, protein expression level is one of the best predictors of $\omega$ and the empirical estimation of $\chi$ in fungi, archaea and bacteria varies in the range $[-0.046;-0.021]$ (table in supplementary materials) \citep{Zhang2015}, which is in the best case one order of magnitude times greater than our theoretical prediction ($\simeq -0.002$).
+Interestingly, estimation of $\chi$ in animals and plants varies in the range $[-0.026;-0.004]$, which is closer to theoretical estimate (table in supplementary materials).
+Altogether, fitness based on protein stability is a compelling model of molecular evolution, but is apparently not a sufficiently comprehensive model to explain the amplitude of variation of $\omega$ empirically observed along a gradient of either effective population size or protein expression level.
+Our theoretical work aligned with recent independent empirical studies providing negative evidence for the hypotheses linking the rate of sequence evolution with the toxicity constraint \citep{Plata2017, Biesiadecka2020}.
 
 % Protein-protein interactions
-	Notably, our theoretical approximations apply more broadly to protein-protein interactions (using a mean-field argument, supplementary materials), where protein may either be in free form or engaged in a non-specific interactions with other proteins.
-	In non-specific interactions, stabilizing amino-acids at the protein surface are hydrophilic and destabilizing amino-acid are hydrophobic (sticking to  hydrophobic residues in other proteins).
-	Fitting this model with empirical structural estimates \cite{Zhang2008}, we obtain an susceptibility of $\chi \simeq -0.2$ thus a much stronger response than under the model based on conformational stability.
-	This effect is due to less sites in the protein being involved in protein-protein interaction than for conformational stability, in addition to the lower energy engaged in the contacts.
-	Comprehensively, the assumption that protein fitness is in part determined such as to avoid non-specific interaction leads to a stronger response of $\omega$ to changes in $\Ne$, is more compatible with empirical data obtained in comparative genomics.
-	% Such path might worth be explored !
+Notably, the response of $\omega$ to changes in expression level has  also theoretically been found to arise due to protein interactions with the proteome, where protein may either be in free form or engaged in non-specific interactions \citep{Yang2012, Zhang2013}.
+In non-specific interactions at the protein surface, stabilizing amino-acids are hydrophilic and destabilizing amino-acids are hydrophobic, sticking to hydrophobic residues in other proteins \citep{Dixit2013,Manhart2015}.
+Our theoretical results can be applied more broadly to protein-protein interactions using a mean-field argument (supplementary materials).
+Fitting this model with empirical structural estimates \citep{Janin1995a, Zhang2008}, we obtain a susceptibility of $\chi \simeq -0.2$ thus a much stronger response than under the model based on conformational stability.
+This effect is due to fewer sites in the protein being involved in protein-protein interaction than for conformational stability, in addition to the lower energy engaged in the contacts.
 
-	Finally, an integrative inference framework of comparative genomics parameterized by structural parameter, expression level and effective population size can in principle allow to robustly test our prediction with empirical data.
-	Orthogonally, this inference framework could allow to estimate unknown variable, for example ancestral effective population size, with enough signal and calibration of the other parameters. 
-	Such inference framework should be careful with some caveats that we will subsequently discuss.
-	\subsection*{Inference models}
-	% Categories of models
-	Models of inference are classified broadly into phenomenological and mechanistics.
-	Mechanistic models dissect the causation chains and construct model for first-principles, while phenomonological models aims to determine the statistical significativity of parameters.
+Taken together, the assumption that protein fitness is in part determined such as to avoid non-specific interaction, instead of being determined by its probability of folding, leads to a stronger susceptibility of $\omega$ to changes in $\Ne$ or protein expression level.
+In conclusion, models of molecular evolution of the protein taking into account destabilizing effect protein-protein interactions, appears to be more compatible with empirical data obtained in comparative genomics, at least in the light of the susceptibility of $\omega$ to changes in $\Ne$ or expression level.
 
-	% Phenomenological models
-	An example of such phenomenological model are the models parametrized directly in $\omega = dN/dS$.
-	They will for example tell us which branch of the tree, or which site of the sequence has a decrease or increase in $\omega$ but not the underlying reason for such changes, which can be either mutation, selection, drift or another evolutionary force.
-	For example, in such models, a mutation and its reverse mutation do not have opposite selection coefficient.
-	As such they can not disentangle the potentially confounding effect of mutation, selection and drift.
-	Neither can they predict the nature of the functional relationship between population genetics or structural physico-chemical parameters to $\omega$.
-	As such, of mechanistic models are necessary to construct an integrative comparative framework.
-
-	% Site specific mechanistic models of evolution 
-	On the other hand, mechanistic models of evolution mapping genotype to fitness have so far assumed site-independent fitness map, mainly for computational rationals.
-	Previous studies have argued that model of protein evolution should consider epistasis for various empirical reasons such as rate heterogeneity along the sequence, rate and time dependence of convergence, role of compensatory substitutions \cite{Goldstein2017}.
-	We argue that epistasis also has an important role in the response of $\omega$ to changes in $\Ne$, both in terms of susceptibility and dynamic of the response.
-	Fundamentally, we argue that any site-specific model implies a slow dynamic and a strong susceptibility, and adding epistasis to the model imply a faster dynamic and a weaker susceptibility.
-	Intuitively, this effect originate that each site is to adapt independently to changes in $\Ne$ leading to overall a slow response (substitutions must  affect all sites), and a strong susceptibility since each site will change its position in the fitness landscape.
-	Taking into account epistasis, the burden of adapting to changes in $\Ne$ is shared by more sites, such that all of them don't have to adapt. 
-	As a conclusion, both mechanistic and phenomenological models of inference are at risk to endure some culprits.
-	In retrospect, these pitfalls have been anticipated thanks to the intuitions developed in the construction of this mechanistic genotype-phenotype-fitness model.
-
-	\subsection*{Theoretical modeling}
+\subsection*{Physically grounded modeling}
 % Analogy to physics
-	This study elaborate the signature on DNA sequences of a long-term evolutionary process as emergent from population-genetics and structural physico-chemical first principles.
+This study describes the signature imprinted on DNA sequences by an evolutionary process by merging equations from population-genetics and from structural physico-chemical first principles.
 Hence, physico-chemical vocabulary and theory are recruited to describe the molecular details.
 More widely, our modeling is imprinted with vocabulary and analogy to physical systems, with wording in the likes of forces, susceptibility and dynamic.
-	We argue analogy between evolutionary and mechanstic system is not restricted to wording, but allows to visualize and intuit of the causality chains and which approximation can be made such as to derive analytical tractable results.
+We argue analogy between evolutionary and physical system is not restricted to wording, but allows to visualize and intuit of the causality chains and which approximation can be made such as to derive analytical tractable results \citep{Bastolla2012, Bastolla2017}.
 Because it is constructive approach, and not solely intuitive, the robustness of theoretical results to approximations can be asserted by computational implementation. 
-	Computational models offers a means to test the validity and robustness, while the mathematical models offers an intuitive mechanistic mental analogy.
+Computational models offer a means to test the validity and robustness, while mathematical models offer an intuitive mechanistic mental analogy.
+
+% Thermodynamic analogy 
+% Analogy to Clapeyron formula ? TdP/dT = Î”H/Î”V, where T is analogous to Ne
+A first analogy between evolutionary biology and physics comes from statistical mechanics, describing how observable macroscopic properties are emerging from microscopic states.
+As an example, in the Ising model the magnetization is a macroscopic resultant of all microscopic atomic spins that can be in one of two states \citep{Brush1967}.
+At equilibrium when system reaches minimum potential energy, the observable magnetization is controlled by variable such as temperature, and by external forces such as a magnetic field forcing the system.
+By analogy, our atoms are sites of the sequence which can be in stabilizing or destabilizing state, and the macroscopic emerging observable is the relative substitution rate of selected mutation ($\omega$), which is controlled by effective population size ($\Ne$) analogous to temperature.
+Ultimately, the susceptibility of macroscopic observable ($\omega$) to a control variable ($\Ne$) can be tied to the microscopic parameters.
+Another example of observable macroscopic property, though not considered in this manuscript, could be for example site entropy, i.e.
+the effective number of observed amino-acid at equilibrium \citep{Goldstein2016, Jimenez2018, Jiang2018}.
 
@@
 % Elasticity ? Susceptibility (dimensionless) ? Compliance (units of metres per newton) ? Flexibility (units of metres per newton) ? 
-	In the mechanical analogy, the phenotype represent a spatial coordinate of the system, which is determined by external evolutionary forces; namely mutation, selection and drift.
-	These evolutionary forces are homologous to springs attached and stretching the phenotype in various directions and with various intensity, and the aim is to find the equilibrium point.
-	Once the equilibrium point is determined, the deformation of the phenotype (called strain) can be studied whenever applying an differential stress to the system.
-	Elasticity refers to a reversible deformation of a system upon a external force, hence its ability to return to its original state when that force is removed.
-	Intuitively, changing the free energy of the reference phenotype ($\Delta G_{\text{min}} = \alpha$) leads in first approximation to changes in the equilibrium point but not the stiffness of the springs, because only the reference is changed, meaning the elastic coefficient is conserved as found in our theoretical results.
-	Conversely, stiffness ot the strings increases with regards to $\Delta \Delta G = \gamma$, meaning a that change in the stress applied to the system ($\Ne$) result in a weak deformation ($\omega$) at the new equilibrium.
-
-	% Thermodynamic analogy 
-	% Analogy to Clapeyron formula ? TdP/dT = ΔH/ΔV, where T is analogous to Ne
-	In the thermodynamic analogy, the phenotype is a macroscopic resultant of microscopic states of each site of the sequence.
-	Such example could be homologous to the Ising model, and each site of the sequence is a magnetic dipole which can be up or down (destabilizing or stabilizing).
-	The phenotype is an aggregate of the microscopic states, simply the proportion of destabilizing states.
-	Under external forces, the macroscopic proportion is statistically determined by the potential energy of the system, and the susceptibility ()analogous to magnetic susceptibility) of the system is the response of macroscopic observable to change in temperature (inverse of $\Ne$).
+A second analogy between evolutionary biology physics comes from mechanics, where phenotype represent a spatial coordinate of the system, which is determined by external evolutionary forces; namely mutation, selection and drift.
+These evolutionary forces are analogous to springs attached and stretching the phenotype in various directions and with various stiffness, resulting in an equilibrium point.
+Once the equilibrium point determined, an elastic (or reversible) deformation of the phenotype can be studied whenever applying a differential stress to the system.
+In this analogy, changing the free energy of the reference phenotype ($\DeltaGmin = \Delta G_{\text{min}}$) leads in first approximation to changes of reference and equilibrium point but not a change in the stiffness of the springs, meaning the elastic coefficient (analog to susceptibility) is approximately conserved.
+Conversely, stiffness ot the springs increases with regards to $\gamma = \Delta \Delta G$, meaning a that change in the stress applied to the system ($\Ne$) result in a weak deformation at the new equilibrium.
 
@@
 Altogether, this framework can be used as a premise % (groundwork, foundation, starting point,...)
-	to integrate other external evolutionnary forces.
-	Additional molecular processes could be included, for instance GC-biased gene conversion inducing a bias in the nucleic composition of genomes. 
-	The change of equilibrium due to the GC-bias, and the magnitude of the response could be investigated, as to describe the signatures observable on molecular sequence, factoring the effect of mutation, selection and drift.
+to integrate other external evolutionary forces.
+Additional molecular processes could be included, for instance mutational bias, or GC-biased gene conversion inducing a bias in the nucleic composition of genomes. 
+The change of equilibrium due to the GC bias, and the magnitude of the response could be investigated, as to describe the signatures observable on molecular sequence, factoring the effect of mutation, selection and drift.
 Finally, we hope the result of this work, and more broadly the framework can foster the understanding of observable signature of a long-term evolutionary process as emergent from ecological parameters and molecular physico-chemical first principles.
 
+\subsection*{Building block of inference models}
+
+% Categories of models
+At the crossroad between empirical data and theoretical models, inference models can compute the likelihood an observation given a model, and also extract interpretable signature from molecular dataset.
+Models of inference are classified broadly into phenomenological and mechanistic \citep{Rodrigue2010a}.
+Mechanistic models dissect the causation chains and construct a model from first principles, while phenomenological models aims to determine the statistical significance of parameters.
+
+% Phenomenological models
+Example of such phenomenological models are the models parameterized directly in $\omega = dN/dS$.
+They will for example tell us which branch of the tree, or which site of the sequence has a decrease or increase in $\omega$ but not the underlying reason for such changes, which can be either mutation, selection, drift or another evolutionary force.
+For example, in such models, a mutation and its reverse mutation does not have the opposite selection coefficient.
+As such they cannot disentangle the potentially confounding effect of mutation, selection and drift.
+Neither can they predict the nature of the functional relationship between population genetics or structural physico-chemical parameters to $\omega$.
+
+% Site specific mechanistic models of evolution 
+Contrarily, mechanistic models relate structural, population-genetic and ecological parameters to the likelihood of the data.
+As such, mechanistic inference models are suitable to construct an integrative framework relating the signal available in molecular sequences to structural parameters ($\Delta \Delta G$), expression level across genes and effective population size across lineages.
+Once such models is fitted to the data, the estimated parameters can be confronted to their independent empirical estimate ($\Delta \Delta G, \Ne, ...$), which allow to robustly test the model since orthogonal estimation of biological and ecological parameters should be congruent.
+Additionally, once the robustness model has been assessed, this inference framework could allow to estimate unknown variable, for example ancestral effective population size, with enough signal and calibration of the other parameters.
+Altogether, we argue that our theoretical results are a building block to construct an integrated inference framework of molecular evolution, since such models are bridges between intrinsic biological parameters and the signal extractable from molecular sequences.
+
 \section*{Materials \& Methods}
-	Protein sequence evolution is simulated under an origin-fixation model \cite{McCandlish2014}, where one sequence represent the whole population.
+Protein sequence evolution is simulated under an origin-fixation model \citep{McCandlish2014}, where one sequence represents the whole population.
 From the resident DNA sequence $\ci$, we define $\setNeighbors$ as the set of all possible mutant that are one nucleotide away from $\ci$, and were mutant sequences containing a stop codon are excluded.
@@
 where $\mu_{\itoj}$ is the mutation rate between $\ci$ and $\cj$, determined by the underlying $4x4$ nucleotide mutation rate matrix, and ${\submatrix_{\itoj}} = \mu_{\itoj}$ in the case of synonymous substitutions.
-	Various optimization are implemented to reduce the computation time of mutant's fitnesses.
+Various optimizations are implemented to reduce the computation time of mutant fitness.
 The simulation starts with a burn-in period to reach mutation-selection-drift equilibrium.
+
 \subsection*{Models of fitness function}
 \label{MatMet:folding}
+
 Under a simulation of protein folding with an additive model of free energy, the protein's difference in free energy between folded and unfolded state is given by:
 \begin{equation*}
-	\deltaG\left(\ci\right) = \alpha + \Nsite \gamma * x\left(\ci\right), 
+\deltaG\left(\ci\right) = \DeltaGmin + \Nsite \gamma * x\left(\ci\right), 
 \end{equation*}
 where $0 \leq x\left(\ci\right) \leq 1$ is the distance of $\ci$ to the optimal sequence.
-	For each site of sequence, the optimal amino-acids are chosen randomly at initialization, and the distance between the current amino-acid and the optimal is scaled by the Grantham amino-acid distance \cite{Grantham1974}.
+For each site of sequence, the optimal amino-acids are chosen randomly at initialization, and the distance between the current amino-acid and the optimal is scaled by the Grantham amino-acid distance \citep{Grantham1974}.
 Wrightian fitness is defined as the probability of our protein to be in the folded state, given by the Fermi-Dirac distribution: 
 \begin{equation}
-	f(\deltaG\left(\ci\right)) = \dfrac{e^{-\beta \deltaG\left(\ci\right) }}{1 + e^{-\beta \deltaG\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }}, 
+f(\ci) = \dfrac{e^{-\beta \deltaG\left(\ci\right) }}{1 + e^{-\beta \deltaG\left(\ci\right) }} = \dfrac{1}{1 + e^{\beta \deltaG\left(\ci\right) }}, 
 \end{equation}
 where $\beta$ is the inverse of the temperature ($\beta=1/kT$).
-
 Under a simulation with site independent fitness profiles, a fitness profile give a fitness for each amino-acid (vector of size $20$).
@@
 \end{equation}
-	where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \cite{Bloom2017}.
+where $G_{\site}$ is the fitness profile at site $\site$, obtained in empirical experiment \citep{Bloom2017}.
 
-	Under simulation with a fixed distribution of fitness effects (DFE), the selection coefficient of the mutant $\cj$ is gamma distributed (shape $\beta > 0$):
+Under simulation with a fixed distribution of fitness effects (DFE), the selection coefficient of the mutant $\cj$ is gamma distributed (shape $k > 0$):
 \begin{equation}
-	- s \left( \ci,\cj\right) \sim \text{Gamma} \left( \bar{|s|}, \beta \right)
+- s \left( \ci,\cj\right) \sim \text{Gamma} \left( \bar{|s|}, k \right)
 \end{equation}
@@
 From the set of mutants $\setNeighbors$ that is one nucleotide away from $\ci$, we define the subsets $\setNonSynNeighbors$ and $\setSynNeighbors$ that are respectively the set of non-synonymous and synonymous mutants, where $\setNonSynNeighbors \cup \setSynNeighbors = \setNeighbors$.
-	As in previous works \cite{Spielman2015a, DosReis2015, Jones2016}, the ratio of non-synonymous over synonymous substitution rates of the sequence is defined as :
+As in previous works \citep{Spielman2015a, DosReis2015, Jones2016}, the ratio of non-synonymous over synonymous substitution rates of the sequence is defined as :
 \begin{align}
@@
 And $\omega$ is taken as the average of the time-dependent $\omega(t)$ along the simulation.
-	\subsection*{Reproducibility}
+
+\section{Supplementary Material}
+The mathematical development of the general case and supplementary figures are compiled in a single file.
 The simulators written in C++ are publicly available under MIT license at \url{https://github.com/ThibaultLatrille/SimuEvol}.
-	The scripts and instructions necessary to reproduce the experiments are available at \url{https://github.com/ThibaultLatrille/GenotypePhenotypeFitness}.
-	\bibliographystyle{apalike}
-	\bibliography{refs-codons,refs-cds}
+The scripts and instructions necessary to reproduce this study are available at \url{https://github.com/ThibaultLatrille/GenotypePhenotypeFitness}.
 
+\section{Acknowledgments}
+
+The authors gratefully acknowledge the help of Nicolas Rodrigue and Laurent Duret for their advice and review concerning this manuscript.
+This work was performed using the computing facilities of the CC LBBE/PRABI.
+A PhD grant of the École normale supérieure de Lyon awarded to T.L. provided a salary during part of the work.
+
+\bibliographystyle{natbib}
+\bibliography{references}
 \end{document}

### 2020-06-16 manuscript/presentation.tex
Manuscript formatted to MBE guidelines

@@
 ~\\
-The equilibrium phenotype ($\x\eq$)	obtained at mutation-selection balance solves the equation:
+The equilibrium phenotype ($x\eq$)	obtained at mutation-selection balance solves the equation:
 \begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}.
+\ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (\Nstate-1) \simeq - \frac{4\Ne}{\Nsite} \frac{ \partial \ln f(x\eq) }{\partial {x\eq}}.
 \end{align}
@@
 \begin{align}
-\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \dfrac{\dfrac{ \partial \ln f(\x\eq) }{\partial {\x\eq}}}{\dfrac{ \partial^2 \ln f(\x\eq) }{\partial {\x\eq}^2}}.
+\frac{ \der \omega\eq}{\der \ln (\Ne)} & \simeq - \dfrac{\dfrac{ \partial \ln f(x\eq) }{\partial {x\eq}}}{\dfrac{ \partial^2 \ln f(x\eq) }{\partial {x\eq}^2}}.
 \end{align}
@@
 \begin{equation}
-f(\x) \propto \dfrac{1}{1 + e^{\beta (\deltaGMin + \Nsite \gamma \x)}}.
+f(x) \propto \dfrac{1}{1 + e^{\beta (\alpha + \Nsite \gamma x)}}.
 \end{equation}
@@
 	\item $\beta=1.686$ mol/kcal is a fixed parameter at room temperature.
-	\item $\deltaGMin < 0$ (in kcal/mol) is the difference in free energy between folded and unfolded state when all sites are stable.
+	\item $\alpha < 0$ (in kcal/mol) is the difference in free energy between folded and unfolded state when all sites are stable.
 	\item $\gamma > 0$ (in kcal/mol) is the expected change in free energy (between folded and unfolded states) for a destabilizing mutation.
 \end{itemize}
-The equilibrium phenotype $\x\eq$ solves the equation: 
+The equilibrium phenotype $x\eq$ solves the equation: 
 \begin{align}
-\ln \left( \frac{1 - \x\eq}{\x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}.
+\ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (\Nstate-1) \simeq 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)}.
 \label{eq:equilibrium}
 \end{align}
-The $\dnds$ elasticity to changes in $\Ne$ simplify to a compact equation: 
+The $\omega$ elasticity to changes in $\Ne$ simplify to a compact equation: 
 \begin{equation}
@@
 ~\\
-The equilibrium $\x\eq$ is obtained when the exponentially increasing function intersect the blue line. 
+The equilibrium $x\eq$ is obtained when the exponentially increasing function intersect the blue line. 
 \begin{itemize}
@@
 	are considered fixed.
-	\item When $\gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $\x\eq$.
-	\item When $\gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $\x\eq$ notably.
+	\item When $\gamma$ is large (red solid line), increasing $\Ne$ (red dotted line) only moves slightly $x\eq$.
+	\item When $\gamma$ is low (green solid line), increasing $\Ne$ (green dotted line) moves $x\eq$ notably.
 \end{itemize}
-The value of $\alpha$ has been chosen given all other parameters such that $\x\eq$ is the same for $\Ne=10^4$.\\
+The value of $\alpha$ has been chosen given all other parameters such that $x\eq$ is the same for $\Ne=10^4$.\\
 
-$\ln(\frac{1 - \x\eq}{\x\eq})=4\Ne \beta \gamma e^{\beta (\deltaGMin + \Nsite \gamma \x\eq)}$
+$\ln(\frac{1 - x\eq}{x\eq})=4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)}$
 ~\\
@@
 	\item $\Ne$ modulates selection, such populations with high $\Ne$ would have stronger purifying selection, due to the decrease of random drift.
-	\item In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\dnds$) with respect to $\Ne$.
+	\item In molecular sequence, this effect translates in the decrease in the substitution rate of selected mutations relative to the substitution rate of neutral mutation ($\omega$) with respect to $\Ne$.
 	\item Such theoretical prediction had been observed in empirical data.
-	\item Computational models of protein folding have observed that $\dnds$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
-	\item Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\dnds$ and then a decrease.
-	\item Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\dnds$ responses to changes in $\Ne$.
+	\item Computational models of protein folding have observed that $\omega$ can be independent of $\Ne$, which can mathematically be proven under certain assumptions.
+	\item Moreover, non-equilibrium properties can imply that an increase of $\Ne$ can result first in an increase of $\omega$ and then a decrease.
+	\item Together, assumptions about the mapping of sequence to fitness can display a variety of behaviors in the $\omega$ responses to changes in $\Ne$.
 \end{itemize}
@@
 \begin{itemize}
-	\item Provide theoretical tools to derive the relationship between $\Ne$ and $\dnds$ in the context of a genotype-phenotype-fitness map.
+	\item Provide theoretical tools to derive the relationship between $\Ne$ and $\omega$ in the context of a genotype-phenotype-fitness map.
 	\item Apply our framework in the special case of fitness proportional to the probability of protein folding.
@@
 	\item Theoretical slope is -0.00198 and observed is -0.00126
-	\item Decreasing $\alpha$ (to more negative values) increases $\dnds$, by shifting the equilibrium to higher $\x\eq$ since more unstable sites are fixed before the protein became "marginally stable".
-	\item With increased $\x\eq$, the $\dnds$ is higher since non-synonymous mutations between unstable states are effectively neutral.
-	\item The slope of the $\dnds$-$\Ne$ relationship is not changed, a predicted in our theoretical model.
-	\item The slope of $\dnds$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
+	\item Decreasing $\alpha$ (to more negative values) increases $\omega$, by shifting the equilibrium to higher $x\eq$ since more unstable sites are fixed before the protein became "marginally stable".
+	\item With increased $x\eq$, the $\omega$ is higher since non-synonymous mutations between unstable states are effectively neutral.
+	\item The slope of the $\omega$-$\Ne$ relationship is not changed, a predicted in our theoretical model.
+	\item The slope of $\omega$-$\Ne$ relationship decreases proportionally to the inverse of $\gamma$, as predicted by our theoretical model.
 	\item Free energy of the folded state is computed using the $3$D folded structure and pairwise contact potential energies between neighboring amino-acid residues.
 	\item Free energy distribution of unfolded states is approximated using $55$ decoy $3$D structures that supposedly represent a sample of possible unfolded states
-	\item Left panel. Using the 3D structure of protein, $\dnds$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally.
+	\item Left panel. Using the 3D structure of protein, $\omega$ at equilibrium is weakly dependent on log-$\Ne$ but is not independent as claimed originally.
 	\item Right panel. Parameters are $\alpha=-118$, $\gamma=1$, $\Nsite=300$, $\beta=1.686$, and for each non-optimal amino-acid, $\gamma$ is scaled by the Grantham distance to the optimal amino-acid.
@@
 	\item Estimation using polymorphism and divergence data observed a slope of $\simeq -0.04$, approximately $20$ times greater than our prediction. 
-	\item Models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\dnds$ relationship.
+	\item Models based on the probability of folding are at odds with empirically results obtained on the $\Ne$-$\omega$ relationship.
 	\item However, theoretical approximations apply more broadly to protein-protein interactions, where stable state are hydrophilic amino-acid and unstable states are hydrophobic amino-acid at the surface of the protein.
@@
 \begin{itemize}
-	\item Importance of epistatic interactions in the $\Ne$-$\dnds$ relationship
+	\item Importance of epistatic interactions in the $\Ne$-$\omega$ relationship
 	\item Epistasis determines to time to reach a new equilibrium, and that models without epistasis might be too slow to be realistic.
@@
 
+\newpage
+~\\
+\begin{gather}
+\Delta  x\eq \gg \Delta  x\eq \\
+\ln \left( \frac{1 - x}{x} \right) + \ln (19) \\
+4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x)}\\
+\ln \left( \frac{1 - x\eq}{x\eq} \right) + \ln (19) = 4\Ne \beta \gamma e^{\beta (\alpha + \Nsite \gamma x\eq)}\\
+x \rightarrow x' \\
+s = \frac{f(x') - f(x)}{f( x)} \\
+p_{\text{fix}} = \frac{ 2 s}{1 - \e^{-4\Ne s}} \\
+x\eq = \E \left[ x \right]
+\\
+\omega = \E \left[ 2 \Ne p_{\text{fix}} \right]\\
+\chi = \frac{ \der \omega}{\der \ln (\Ne)} 
+\end{gather}
+
 \end{document}

### 2020-06-16 manuscript/supp-mat.tex
Manuscript formatted to MBE guidelines

@@
 \documentclass{article}
-\usepackage{url}
+\usepackage[bookmarks]{hyperref}
+\hypersetup{
+    colorlinks,
+	citecolor=black,
+	filecolor=black,
+	linkcolor=black,
+	urlcolor=black,
+	linktoc=all,
+}
+
+\usepackage[margin=60pt]{geometry}
 \usepackage{amssymb,amsfonts,amsmath,amsthm,mathtools}
-\usepackage{adjustbox}
-\usepackage{float}
-\usepackage{caption}
 \usepackage{lmodern}
 \usepackage{bm,bbold}
+\pdfstringdefDisableCommands{%
+	\renewcommand*{\bm}[1]{#1}%
+	% any other necessary redefinitions 
+}
 \usepackage{xfrac, nicefrac}
 \usepackage{lmodern}
-\usepackage{enumitem}
-\usepackage[margin=60pt]{geometry}
+\usepackage{natbib}
 \pdfinclusioncopyfonts=1
-\captionsetup{width=0.85\textwidth}
 \include{notations}
+
 \begin{document}
@@
 \tableofcontents
-\section{$\bm{\omega}$ elasticity after a change in $\bm{\Ne}$}
+\section{$\bm{\omega}$ susceptibility after a change in $\bm{\Ne}$}
 \subsection{Genotype to phenotype map}
@@
 \begin{align}
-p_{\text{fix}}(x, \dx) &= \frac{ 2 s(x, \dx)}{1 - \e^{-4\Ne s(x, \dx)}}, \\
+p_{\text{fix}}(x, \dx) &= \frac{ 1 - \e^{-2 s(x, \dx)}}{1 - \e^{-4\Ne s(x, \dx)}}, \\
+ &\simeq \frac{ 2 s(x, \dx)}{1 - \e^{-4\Ne s(x, \dx)}}, \\
  &= \frac{ 2 s(x, \dx)}{1 - \e^{-S(x, \dx)}}. \label{eq:pfix}
@@
 \end{align}
-\subsection{$\bm{\omega}$ elasticity after a change in $\bm{\Ne}$}
+\subsection{$\bm{\omega}$ susceptibility after a change in $\bm{\Ne}$}
 Define the function $G(x, \Ne )$ as:
@@
 \end{gather}
-Together, these approximations leads to the following elasticity in equilibrium $\omega$ after change in $\Ne$ as:
+Together, these approximations leads to the following susceptibility in equilibrium $\omega$ after change in $\Ne$ as:
 \begin{align}
@@
 \subsection{Fitness equal to folded fraction}
-A first model is to assume that the fitness is equal to the folded fraction \cite{Goldstein2013}:
+A first model is to assume that the fitness is equal to the folded fraction \citep{Goldstein2013}:
 \begin{gather}
@@
 \end{align}
-Finally, $\omega$ elasticity after a change in $\Ne$ is simply:
+Finally, $\omega$ susceptibility after a change in $\Ne$ is simply:
 \begin{align}
@@
 \subsection{Selective cost proportional to amount of misfolded protein}
-A slight variation is to assume that the selective cost itself is proportional to the total amount of misfolded protein. For a given protein with expression level $y$:
+A slight variation is to assume that the selective cost itself is proportional to the total amount of misfolded protein  \citep{Drummond2005a, Wilke2006, Drummond2008, Serohijos2012}.
+For a given protein with expression level $y$:
 \begin{eqnarray}
@@
 \begin{eqnarray}
-\frac{\partial \ln (f(x))}{\partial x} &\simeq& -A \beta \gamma \Nsite y  e^{\beta(\alpha + \gamma \Nsite x)}
+\frac{\partial \ln (f(x))}{\partial x} &\simeq& -A y \beta \gamma \Nsite e^{\beta(\alpha + \gamma \Nsite x)}.
 \end{eqnarray}
-or equivalently
+Under this model, the phenotype at equilibrium is given by:
 \begin{eqnarray}
-4 \Ne \frac{\partial \ln (f(x))}{\partial x} &\propto& -4 \Ne y e^{\beta(\alpha + \gamma \Nsite x)}
+\lambda_{K}(x\eq) = 4\Ne y A \beta \gamma \Nsite e^{\beta(\alpha + \gamma \Nsite x\eq)} \text{ from eq. \ref{eq:equilibrium_lambda} and \ref{eq:df_dx}}.
 \end{eqnarray}
-Thus, under this model, an increase in the effective population size $\Ne$ or the expression level $y$ both have the same effect on the molecular evolutionary process undergone by the protein.
-Moreover, $\omega$ elasticity after a change in $\Ne$ is the same as before:
+And the susceptibility of $\omega$ after a change in $\Ne$ is the same as before:
 \begin{align}
@@
 \end{align}
+Since $\Ne$ and $y$ are confounded factors, meaning they only appear in the equation as a product between the two, implicit derivation leads to the same result whenever the derivation is w.r.t $\Ne$ or $y$, leading to same compact susceptibility:
+\begin{eqnarray}
+\frac{ \der \omega\eq}{\der \ln (y)} = \frac{ \der \omega\eq}{\der \ln (\Ne)} \simeq - \frac{1}{\beta \Nsite \gamma}.
+\end{eqnarray}
+
 \subsection{Translational errors}
@@
 \end{eqnarray}
-Moreover, $\omega$ elasticity after a change in $\Ne$ is again the same as before:
+Moreover, $\omega$ susceptibility after a change in $\Ne$ is again the same as before:
 \begin{align}
@@
 \subsection{Cost-benefit argument}
-This models assumes that the :
+The cost-benefit argument \citep{Beaulieu2018} is based on two assumptions 
 \begin{enumerate}
  \item the expression level is regulated so that the total number of \emph{functional} macromolecules is maintained at a target level $y$;
- \item the log-fitness is proportional to the ratio of the \emph{total} cost of expression over the benefit contributed by the protein (Beaulieu et al).
+ \item the log-fitness is proportional to the ratio of the \emph{total} cost of expression over the benefit contributed by the protein .
 \end{enumerate}
@@
 \end{eqnarray}
-Alternative cost-exp models could also be used, allowing for a non-linear cost function for expression or for some elasticity of the realized equilibrium expression level, as a function of the number of mutations (Gout et al).
+Alternative cost-expression models could also be used, allowing for a non-linear cost function for expression or for some susceptibility of the realized equilibrium expression level, as a function of the number of mutations.
 Under these models, the strength of selection is still expected to be an increasing function of $y$, although not linear, e.g.
@@
 where $g$ is some function of $y$.
-Moreover, $\omega$ elasticity after a change in $\Ne$ is again the same as before:
+Moreover, $\omega$ susceptibility after a change in $\Ne$ is again the same as before:
 \begin{align}
@@
 where only the last term depends on $x_i$.
-Finally, applying the main result of this work to the present case allows us to express the elasticity of $\omega$ as a function of $\Ne$ as:
+Finally, applying the main result of this work to the present case allows us to express the susceptibility of $\omega$ as a function of $\Ne$ as:
 \begin{eqnarray}
@@
 \subsection{Empirical calibration}
-Based on empirical estimates obtained from Zhang et al.
+Based on empirical estimates \citep{Zhang2008}.
 The mean fraction of hydrophobic residues at the surface of proteins is $0.22 \pm 0.06$.
@@
 
-\section{Simulation of protein-coding DNA sequences evolution}
-\subsection{Probability of folding using the 3d structure of protein}
+\section{Empirical estimation of $\bm{\chi}$}
+Substitution rate as a function of expression level compiled by \citep{Zhang2015}.
+\begin{table}[ht]
+	\centering
+	\begin{tabular}{|l|l|c|c|}
+		\hline
+		Type & Specie & $\hat{\chi}$ & $r^2$ \\ 
+		\hline
+		Plant & Oryza sativa & -0.008 & 0.047 \\ 
+		Plant & Arabidopsis thaliana & -0.012 & 0.128 \\ 
+		Archaea & Sulfolobus solfataricus & -0.037 & 0.097 \\ 
+		Archaea & Thermococcus kodakarensis & -0.026 & 0.058 \\ 
+		Fungi & Saccharomyces cerevisiae & -0.029 & 0.211 \\ 
+		Fungi & Aspergillus nidulans & -0.034 & 0.124 \\ 
+		Bacteria & Escherichia coli & -0.021 & 0.151 \\ 
+		Bacteria & Bacillus subtilis & -0.046 & 0.151 \\ 
+		Animal & Caenorhabditis elegans & -0.026 & 0.039 \\ 
+		Animal & Drosophila melanogaster & -0.005 & 0.021 \\ 
+		Animal & Mus musculus & -0.008 & 0.085 \\ 
+		Animal & Homo sapiens & -0.004 & 0.031 \\ 
+		\hline
+	\end{tabular}
+\end{table}
+In \citep{Brevet2019}, the covariance matrix with $\ln(\Ne)$ and $\ln(\omega)$ as entries allow us to approximate $\chi$:
+\begin{align}
+	\widehat{\frac{\der \ln (\omega)}{\der \ln (\Ne)}} & = \dfrac{\text{Cov}[\ln(\omega), \ln(\Ne)]}{\text{Var}[\ln(\Ne)]}\text{ from linear regression.} \\
+	\Rightarrow \widehat{\frac{\der \omega}{\omega \der \ln (\Ne)}} & = \dfrac{\text{Cov}[\ln(\omega), \ln(\Ne)]}{\text{Var}[\ln(\Ne)]} \\
+	\Rightarrow \widehat{\chi} & \simeq \widehat{\omega} \dfrac{\text{Cov}[\ln(\omega), \ln(\Ne)]}{\text{Var}[\ln(\Ne)]} \\
+	\Rightarrow \chi & \simeq 0.2 \dfrac{-0.45}{4.45} \\
+	\Rightarrow \chi & \simeq - 0.02 
+\end{align}
+\section{Simulation using the 3d structure of protein}
 We simulated substitutions in the protein phosphatase ($\Nsite=300$ codon sites).
@@
 
-\subsection{Simulated $\bm{\omega}$ elasticity to changes in $\bm{\Ne}$}
+\section{Simulated $\bm{\omega}$ susceptibility to changes in $\bm{\Ne}$}
+\subsection{$\bm{\omega}$ susceptibility with gamma distributed selection coefficient}
 \begin{center}
@@
 \end{center}
-\textbf{$\bm{\omega}$ elasticity with gamma distributed selection coefficient}.
 $\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
 For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-In the model of gamma distributed fitness effect, $\omega$ at equilibrium is strongly dependent on log-$\Ne$ where the slope correlation is proportional to the inverse of the shape parameter of the gamma distribution \cite{Welch2008}.
+In the model of gamma distributed fitness effect, $\omega$ at equilibrium is strongly dependent on log-$\Ne$ where the slope correlation is proportional to the inverse of the shape parameter of the gamma distribution \citep{Welch2008}.
+
+\subsection{$\bm{\omega}$ susceptibility with amino-acid fitness profiles}.
 \begin{center}
@@
 \end{center}
-\textbf{$\bm{\omega}$ elasticity with amino-acid fitness profiles}.
 $\omega$ at equilibrium as a function of $\bm{\Ne}$ (relative).
 For each population size, $200$ simulations were performed and the average (solid line) and $90\%$ confidence interval (shaded area) are shown.
-In the model of site-wise amino-acid fitness profiles taken from \cite{Bloom2017}, $\omega$ at equilibrium is strongly dependent on log-$\Ne$. 
+In the model of site-wise amino-acid fitness profiles taken from \citep{Bloom2017}, $\omega$ at equilibrium is strongly dependent on log-$\Ne$.
+
+\subsection{$\bm{\omega}$ susceptibility with additive free energy of folding}.
 \begin{center}
@@
 \end{center}
-\textbf{$\bm{\omega}$ elasticity with additive free energy of folding}.
 $\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
@@
 The simulations of our additive free energy model match the theoretical prediction that the slope of the linear relation (dashed line) is equal to $\beta \Nsite \gamma)^{-1} = 0.00198 \simeq = 0.00199$.
-%The non-monotony is suspected to be due to the discrete number of sites and states, such that the changes in $\deltaG$ after a mutation is either $-1$, $0$ or $1$.
-Such non-monotony is not observed with the Grantham model, in which the $\omega$ is lower and the slope of elasticity lower, closer to the empirical 3D model of Golstein \& Pollock.
+The non-monotony is suspected to be due to the discrete number of sites and states, such that the changes in $\deltaG$ after a mutation is either $-1$, $0$ or $1$.
+Such non-monotony is not observed with the Grantham model, in which the $\omega$ is lower and the slope of susceptibility lower, closer to the empirical 3D model.
+
+%\subsection{Effect of $\bm{\beta}$ on the susceptibility}.
 %\begin{center}
@@
 %\end{center}
-%\textbf{Effect of $\bm{\beta}$ on the elasticity}.
 %$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
@@
 %Increasing $\beta$ increases the slope of the $\omega$-$\Ne$ relationship, as predicted in our theoretical model.
+
+%\subsection{Effect of site variance on the susceptibility}.
 %\begin{center}
@@
 %\end{center}
-%\textbf{$\bm{\omega}$ elasticity with additive free energy of folding}.
 %$\omega$ at equilibrium as a function of $\bm{\Ne}$ (log scale).
@@
 However the slope of the $\omega$-$\Ne$ relationship is not sensibly changed.
-\subsection{Simulated relaxation time of $\bm{\omega}$}
+
+\section{Simulated relaxation time of $\bm{\omega}$}
+
+\subsection{Relaxation time of $\bm{\omega}$ dependence on $\Nsites$}.
 \begin{center}
@@
 \end{center}
-\textbf{Relaxation time of $\bm{\omega}$ dependence on $\Nsites$}.
+
 $\omega$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
@@
 The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Increasing $\Nsite$ implies a higher $\omega$ at equilibrium, a lower elasticity of the $\omega$ to changes in $\Ne$ and a higher rate of relaxation.
+Increasing $\Nsite$ implies a higher $\omega$ at equilibrium, a lower susceptibility of the $\omega$ to changes in $\Ne$ and a higher rate of relaxation.
+
+\subsection{Relaxation time of $\bm{\omega}$ dependence on $\Nsites$, while correction for $\gamma$}.
 \begin{center}
@@
 \end{center}
-\textbf{Relaxation time of $\bm{\omega}$ dependence on $\Nsites$, while correction for $\gamma$}.
 $\omega$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
@@
 The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\omega$ to changes in $\Ne$ is kept constant.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  susceptibility of the $\omega$ to changes in $\Ne$ is kept constant.
 Increasing $\Nsite$ implies a higher $\omega$ at equilibrium, and a higher rate of relaxation.
+
+\subsection{Relaxation time of $\bm{\omega}$ for the Grantham model}.
 \begin{center}
@@
 \end{center}
-\textbf{Relaxation time of $\bm{\omega}$ for the Grantham model}.
 $\omega$ Relaxation after a brutal change in $\Ne$, the left and right panel correspond to low $\Ne$ ($1e^{5}$) and the middle panel corresponds to high $\Ne$ ($2e^{6}$). 
@@
 The number of sites is changed from $\Nsite=15$ to $\Nsite=158$, and the number of replicates is changed accordingly such that the total number of sites ($\Nsite*r$) is kept constant.
-Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  elasticity of the $\omega$ to changes in $\Ne$ is kept constant.
+Moreover, $\gamma$ is changed according to $\Nsite$ such that the product $\gamma\Nsite$ is kept constant, thus the  susceptibility of the $\omega$ to changes in $\Ne$ is kept constant.
 Finally, $\alpha$ is changed according to $\Nsite$ and $\gamma$ such that the equilibrium value $x\eq$ is kept constant, by solving numerically equation \ref{eq:equilibrium}.
@@
 
-\subsection*{Distribution of fitness effects}
+\section{Distribution of fitness effects}
 DNA mutations changing a genotype can result in a change of phenotype, and ultimately a change in fitness.
@@
 Empirically, these distributions are of particular importance since they can be obtained experimentally or inferred with other data.
-As an example, DFE can be inferred from polymorphism dataset \cite{Eyre-walker2007, Galtier2016}. 
-Moreover, the distribution of $\deltadeltaG=\gamma \Nsite \dx$ of novel mutations can be obtained experimentally.
+As an example, DFE can be inferred from polymorphism dataset \citep{Eyre-walker2007, Galtier2016}. 
+Moreover, the distribution of $\deltadeltaG$ for novel mutations can be obtained experimentally.
 
@@
 
-\bibliographystyle{apalike}
-\bibliography{refs-codons,refs-cds}
+\bibliographystyle{natbib}
+\bibliography{references}
 \end{document}
